### 2024-05-19 README.Rmd
Started writing this commentary

---
output: github_document
bibliography: bibliography/bibliography.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r include=FALSE}
library(dplyr)
library(ggplot2)
```

# Use of AI in Teaching and Learning: Some Considerations

<!-- badges: start -->

<!-- badges: end -->

>> Mission Statement  
>> Our organization seeks to level up on all matters related.

## Background: AI roams the World

On November 30, 2022, OpenAI released what has become its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models] that mimics natural language. The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and was rapidly panned as "the best artificial intelligence chatbot ever released to the general public" [@roose_brilliance_2023], and able to generate "impressively detailed" and "human-like" text [@lock_what_2022].

<!--impressive feat of marketing: think about what it does well, what it does poorly, and how the user interface is essential to obscure its uselessness.-->

Universities have had to respond in a compressed time period to the release of generative AI tools by a handful of large American corporations, notoriously [Meta](https://en.wikipedia.org/wiki/Meta_Platforms) (or were we to deadname it, facebook), [OpenAI](https://en.wikipedia.org/wiki/OpenAI), [Google](https://en.wikipedia.org/wiki/Alphabet_Inc.), and [Microsoft](https://en.wikipedia.org/wiki/Microsoft).

McMaster developed a set of provisional guidelines (when where they developed and released? It is possible that this was done as a relatively urgent matter). These [guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) where released under a Creative Commons Licence and have since inspired those of other post-secondary institutions including Trent University [-@trent_university_generative_2024] and University of Guelph [-@office_of_teaching_and_learning_provisional_2023].

There is now an effort underway to update the preliminary guidelines (to make the permanent?) McMaster [launched a committee](#ai-committee-announced) to explore the use of AI across campus. This committee is composed of three sub-committees that examine the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in [our work] (the Provost's words)(probably meaning the university's work; not sure if the Provost themself uses AI).

The sub-committees, in turn, convened panels to:

1. Provide [provide recommendations]({#ai-panel-tasks}) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/)

2. And also to do (what? consultation?).

3. Q: What are the terms of reference of the panels? Who will decide which recommendations are adopted and which are ignored?

According to the announcement of the AI Committee:

> AI is a powerful tool and I believe that universities have an important role to play in its adoption.  
> -- Susan Tighe, Provost, McMaster University

Hammers are powerful tools, sometimes even to conduct research or to teach. Word processors are powerful tools. What makes AI different from other tools that it requires a pan-university consultation on an "_ethical approach to shaping policies and developing new ways of using AI to ensure a responsible and beneficial integration where it makes sense_"?

- Where does it make sense?

To understand why many organizations, including McMaster University, have decided to spend considerable resources considering the adoption and responsible use of AI, it is essential to have a working understanding of what AI does differently from other tools. We also need to understand what are the goals of the organization. When does it make sense for a university like McMaster to "integrate" AI (pressumably as part of the core mission of the organization)?

More generally, we must be alert to the fact that technologies are never politically neutral: it is important to ask two questions:

- what does a technology do for whom?

- And what does it do to whom?

**McMaster's Mission**

https://president.mcmaster.ca/mission-vision/

The mission of the university is "the discovery, communication and preservation of knowledge," accomplished with a commitment to "creativity, innovation and excellence". These are aspirations that must respond to some underlying values, and indeed, the university values "integrity, quality, inclusiveness and teamwork". Ultimately, the university aims to "inspire critical thinking, personal growth, and a passion for lifelong learning" and to serve "the social, cultural, and economic needs of our community and our society."

From the Overview on Generative AI document:

> Some of those uses in our context could include qualitative and quantitative data analysis, writing text (e.g. reference letters, grant applications, job descriptions, report summaries), translation, captioning, multiple formats for learning and engagement, and personalized support.

## What are LLMs?

To understand why many organizations have decided to spend considerable resources considering the adoption and responsible use of AI, it is essential to have a working understanding of what AI does differently from other tools.

Large Language Models are models composed of multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].

<!-- 
IBM uses a dataset called Project CodeNet Dataset, which is sourced from two online judge web sites, AIZU Online Judge and AtCoder. https://research.ibm.com/blog/codenet-ai-neurips-2021 According to IBM "[the] end goal of CodeNet is to enable developers to create systems that can modernize existing codebases, as well as fix errors and security vulnerabilities in code."

From Aizu Terms of Use: https://onlinejudge.u-aizu.ac.jp/term_of_use

Handling of Submitted and Registered Codes

Commercial uses of the registered codes are prohibited.

You acknowledge and agree that users of this site may use copies or modifications of the public codes only for educational or research purposes.

You acknowledge and agree that public source codes may be opened collectively to persons authorized to access only for educational or research purposes.

You acknowledge and agree that source code submitted may be used by University of Aizu and related research institutes only for educational or research purposes.

IBM studiously avoids the word "commercial" in its posts about LLM, foundational models, and CodeNet. Is IBM doing all this research as a non-commercial effort? Will those developers creating systems, modernizing databases, fixing errors, etc., be able to access this for free? What does "commercial uses" even mean?
-->

What are neural networks? Here are some resources to learn some elemental facts about neural networks.

https://medium.com/@asifurrahmanaust/lesson-3-neural-network-is-nothing-but-a-linear-regression-e05a328a0f23

https://www.ibm.com/topics/large-language-models

https://en.wikipedia.org/wiki/Large_language_model#Further_reading

Behind all the jargon, neural networks (LLMs are neural networks) are regression models [@ripley_neural_1994]. Regression of course refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean). <!-- **NOTE**: also called "reversion to mediocrity". -->

**NOTE**: Labeling data is an essential, yet extremely expensive aspect of training GenAIs.

## What do GenAI tools do for students at McMaster?

## What do GenAI tools do to students at McMaster?

### Persistence of skills (how long will learning to use a chatbot be useful?)

How useful will be AIs in the future? Think of model poisoning, and the possibility that we are going to see (or may already have seen) Peak ChatGPT.

De-duplication of training data improves LLMs [@lee2022deduplicating]

No wonder: it amounts to having more different information.

But there is the following implication: once human-generated content is exhausted and/or has been overwhelmed by chatbot generated content, de-duplication becomes increasingly challenging.

```{r simulate-simple-example}
set.seed(13253)

b1 <- 1.5

df <- data.frame(x = b1 * runif(100, 
                                0, 
                                10),
                 e = rnorm(100),
                 type = "training data") |>
  mutate(y = x + e)

mod <- lm(y ~ x, 
          data = df)
```

```{r plot-simulated-data, fig.cap="Simple Example of a Regression Model"}
ggplot(df,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```

Predict:
```{r data-vs-predictions-comparison, fig.cap="Training data and model output compared"}
df_pred <- data.frame(x = b1 * runif(100, 
                                0, 
                                10))

y_hat <- predict(mod, df_pred)

df_pred <- df_pred |> 
  mutate(y = y_hat,
         type = "model output")

df_pred <- rbind(df |>
                   select(x, y, type),
                 df_pred |>
                   select(x, y, type))

ggplot() + 
  geom_point(data = df_pred,
  aes(x = x,
      y = y,
      color = type,
      shape = type)) +
  theme_minimal()
```

```{r data-vs-prediction-variance}
df_pred |> 
  group_by(type) |> 
  summarize(variance = var(y))
```

## What do GenAIs do for the mega-corporations pushing these tools?

## What do GenAIs to to everyone else?

### Use of materials on the web

In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )

### Copyright issues

LLM are known to regurgitate verbatim chunks of the training datasets. This is called _memorization_, another choice of words that tends to antropomorphize what is essentially a statistical model. Memorization, in fact, is just a fancy way of saying "model overfitting"--in other words, an extreme regression to the mean that excels at predicting data in the training set, but is poor at projecting new data points. This issue can be as high 7% in the cases of some LLMs [@peng_near_duplicate_2023, p. 8].

### Environmental impacts

Microsoft: Here's how burning the planet will help to save it.

https://blogs.microsoft.com/on-the-issues/2024/05/15/microsoft-environmental-sustainability-report-2024/ [@smith_our_2024]

https://blogs.microsoft.com/on-the-issues/2023/11/16/accelerating-sustainability-ai-playbook/ [@nakagawa_accelerating_2023]

ChatGPT consumes substantially more than a search...are the results better? [@https://www.brusselstimes.com/1042696/chatgpt-consumes-25-times-more-energy-than-google]

Compare the results:

![A search using Duck Duck Go](images/logical-puzzle-web-search.png)

![A conversation with ChatGPT-4](images/logical-puzzle-chat-gpt-4.png)

<!--

<iframe src="https://assemblag.es/@hugo/112450401657084750/embed" width="400" allowfullscreen="allowfullscreen" sandbox="allow-scripts allow-same-origin allow-popups allow-popups-to-escape-sandbox allow-forms"></iframe>

https://assemblag.es/@hugo/112450401657084750 -->

### Replication of unjust structures of power

When using a Generative AI we must ask whose perspectives that tool represents.

## As an aside

On antropomorphizing GenAI tools. This happens in ways that are barely perceptively. For example, in the document "Overview of AI at McMaster" we read that GenAI's somehow have an "ability to engage in nuanced conversation". This phrasing could be simply a literary device, but even so is indicative of a tendency to ascribe human attributes to a statistical model. Doing this carelessly risks amplifying existing biases. The effort to wrap a statistical model in a human-like skin may actually be more devious, as in the new, sexy ChatGPT interface:

<video width="320" height="240" controls>
<source src="video/open-ai-commercial.mp4" type="video/mp4">
</video>

**NOTES:** Need to research more the argument that UX is so important in the case of these tools because what they do well is trivial and what they do poorly is what is being hyped.

## Concluding remarks

ChatGPT was an awesome feat of marketing [@cowen_chatgpt_2023], but much coverage of the relevant issues in the media has been so credulous it has stretched credulity past Pluto [@burneko_if_2024]^[Bob woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called.]. There are now calls to press pause on the hype [@angwin_opinion_2024] but these are still drown by the credulous coverage.

I suggest that universities should avoid the mistakes of the media.

Why? The reputational risk...universities are under siege from unsympathetic or actively hostile actors. Universities main source of power is their reputation as places that preserve and expand knowledge in a principled way. "This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an instution of higher-education is not needed to create spam and dissinformation.

## Loose thoughts

Are GenAI reality-warping machines? Think about the small and large ways in which these tools can be used to rewrite history, for relatively banal commercial objectives (monetizing clicks) or worse, for malicious propaganda campaigns.

**NOTE** Russia copied millions of articles from wikipedia wholesale, to replace the open encyclopedia with a home-grown version without all the bits that the current regime finds objectionable [@corfield_russia_2023; @jankowicz_russia_2023].

> “The first step in liquidating a people,' said Hubl, 'is to erase its memory. Destroy its books, its culture, its history. Then have somebody write new books, manufacture a new culture, invent a new history. Before long the nation will begin to forget what it is and what it was. The world around it will forget even faster.”
> -― Milan Kundera, The Book of Laughter and Forgetting

Potential for weaponization of GenAI. **Farfetched (?) idea** Could it be that GenAI will be at some point declared a military-grade technology with restricted applications?

## Notes

### AI Committee announcement {#ai-committee-announced}

This is the announcement of the launch of the AI Committee (see email from the Office of the Provost, dated March 20, 2024):

<!--https://app.fastmail.com/mail/search:expert+panels/T2d9b9367655ae218.M6205d28844daa67dfd3432ff-->

> "McMaster launches AI committee
>
> Whether you are a researcher, faculty member, student, administrator, or have another role on campus, you are likely thinking about Artificial Intelligence (AI) and its challenges and potential.
>
> AI is a powerful tool and I believe that universities have an important role to play in its adoption. At McMaster, we are well placed to prioritize an ethical approach to shaping policies and developing new ways of using AI to ensure a responsible and beneficial integration where it makes sense.
>
> I’m pleased to announce that McMaster has launched an AI Advisory Committee to explore the use of AI across campus. Three subcommittees have also been started to examine the use of AI in teaching and learning, research and operational excellence in our work. It is a strategy that makes McMaster one of the few institutions taking a pan-university approach to AI.
>
> Our earlier work to develop provisional guidelines for the use of generative AI in teaching and learning drew interest from other post-secondary institutions, which adopted McMaster's guidelines after we made them available through a Creative Commons licence.
>
> The committee will be led by three co-chairs, each bringing different expertise. They are Matheus Grasselli, deputy provost, Gayleen Gray, associate vice-president and chief technology officer and Gianni Parise, acting deputy vice-president, Research. They will be supported by Erin Aspenlieder, special advisor to the provost on generative AI.
>
> The three subcommittees will each by led by a subject matter expert. Kim Dej, vice-provost, Teaching and Learning, will lead the exploration of AI in teaching and learning, Maggie Pooran, executive director, Health, Safety, Well-Being & Labour Relations, will explore its use in operational excellence, and Martin Horn, associate dean, Faculty of Humanities, will look at implications for research.
>
> Each expert subcommittee will have a pool of contributors made up of members of the McMaster community. Depending on the initiative, some or all members will be invited to complete the specific work based on their expertise, interest and availability.
>
> If you would like to be considered for the expert panels, please complete this expression of interest form."

### Panel tasks{#ai-panel-tasks}

The task of the Artificial Intelligence Advisory Committee Expert Panel on Teaching and Learning was initially defined as follows by the Vice-Provost Teaching and Learning in an email dated April 16:

<!--https://app.fastmail.com/mail/search:erin/T6fae26856703ef40.Mfe5dff646c6ae452b84ef4bf-->

> [T]he first task of this Expert Panel will be to provide recommendations to the AI Advisory Committee on updates to the Provisional Guidelines on the Use of Generative AI in Teaching and Learning.

<!--This does not say anything about what the Committee will do with the recommendations; I cannot find anything about this.-->

## References

### 2024-05-19 README.Rmd
Edits

@@
 
->> Mission Statement  
->> Our organization seeks to level up on all matters related.
+> Mission Statement: Our organization seeks to level up on all matters related.  
+> -- An Organization Run by its Marketing Department
 
-## Background: AI roams the World
+## Background: An AI roams the University
 
@@
 
-- Where does it make sense?
+- Where does it make sense? Presumably this is what the Committee on AI will try to answer.
 
@@
 
-- what does a technology do for whom?
+- What does technology X do for whom?
 
@@
 
+## Not learning
+
+Words go here.
+
+## Learning the wrong stuff
+
+Words go here.
+
+## Deskilling
+
+Words go here.
+
 ### Persistence of skills (how long will learning to use a chatbot be useful?)
@@
 
-```{r plot-simulated-data, fig.cap="Simple Example of a Regression Model"}
+```{r plot-simulated-data, out.width="50%", fig.cap="Simple Example of a Regression Model"}
 ggplot(df,
@@
 Predict:
-```{r data-vs-predictions-comparison, fig.cap="Training data and model output compared"}
+```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Training data and model output compared"}
 df_pred <- data.frame(x = b1 * runif(100, 
@@
 
+Words go here.
+
 ## What do GenAIs to to everyone else?
@@
 
-ChatGPT consumes substantially more than a search...are the results better? [@https://www.brusselstimes.com/1042696/chatgpt-consumes-25-times-more-energy-than-google]
+ChatGPT consumes substantially more energy than a regular web search...are the results better? [@https://www.brusselstimes.com/1042696/chatgpt-consumes-25-times-more-energy-than-google]
 
@@
 
-![A search using Duck Duck Go](images/logical-puzzle-web-search.png)
+![A search using Duck Duck Go](images/logical-puzzle-web-search.png){width=50%}
 
-![A conversation with ChatGPT-4](images/logical-puzzle-chat-gpt-4.png)
+![A conversation with ChatGPT-4](images/logical-puzzle-chat-gpt-4.png){width=50%}
+
+https://assemblag.es/@hugo/112450401657084750
 
 <!--
-
 <iframe src="https://assemblag.es/@hugo/112450401657084750/embed" width="400" allowfullscreen="allowfullscreen" sandbox="allow-scripts allow-same-origin allow-popups allow-popups-to-escape-sandbox allow-forms"></iframe>
-
-https://assemblag.es/@hugo/112450401657084750 -->
+ -->
 
@@
 
-> “The first step in liquidating a people,' said Hubl, 'is to erase its memory.
+>The first step in liquidating a people,' said Hubl, 'is to erase its memory.
 Destroy its books, its culture, its history.
@@
 Before long the nation will begin to forget what it is and what it was.
-The world around it will forget even faster.”
-> -― Milan Kundera, The Book of Laughter and Forgetting
+The world around it will forget even faster.
+>-― Milan Kundera, The Book of Laughter and Forgetting
 

### 2024-05-20 README.Rmd
Further edits

@@
 
-## Background: An AI roams the University
+By: Antonio Paez
+
+All opinions expressed here are mine and do not necessarily represent those of my employer(s).
+
+## Background: An AI is Haunting Campus^[Karl Marx wrote that "A spectre is haunting Europe--the spectre of communism"; in this case, AI is the shade of another spectre, that of unregulated surveillance capitalism.]
 
@@
 
-Universities have had to respond in a compressed time period to the release of generative AI tools by a handful of large American corporations, notoriously [Meta](https://en.wikipedia.org/wiki/Meta_Platforms) (or were we to deadname it, facebook), [OpenAI](https://en.wikipedia.org/wiki/OpenAI), [Google](https://en.wikipedia.org/wiki/Alphabet_Inc.), and [Microsoft](https://en.wikipedia.org/wiki/Microsoft).
+Universities have had to respond in a compressed span of time to the release of generative AI tools by a handful of large American corporations, notoriously [Meta](https://en.wikipedia.org/wiki/Meta_Platforms) (or were we to deadname it, facebook), [OpenAI](https://en.wikipedia.org/wiki/OpenAI), [Google](https://en.wikipedia.org/wiki/Alphabet_Inc.), and [Microsoft](https://en.wikipedia.org/wiki/Microsoft).
 
@@
 
+The variance of the model's output is substantially reduced: the output will, of necessity, be less varied, less diverse, and more norm-conforming:
 ```{r data-vs-prediction-variance}
@@
 
-### Copyright issues
+### Copyright (and privacy) issues
 
 LLM are known to regurgitate verbatim chunks of the training datasets.
-This is called _memorization_, another choice of words that tends to antropomorphize what is essentially a statistical model.
+This is called _memorization_, a word that contributes to antropomorphize what is essentially a statistical model.
 Memorization, in fact, is just a fancy way of saying "model overfitting"--in other words, an extreme regression to the mean that excels at predicting data in the training set, but is poor at projecting new data points.
-This issue can be as high 7% in the cases of some LLMs [@peng_near_duplicate_2023, p.
+Overfitting happens when a model with too many parameters is fitted to a relatively small data set (Models have [billions of parameters](https://en.wikipedia.org/wiki/Large_language_model#Training_cost) and are trained on [billions of pages of text](https://www.ibm.com/topics/large-language-models])).
+Data memorization--the verbatim or quasi-verbatim output of training data--can be as high as 7% in the case of some LLMs [@peng_near_duplicate_2023, p.
 8].
+
+Where will new data come from? [@murgia_why_2023]
+
+Note how several organizations have responded to the AIfication of all the things: by banning it.
+The concerns cited are seldom ethical or moral, but mostly about the risk of sensitive information leaking.
+As LLMs begin to be increasingly trained on model-produce data, overfitting (or memorization) can become a larger risk.
+
+And this is just a result of using an overfitted model. What about data leaks? Some stories regarding Microsoft's security track record:
+
+https://www.theregister.com/2024/04/12/microsoft_cisa_order/
+
+https://techreport.com/news/microsoft-azure-hit-with-the-largest-data-breach-in-its-history-hundreds-of-executive-accounts-compromised/
+
+Also, how confident can the public be about assurances that a big tech corporation will not data mine and commercialize data by the companies that sell the services of LLMs?
+Microsoft is so big that it can stand up to the US government:
+
+https://www.theregister.com/2024/04/05/microsoft_government_contracts/
+
+<!-- "Nice operation you have there...be a shame if your systems stopped working all of a sudden."-->
+
+Organization ranging from Free Open Source projects to elements of the federal government in the United States have limited/restricted the use of AI.
+
+NetBSD (a project that produces a free, Unix-like Open Source operating system) states in its commit guidelines (contributions to code) that "[c]ode generated by a large language model or similar technology...is presumed to be tainted code, and must not be committed without prior written approval by core." [@netbsd_netbsd_nodate].
+
+The National Archives and Records Administration (USA) bans use of ChatGPT [@koebler_national_2024]
+
+US Congress bans use of ChatGPT and Copilot for staff [@singh_after_2024].
+This article also note how companies like Samsung and Apple have restricted the use of chatbots citing concerns with data security.
 
@@
 
-ChatGPT consumes substantially more energy than a regular web search...are the results better? [@https://www.brusselstimes.com/1042696/chatgpt-consumes-25-times-more-energy-than-google]
+ChatGPT consumes substantially more energy than a regular web search...are the results better? [https://www.brusselstimes.com/1042696/chatgpt-consumes-25-times-more-energy-than-google]
 
@@
 
+This problem may seem trivial, even amusing in the case above.
+But what when an AI replicates the both-sides perspective of much of the media when trying to learn about an existential crisis like climate changes?
+[@gordon_multiple_2024].
+Or when trying to obtain accurate voting information in the context of voter supression?
+[@angwin_seeking_2024]
+
 ### Replication of unjust structures of power
@@
 On antropomorphizing GenAI tools.
-This happens in ways that are barely perceptively.
+This happens mostly in ways that are barely liminal.
 For example, in the document "Overview of AI at McMaster" we read that GenAI's somehow have an "ability to engage in nuanced conversation".
@@
 Doing this carelessly risks amplifying existing biases.
-The effort to wrap a statistical model in a human-like skin may actually be more devious, as in the new, sexy ChatGPT interface:
+But in other cases, the effort is more blatant.
+Wrapping a statistical model in a human-like skin may actually be more devious, as in the new, sexy ChatGPT interface:
 
@@
 
-**NOTES:** Need to research more the argument that UX is so important in the case of these tools because what they do well is trivial and what they do poorly is what is being hyped.
+**NOTES:** Need to research more the argument that the interface is so important in the case of these tools because what they do well is trivial and what they do poorly is what is being hyped.
+Found the article [@farrugia_known_2024]; see here: https://fasterandworse.com/known-purpose-and-trusted-potential/
 
@@
 
-ChatGPT was an awesome feat of marketing [@cowen_chatgpt_2023], but much coverage of the relevant issues in the media has been so credulous it has stretched credulity past Pluto [@burneko_if_2024]^[Bob woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called.].
-There are now calls to press pause on the hype [@angwin_opinion_2024] but these are still drown by the credulous coverage.
+ChatGPT was an awesome feat of marketing [@cowen_chatgpt_2023], but much coverage of the relevant issues in the media has been so credulous it has stretched credulity past Pluto [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called.].
+There are now calls to press pause on the hype [@angwin_press_2024] but these are still drown by the credulous coverage.
 
-I suggest that universities should avoid the mistakes of the media.
+I suggest that universities should avoid the mistakes of the media, an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) (also see the "pivot-to-video" and how tech companies suckered many media companies, including ).
 
@@
 Universities main source of power is their reputation as places that preserve and expand knowledge in a principled way.
-"This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an instution of higher-education is not needed to create spam and dissinformation.
+"This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an instution of higher-education is not needed to create spam and misinformation.
+
+"McMaster is best known for our pioneering work in problem-based learning, which began in our medical school in 1969 and has been adopted across the world." https://provost.mcmaster.ca/teaching-learning/
+
+Do we really want to be best known in 40 years as "McMaster, the institution that pioneered the use of ChatGPT"?
+
+Recall that [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean) is also called "reversion to mediocrity" [@barnett_regression_2017].
 

### 2024-05-21 README.Rmd
More edits

@@
 ```{r include=FALSE}
-library(dplyr)
-library(ggplot2)
+library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2'
+library(dplyr) # A Grammar of Data Manipulation
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(magick) # Advanced Graphics and Image-Processing in R
 ```
@@
 
-All opinions expressed here are mine and do not necessarily represent those of my employer(s).
+Disclaimner: _All opinions expressed here are mine and do not necessarily represent those of my employer(s)_.
 
-## Background: An AI is Haunting Campus^[Karl Marx wrote that "A spectre is haunting Europe--the spectre of communism"; in this case, AI is the shade of another spectre, that of unregulated surveillance capitalism.]
+## Background: An AI is Haunting Campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; in this case, GenAI is the shade of a very different spectre, that of unregulated surveillance capitalism.]
 
-On November 30, 2022, OpenAI released what has become its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models] that mimics natural language.
-The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and was rapidly panned as "the best artificial intelligence chatbot ever released to the general public" [@roose_brilliance_2023], and able to generate "impressively detailed" and "human-like" text [@lock_what_2022].
+On November 30, 2022, OpenAI released what has become its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language.
+The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels: by early December, ChatGPT was being panned as "the best artificial intelligence chatbot ever released to the general public" [@roose_brilliance_2023], and a tool able to generate "impressively detailed" and "human-like" text, leading some to wonder whether it could "replace humans" [@lock_what_2022].
 
-<!--impressive feat of marketing: think about what it does well, what it does poorly, and how the user interface is essential to obscure its uselessness.-->
+Universities, much like every other organization dealing with information, have had to respond in a compressed span of time to the rapid release of generative AI (GenAI) tools, particularly those by a handful of large American corporations including [Meta](https://en.wikipedia.org/wiki/Meta_Platforms) (or were we to deadname it, facebook), [OpenAI](https://en.wikipedia.org/wiki/OpenAI), [Google](https://en.wikipedia.org/wiki/Alphabet_Inc.), and [Microsoft](https://en.wikipedia.org/wiki/Microsoft).
 
-Universities have had to respond in a compressed span of time to the release of generative AI tools by a handful of large American corporations, notoriously [Meta](https://en.wikipedia.org/wiki/Meta_Platforms) (or were we to deadname it, facebook), [OpenAI](https://en.wikipedia.org/wiki/OpenAI), [Google](https://en.wikipedia.org/wiki/Alphabet_Inc.), and [Microsoft](https://en.wikipedia.org/wiki/Microsoft).
+The release of GenAIs to the masses was accompanied by numerous lofty promises, mostly by entities that had a stake in their adoption.
 
-McMaster developed a set of provisional guidelines (when where they developed and released?
-It is possible that this was done as a relatively urgent matter).
-These [guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) where released under a Creative Commons Licence and have since inspired those of other post-secondary institutions including Trent University [-@trent_university_generative_2024] and University of Guelph [-@office_of_teaching_and_learning_provisional_2023].
+The public was promised that GenAI would "accelerate sustainability" [@nakagawa_accelerating_2023]; that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" by "automating customer self-service, accelerating response times on an increasing number of tasks as well as providing greater accuracy, enhanced routing and intelligent context gathering" [@ibm_what_2023]; that AI-powered self-driving cars would reduce accidents, car ownership, reduce pollution and noise, automate logistics, and make parking tickets obsolete, if artificial intelligence company Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.] is to be believed [@joshi_5_2022; @garsten_what_2024].
+The list of promises has grown to encompass pretty much everything, from saving the planet all the way to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably are also using chatbots) [@khalatian_matchmaking_2023].
 
-There is now an effort underway to update the preliminary guidelines (to make the permanent?) McMaster [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
-This committee is composed of three sub-committees that examine the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in [our work] (the Provost's words)(probably meaning the university's work; not sure if the Provost themself uses AI).
+Higher-education was not spared the onslaught of promises.
+
+@hie_how_2023
+
+@power_3_2024
+
+It seemed inevitable that universities _had to do something_ in the face of all this innovative and perhaps only slightly scary technology.
+
+In response to the rapid, and disruptive, release of tools that were announced in revolutionary terms, McMaster developed a set of provisional [guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) (when where they developed and released?
+It is possible that this was done as a relatively urgent matter) to inform the university community about the use of AI in "operational excellence".
+These guidelines where released under a Creative Commons Licence and have since inspired those of other post-secondary institutions in Canada, including Trent University [-@trent_university_generative_2024] and the University of Guelph [-@office_of_teaching_and_learning_provisional_2023].
+
+There is now an effort underway to update the preliminary guidelines (to make them permanent?) and McMaster, via the Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
+This committee is composed of three sub-committees that examine the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work" (the Provost's words)(probably meaning the university's work; not sure if the Provost themself uses AI).
 
@@
 
-Hammers are powerful tools, sometimes even to conduct research or to teach.
+Acknowledging that AI are powerful tools, right off the bat, is in line with the hype observed so far.
+Hammers are powerful tools.
 Word processors are powerful tools.
+Lasers are powerful tools.
+And yet we do not have a high-level committee for each of them.
 What makes AI different from other tools that it requires a pan-university consultation on an "_ethical approach to shaping policies and developing new ways of using AI to ensure a responsible and beneficial integration where it makes sense_"?
 
-- Where does it make sense? Presumably this is what the Committee on AI will try to answer.
+The last statement carries a lot of weight here: where does adoption of AI make sense? Presumably this is what the Committee on AI will try to answer.
+
+<!--As I argue here, AI does not make a lot of sense outside of some limited applications that are nowhere near the scope of the promises of the merchants of hype.-->
 
@@
 
-To understand why many organizations have decided to spend considerable resources considering the adoption and responsible use of AI, it is essential to have a working understanding of what AI does differently from other tools.
+To understand why many organizations have decided to spend considerable resources considering the adoption and responsible use of GenAI, it is essential to have a working understanding of what AI does differently from other tools.
 
-Large Language Models are models composed of multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].
+GenAIs are  [@ibm_what_2021]. Large Language Models are models composed of multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].
 
@@
 
-Behind all the jargon, neural networks (LLMs are neural networks) are regression models [@ripley_neural_1994].
-Regression of course refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean).
-<!-- **NOTE**: also called "reversion to mediocrity".
--->
+Behind all the jargon, the neural networks that underpin GenAI are models designed to produce "statistically probable outputs" when prompted [@ibm_what_2021].
+In more conventional terms, neural networks are regression models [@ripley_neural_1994]--much more sophisticated and powerful than linear regression, but regression models nonetheless.
+Here, it is important to note that the term regression, of course, refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing a prompt to a "statistically probable" outcome.
 
@@
 
-ChatGPT consumes substantially more energy than a regular web search...are the results better? [https://www.brusselstimes.com/1042696/chatgpt-consumes-25-times-more-energy-than-google]
+"Environmental costs of AI are soaring" [@crawford_generative_2024]
+
+ChatGPT consumes substantially more energy than a regular web search, up to 25 times more, and according to some estimates, the whole of AI may consume twice as much energy as the whole of France by 2030 [@brussels_chatgpt_2024].
+
+...are the results better?
 
@@
 
-![A search using Duck Duck Go](images/logical-puzzle-web-search.png){width=50%}
+```{r puzzle-web-search, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}A search using Duck Duck Go"}
+ggdraw() + 
+  draw_image(image = "images/logical-puzzle-web-search.png")
+```
 
-![A conversation with ChatGPT-4](images/logical-puzzle-chat-gpt-4.png){width=50%}
+```{r puzzle-chatgpt, out.width="50%", fig.cap="\\label{fig:puzzle-chatgpt}A conversation with ChatGPT-4"}
+puzzle_plot <- ggplot() +
+  labs(caption = "https://assemblag.es/@hugo/112450401657084750") +
+  theme_void()
 
-https://assemblag.es/@hugo/112450401657084750
+ggdraw() + 
+  draw_plot(puzzle_plot) +
+  draw_image(image = "images/logical-puzzle-chat-gpt-4.png")
 
-<!--
-<iframe src="https://assemblag.es/@hugo/112450401657084750/embed" width="400" allowfullscreen="allowfullscreen" sandbox="allow-scripts allow-same-origin allow-popups allow-popups-to-escape-sandbox allow-forms"></iframe>
- -->
+```
 
-This problem may seem trivial, even amusing in the case above.
+The example above may seem trivial, even amusing in the case above.It was not so funny for
+
 But what when an AI replicates the both-sides perspective of much of the media when trying to learn about an existential crisis like climate changes?
@@
 
-Potential for weaponization of GenAI. **Farfetched (?) idea** Could it be that GenAI will be at some point declared a military-grade technology with restricted applications?
+Potential for weaponization of GenAI.
+**Farfetched (?) idea** Could it be that GenAI will be at some point declared a military-grade technology with restricted applications?
+(see https://www.defenseone.com/technology/2024/05/ai-and-other-tech-creating-special-operations-renaissance/396388/)
 

### 2024-05-22 README.Rmd
Still trying to pull a bunch of separate strands to obtain a coherent narrative

@@
 
-@hie_how_2023
+@power_3_2024,^[Rhett Power is a motivational speaker and executive coach] for example, presents three innovative use cases for AI in higher education.
+First, AI can take over manual tasks and improve the budget by releasing administrative staff.
+AI can evaluate applications for admissions, including traits like the "grit" and "empathy" of applicants.
+AI can personalize the experience of consumers (also called students) by predicting their final grade with 60-70% accuracy, and by providing corrective and instructive feedback.
+<!--release them in a very literal sense, as one advantage is to do _more_ with limited resources (which in the current fiscal environment means laying people off).
+The cases discussed by Mr.
+Power are questionable: there are already non-AI systems that deploy paperwork; the kind of information needed to identify students at risk of dropping out is probably quite invasive, and therefore impractical or unethical to collect; I very much doubt that it is legal to use AI in hiring and admissions.
+--> Mr.
+Power's message to the higher-eduation sector?
+It is time for institutions to level up: jump on the bandwagon, fellows, don't let it let you behind!
 
-@power_3_2024
+@hie_how_2023 argue that AI will likely be the primary way to access information in the future.
+For this reason, institutions must equip students with the skills needed to use AI effectively, in particular to perfect the art of _prompt engineering_, that is, the ability to craft queries that prompt an AI to provide the most useful outputs.
+Using AIs may improve learning and challenge the critical thinking of students.
+Faculty will also benefit from AI: it will be used to evaluate their students<!-- is this even legal?
+-->, and will handle repetitive tasks like writing course syllabi.
+@hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school).
+Writing in 2022, Mr.
+Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9).
+In October 2023 facebook [slashed its metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/).
+Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team four months after launching it.
+Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence".] advice that "AI is here to stay", that "[faculty] might not be able to stop the AI trend from growing", so they might as well get on with the program and try to shape it as best they can.
 
-It seemed inevitable that universities _had to do something_ in the face of all this innovative and perhaps only slightly scary technology.
+It seemed inevitable that universities _had to do something_ in the face of all this unavoidable, innovative technology that many feared might leave them behind.
 
-In response to the rapid, and disruptive, release of tools that were announced in revolutionary terms, McMaster developed a set of provisional [guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) (when where they developed and released?
-It is possible that this was done as a relatively urgent matter) to inform the university community about the use of AI in "operational excellence".
-These guidelines where released under a Creative Commons Licence and have since inspired those of other post-secondary institutions in Canada, including Trent University [-@trent_university_generative_2024] and the University of Guelph [-@office_of_teaching_and_learning_provisional_2023].
+The fear of missing out, combined with a modicum of concern for the ethical implications, has led to institutional responses that have been less than coherent.
+An example is the case of Emory University, where the university gave a $10,000 prize to a group of students participating in an Entrepreneurship Summit, only for the the students to be suspended later by the school's Honor Council for the potential of their tool for cheating^[There was no evidence that cheating had happened.] [@koebler__university_2024].
 
-There is now an effort underway to update the preliminary guidelines (to make them permanent?) and McMaster, via the Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
+For its part, McMaster was a leader in responding to the rapid, and disruptive, release of _revolutionary_ tools, and its provisional [guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) (released under a Creative Commons Licence) have since inspired those of other post-secondary institutions in Canada, including Trent University [-@trent_university_generative_2024] and the University of Guelph [-@office_of_teaching_and_learning_provisional_2023].
+
+The aim of the provisional guidelines was to inform the university community about the use of AI in "operational excellence".
+But the time has come to update them, and the institution, via its Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
 This committee is composed of three sub-committees that examine the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work" (the Provost's words)(probably meaning the university's work; not sure if the Provost themself uses AI).
@@
 
-De-duplication of training data improves LLMs [@lee2022deduplicating]
-
-No wonder: it amounts to having more different information.
+De-duplication of training data improves LLMs [@lee2022deduplicating]. No wonder: it amounts to having more different information.
 
@@
 
-### Use of materials on the web
+### Extractive
 
 In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )
+
+@germain_google_2023
+@norem_intels_2023
+
+### Exploitative
+
+Above, the ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted.
+But AI companies do not spare the rich either.
+OpenAI's Sam Altman, a self-avowed admirer of the movie _Her_, featuring Scarlet Johansson as an AI voice assistant, reportedly tried to hire Ms.
+Johannson to be the voice of ChatGPT 4.0 [@kastrenakes_scarlett_2024].
+After Ms.
+Johannson decline, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways, which was retired after Ms.
+Johansson threatened legal action^[Altman's denials that OpenAI had not used Ms.
+Johansson's voice sounded hollow, after he [posted](https://x.com/sama/status/1790075827666796666) on social media the word "her" at the time the new version of ChatGPT was announced.] [@allyn_scarlett_2024].
+Alas, very few have the social standing of Scarlett Johansson to challenge OpenAI.
+Most people in the world lack that power.
+
+@williams_exploited_2022 (Summary needed)
+
+@perrigo_2_2023 (Follows summary)
+
+ChatGPT was trained on vast amounts of data scooped from the internet, where many societal problems are amplified.
+Trained on horrible content, it was hard to sell ChatGPT when it was prone to racist, sexist, obscene outbursts.
+But companies like facebook, who did not want to pay for moderation, had hit on a solution of sorts <!--: train an AI to detect content that was unpalatable.-->
+
+The solution?
+
+> [F]eed an AI with labeled examples of violence, hate speech, and sexual abuse, and that tool could learn to detect those forms of toxicity in the wild.
+That detector would be built into ChatGPT to check whether it was echoing the toxicity of its training data, and filter it out before it ever reached the user.
+It could also help scrub toxic text from the training datasets of future AI models.
+
+The key here is "labeled examples".
+What are those labels?
+Many machine algorithms do not learn by themselves...they learn from humans, and need what is called "labelled" data: a images that humans have labelled as "dog", "cats", etc.
+Or bits of text that have been labelled as...well...questionable.
+Labelling data is extremely time consuming and labor intensive.
+What's a corporation to do?
+
+> To get those labels, OpenAI sent tens of thousands of snippets of text to an outsourcing firm in Kenya, beginning in November 2021.
+Much of that text appeared to have been pulled from the darkest recesses of the internet.
+Some of it described situations in graphic detail like child sexual abuse, bestiality, murder, suicide, torture, self harm, and incest.
+"
+
+What did this mean to the people doing all this labeling in one of the poorest regions in the world?
+First of all, an income: "data labelers employed...on behalf of OpenAI were paid a take-home wage of between around $1.32 and $2 per hour depending on seniority and performance." But at what cost?
+
+> One Sama worker tasked with reading and labeling text for OpenAI told TIME he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child.
+“That was torture,” he said.
+“You will read a number of statements like that all through the week.
+By the time it gets to Friday, you are disturbed from thinking through that picture."
+
+@hao_hidden_2023
+
+Quotes from Hao's podcast:
+
+> Karen Hao: One of the reasons why ChatGPT was able to become so virally popular and continued to sustain popularity is because it is largely not spewing really awful things.
+People feel comfortable using the product knowing that it's not going to do that.
+
+> Annie Minoff: At least it won't do it in English. If Alex wanted to use ChatGPT in his native language, Swahili, would he be able to do that?
+
+> Karen Hao: You can interact with ChatGPT in Swahili, but ChatGPT was developed primarily to work in English.
+So a lot of the scrubbing, the content moderation, the important safety measures within the chatbot were done in English.
+So when you prompt it in Swahili, you'll get more misinformation.
+You'll get more confusing sentences that don't make sense, and you will potentially get more of this content that they worked so hard to filter out because they were only filtering it in English.
+
+Think about linguistic hegemony, and how the models are only as good (or as bad) as their inputs.
 
@@
 
-The example above may seem trivial, even amusing in the case above.It was not so funny for
+The example above may seem like an amusement, but a GenAI is not a toy, as Air Canada discover when the company was found liable for "negligent misrepresentation", after a court denied claims by the airline that the chatbot was somehow responsible for its own actions as an effectively separate legal entity [@garcia_what_2024].
+
+Similar GenAI-generated mishaps abound.
 
@@
 
-## As an aside
+## As an aside {#as-an-aside}
 
@@
 For example, in the document "Overview of AI at McMaster" we read that GenAI's somehow have an "ability to engage in nuanced conversation".
-This phrasing could be simply a literary device, but even so is indicative of a tendency to ascribe human attributes to a statistical model.
-Doing this carelessly risks amplifying existing biases.
+This phrasing could be simply a literary device, but even so it is indicative of a tendency to ascribe human attributes to a statistical model.
+Doing this risks amplifying existing biases.
 But in other cases, the effort is more blatant.
@@
 Universities main source of power is their reputation as places that preserve and expand knowledge in a principled way.
-"This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an instution of higher-education is not needed to create spam and misinformation.
+"This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an institution of higher-education is not needed when there are cheaper alternatives for such an education ("act now before you are laid off!"):
+
+```{r ligemy-ai-bundle, out.width="50%", fig.cap="\\label{fig:ai-success-bundle}An AI success bundle from the \"The Real People of Ligency PR and Marketing Team\""}
+bundle_plot <- ggplot() +
+  labs(caption = "https://www.udemy.com/user/ligency-team/") +
+  theme_void()
+
+ggdraw() + 
+  draw_plot(bundle_plot) +
+  draw_image(image = "images/trail-blazing-ai-bundle.png")
+```
 

### 2024-05-26 README.Rmd
Testing textbox

@@
 
+```{css textbox}
+.blackbox {
+  padding: 1em;
+  background: black;
+  color: white;
+  border: 2px solid orange;
+  border-radius: 10px;
+}
+.center {
+  text-align: center;
+}
+```
+
 ```{r include=FALSE}
@@
 library(magick) # Advanced Graphics and Image-Processing in R
+library(neuralnet) # Training of Neural Networks
 ```
@@
 
+By: Antonio Páez (paezha@mcmaster.ca)
+
+Disclaimer: _All opinions expressed here are mine and any resemblance to those of my employer(s) is purely coincidental_.
+
+## Background: An AI is Haunting Campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the shade of the spectre of unregulated surveillance capitalism.]
+
 > Mission Statement: Our organization seeks to level up on all matters related.  
@@
 
-By: Antonio Paez
+:::: {.blackbox data-latex=""}
+::: {.center data-latex=""}
+**NOTICE!**
+:::
 
-Disclaimner: _All opinions expressed here are mine and do not necessarily represent those of my employer(s)_.
+Thank you for noticing this **new notice**! Your noticing it has
+been noted, and _will be reported to the authorities_!
+::::
 
-## Background: An AI is Haunting Campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; in this case, GenAI is the shade of a very different spectre, that of unregulated surveillance capitalism.]
-
-On November 30, 2022, OpenAI released what has become its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language.
+On November 30, 2022, OpenAI released what became its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language.
 The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels: by early December, ChatGPT was being panned as "the best artificial intelligence chatbot ever released to the general public" [@roose_brilliance_2023], and a tool able to generate "impressively detailed" and "human-like" text, leading some to wonder whether it could "replace humans" [@lock_what_2022].
 
-Universities, much like every other organization dealing with information, have had to respond in a compressed span of time to the rapid release of generative AI (GenAI) tools, particularly those by a handful of large American corporations including [Meta](https://en.wikipedia.org/wiki/Meta_Platforms) (or were we to deadname it, facebook), [OpenAI](https://en.wikipedia.org/wiki/OpenAI), [Google](https://en.wikipedia.org/wiki/Alphabet_Inc.), and [Microsoft](https://en.wikipedia.org/wiki/Microsoft).
+Universities, much like every other organization dealing with information, had to respond in a compressed span of time to the rapid release of generative AI (GenAI) tools like ChatGPT, most of them created and unleashed in the wild by a handful of large American corporations including [Meta](https://en.wikipedia.org/wiki/Meta_Platforms) (or were we to deadname it, facebook), [OpenAI](https://en.wikipedia.org/wiki/OpenAI), [Google](https://en.wikipedia.org/wiki/Alphabet_Inc.), and [Microsoft](https://en.wikipedia.org/wiki/Microsoft).
 
-The release of GenAIs to the masses was accompanied by numerous lofty promises, mostly by entities that had a stake in their adoption.
+GenAI was announced as nothing short of revolutionary.
+McMaster, for example, [states](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/):
 
-The public was promised that GenAI would "accelerate sustainability" [@nakagawa_accelerating_2023]; that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" by "automating customer self-service, accelerating response times on an increasing number of tasks as well as providing greater accuracy, enhanced routing and intelligent context gathering" [@ibm_what_2023]; that AI-powered self-driving cars would reduce accidents, car ownership, reduce pollution and noise, automate logistics, and make parking tickets obsolete, if artificial intelligence company Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.] is to be believed [@joshi_5_2022; @garsten_what_2024].
-The list of promises has grown to encompass pretty much everything, from saving the planet all the way to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably are also using chatbots) [@khalatian_matchmaking_2023].
+> Understanding how generative AI works, and how it can support work within organizations represents an ongoing area of exploration and innovation.
+The opportunities of generative AI are exciting: creating workplace efficiencies to allow for different kinds of work, adding capabilities for individuals and teams, and offering personalized uses.
+
+AI was quickly perceived as an _innovative_ and _exciting_ technology, that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p.
+64].].
+It seemed inevitable that universities _had to do something_ or risk being let behind.
+
+The fear of missing out, combined with a modicum of concern for the ethical implications of GenAIs^[Ethical concerns for the most part have had to do with policing academic dishonesty among students], has led to institutional responses that have been less than coherent.
+An example is the case of Emory University, where the university gave a $10,000 prize to a group of students who proposed developing an AI tool for an Entrepreneurship Summit, only for the students to be subsequently suspended by the school's Honor Council for the potential of their tool for cheating^[There was no evidence that cheating had happened.] [@koebler__university_2024].
+
+For its part, McMaster was a leader in responding to the rapid, and disruptive, release of these _revolutionary_ tools, and its provisional [guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) (released under a Creative Commons Licence) have since inspired those of other post-secondary institutions in Canada, including the University of Guelph [-@office_of_teaching_and_learning_provisional_2023] and Trent University [-@trent_university_generative_2024].
+
+The aim of the provisional guidelines was to inform the university community about the use of AI in "operational excellence".
+But the time has come to update the guidelines, and the institution, via its Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
+This committee is composed of three sub-committees that examine the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work" (the Provost's words)(probably meaning the university's work; not sure if the Provost themself uses AI).
+
+The sub-committees, in turn, convened panels to:
+
+1.
+[Provide recommendations]({#ai-panel-tasks}) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/)
+
+2. And also to do (what? consultation?).
+
+3. Q: What are the terms of reference of the panels? Who will decide which recommendations are adopted and which are ignored?
+
+According to the announcement of this AI Committee at McMaster:
+
+> AI is a powerful tool and I believe that universities have an important role to play in its adoption.  
+> -- Susan Tighe, Provost
+
+Accepting that AIs are powerful tools, right off the bat, is consistent with the hype observed so far [@angwin_press_2024].
+Hammers are powerful tools.
+Word processors are powerful tools.
+Lasers are powerful tools.
+And yet we do not have high-level committees for each of and every tool that the university uses.
+What makes AIs different from other tools that they require a pan-university consultation on an "_ethical approach to shaping policies and developing new ways of using AI to ensure a responsible and beneficial integration where it makes sense_"?
+
+(The last sentence in the statement above carries quite a bit of weight here: where does adoption of AI make sense? Presumably this is what the Committee on AI will try to answer.)
+
+<!--As I argue here, AI does not make a lot of sense outside of some limited applications that are nowhere near the scope of the promises of the merchants of hype.-->
+
+To understand why many organizations, including McMaster University, have decided to spend considerable resources studying the adoption and responsible use of AI, it is essential to have a working understanding of what AI does that sets it apart from other tools.
+
+In doing this, we must be alert to the fact that technologies are never politically neutral: it is important to ask two questions:
+
+- What does a specific technology do for whom?
+
+- And what does it do to whom?
+
+We also need to understand the goals of the organization that considers the technology.
+When does it make sense for a university like McMaster to "integrate" AI as part of the core mission of the organization, accounting for what it does to some people and what it does for others?
+
+Let us begin with the mission of McMaster as an institution of higher education.
+
+## The [mission](https://president.mcmaster.ca/mission-vision/)
+
+McMaster's mission is "the discovery, communication and preservation of knowledge," accomplished with a commitment to "creativity, innovation and excellence".
+These aspirations must respond to some underlying values, and indeed, the university explicitly values "integrity, quality, inclusiveness and teamwork".
+Ultimately, the university aims to "inspire critical thinking, personal growth, and a passion for lifelong learning" and to serve "the social, cultural, and economic needs of our community and our society."
+
+A mission statement is like an onion, and one needs to peel multiple layers to try to understand the core principles enunciated there.
+Why are things like "integrity" and "critical thought" valued?
+Or, what happens when "integrity", "quality", and/or "excellence" clash with "the economic needs of our society" (say, a mandate by democratically elected governments to spend as little as possible in post-secondary education^[Between 2000-2020, Ontario had the lowest spending on post-secondary education of all provinces in Canada, relative to its share of spending [@caut_provincial].
+Public funding of universities in Ontario as percent of revenue fell from 45.6% in 2001/2 to 35.5% in 2019/20 [@statistics_canada_college_2022].
+It is little wonder that universities are seeking "efficiencies"])?
+
+Clearly, there will be some tension implicit in the mission of a university.
+But being generous, let us take the mission statement at face value and assume that the highest priorities are those related to knowledge obtained, preserved, and communicated with integrity and quality.
+
+## What do AIs do?
+
+At this point, we might as well ask "what do AIs _not_ do"?
+
+A possible reason why many organizations have decided to spend substantial resources considering the adoption of GenAI is that, unlike most other tools, they are seen as _everything tools_.
+Indeed, the release of GenAIs to the masses was accompanied by innumerable lofty promises, albeit most of them made by entities that had a stake in the adoption of the technology.
+
+The public was promised that GenAI would "accelerate sustainability" [@nakagawa_accelerating_2023]; that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" by "automating customer self-service, accelerating response times on an increasing number of tasks as well as providing greater accuracy, enhanced routing and intelligent context gathering" [@ibm_what_2023]; according to Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.], AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024].
+In mental health care, chatbots could complement human therapists "via comprehensive data access and [analysis of] behavioral patterns" to "mimic practitioner questions and subsequently make recommendations based on a user’s inputs" [@silva_4_2023].
+
+The list of promises has grown to encompass pretty much everything, from saving the planet all the way to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023].
 
@@
 @power_3_2024,^[Rhett Power is a motivational speaker and executive coach] for example, presents three innovative use cases for AI in higher education.
-First, AI can take over manual tasks and improve the budget by releasing administrative staff.
-AI can evaluate applications for admissions, including traits like the "grit" and "empathy" of applicants.
-AI can personalize the experience of consumers (also called students) by predicting their final grade with 60-70% accuracy, and by providing corrective and instructive feedback.
+First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies"].
+AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants.
+AI could also personalize the experience of consumers (also called students) by predicting their final grade with 60-70% accuracy, and by providing corrective and instructive feedback.
 <!--release them in a very literal sense, as one advantage is to do _more_ with limited resources (which in the current fiscal environment means laying people off).
@@
 --> Mr.
-Power's message to the higher-eduation sector?
-It is time for institutions to level up: jump on the bandwagon, fellows, don't let it let you behind!
+Power's message to the higher-education sector?
+It is time for institutions to "level up": jump on the bandwagon, fellas, don't let it let you behind!
 
-@hie_how_2023 argue that AI will likely be the primary way to access information in the future.
-For this reason, institutions must equip students with the skills needed to use AI effectively, in particular to perfect the art of _prompt engineering_, that is, the ability to craft queries that prompt an AI to provide the most useful outputs.
-Using AIs may improve learning and challenge the critical thinking of students.
-Faculty will also benefit from AI: it will be used to evaluate their students<!-- is this even legal?
--->, and will handle repetitive tasks like writing course syllabi.
+In another instance, @hie_how_2023 argue that AIs will likely be the primary way to access information in the future.
+For this reason, institutions must equip students with the skills needed to use AI competently, in particular to perfect the art of _prompt engineering_, that is, the ability to craft effective queries that prompt a GenAI to provide the most useful outputs.
+Universities will use AIs to improve learning and challenge the critical thinking of students.
+Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been essential to understand my own courses; automating the task is another way of cheapening the quality of the learning experience for students].
 @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school).
@@
 Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team four months after launching it.
-Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence".] advice that "AI is here to stay", that "[faculty] might not be able to stop the AI trend from growing", so they might as well get on with the program and try to shape it as best they can.
+Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence".] warn that "AI is here to stay", that "[faculty] might not be able to stop the AI trend from growing", and so they might as well get on with the program and try to shape it as best they can.
 
-It seemed inevitable that universities _had to do something_ in the face of all this unavoidable, innovative technology that many feared might leave them behind.
+What does McMaster hope to achieve with AI? From the Overview on Generative AI document:
 
-The fear of missing out, combined with a modicum of concern for the ethical implications, has led to institutional responses that have been less than coherent.
-An example is the case of Emory University, where the university gave a $10,000 prize to a group of students participating in an Entrepreneurship Summit, only for the the students to be suspended later by the school's Honor Council for the potential of their tool for cheating^[There was no evidence that cheating had happened.] [@koebler__university_2024].
-
-For its part, McMaster was a leader in responding to the rapid, and disruptive, release of _revolutionary_ tools, and its provisional [guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) (released under a Creative Commons Licence) have since inspired those of other post-secondary institutions in Canada, including Trent University [-@trent_university_generative_2024] and the University of Guelph [-@office_of_teaching_and_learning_provisional_2023].
-
-The aim of the provisional guidelines was to inform the university community about the use of AI in "operational excellence".
-But the time has come to update them, and the institution, via its Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
-This committee is composed of three sub-committees that examine the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work" (the Provost's words)(probably meaning the university's work; not sure if the Provost themself uses AI).
-
-The sub-committees, in turn, convened panels to:
-
-1.
-Provide [provide recommendations]({#ai-panel-tasks}) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/)
-
-2. And also to do (what? consultation?).
-
-3. Q: What are the terms of reference of the panels? Who will decide which recommendations are adopted and which are ignored?
-
-According to the announcement of the AI Committee:
-
-> AI is a powerful tool and I believe that universities have an important role to play in its adoption.  
-> -- Susan Tighe, Provost, McMaster University
-
-Acknowledging that AI are powerful tools, right off the bat, is in line with the hype observed so far.
-Hammers are powerful tools.
-Word processors are powerful tools.
-Lasers are powerful tools.
-And yet we do not have a high-level committee for each of them.
-What makes AI different from other tools that it requires a pan-university consultation on an "_ethical approach to shaping policies and developing new ways of using AI to ensure a responsible and beneficial integration where it makes sense_"?
-
-The last statement carries a lot of weight here: where does adoption of AI make sense? Presumably this is what the Committee on AI will try to answer.
-
-<!--As I argue here, AI does not make a lot of sense outside of some limited applications that are nowhere near the scope of the promises of the merchants of hype.-->
-
-To understand why many organizations, including McMaster University, have decided to spend considerable resources considering the adoption and responsible use of AI, it is essential to have a working understanding of what AI does differently from other tools.
-We also need to understand what are the goals of the organization.
-When does it make sense for a university like McMaster to "integrate" AI (pressumably as part of the core mission of the organization)?
-
-More generally, we must be alert to the fact that technologies are never politically neutral: it is important to ask two questions:
-
-- What does technology X do for whom?
-
-- And what does it do to whom?
-
-**McMaster's Mission**
-
-https://president.mcmaster.ca/mission-vision/
-
-The mission of the university is "the discovery, communication and preservation of knowledge," accomplished with a commitment to "creativity, innovation and excellence".
-These are aspirations that must respond to some underlying values, and indeed, the university values "integrity, quality, inclusiveness and teamwork".
-Ultimately, the university aims to "inspire critical thinking, personal growth, and a passion for lifelong learning" and to serve "the social, cultural, and economic needs of our community and our society."
-
-From the Overview on Generative AI document:
-
-> Some of those uses in our context could include qualitative and quantitative data analysis, writing text (e.g.
+> Some of [the uses of AI] in our context could include qualitative and quantitative data analysis, writing text (e.g.
 reference letters, grant applications, job descriptions, report summaries), translation, captioning, multiple formats for learning and engagement, and personalized support.
 
-## What are LLMs?
+## But really, what do AIs do?
 
-To understand why many organizations have decided to spend considerable resources considering the adoption and responsible use of GenAI, it is essential to have a working understanding of what AI does differently from other tools.
+GenAIs are _deep learning models_ [@ibm_what_2021], so called because they are neural networks with _depth_, meaning many, many hidden layers.
+As an example, Large Language Models (such as those behind OpenAI's ChatGPT, Microsoft's Copilot, Google's Gemini, and facebook's LLaMA) are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].
 
-GenAIs are  [@ibm_what_2021]. Large Language Models are models composed of multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].
+But, what are [neural](https://en.wikipedia.org/wiki/Large_language_model#Further_reading) [networks](https://medium.com/@asifurrahmanaust/lesson-3-neural-network-is-nothing-but-a-linear-regression-e05a328a0f23)?
+
+Behind all the jargon, the neural networks that underpin GenAI are models designed to produce "statistically probable outputs" when prompted [@ibm_what_2021].
+In more conventional terms, neural networks are regression models [@ripley_neural_1994]--much more sophisticated and powerful than linear regression to be sure, but regression models nonetheless.
+Here, it is important to note that the term regression refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing a prompt to a "statistically probable" outcome.
+
+A regression model takes data as inputs, typically a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable).
+With these inputs, the model finds a conditional mean for the output $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables).
+The process of finding that conditional mean is to satisfy some critarion (e.g., that the conditional mean minimizes the error of the model--the distance between $y$ and $\hat{y}$).
+
+Below is an example of the simplest regression model with only one independent variable (i.e., a bivariate model).
+The plot shows the pairs of $y$ and $x$ values, and the line is the model, the conditional mean $\hat{y}$.
+This model needs only two parameters: an intercept (the value of &\hat{y}$ when $x = 0$), and the slope of the line, in other words, the rate of change of $\hat{y}$ with respect to $x$.
+Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.
+```{r simulate-simple-example}
+set.seed(13253)
+
+b0 <- 1.0
+b1 <- 1.5
+
+df <- data.frame(x = b0 + b1 * runif(100, 
+                                0, 
+                                10),
+                 e = rnorm(100),
+                 type = "training data") |>
+  mutate(y = x + e)
+
+x <- df |> pull(x)
+y <- df |> pull(y)
+
+mod <- lm(y ~ x, 
+          data = df)
+```
+
+```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Simple Example of a Regression Model"}
+ggplot(df,
+       aes(x = x,
+           y = y)) +
+  geom_point() +
+  geom_smooth(method = "lm") + 
+  theme_minimal()
+```
+
+Predict:
+```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Training data and model output compared"}
+
+df_pred <- data.frame(x = x)
+
+y_hat <- predict(mod, df_pred)
+
+df_pred_lr <- df_pred |> 
+  mutate(y = y_hat,
+         type = "regression output")
+
+df_pred_all <- rbind(df |>
+                   select(x, y, type),
+                 df_pred_lr |>
+                   select(x, y, type))
+
+ggplot() + 
+  geom_point(data = df_pred_all,
+  aes(x = x,
+      y = y,
+      color = type,
+      shape = type,
+      size = type)) +
+  scale_color_manual(values = c("training data" = "black",
+                                "regression output" = "blue",
+                                "nn1 output" = "chocolate1",
+                                "nn2 output" = "brown")) +
+  scale_shape_manual(values = c("training data" = 16,
+                                "regression output" = 17,
+                                "nn1 output" = 18,
+                                "nn2 output" = 19)) +
+  scale_size_manual(values = c("training data" = 1,
+                                "regression output" = 2,
+                                "nn1 output" = 2,
+                                "nn2 output" = 2)) +
+  theme_minimal()
+```
+
+Neural networks do something similar but using a net-like architecture that is more flexible at identifying the characteristics of the data.
+For instance, the following figure shows a relatively simple neural network trained using using the same data.
+This network is not "deep", as it has only one so called "hidden layer", comprised of two neurons between the input and the output layers.
+In comparison to the linear regression and its two parameters, this neural network uses seven parameters.
+
+```{r}
+# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
+f <- as.formula("y ~ x")
+```
+
+```{r train-shallow-neural-network}
+# Train a neural network using only two neurons in one hidden layer.
+The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node.
+The default activation function is the logistic function.
+set.seed(33225433)
+
+nn1 <- neuralnet(f, 
+                data = df, 
+                hidden = c(2), 
+                linear.output = T) 
+```
+
+```{r save-shallow-neural-network}
+# For some reason, the figure cannot be saved directly, so I saved it manually using screen capture
+plot(nn1)
+```
+
+```{r plot-shallow-neural-network, out.width="50%", fig.cap="Simple Example of a Shallow Neural Network"}
+knitr::include_graphics("images/nn1.png")
+```
+
+```{r shallow-neural-network-predictions}
+y_hat <- predict(nn1,
+                 df_pred)
+
+df_pred_nn1 <- df_pred |> 
+  mutate(y = y_hat,
+         type = "nn1 output")
+
+df_pred_all <- rbind(df_pred_all |>
+                   select(x, y, type),
+                 df_pred_nn1 |>
+                   select(x, y, type))
+```
+
+The plot below illustrates the predictions of this neural network, where we can see that it is not limited to a linear response; neural networks are more flexible than linear regression, and they achieve this by being less parsimonious (they use more parameters).
+
+```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Training data and model outputs compared: linear regression and shallow neural network"}
+ggplot() + 
+  geom_point(data = df_pred_all,
+  aes(x = x,
+      y = y,
+      color = type,
+      shape = type,
+      size = type)) +
+  scale_color_manual(values = c("training data" = "black",
+                                "regression output" = "blue",
+                                "nn1 output" = "chocolate1",
+                                "nn2 output" = "brown")) +
+  scale_shape_manual(values = c("training data" = 16,
+                                "regression output" = 17,
+                                "nn1 output" = 18,
+                                "nn2 output" = 19)) +
+  scale_size_manual(values = c("training data" = 1,
+                                "regression output" = 2,
+                                "nn1 output" = 2,
+                                "nn2 output" = 2)) +
+  theme_minimal()
+```
+
+More complex neural network can be designed.
+The following plot is of a deeper network than the previous one, with two hidden layers, each consisting of five neurons.
+This model uses forty-six parameters.
+```{r train-deeper-neural-network}
+# Train a neural network using only two neurons in one hidden layer.
+The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node.
+The default activation function is the logistic function.
+set.seed(242)
+nn2 <- neuralnet(f, 
+                data = df, 
+                hidden = c(5, 5), 
+                linear.output = T)
+```
+
+```{r save-plot-deeper-neural-network}
+# For some reason, the figure cannot be saved directly, so I saved it manually using screen capture
+plot(nn2)
+```
+```{r plot-deeper-neural-network, out.width="50%", fig.cap="Example of a Deeper Neural Network"}
+knitr::include_graphics("images/nn2.png")
+```
+
+```{r}
+y_hat <- predict(nn2,
+                 df_pred)
+
+df_pred_nn2 <- df_pred |> 
+  mutate(y = y_hat,
+         type = "nn2 output")
+
+df_pred_all <- rbind(df_pred_all |>
+                   select(x, y, type),
+                 df_pred_nn2 |>
+                   select(x, y, type))
+
+```
+
+The increased complexity leads to greater flexibility, and we can see that the model makes predictions that resemble more closely the actual data.
+
+```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Training data and model outputs compared: shallow neural network and deeper neural network"}
+ggplot() + 
+  geom_point(data = df_pred_all |>
+               filter(type != "regression output"),
+  aes(x = x,
+      y = y,
+      color = type,
+      shape = type,
+      size = type)) +
+  scale_color_manual(values = c("training data" = "black",
+                                "regression output" = "blue",
+                                "nn1 output" = "chocolate1",
+                                "nn2 output" = "brown")) +
+  scale_shape_manual(values = c("training data" = 16,
+                                "regression output" = 17,
+                                "nn1 output" = 18,
+                                "nn2 output" = 19)) +
+  scale_size_manual(values = c("training data" = 1,
+                                "regression output" = 2,
+                                "nn1 output" = 2,
+                                "nn2 output" = 2)) +
+  theme_minimal()
+```
+
+```{r}
+# How do these models perform?
+mad_lr <- data.frame(df |>
+             mutate(y_hat = df_pred_lr$y)) |>
+  summarize(mse = sum(abs(y - y_hat)/n()))
+
+mad_nn1 <- data.frame(df |>
+             mutate(y_hat = df_pred_nn1$y)) |>
+  summarize(mad = sum(abs(y - y_hat)/n()))
+
+mad_nn2 <- data.frame(df |>
+             mutate(y_hat = df_pred_nn2$y)) |>
+  summarize(mad = sum(abs(y - y_hat)/n()))
+
+delta_mad_nn1 <- (mad_nn1 - mad_lr)/mad_lr * 100
+
+delta_mad_nn2 <- (mad_nn2 - mad_lr)/mad_lr * 100
+```
+
+Indeed, the first neural network is "better" in that it reduces the typical error by `r paste0(prettyNum(delta_mad_nn2, digits = 2), "%")` compared to the linear regression, whereas the deeper network reduces the typical error by `r paste0(prettyNum(delta_mad_nn2, digits = 2), "%")`, also with respect to the linear regression^[The "typical error" here is the mean absolute deviation of the predictions.].
+
+There is a limit to how flexible any model can be, and this is as true of linear regression as it is of neural networks.
+Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than we have data to train the model.
+Roughly, every additional parameter needs a data input to train on, and when there are as many parameters as data inputs the model is overfit and will predict each value of the independent variable in the training dataset perfectly.
 
@@
 -->
-
-What are neural networks? Here are some resources to learn some elemental facts about neural networks.
-
-https://medium.com/@asifurrahmanaust/lesson-3-neural-network-is-nothing-but-a-linear-regression-e05a328a0f23
-
-https://www.ibm.com/topics/large-language-models
-
-https://en.wikipedia.org/wiki/Large_language_model#Further_reading
-
-Behind all the jargon, the neural networks that underpin GenAI are models designed to produce "statistically probable outputs" when prompted [@ibm_what_2021].
-In more conventional terms, neural networks are regression models [@ripley_neural_1994]--much more sophisticated and powerful than linear regression, but regression models nonetheless.
-Here, it is important to note that the term regression, of course, refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing a prompt to a "statistically probable" outcome.
 
@@
 
-```{r simulate-simple-example}
-set.seed(13253)
-
-b1 <- 1.5
-
-df <- data.frame(x = b1 * runif(100, 
-                                0, 
-                                10),
-                 e = rnorm(100),
-                 type = "training data") |>
-  mutate(y = x + e)
-
-mod <- lm(y ~ x, 
-          data = df)
-```
-
-```{r plot-simulated-data, out.width="50%", fig.cap="Simple Example of a Regression Model"}
-ggplot(df,
-       aes(x = x,
-           y = y)) +
-  geom_point() +
-  geom_smooth(method = "lm") + 
-  theme_minimal()
-```
-
-Predict:
-```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Training data and model output compared"}
-df_pred <- data.frame(x = b1 * runif(100, 
-                                0, 
-                                10))
-
-y_hat <- predict(mod, df_pred)
-
-df_pred <- df_pred |> 
-  mutate(y = y_hat,
-         type = "model output")
-
-df_pred <- rbind(df |>
-                   select(x, y, type),
-                 df_pred |>
-                   select(x, y, type))
-
-ggplot() + 
-  geom_point(data = df_pred,
-  aes(x = x,
-      y = y,
-      color = type,
-      shape = type)) +
-  theme_minimal()
-```
-
 The variance of the model's output is substantially reduced: the output will, of necessity, be less varied, less diverse, and more norm-conforming:
 ```{r data-vs-prediction-variance}
-df_pred |> 
+df_pred_all |> 
   group_by(type) |> 
@@
 Memorization, in fact, is just a fancy way of saying "model overfitting"--in other words, an extreme regression to the mean that excels at predicting data in the training set, but is poor at projecting new data points.
-Overfitting happens when a model with too many parameters is fitted to a relatively small data set (Models have [billions of parameters](https://en.wikipedia.org/wiki/Large_language_model#Training_cost) and are trained on [billions of pages of text](https://www.ibm.com/topics/large-language-models])).
+Overfitting happens when a model has too many parameters or too few data inputs (Models have [billions of parameters](https://en.wikipedia.org/wiki/Large_language_model#Training_cost) and are trained on [billions of pages of text](https://www.ibm.com/topics/large-language-models])).
 Data memorization--the verbatim or quasi-verbatim output of training data--can be as high as 7% in the case of some LLMs [@peng_near_duplicate_2023, p.

### 2024-05-27 README.Rmd
Continued editing and reorganizing the essay

@@
 knitr::opts_chunk$set(
-  echo = FALSE,
+  cache = TRUE,
   collapse = TRUE,
-  comment = "#>"
+  comment = "#>",
+  echo = FALSE
 )
-```
-
-```{css textbox}
-.blackbox {
-  padding: 1em;
-  background: black;
-  color: white;
-  border: 2px solid orange;
-  border-radius: 10px;
-}
-.center {
-  text-align: center;
-}
 ```
@@
 
-Disclaimer: _All opinions expressed here are mine and any resemblance to those of my employer(s) is purely coincidental_.
+Disclaimer: _All opinions expressed here are mine and any resemblance to those of my employer(s) is purely coincidental_. _No generative AI tool was used to research or write this essay_.
 
@@
 > -- An Organization Run by its Marketing Department
-
-:::: {.blackbox data-latex=""}
-::: {.center data-latex=""}
-**NOTICE!**
-:::
-
-Thank you for noticing this **new notice**! Your noticing it has
-been noted, and _will be reported to the authorities_!
-::::
 
@@
 GenAI was announced as nothing short of revolutionary.
-McMaster, for example, [states](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/):
+McMaster, for example, [states](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) that:
 
-> Understanding how generative AI works, and how it can support work within organizations represents an ongoing area of exploration and innovation.
+> [u]nderstanding how generative AI works, and how it can support work within organizations represents an ongoing area of exploration and innovation.
 The opportunities of generative AI are exciting: creating workplace efficiencies to allow for different kinds of work, adding capabilities for individuals and teams, and offering personalized uses.
 
-AI was quickly perceived as an _innovative_ and _exciting_ technology, that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p.
-64].].
-It seemed inevitable that universities _had to do something_ or risk being let behind.
+AI was quickly painted as an _innovative_ and _exciting_ technology, that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p.
+64].
+Cost-cutting is usually implemented by finding "redundancies" rather than finding other things for people to do.].
+It seemed inevitable that universities _had to do something_ with it, at the risk of being let behind.
 
 The fear of missing out, combined with a modicum of concern for the ethical implications of GenAIs^[Ethical concerns for the most part have had to do with policing academic dishonesty among students], has led to institutional responses that have been less than coherent.
-An example is the case of Emory University, where the university gave a $10,000 prize to a group of students who proposed developing an AI tool for an Entrepreneurship Summit, only for the students to be subsequently suspended by the school's Honor Council for the potential of their tool for cheating^[There was no evidence that cheating had happened.] [@koebler__university_2024].
+A case in point is Emory University, where the university gave a $10,000 prize to a group of students who proposed developing an AI tool for an Entrepreneurship Summit, only for the students to be subsequently suspended by the school's Honor Council for the potential of their tool for cheating^[There was no evidence that cheating had happened.] [@koebler__university_2024].
 
@@
 The aim of the provisional guidelines was to inform the university community about the use of AI in "operational excellence".
-But the time has come to update the guidelines, and the institution, via its Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
-This committee is composed of three sub-committees that examine the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work" (the Provost's words)(probably meaning the university's work; not sure if the Provost themself uses AI).
+As we approach the Fall Term of 2024, with almost two years of cohabitation experience with AI for the masses at our backs, the time has come to update these guidelines.
+Pertinently, the institution, via its Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
+This committee is composed of three sub-committees that examine the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work" (the Provost's words)(in all likelihood meaning the university's work; not sure if the Provost herself uses AI).
 
@@
 
-Accepting that AIs are powerful tools, right off the bat, is consistent with the hype observed so far [@angwin_press_2024].
-Hammers are powerful tools.
+Notice that this statement takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption.
+This perspective is almost fatalistic in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one with chatbots in the pole position is just the most recent of several hype cycles [see @siegel_ai_2023]] [@angwin_press_2024].
+To put things in perspective, hammers are powerful tools.
 Word processors are powerful tools.
 Lasers are powerful tools.
-And yet we do not have high-level committees for each of and every tool that the university uses.
-What makes AIs different from other tools that they require a pan-university consultation on an "_ethical approach to shaping policies and developing new ways of using AI to ensure a responsible and beneficial integration where it makes sense_"?
+And yet we do not have high-level university committees for each and every tool that is used at the institution.
+What is it about AIs that, unlike many other tools, requires a pan-university consultation on an "_ethical approach to shaping policies and developing new ways of using AI to ensure a responsible and beneficial integration where it makes sense_"?
 
@@
 
-<!--As I argue here, AI does not make a lot of sense outside of some limited applications that are nowhere near the scope of the promises of the merchants of hype.-->
+To understand why many organizations, including McMaster University, have decided to spend considerable resources studying the adoption and responsible use of AIs, it is essential to have a working understanding of what the technology claims to do that sets it apart from other tools.
 
-To understand why many organizations, including McMaster University, have decided to spend considerable resources studying the adoption and responsible use of AI, it is essential to have a working understanding of what AI does that sets it apart from other tools.
-
-In doing this, we must be alert to the fact that technologies are never politically neutral: it is important to ask two questions:
+In doing this, it is essential that we remain alert to the fact that technologies are _never_ politically neutral, and therefore it is important to ask two questions when considering their adoption:
 
@@
 
-We also need to understand the goals of the organization that considers the technology.
-When does it make sense for a university like McMaster to "integrate" AI as part of the core mission of the organization, accounting for what it does to some people and what it does for others?
+We also need to understand the goals of the organization that considers using the technology.
+When does it make sense for a university like McMaster to "integrate" AIs as part of the core mission of the organization, accounting for what they do for some people and what they do to others?
 
@@
 Public funding of universities in Ontario as percent of revenue fell from 45.6% in 2001/2 to 35.5% in 2019/20 [@statistics_canada_college_2022].
-It is little wonder that universities are seeking "efficiencies"])?
+As noted by @usher_state_2023, Ontario is "a province where underfunding [higher-education] institutions is a pan-partisan affair".
+Is it any wonder that universities in the province are pushed to find "efficiencies"?])?
 
-Clearly, there will be some tension implicit in the mission of a university.
-But being generous, let us take the mission statement at face value and assume that the highest priorities are those related to knowledge obtained, preserved, and communicated with integrity and quality.
+Clearly, there is some tension implicit in the mission of a university.
+But being generous, we can take the mission statement at face value and assume that the highest priorities are those related to knowledge obtained, preserved, and communicated with integrity and quality.
 
@@
 
-A possible reason why many organizations have decided to spend substantial resources considering the adoption of GenAI is that, unlike most other tools, they are seen as _everything tools_.
+A possible reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they are seen as _everything tools_.
 Indeed, the release of GenAIs to the masses was accompanied by innumerable lofty promises, albeit most of them made by entities that had a stake in the adoption of the technology.
@@
 
-In another instance, @hie_how_2023 argue that AIs will likely be the primary way to access information in the future.
+In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school).
+Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence", whatever that means.] argue that AIs will likely be the primary way to access information in the future.
 For this reason, institutions must equip students with the skills needed to use AI competently, in particular to perfect the art of _prompt engineering_, that is, the ability to craft effective queries that prompt a GenAI to provide the most useful outputs.
@@
 Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been essential to understand my own courses; automating the task is another way of cheapening the quality of the learning experience for students].
-@hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school).
-Writing in 2022, Mr.
+
+@hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr.
 Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9).
-In October 2023 facebook [slashed its metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/).
-Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team four months after launching it.
-Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence".] warn that "AI is here to stay", that "[faculty] might not be able to stop the AI trend from growing", and so they might as well get on with the program and try to shape it as best they can.
-
-What does McMaster hope to achieve with AI? From the Overview on Generative AI document:
-
-> Some of [the uses of AI] in our context could include qualitative and quantitative data analysis, writing text (e.g.
-reference letters, grant applications, job descriptions, report summaries), translation, captioning, multiple formats for learning and engagement, and personalized support.
+On February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it.
+Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], that "[faculty] might not be able to stop the AI trend from growing", and so they might as well get on with the program and try to shape it as best they can.
 
@@
 Behind all the jargon, the neural networks that underpin GenAI are models designed to produce "statistically probable outputs" when prompted [@ibm_what_2021].
-In more conventional terms, neural networks are regression models [@ripley_neural_1994]--much more sophisticated and powerful than linear regression to be sure, but regression models nonetheless.
+In more conventional terms, neural networks are regression models--much more sophisticated and powerful than linear regression to be sure, but regression models nonetheless [@ripley_neural_1994].
 Here, it is important to note that the term regression refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing a prompt to a "statistically probable" outcome.
@@
 With these inputs, the model finds a conditional mean for the output $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables).
-The process of finding that conditional mean is to satisfy some critarion (e.g., that the conditional mean minimizes the error of the model--the distance between $y$ and $\hat{y}$).
+The process of finding that conditional mean is to satisfy some criterion (e.g., that the conditional mean minimizes the error of the model--the distance between $y$ and $\hat{y}$).
 
@@
 
-```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Simple Example of a Regression Model"}
+```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 1. Simple Example of a Regression Model"}
 ggplot(df,
@@
 
-Predict:
-```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Training data and model output compared"}
+```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 2. Training data and model output compared"}
 
@@
                                 "nn2 output" = 2)) +
-  theme_minimal()
+  theme_minimal() +
+  theme(legend.position = "bottom")
 ```
@@
 
-```{r plot-shallow-neural-network, out.width="50%", fig.cap="Simple Example of a Shallow Neural Network"}
+```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 3. Simple Example of a Shallow Neural Network"}
 knitr::include_graphics("images/nn1.png")
@@
 
-The plot below illustrates the predictions of this neural network, where we can see that it is not limited to a linear response; neural networks are more flexible than linear regression, and they achieve this by being less parsimonious (they use more parameters).
+The plot below illustrates the predictions of this neural network, where we can see that the model is not limited to a linear response; neural networks are more flexible than linear regression, and they achieve this by being less parsimonious (they use more parameters).
 
-```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Training data and model outputs compared: linear regression and shallow neural network"}
+```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 4. Training data and model outputs compared: linear regression and shallow neural network"}
 ggplot() + 
@@
 ```
-```{r plot-deeper-neural-network, out.width="50%", fig.cap="Example of a Deeper Neural Network"}
+
+```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 5. Example of a Deeper Neural Network"}
 knitr::include_graphics("images/nn2.png")
@@
 
-```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Training data and model outputs compared: shallow neural network and deeper neural network"}
+```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 6. Training data and model outputs compared: shallow neural network and deeper neural network"}
 ggplot() + 
@@
 
-Indeed, the first neural network is "better" in that it reduces the typical error by `r paste0(prettyNum(delta_mad_nn2, digits = 2), "%")` compared to the linear regression, whereas the deeper network reduces the typical error by `r paste0(prettyNum(delta_mad_nn2, digits = 2), "%")`, also with respect to the linear regression^[The "typical error" here is the mean absolute deviation of the predictions.].
+Indeed, the first neural network is "better" than the linear regression in that it reduces the typical error by `r paste0(prettyNum(delta_mad_nn1, digits = 2), "%")`, whereas the deeper network reduces the typical error by `r paste0(prettyNum(delta_mad_nn2, digits = 2), "%")`, also with respect to the linear regression^[The "typical error" here is the mean absolute deviation of the predictions.].
 
-There is a limit to how flexible any model can be, and this is as true of linear regression as it is of neural networks.
-Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than we have data to train the model.
+There is a limit as to how flexible a model can be, and this is true of linear regression and of neural networks.
+Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than the number of data points that exist in the dataset used to train the model.
 Roughly, every additional parameter needs a data input to train on, and when there are as many parameters as data inputs the model is overfit and will predict each value of the independent variable in the training dataset perfectly.
+The model will fail to train if there are more parameters than data points.
+
+Here are some key ideas to keep in mind:
+
+- AIs are neural networks, which is to say, regression models.
+- A regression model generates outputs that are regressions to the mean (i.e., statistically probable outputs).
+- AIs are _very_ flexible non-linear regression models.
+- Their flexibility comes from using "deep" architectures, which means many layers with many, many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters.
+OpenAI no longer reports the number of parameters, but in the latest version this is estimated to be trillions.].
+- Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think).
+The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data was used to train LLMs.
+The increasingly large datasets  essentially expanded the space where interpolation was possible.].
 
@@
 
-## What do GenAI tools do for students at McMaster?
+## AI in teaching and learning
 
-## What do GenAI tools do to students at McMaster?
+Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it makes sense to use this technology.
+To begin, we must note the sheer number of examples of the use of AI by students.
+A report by @tyton_partners_generative_2023 that was sponsored by Turnitin found that 50% out of 1,600 post-secondary students who participated in the study, were users of GenAI (but only 22% of 1,000 faculty in the study).
+Use of AI by students is one of the reasons cited by some for universities to embrace AI [e.g., @power_3_2024].
+McMaster's own internal research indicates that at least 61% of over 1,300 students polled, used AI during the Fall Term in 2023, 75% of them for school work [@yachouh_student_2024].
 
-## Not learning
+But embracing AI just because the students use it is like putting the cart before the horse.
+After all, smoking is banned on campus, and I do not recall the university asking the students if they smoked^["A tobacco and smoke-free campus is the next important step towards fulfilling our responsibilities as educators, healthcare professionals and to the communities we serve" said then-president Patrick Deane [@carter_mcmaster_2017]].
+
+What does McMaster hope to achieve with AI? From the Overview on Generative AI document:
+
+> Some of [the uses of AI] in our context could include qualitative and quantitative data analysis, writing text (e.g.
+reference letters, grant applications, job descriptions, report summaries), translation, captioning, multiple formats for learning and engagement, and personalized support.
+
+## What can GenAI tools do for students?
+
+<!-- From Review Report in private folder: May 6 Review Report_AI Expert Panel on Teaching and Learning.docx -->
+
+- Students want concepts explained to them differently.
+
+Relatively common concepts have already been explained by humans in dozens if not hundreds of different ways.
+A search of McMaster's library catalogue, looking only for resources available online (for convenience), reveals that there are hundreds, thousands, and sometimes tens of thousands of resources available to explore the same concept from multiple perspectives.
+
+```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 7. Result of a search for online 'econometrics' resources in McMaster's Library catalogue"}
+knitr::include_graphics("images/mac-library-econometrics.png")
+```
+
+```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 8. Result of a search for online 'cell biology' resources in McMaster's Library catalogue"}
+knitr::include_graphics("images/mac-library-cell-biology.png")
+```
+
+```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 9. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
+knitr::include_graphics("images/mac-library-euphemism.png")
+```
+
+Is an AI tool essential to find alternative explanations for common concepts?
+Probably only at the expense of research skills.
+As for more obscure concepts, given their paucity in training datasets, those would tend to be uncommon, and therefore improbable outputs (i.e.
+for reasons of lying far from the mean).
+This strongly suggests that any outputs about highly specialized queries should be treated with caution^[The Overview on Generative AI notes that "if you ask [a chatbot] about yourself or a more obscure topic, [it] will almost certainly make up nonsense." This is true even when asking about a person with an extensive online presence (the internet being the likely source of the training datasets); as recently as 2023, a famous law professor discovered when ChatGPT produced output stating that he had been accused of sexual harassment (in the real world he had not been) [@verma_chatgpt_2023]].
+
+In summary: there is no dearth of materials to learn from about relatively common concepts, and AIs should not be used to try to learn specialized concepts.
+
+- Practice questions
+
+Most textbooks include plenty of practice questions, but it may signal something worse: preparation for the examination.
+
+- Immediate feedback on drafted work
+
+From @mollick_instructors_2024
+
+-
+-
+-
+
+## What do GenAI tools do to students?
+
+### Not learning
 
@@
 
-## Learning the wrong stuff
+### Learning the wrong stuff
 
@@
 
-## Deskilling
+### Deskilling
 

### 2024-05-28 README.Rmd
More edits, it feels like I am about 70% done

@@
 
-# Use of AI in Teaching and Learning: Some Considerations
+# An AI is haunting campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the shade of the spectre of unregulated surveillance capitalism.]: Some considerations about the use of AI for teaching and learning in post-secondary education
 
@@
 
-## Background: An AI is Haunting Campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the shade of the spectre of unregulated surveillance capitalism.]
+## Background
 
@@
 GenAI was announced as nothing short of revolutionary.
-McMaster, for example, [states](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) that:
+McMaster, for example, [stated](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) that:
 
@@
 Pertinently, the institution, via its Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
-This committee is composed of three sub-committees that examine the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work" (the Provost's words)(in all likelihood meaning the university's work; not sure if the Provost herself uses AI).
+This committee is composed of three sub-committees tasked with examining the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work" (the Provost's words)(in all likelihood meaning the university's work; not sure if the Provost herself uses AI in addition to the majestic plural).
 
-The sub-committees, in turn, convened panels to:
-
-1.
-[Provide recommendations]({#ai-panel-tasks}) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/)
-
-2. And also to do (what? consultation?).
-
-3. Q: What are the terms of reference of the panels? Who will decide which recommendations are adopted and which are ignored?
+The sub-committees, in turn, convened panels to [provide recommendations]({#ai-panel-tasks}) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/).
+Participants in the panel were told that they would help to shape the guidelines for teaching and learning at McMaster, a consequential task "for how educators take up generative AI in their course and assessment design and how students approach the use of this technology"^[I have been unable to find the terms of reference of the panels.
+Who will decide which recommendations of the panel are adopted and which are ignored?].
 
@@
 
-Notice that this statement takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption.
-This perspective is almost fatalistic in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one with chatbots in the pole position is just the most recent of several hype cycles [see @siegel_ai_2023]] [@angwin_press_2024].
+Notice that this statement (like the definition of the work before the panels), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption.
+This attitude is almost fatalistic in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one with chatbots in the pole position is just the most recent of several hype cycles [see @siegel_ai_2023]] [@angwin_press_2024].
+
 To put things in perspective, hammers are powerful tools.
@@
 We also need to understand the goals of the organization that considers using the technology.
-When does it make sense for a university like McMaster to "integrate" AIs as part of the core mission of the organization, accounting for what they do for some people and what they do to others?
+When does it make sense for a university like McMaster to "integrate" AIs as part of the core mission of the organization, accounting for what the technology does for some people and what it does to others?
 
@@
 A possible reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they are seen as _everything tools_.
-Indeed, the release of GenAIs to the masses was accompanied by innumerable lofty promises, albeit most of them made by entities that had a stake in the adoption of the technology.
+While hammers are understood to have a limited range of uses having to do with concentrating force, and word processors are mostly used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
 
@@
 Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], that "[faculty] might not be able to stop the AI trend from growing", and so they might as well get on with the program and try to shape it as best they can.
+
+## A note on the terms of the debate
+
+A benefit of portraying AI as an _everything tool_ (for the proponents of the technology, at least), is that conversations around it quickly turn into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength.
+In the case of AI, critics are forced to respond to every single propounded use case with in-depth analysis, which usually is more time consuming than making all sorts of claims to begin with^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed [flooding the zone with shit](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation).].
+
+A second discoursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.
+
+In what follows, I do not plan to confront every single possible use case of AI as doing so plays into the intent of the Gish Gallop^[I have already spent close to 50 hours working on this essay, and it is still nowhere near an exhaustive apology or indictment of AI.], which is to defeat criticism by flooding the zone.
+As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are other ways for teaching and learning to be excellent that do not require AI.
 
@@
 In comparison to the linear regression and its two parameters, this neural network uses seven parameters.
-
+Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the model contigent on the starting conditions.
 ```{r}
@@
 
+In a neural network trained to work with natural language, the "predictors" are words that come before and after a certain word.
+Several approaches exist to _embed_ a bag of words, using large text corpora the co-occurrence of words is used to convert words to numerical vectors.
+These vectors are decontextualized, that is, they only reflect the position of a given word with respect to other words in the training corpora; however, the numeric values are static even if the surrounding words in a specific prompt change.
+A recent breakthrough in this field is the development of _transformers_, a network architecture architecture that updates the numeric values of a word depending on the words that come before and after.
+
 Here are some key ideas to keep in mind:
@@
 OpenAI no longer reports the number of parameters, but in the latest version this is estimated to be trillions.].
+- AIs are trained using an initial set of random values as parameters. The same training dataset can result in different models due to the randomness in the initial conditions.
+- The flexibility of AIs means that outputs will vary seemingly at random with relatively small changes in the prompt. 
 - Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think).
-The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data was used to train LLMs.
-The increasingly large datasets  essentially expanded the space where interpolation was possible.].
+The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data were used to train LLMs.
+The increasingly large datasets  essentially expanded the space where the model could interpolate.].
 
 <!--
-IBM uses a dataset called Project CodeNet Dataset, which is sourced from two online judge web sites, AIZU Online Judge and AtCoder.
-https://research.ibm.com/blog/codenet-ai-neurips-2021 According to IBM "[the] end goal of CodeNet is to enable developers to create systems that can modernize existing codebases, as well as fix errors and security vulnerabilities in code."
 
-From Aizu Terms of Use: https://onlinejudge.u-aizu.ac.jp/term_of_use
-
-Handling of Submitted and Registered Codes
-
-Commercial uses of the registered codes are prohibited.
-
-You acknowledge and agree that users of this site may use copies or modifications of the public codes only for educational or research purposes.
-
-You acknowledge and agree that public source codes may be opened collectively to persons authorized to access only for educational or research purposes.
-
-You acknowledge and agree that source code submitted may be used by University of Aizu and related research institutes only for educational or research purposes.
-
-IBM studiously avoids the word "commercial" in its posts about LLM, foundational models, and CodeNet.
-Is IBM doing all this research as a non-commercial effort?
-Will those developers creating systems, modernizing databases, fixing errors, etc., be able to access this for free?
-What does "commercial uses" even mean?
 -->
@@
 
+> Anecdotally students at McMaster report wanting to use generative AI not to complete assignments for them, but to help enhance learning in their courses by having concepts explained differently, by working on practice questions or by getting immediate feedback on drafted work.
+
 <!-- From Review Report in private folder: May 6 Review Report_AI Expert Panel on Teaching and Learning.docx -->
@@
 
-Relatively common concepts have already been explained by humans in dozens if not hundreds of different ways.
-A search of McMaster's library catalogue, looking only for resources available online (for convenience), reveals that there are hundreds, thousands, and sometimes tens of thousands of resources available to explore the same concept from multiple perspectives.
+There are many traditional ways for students to get different perspectives on concepts.
+They can form study groups, attend office hours to engage with their instructors, ask their Teaching Assistants, ask relatives and friends, and so on.
+But even for a student who for a whatever reason cannot or will not do any of these things, the fact remains that relatively common concepts have already been explained by humans in dozens if not hundreds of different ways, often in writing that is available through an excellent university library.
+A search of McMaster's library catalog, looking only for resources available online (for convenience), reveals that there are hundreds, thousands, and sometimes tens of thousands of resources available to explore the same concept from multiple perspectives.
 
-```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 7. Result of a search for online 'econometrics' resources in McMaster's Library catalogue"}
+```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 7. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
 knitr::include_graphics("images/mac-library-econometrics.png")
@@
 
-```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 8. Result of a search for online 'cell biology' resources in McMaster's Library catalogue"}
+```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 8. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
 knitr::include_graphics("images/mac-library-cell-biology.png")
@@
 Is an AI tool essential to find alternative explanations for common concepts?
-Probably only at the expense of research skills.
+Only as a shortcut for a good library search (or to make up for the lack of an academic network, not taking advantage of office hours, etc.), and at the expense of research skills.
 As for more obscure concepts, given their paucity in training datasets, those would tend to be uncommon, and therefore improbable outputs (i.e.
@@
 
-In summary: there is no dearth of materials to learn from about relatively common concepts, and AIs should not be used to try to learn specialized concepts.
+In summary: there is no dearth of resources to learn about relatively common concepts, and AIs should not be used to try to learn more specialized concepts.
 
@@
 
-Most textbooks include plenty of practice questions, but it may signal something worse: preparation for the examination.
+Most textbooks already include plenty of practice questions, especially for relatively common topics.
+For more specialized topics, asking interesting questions should be a skill in and of itself.
+An emphasis on practice questions, however, signals an underlying weakness of the process, namely, a focus on learning for the examination, as opposed to learning for its own sake.
+A growing body of evidence points at the effectiveness of alternative assessment schemes, such as continuous assessment through course work, and even seen examinations instead of the traditional unseen examinations [@turner_see_2018], the latter form of assessment continues to dominate the evaluation landscape [@buckley_are_2023].
+Ungrading, in particular, is increasingly recognized as its potential to internalize the motivation to learn, and for fostering adaptability, creative thinking, and self-management [@gorichanaz_it_2024].
+
+Two relevant issues appear to be relevant.
+
+First, posing interesting questions for students to work on is an essential part of an instructor's job, but this part of the job becomes compromised when instructors operate in massive classrooms situations^[Average class sizes at McMaster (with a full time under-graduate student population of 30,400) are 127.8 in first/second year, and 52.3 in third/fourth year; this places McMaster dead last among all Canadian institutions in class sizes in first/second year, and second to last after UBC in third/fourth year [@macleans_canadas_2022]].
+In this case, AI is a substitute, and demonstrably not a good one, for a qualified instructor who can intellectually challenge their students.
+
+The second issue is traditional evaluation.
+As noted, studying for the examination often detracts from genuine learning.
+Why are other assessment approaches seldom considered?
+Research by @vahed_continuous_2023 suggests that instructors need to develop assessment literacy, that is, an understanding of "the fundamental assessment concepts and procedures, and to engage in the ongoing process of adjusting instructional strategies to enrich students learning experiences".
+There is a need for innovative approaches in assessment, but it is unlikely that AI is the tool to develop the assessment literacy that they require.
 
@@
 
-From @mollick_instructors_2024
+This is an unrealistic expectation.
+More realistically, students could have the opportunity to discuss their drafted work with instructors before final submission.
+A conscientious instructor will invest time reading and understanding the draft work, to be able to provide quality feedback.
+But there is a more insidious aspect about getting feedback from an AI...who, actually, is providing the feedback?
+What data were used to train the machine that provides the feedback?
+Who fine-tuned the machine who provides the feedback?
+Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided?^[For instance, [@verma_chatgpt_2023]] (More on this below).
 
--
--
--
+- Personalized
+
+@mollick_instructors_2024^[Drs.
+Eric Mollick and Lilach Mollick are members of Wharton Interactive, in the business school of the University of Pennsylvania.
+Wharton Interactive offers to "democratize education" so to level up your learning!
+[@kelner_level_2022]] envision a future in which educators are innovators thanks to their adroit use of prompts for AIs.
+According to these authors, GenAI stands alone among educational technologies in that it can be "'programmed' through prompts alone", which should allow even instructors "without extensive technology or coding experience" to more easily create classroom applications.
 
@@
 
-### Not learning
+### Distract from learning
 
-Words go here.
+Reliance on AI can short-circuit the learning process.
+Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] convey the idea that prompt engineering will become an essential skill; perhaps the skill to end all skills.
+
+However, this skill is tied to a singular technology.
+How is prompt engineering a transferable skill?
+As @mollick_instructors_2024 note, prompt engineering is not even transferable _between_ different LLMs (see p.
+5 and p.
+36).
+Good research skills transfer between contexts.
+Learning how to form an effective query to find bibliographical resources would work at McMaster, at the University of Toronto, and at the New York Public Library.
+Learning to identify relevant content would work while reading a novel, a book of history, or a text of econometrics.
+Writing eventually helps to create a style.
+
+In a sense, an AI is like a crutch.
+It can help someone to accomplish a task more easily, but it will not help if used constantly.
+However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle that the end user cannot hope to learn^[Reports of people getting their fingers crushed by Tesla's cybertruck could be a result of 1) people trusting the technology more than they should; and 2) not understanding the technology: apparently the algorithm uses increasing force to close the trunk if it fails to latch on back-to-back cycles.
+Hence, the trunk will not slash a banana on its first attemt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature (in the sense that neural networks are blackboxes, and the development of AIs itself is a blackbox where things change based on the interests^[Sometimes the interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that those interests will always be align with the public good, and in fact, they almost never will be as long as there are only a handful of American companies that offer an "essential" service.] of their real developers, not those who fancy themselves as able to create apps without the need to learn how to code.).
+
+My perspective is that it is better to do things more slowly at the beginning, to really exercise the brain, so that the brain can do more exceptional things later (perhaps even more efficiently).
+
+But let us assume for a moment that AI in its current form really becomes essential. What is the expected longevity of skills like prompt engineering?
+
+Apart from a few methodological developments (like transformers), improvement in the performance of AIs (say, from ChatGPT-2 to the latest version), mostly comes from using bigger datasets.
+There has been a rush for data, as developers try to improve their models by training them using more and more data.
+But, despite the large number of texts available in digital form, there is still only a finite amount of data to be had.
+
+Big corporations are mostly done grabbing all data that they could lay their hands on, sometimes by morally dubious, and at times by possibly illegal means.
+
+IBM, for example, uses data drawn from the Project CodeNet Dataset.
+The data are "sourced" from two online judge web sites, AIZU Online Judge and AtCoder [@ibm_automatic_2021].
+According to IBM "[the] end goal of CodeNet is to enable developers to create systems that can modernize existing codebases, as well as fix errors and security vulnerabilities in code." When we inspect the Terms of Use of AIZU (a Japanese academic project), "Commercial uses of the registered codes are prohibited".
+IBM studiously avoids the word "commercial" in its posts about LLM, foundational models, and CodeNet.
+One might ask if IBM is doing all this research as a non-commercial effort?
+Will those developers creating systems, modernizing databases, fixing errors, etc., be able to access this for free, and use the systems for education and research only?
+What does "commercial uses" even mean?
+
+As the sources of human-generated natural language data dry up, the corporations developing AIs have turned to what might be the last juicy nuggets left on the internet.
+
+OpenAI turned to Stack Overflow [@stack_overflow_stack_2024] for more data.
+Stack Overflow [bills itself](https://stackoverflow.com) as "A community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world." The community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) with the way Stack Overflow dealt with OpenAI, and believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution.
+Some members of the community have protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/).
+It is an open question how long Stack Overflow will remain a community-based space, when it treats the members of the community as unwitting providers of data for OpenAI.
+
+Google, for its part, turned to Reddit for training data [@roth_google_2024] to the tune of \$60 million per year.
+The results of this deal have been less than stellar, as Reddit content begins to creep into Google's AI-assisted search.
+Reports of this include a suggestion by Google search to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024]^[The headline of the 404 Media report was "Google Is Paying Reddit $60 Million for Fucksmith to Tell Its Users to Eat Glue".].
+
+```{r google-reddit, out.width="50%", fig.cap="Figure 10. Google's AI overview result was traced back to a post by redditor Fucksmith"}
+knitr::include_graphics("images/google-cheese-glue.png")
+```
+
+Other AI Overview results include a recommendation by geologists to eat at least one small rock a day, a piece of advice tracked back to an article in the well-known satirical periodical The Onion.
+
+```{r google-the-onion, out.width="50%", fig.cap="Figure 10. Google's AI overview result was traced back to an article in The Onion"}
+knitr::include_graphics("images/google-the-rock.jpg")
+```
+
+Mind you, these, and numerous other [hapless results](https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301) (as of May, 2023) are produce by AI _after_ Google has already spent tens of billions of dollars on the technology [@vynck_after_2024].
+To what extent can more money improve AI?
+Probably not by much, since the key constraint is the amount of data available for training.
+
+NEXT:
+
+Peak ChatGPT.
+
+Model poisoning.
+
+Develop a garbage skill https://arstechnica.com/tech-policy/2023/05/lawyer-cited-6-fake-cases-made-up-by-chatgpt-judge-calls-it-unprecedented/
 
@@
 Words go here.
+
+Automation bias: https://en.wikipedia.org/wiki/Automation_bias
+
+It is likely that there is a similar authority bias in which students tend to be more positive
 
@@
 
-Words go here.
+So far, not money.
+But since corporations are not in the development of AI for charitable motives, at some point they will want to make money out of them.
+Who will pay?
+And how much will they need to pay?
+Or have to pay, if their skills are inexorable dependent on this one technology?
 
@@
 
+<!-- AI is a way to cheapen university education; it compromises the ideals of quality, integrity, and excellence. -->
+
 ChatGPT was an awesome feat of marketing [@cowen_chatgpt_2023], but much coverage of the relevant issues in the media has been so credulous it has stretched credulity past Pluto [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called.].

### 2024-05-28 README.Rmd
More edits!!

@@
 
-## Background
+---
+
+## Contents
+
+- [Background](#background)  
+- [The university's mission](#the-mission)  
+- [What do AIs do?](#what-do-ais-do)  
+- [A note on the terms of the conversation](#terms-of-debate)  
+- [But really, what do AIs do?](#really-what-do-they-do)  
+- [AI in teaching and learning](#ai-in-teaching-and-learning)  
+- [What can GenAI tools do for students?](#what-can-ais-do-for-students)  
+- [What do GenAI tools do to students?](#what-can-ais-do-to-students)  
+- [What do GenAIs do for the mega-corporations pushing these tools?](#what-do-ais-to-corporations)  
+- [What do GenAIs to to everyone else?](#what-do-ais-do-to-everyone-else)
+
+---
+
+## Background {#background}
 
@@
 
-## The [mission](https://president.mcmaster.ca/mission-vision/)
+## The university's mission {#the-mission}
 
-McMaster's mission is "the discovery, communication and preservation of knowledge," accomplished with a commitment to "creativity, innovation and excellence".
+McMaster's [mission](https://president.mcmaster.ca/mission-vision/) is "the discovery, communication and preservation of knowledge," accomplished with a commitment to "creativity, innovation and excellence".
 These aspirations must respond to some underlying values, and indeed, the university explicitly values "integrity, quality, inclusiveness and teamwork".
@@
 
-## What do AIs do?
+## What do AIs do?{#what-do-ais-do}
 
@@
 
-## A note on the terms of the debate
+## A note on the terms of the conversation {#terms-of-debate}
 
@@
 
-## But really, what do AIs do?
+## But really, what do AIs do? {#really-what-do-they-do}
 
@@
 
-## AI in teaching and learning
+## AI in teaching and learning {#ai-in-teaching-and-learning}
 
@@
 
-## What can GenAI tools do for students?
+## What can GenAI tools do for students?{#what-can-ais-do-for-students}
 
@@
 
-## What do GenAI tools do to students?
+## What do GenAI tools do to students? {#what-can-ais-do-to-students}
 
@@
 
-Mind you, these, and numerous other [hapless results](https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301) (as of May, 2023) are produce by AI _after_ Google has already spent tens of billions of dollars on the technology [@vynck_after_2024].
+Mind you, these, and numerous other hapless results^[For instance, ; or returning misinformation about depresion [@tenbarge_glue_2024]] (as of May, 2024) are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
 To what extent can more money improve AI?
-Probably not by much, since the key constraint is the amount of data available for training.
+Probably not by much, since the key constraint is the amount of data available for training, and as numerous examples
+
+a search for “how to rotate text in ms paint,” referring to the Microsoft Paint application.
+The top Google search result, viewed by NBC News, directs the reader to a sarcastic Reddit comment that says to press the “Flubblegorp” key on your keyboard.
+This key does not exist.
 
@@
 
-## What do GenAIs do for the mega-corporations pushing these tools?
+## What do GenAIs do for the mega-corporations pushing these tools? {#what-do-ais-to-corporations}
 
@@
 
-## What do GenAIs to to everyone else?
+## What do GenAIs to to everyone else? {#what-do-ais-do-to-everyone-else}
 
@@
 
-Above, the ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted.
-But AI companies do not spare the rich either.
-OpenAI's Sam Altman, a self-avowed admirer of the movie _Her_, featuring Scarlet Johansson as an AI voice assistant, reportedly tried to hire Ms.
-Johannson to be the voice of ChatGPT 4.0 [@kastrenakes_scarlett_2024].
-After Ms.
-Johannson decline, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways, which was retired after Ms.
+ChatGPT 4.o was announced on May 13, 2024, with a flashy display of the model's new capabilities, which included the ability to respond vocally to prompts, and to identify images via a mobile device's camera.
+4.o's voice, in particular, caused some titillation in the pants of even its creators [@roose_is_2024].
+The voice was modeled after the digital personal assistant of the movie _Her_, played by Scarlett Johannson.
+This film (to approximately no one's surprise) is, by his own account, one of OpenAI's Altman's favorites.
+The background story of 4.o's voice is somewhat sordid: Altman reportedly tried to hire Ms.
+Johannson to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when Ms.
+Johannson declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways.
+This voice was retired (probably to the profound sadness of many an AI engineer) after Ms.
 Johansson threatened legal action^[Altman's denials that OpenAI had not used Ms.
 Johansson's voice sounded hollow, after he [posted](https://x.com/sama/status/1790075827666796666) on social media the word "her" at the time the new version of ChatGPT was announced.] [@allyn_scarlett_2024].
-Alas, very few have the social standing of Scarlett Johansson to challenge OpenAI.
-Most people in the world lack that power.
+
+Alas, very few have the social standing of Scarlett Johansson to challenge OpenAI. Most people in the world lack that power.
+
+The ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted. But AI companies do not spare the rich either.
 
@@
 Think about linguistic hegemony, and how the models are only as good (or as bad) as their inputs.
+
+@roberts_commercial_2016 Commercial {Content} {Moderation}: {Digital} {Laborers}' {Dirty} {Work}
 
@@
 
-### Replication of unjust structures of power
+### Reproduction of unjust structures of power
 
-When using a Generative AI we must ask whose perspectives that tool represents.
+When using a Generative AI we must ask whose perspectives the tool will tend to reproduce and amplify.
+
+Training data, much like technology, are not politically neutral, and several processes contribute to reproduce malignant biases, A way to explain this is by means of the concept of _platformed racism_ [@matamoros-fernandez_platformed_2017].
+Platformed racism is the process whereby the functionalities of platforms are exploited to create and spread racist ideology (liking and sharing), often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has had a wide reach [@treyger_russian_2022] and furthermore is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S.
+politics [@us_department_of_state_gec_2024, p.
+51].
+Given how cagey corporations are about the data they use, it is anyone's guess how much dis- and misinformation is in the datasets used to train their AIs (the Unfathomable Data problem discussed by @bender_dangers_2021).].
+The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and [arbitrary moderation practices](https://www.bbc.com/news/technology-58058428) that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].
+
+Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it is too mych to expecet that AI-augmented search engines do better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
+In addition to being misleading to the unadvised, it is also a stupendous tool ripe for abuse by the advised with an agenda.].
+In another wretched incident, Google's AI overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long right-wing disinformation campaign to otherize Obama [@tenbarge_glue_2024].
+Mishaps like this, and others, are dismissed by Google as attempts to "to trip up the technology with uncommon questions" [@tenbarge_glue_2024], which only reinforces the notion that AIs only are useful when [regressing to the mean](#really-what-do-they-do).
+Responses like fiddling with the model (fine tune it) to provide more palatable outputs (like in the case of **FIND EXAMPLE**) are not more reassuring.
+Who is to say what a handful of unaccountable corporations will find palatable tomorrow, let alone in the face of powerful authoritarian regimes?
+
+```{r google-obama, out.width="50%", fig.cap="\\label{fig:google-ai-overview-of-obama}Google's AI overview of Muslim US presidents"}
+google_plot <- ggplot() +
+  labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
+  theme_void()
+
+ggdraw() + 
+  draw_plot(google_plot) +
+  draw_image(image = "images/google-obama.png")
+```
 
@@
 
-ChatGPT was an awesome feat of marketing [@cowen_chatgpt_2023], but much coverage of the relevant issues in the media has been so credulous it has stretched credulity past Pluto [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called.].
+ChatGPT was an awesome feat of marketing [@cowen_chatgpt_2023], but much coverage of the relevant issues in the media has been so credulous it has stretched credulity past Pluto [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.].
 There are now calls to press pause on the hype [@angwin_press_2024] but these are still drown by the credulous coverage.

### 2024-05-29 README.Rmd
Continued writing here and there

@@
 Notice that this statement (like the definition of the work before the panels), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption.
-This attitude is almost fatalistic in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one with chatbots in the pole position is just the most recent of several hype cycles [see @siegel_ai_2023]] [@angwin_press_2024].
+This attitude is almost fatalistic^[In the sense of "fated to"] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one with chatbots in the pole position is just the most recent of several hype cycles [see @siegel_ai_2023]] [@angwin_press_2024].
 
@@
 
-A possible reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they are seen as _everything tools_.
+A key reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they are seen as _everything tools_.
 While hammers are understood to have a limited range of uses having to do with concentrating force, and word processors are mostly used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
@@
 In mental health care, chatbots could complement human therapists "via comprehensive data access and [analysis of] behavioral patterns" to "mimic practitioner questions and subsequently make recommendations based on a user’s inputs" [@silva_4_2023].
+
+AI would even write code, perhaps helping itself to some self-programming (AI self-help?) as an outcome.
+
+Figure @\ref(fig:ai-writing-code)
+
+```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure X. Example of an AI writing code"}
+knitr::include_graphics("images/AI-bug-fix.png")
+```
 
@@
 
+@contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students".
+Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].] **MORE HYPE**
+
 ## A note on the terms of the conversation {#terms-of-debate}
@@
 A benefit of portraying AI as an _everything tool_ (for the proponents of the technology, at least), is that conversations around it quickly turn into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength.
-In the case of AI, critics are forced to respond to every single propounded use case with in-depth analysis, which usually is more time consuming than making all sorts of claims to begin with^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed [flooding the zone with shit](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation).].
+In the case of AI, critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making all sorts of claims to begin with^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed [flooding the zone with shit](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation).
+Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot possibly match [@maiberg__google_2024].].
 
@@
 In what follows, I do not plan to confront every single possible use case of AI as doing so plays into the intent of the Gish Gallop^[I have already spent close to 50 hours working on this essay, and it is still nowhere near an exhaustive apology or indictment of AI.], which is to defeat criticism by flooding the zone.
-As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are other ways for teaching and learning to be excellent that do not require AI.
+As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple other ways for teaching and learning to be excellent that do not require AI, and do not require the moral, environmental, social compromises demanded by AI[^Most discussions of AI as an everything tool acknowledge ].
 
@@
 Training data, much like technology, are not politically neutral, and several processes contribute to reproduce malignant biases, A way to explain this is by means of the concept of _platformed racism_ [@matamoros-fernandez_platformed_2017].
-Platformed racism is the process whereby the functionalities of platforms are exploited to create and spread racist ideology (liking and sharing), often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has had a wide reach [@treyger_russian_2022] and furthermore is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S.
+Platformed racism is the process whereby the functionalities of platforms are exploited to create and spread racist ideology (liking and sharing), often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has had a wide reach [@treyger_russian_2022] and furthermore is known to seed distrust between ethnic communities [@svedkauskas_analysis_2020].
+As the U.S.
+Department of State Reports, Russian intelligence is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S.
 politics [@us_department_of_state_gec_2024, p.
-51].
-Given how cagey corporations are about the data they use, it is anyone's guess how much dis- and misinformation is in the datasets used to train their AIs (the Unfathomable Data problem discussed by @bender_dangers_2021).].
+51].].
+Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs (this is part of the Unfathomable Data problem discussed by @bender_dangers_2021).
+In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].].
 The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and [arbitrary moderation practices](https://www.bbc.com/news/technology-58058428) that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].
 
-Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it is too mych to expecet that AI-augmented search engines do better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
-In addition to being misleading to the unadvised, it is also a stupendous tool ripe for abuse by the advised with an agenda.].
-In another wretched incident, Google's AI overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long right-wing disinformation campaign to otherize Obama [@tenbarge_glue_2024].
-Mishaps like this, and others, are dismissed by Google as attempts to "to trip up the technology with uncommon questions" [@tenbarge_glue_2024], which only reinforces the notion that AIs only are useful when [regressing to the mean](#really-what-do-they-do).
-Responses like fiddling with the model (fine tune it) to provide more palatable outputs (like in the case of **FIND EXAMPLE**) are not more reassuring.
+Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
+In addition to being misleading to the unadvised, AI is also a stupendous tool ripe for abuse by those who are advised _and_ have an agenda.
+Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than using AI to identify misinformation [@maiberg__google_2024].].
+In another wretched incident, Google's AI overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Obama [@tenbarge_glue_2024].
+Mishaps like this, and others, are dismissed by Google as attempts to "to trip up the technology with uncommon questions" [@tenbarge_glue_2024].
+This dismissal, though, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do).
+Responses like fiddling with the model (fine tuning it) to provide more palatable outputs (like in the case of **FIND EXAMPLE**) are not more reassuring.
 Who is to say what a handful of unaccountable corporations will find palatable tomorrow, let alone in the face of powerful authoritarian regimes?
@@
 Recall that [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean) is also called "reversion to mediocrity" [@barnett_regression_2017].
+
+What do AIs do? For the province, the main patron and effectively the boss of McMaster, it provides a rationale (an excuse even) to force the institution to find "efficiencies".
 

### 2024-05-29 README.Rmd
I could get some more writing in today

@@
 Pertinently, the institution, via its Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
-This committee is composed of three sub-committees tasked with examining the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work" (the Provost's words)(in all likelihood meaning the university's work; not sure if the Provost herself uses AI in addition to the majestic plural).
+This committee is composed of three sub-committees tasked with examining the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work".
 
@@
 
-Notice that this statement (like the definition of the work before the panels), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption.
+Notice that this statement (like the definition of the work before the panels above), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption.
 This attitude is almost fatalistic^[In the sense of "fated to"] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one with chatbots in the pole position is just the most recent of several hype cycles [see @siegel_ai_2023]] [@angwin_press_2024].
@@
 The public was promised that GenAI would "accelerate sustainability" [@nakagawa_accelerating_2023]; that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" by "automating customer self-service, accelerating response times on an increasing number of tasks as well as providing greater accuracy, enhanced routing and intelligent context gathering" [@ibm_what_2023]; according to Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.], AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024].
+But there is no reason to stop at self-driving vehicles: McMaster University's Canada Excellence Research Chair Laureate Ali Emadi "hopes to help revolutionize and influence everything from personal travel to city planning" by putting AI directly into our transportation systems [@dorey_driving_2023].
 In mental health care, chatbots could complement human therapists "via comprehensive data access and [analysis of] behavioral patterns" to "mimic practitioner questions and subsequently make recommendations based on a user’s inputs" [@silva_4_2023].
 
-AI would even write code, perhaps helping itself to some self-programming (AI self-help?) as an outcome.
+AI would even write code, perhaps helping itself to some self-programming along the way: AIs that write themselves, or AI self-help?
 
@@
 
-```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure X. Example of an AI writing code"}
+```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure X. An example of Atlassian AI writing code"}
 knitr::include_graphics("images/AI-bug-fix.png")
@@
 
-The list of promises has grown to encompass pretty much everything, from saving the planet all the way to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023].
+The list of promises about what AI will do has grown to encompass pretty much everything, from saving the planet from ourselves, all the way down to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023].
 
-Higher-education was not spared the onslaught of promises.
+Higher-education has not spared the onslaught of promises.
 
@@
 
-## What can GenAI tools do for students?{#what-can-ais-do-for-students}
+## What do GenAI tools do for students?{#what-can-ais-do-for-students}
 
@@
 
-## What do GenAIs do for the mega-corporations pushing these tools? {#what-do-ais-to-corporations}
+## What do GenAIs do for the mega-corporations pushing them? {#what-do-ais-to-corporations}
 
@@
 
-## What do GenAIs to to everyone else? {#what-do-ais-do-to-everyone-else}
+To read:
+
+@sriram_ai_2023
+
+## What do GenAIs do to everyone else? {#what-do-ais-do-to-everyone-else}
 
@@
 In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].].
-The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and [arbitrary moderation practices](https://www.bbc.com/news/technology-58058428) that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].
+The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].
 

### 2024-06-01 README.Rmd
I am now polishing the first part, and tying up the second.

@@
 
-Disclaimer: _All opinions expressed here are mine and any resemblance to those of my employer(s) is purely coincidental_. _No generative AI tool was used to research or write this essay_.
+Disclaimer: _All opinions expressed here are mine and any resemblance to those of my employer(s) is purely coincidental_.
+_No generative AI tool was used to research or write this essay, only good old-fashioned library and internet search skills_.
 
@@
 - [AI in teaching and learning](#ai-in-teaching-and-learning)  
-- [What can GenAI tools do for students?](#what-can-ais-do-for-students)  
-- [What do GenAI tools do to students?](#what-can-ais-do-to-students)  
+- [What do GenAIs do for students?](#what-can-ais-do-for-students)  
+- [What do GenAIs do to students?](#what-can-ais-do-to-students)  
 - [What do GenAIs do for the mega-corporations pushing these tools?](#what-do-ais-to-corporations)  
-- [What do GenAIs to to everyone else?](#what-do-ais-do-to-everyone-else)
+- [What do GenAIs do to everyone else?](#what-do-ais-do-to-everyone-else)
 
@@
 
-GenAI was announced as nothing short of revolutionary.
+GenAI was hailed as nothing short of revolutionary.
 McMaster, for example, [stated](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) that:
@@
 64].
-Cost-cutting is usually implemented by finding "redundancies" rather than finding other things for people to do.].
+Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.].
 It seemed inevitable that universities _had to do something_ with it, at the risk of being let behind.
 
-The fear of missing out, combined with a modicum of concern for the ethical implications of GenAIs^[Ethical concerns for the most part have had to do with policing academic dishonesty among students], has led to institutional responses that have been less than coherent.
+The fear of missing out, combined with a modicum of concern for the ethical implications of GenAIs^[Ethical concerns for the most part have had to do with policing academic dishonesty among students.], has led to institutional responses that have been less than coherent.
 A case in point is Emory University, where the university gave a $10,000 prize to a group of students who proposed developing an AI tool for an Entrepreneurship Summit, only for the students to be subsequently suspended by the school's Honor Council for the potential of their tool for cheating^[There was no evidence that cheating had happened.] [@koebler__university_2024].
@@
 
-The sub-committees, in turn, convened panels to [provide recommendations]({#ai-panel-tasks}) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/).
-Participants in the panel were told that they would help to shape the guidelines for teaching and learning at McMaster, a consequential task "for how educators take up generative AI in their course and assessment design and how students approach the use of this technology"^[I have been unable to find the terms of reference of the panels.
-Who will decide which recommendations of the panel are adopted and which are ignored?].
+The sub-committees, in turn, convened panels to [provide recommendations](#ai-panel-tasks) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/).
+The [priority initiatives](#ai-panel-tasks) for the members of the AI Teaching and Learning Panel are to help shape the guidelines for teaching and learning at McMaster, a consequential task "for how educators take up generative AI in their course and assessment design and how students approach the use of this technology".
 
@@
 Notice that this statement (like the definition of the work before the panels above), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption.
-This attitude is almost fatalistic^[In the sense of "fated to"] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one with chatbots in the pole position is just the most recent of several hype cycles [see @siegel_ai_2023]] [@angwin_press_2024].
+This attitude is almost fatalistic^[In the sense of "fated to".] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn of the bandwagon, is just the most recent of several cycles of hype [see @siegel_ai_2023].] [@angwin_press_2024].
 
@@
 
-(The last sentence in the statement above carries quite a bit of weight here: where does adoption of AI make sense? Presumably this is what the Committee on AI will try to answer.)
-
-To understand why many organizations, including McMaster University, have decided to spend considerable resources studying the adoption and responsible use of AIs, it is essential to have a working understanding of what the technology claims to do that sets it apart from other tools.
+(The last part of the statement above carries quite a bit of weight here: where does adoption of AI make sense? Presumably this is what the Committee on AI will try to answer.)
+
+To understand why many organizations, including McMaster University, have decided to spend considerable resources studying the adoption and responsible use of AIs, it is essential to understand what the technology claims to do that sets it apart from other tools, and what it in actuality does.
 
@@
 Clearly, there is some tension implicit in the mission of a university.
-But being generous, we can take the mission statement at face value and assume that the highest priorities are those related to knowledge obtained, preserved, and communicated with integrity and quality.
+But being generous, we can take the mission statement at face value and assume that the highest priorities are those related to knowledge obtained, preserved, and communicated with creativity, integrity, excellence, and quality.
 
@@
 
-A key reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they are seen as _everything tools_.
-While hammers are understood to have a limited range of uses having to do with concentrating force, and word processors are mostly used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
-
-The public was promised that GenAI would "accelerate sustainability" [@nakagawa_accelerating_2023]; that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" by "automating customer self-service, accelerating response times on an increasing number of tasks as well as providing greater accuracy, enhanced routing and intelligent context gathering" [@ibm_what_2023]; according to Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.], AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024].
+A key reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they have been presented quite deliberately as _everything tools_^[Almost as flexible as, well, a naturally intelligent human, but faster and supposedly cheaper.].
+While hammers are understood to have a limited range of uses that rely on concentrating force, and word processors are mostly used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
+
+The public was promised that GenAI would "accelerate sustainability" [@nakagawa_accelerating_2023].
+
+Also, that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" by "automating customer self-service, accelerating response times on an increasing number of tasks as well as providing greater accuracy, enhanced routing and intelligent context gathering" [@ibm_what_2023].
+
+According to Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.], AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024].
 But there is no reason to stop at self-driving vehicles: McMaster University's Canada Excellence Research Chair Laureate Ali Emadi "hopes to help revolutionize and influence everything from personal travel to city planning" by putting AI directly into our transportation systems [@dorey_driving_2023].
+
 In mental health care, chatbots could complement human therapists "via comprehensive data access and [analysis of] behavioral patterns" to "mimic practitioner questions and subsequently make recommendations based on a user’s inputs" [@silva_4_2023].
 
-AI would even write code, perhaps helping itself to some self-programming along the way: AIs that write themselves, or AI self-help?
-
-Figure @\ref(fig:ai-writing-code)
-
-```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure X. An example of Atlassian AI writing code"}
+AI will even write code, perhaps helping itself to some self-programming along the way: AIs that write themselves, or AI self-help?
+
+```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 1. An example of Atlassian AI writing code"}
 knitr::include_graphics("images/AI-bug-fix.png")
@@
 
-The list of promises about what AI will do has grown to encompass pretty much everything, from saving the planet from ourselves, all the way down to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023].
-
-Higher-education has not spared the onslaught of promises.
-
-@power_3_2024,^[Rhett Power is a motivational speaker and executive coach] for example, presents three innovative use cases for AI in higher education.
-First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies"].
+The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves, all the way down to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023].
+
+Higher-education has not been spared the onslaught of promises.
+
+@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.] for example, presents three innovative use cases for AI in higher education.
+First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies".].
 AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants.
-AI could also personalize the experience of consumers (also called students) by predicting their final grade with 60-70% accuracy, and by providing corrective and instructive feedback.
-<!--release them in a very literal sense, as one advantage is to do _more_ with limited resources (which in the current fiscal environment means laying people off).
-The cases discussed by Mr.
-Power are questionable: there are already non-AI systems that deploy paperwork; the kind of information needed to identify students at risk of dropping out is probably quite invasive, and therefore impractical or unethical to collect; I very much doubt that it is legal to use AI in hiring and admissions.
---> Mr.
-Power's message to the higher-education sector?
-It is time for institutions to "level up": jump on the bandwagon, fellas, don't let it let you behind!
+AI could also personalize the experience of consumers (also called students) by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.
+
+Mr. Power's message to the higher-education sector? It is time for institutions to "level up": jump on the bandwagon, fellas, don't let it leave you behind!
 
@@
 Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence", whatever that means.] argue that AIs will likely be the primary way to access information in the future.
-For this reason, institutions must equip students with the skills needed to use AI competently, in particular to perfect the art of _prompt engineering_, that is, the ability to craft effective queries that prompt a GenAI to provide the most useful outputs.
+For this reason, institutions must equip students with the skills needed to use AI competently, in particular train them to perfect the art of _prompt engineering_ (the ability to craft effective queries that prompt a GenAI to provide the most useful outputs).
 Universities will use AIs to improve learning and challenge the critical thinking of students.
-Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been essential to understand my own courses; automating the task is another way of cheapening the quality of the learning experience for students].
+Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience.], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been essential to understanding my own courses, what I teach and how I teach it; automating the task of writing syllabi is another way of degrading the quality of instruction that students can expect to receive.].
 
@@
 On February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it.
-Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], that "[faculty] might not be able to stop the AI trend from growing", and so they might as well get on with the program and try to shape it as best they can.
-
-@contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students".
-Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].] **MORE HYPE**
+Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.
+
+Similarly hyperbolic claims aoput AI in higher education can be found peppered throughout the landscape, including by @johnson_ai_2023^[[Gridserve](https://www.gridserve.co.uk/about-us/) is a UK company "dedicated to enhancing the technological capabilities of the education sector through customised information and communication technology (ICT) services".] and @contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students".
+Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].].
+A fairly comprehensive list [from @johnson_ai_2023] is as follows (notice the use of "can" in the original, not "could", or "maybe will"):
+
+- AI can provide personalised learning.  
+- AI-powered systems can provide smart tutoring.  
+- AI can automate grading.  
+- AI can enable virtual classrooms.  
+- AI can provide data insights for informed decisions.  
+- Natural Language Processing (NLP) can enhance language learning.  
+- AI can create interactive and adaptive learning materials.  
+- AI can identify early signs of learning difficulties.  
+- AI can aid in providing personalised professional development opportunities for teachers.
+- AI can streamline administrative tasks.
+
+Is it possible that AIs can really do all this?
 
@@
 A benefit of portraying AI as an _everything tool_ (for the proponents of the technology, at least), is that conversations around it quickly turn into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength.
-In the case of AI, critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making all sorts of claims to begin with^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed [flooding the zone with shit](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation).
-Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot possibly match [@maiberg__google_2024].].
-
-A second discoursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.
-
-In what follows, I do not plan to confront every single possible use case of AI as doing so plays into the intent of the Gish Gallop^[I have already spent close to 50 hours working on this essay, and it is still nowhere near an exhaustive apology or indictment of AI.], which is to defeat criticism by flooding the zone.
-As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple other ways for teaching and learning to be excellent that do not require AI, and do not require the moral, environmental, social compromises demanded by AI[^Most discussions of AI as an everything tool acknowledge ].
+To be sure, AIs _can_ do some things well, but critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed ["flooding the zone with shit"](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation).
+Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot hope to counteract [@maiberg__google_2024].].
+Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they do?
+
+A second discursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.
+
+In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 65 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone.
+As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems.
+For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training.
+Models are trained on biased data and can produce either subtly or overtly biased results.
+And because these biases seem to come from an objective machine, they can be especially pernicious.
+Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems".
+This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highligths the importance of institutions to provide the moral and ethical clarity that may elude individual instructors.].
 
@@
 
-GenAIs are _deep learning models_ [@ibm_what_2021], so called because they are neural networks with _depth_, meaning many, many hidden layers.
+GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning many, many hidden layers.
 As an example, Large Language Models (such as those behind OpenAI's ChatGPT, Microsoft's Copilot, Google's Gemini, and facebook's LLaMA) are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].
@@
 A regression model takes data as inputs, typically a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable).
-With these inputs, the model finds a conditional mean for the output $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables).
+With these inputs, the model finds a conditional mean that becomes the output of the model, or $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables).
 The process of finding that conditional mean is to satisfy some criterion (e.g., that the conditional mean minimizes the error of the model--the distance between $y$ and $\hat{y}$).
@@
 The plot shows the pairs of $y$ and $x$ values, and the line is the model, the conditional mean $\hat{y}$.
-This model needs only two parameters: an intercept (the value of &\hat{y}$ when $x = 0$), and the slope of the line, in other words, the rate of change of $\hat{y}$ with respect to $x$.
+This model needs only two parameters: an intercept (the conditianl mean $\hat{y}$ when $x = 0$), and the slope of the line, in other words, the rate of change of $\hat{y}$ with respect to $x$.
 Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.
@@
 
-```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 1. Simple Example of a Regression Model"}
+```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 2. Simple Example of a Regression Model"}
 ggplot(df,
@@
 
-```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 2. Training data and model output compared"}
+```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 3. Training data and model output compared"}
 
@@
 For instance, the following figure shows a relatively simple neural network trained using using the same data.
-This network is not "deep", as it has only one so called "hidden layer", comprised of two neurons between the input and the output layers.
+This network is not "deep", as it has only one "hidden layer" comprised of two neurons between the input and the output layers.
 In comparison to the linear regression and its two parameters, this neural network uses seven parameters.
-Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the model contigent on the starting conditions.
+Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the results of the model contigent on the starting conditions of the training process.
 ```{r}
@@
 
-```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 3. Simple Example of a Shallow Neural Network"}
+```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 4. Simple Example of a Shallow Neural Network"}
 knitr::include_graphics("images/nn1.png")
@@
 
-```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 4. Training data and model outputs compared: linear regression and shallow neural network"}
+```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 5. Training data and model outputs compared: linear regression and shallow neural network"}
 ggplot() + 
@@
 More complex neural network can be designed.
-The following plot is of a deeper network than the previous one, with two hidden layers, each consisting of five neurons.
+The following plot is of a network deeper than the previous one, with two hidden layers, each consisting of five neurons.
 This model uses forty-six parameters.
@@
 
-```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 5. Example of a Deeper Neural Network"}
+```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 6. Example of a Deeper Neural Network"}
 knitr::include_graphics("images/nn2.png")
@@
 
-```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 6. Training data and model outputs compared: shallow neural network and deeper neural network"}
+```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 7. Training data and model outputs compared: shallow neural network and deeper neural network"}
 ggplot() + 
@@
 
-Here are some key ideas to keep in mind:
+Here are some key ideas to keep in mind when thinking about AIs:
 
@@
 - Their flexibility comes from using "deep" architectures, which means many layers with many, many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters.
-OpenAI no longer reports the number of parameters, but in the latest version this is estimated to be trillions.].
+OpenAI no longer reports the number of parameters, but in the latest version this is estimated to be in the trillions.].
 - AIs are trained using an initial set of random values as parameters. The same training dataset can result in different models due to the randomness in the initial conditions.
@@
 - Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think).
-The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data were used to train LLMs.
+The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data were used to train AIs.
 The increasingly large datasets  essentially expanded the space where the model could interpolate.].
-
-<!--
-
--->
-
-**NOTE**: Labeling data is an essential, yet extremely expensive aspect of training GenAIs.
 
@@
 To begin, we must note the sheer number of examples of the use of AI by students.
-A report by @tyton_partners_generative_2023 that was sponsored by Turnitin found that 50% out of 1,600 post-secondary students who participated in the study, were users of GenAI (but only 22% of 1,000 faculty in the study).
-Use of AI by students is one of the reasons cited by some for universities to embrace AI [e.g., @power_3_2024].
+A report sponsored by Turnitin [@tyton_partners_generative_2023] found that 50% out of 1,600 post-secondary students in the study were users of GenAI (but only 22% of 1,000 faculty).
+Use of AI by students is a commonly cited reason for universities to embrace AI [e.g., @power_3_2024].
 McMaster's own internal research indicates that at least 61% of over 1,300 students polled, used AI during the Fall Term in 2023, 75% of them for school work [@yachouh_student_2024].
@@
 But embracing AI just because the students use it is like putting the cart before the horse.
-After all, smoking is banned on campus, and I do not recall the university asking the students if they smoked^["A tobacco and smoke-free campus is the next important step towards fulfilling our responsibilities as educators, healthcare professionals and to the communities we serve" said then-president Patrick Deane [@carter_mcmaster_2017]].
+After all, smoking is banned on campus, and I do not recall the university asking the students if they smoked^["A tobacco and smoke-free campus is the next important step towards fulfilling our responsibilities as educators, healthcare professionals and to the communities we serve" said then-president Patrick Deane [@carter_mcmaster_2017].].
+A university is supposed to be a place of education (higher, at that), and not a fast-food locale where customers get what they want.
 
@@
 
-## What do GenAI tools do for students?{#what-can-ais-do-for-students}
+## What do GenAIs do for students?{#what-can-ais-do-for-students}
+
+According to McMaster's internal research:
 
@@
 They can form study groups, attend office hours to engage with their instructors, ask their Teaching Assistants, ask relatives and friends, and so on.
-But even for a student who for a whatever reason cannot or will not do any of these things, the fact remains that relatively common concepts have already been explained by humans in dozens if not hundreds of different ways, often in writing that is available through an excellent university library.
+But even for a student who for whatever reason cannot or will not do any of these things, the fact remains that relatively common concepts have already been explained by humans in dozens if not hundreds of different ways, often in writing that is available through an excellent university library.
 A search of McMaster's library catalog, looking only for resources available online (for convenience), reveals that there are hundreds, thousands, and sometimes tens of thousands of resources available to explore the same concept from multiple perspectives.
 
-```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 7. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
+```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 8. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
 knitr::include_graphics("images/mac-library-econometrics.png")
@@
 
-```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 8. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
+```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 9. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
 knitr::include_graphics("images/mac-library-cell-biology.png")
@@
 
-```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 9. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
+```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 10. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
 knitr::include_graphics("images/mac-library-euphemism.png")
@@
 for reasons of lying far from the mean).
-This strongly suggests that any outputs about highly specialized queries should be treated with caution^[The Overview on Generative AI notes that "if you ask [a chatbot] about yourself or a more obscure topic, [it] will almost certainly make up nonsense." This is true even when asking about a person with an extensive online presence (the internet being the likely source of the training datasets); as recently as 2023, a famous law professor discovered when ChatGPT produced output stating that he had been accused of sexual harassment (in the real world he had not been) [@verma_chatgpt_2023]].
+This strongly suggests that any outputs about highly specialized queries should be treated with caution^[The Overview on Generative AI notes that "if you ask [a chatbot] about yourself or a more obscure topic, [it] will almost certainly make up nonsense." This is true even when asking about a person with an extensive online presence (the internet being the likely source of the training datasets).
+As recently as 2023, a famous law professor discovered that he was being named as a sexual harasser by ChatGPT, when in reality he had never been accused of such conduct [@verma_chatgpt_2023].].
 
@@
 An emphasis on practice questions, however, signals an underlying weakness of the process, namely, a focus on learning for the examination, as opposed to learning for its own sake.
-A growing body of evidence points at the effectiveness of alternative assessment schemes, such as continuous assessment through course work, and even seen examinations instead of the traditional unseen examinations [@turner_see_2018], the latter form of assessment continues to dominate the evaluation landscape [@buckley_are_2023].
-Ungrading, in particular, is increasingly recognized as its potential to internalize the motivation to learn, and for fostering adaptability, creative thinking, and self-management [@gorichanaz_it_2024].
-
-Two relevant issues appear to be relevant.
-
-First, posing interesting questions for students to work on is an essential part of an instructor's job, but this part of the job becomes compromised when instructors operate in massive classrooms situations^[Average class sizes at McMaster (with a full time under-graduate student population of 30,400) are 127.8 in first/second year, and 52.3 in third/fourth year; this places McMaster dead last among all Canadian institutions in class sizes in first/second year, and second to last after UBC in third/fourth year [@macleans_canadas_2022]].
+A growing body of evidence points at the effectiveness of alternative assessment schemes, such as continuous assessment through course work, and even seen examinations instead of the traditional unseen examinations [@turner_see_2018].
+Alas, unseen examinations continue to dominate the evaluation landscape [@buckley_are_2023].
+Ungrading, in particular, is increasingly recognized for its potential to help learners internalize the motivation to learn, and for fostering adaptability, creative thinking, and self-management [@gorichanaz_it_2024].
+
+Two issues appear to be relevant.
+
+First, posing interesting questions for students to work on is an essential part of an instructor's job, but this part of the job becomes compromised when instructors operate in massive classrooms situations^[Average class sizes at McMaster (with a full time under-graduate student population of 30,400) are 127.8 in first/second year, and 52.3 in third/fourth year; this places McMaster dead last among all Canadian institutions in class sizes in first/second year, and second to last after UBC in third/fourth year [@macleans_canadas_2022].].
 In this case, AI is a substitute, and demonstrably not a good one, for a qualified instructor who can intellectually challenge their students.
@@
 Research by @vahed_continuous_2023 suggests that instructors need to develop assessment literacy, that is, an understanding of "the fundamental assessment concepts and procedures, and to engage in the ongoing process of adjusting instructional strategies to enrich students learning experiences".
-There is a need for innovative approaches in assessment, but it is unlikely that AI is the tool to develop the assessment literacy that they require.
+There is a need for innovative approaches in assessment, but it is unlikely that AI is the tool to develop the literacy that they require.
 
@@
 
-This is an unrealistic expectation.
-More realistically, students could have the opportunity to discuss their drafted work with instructors before final submission.
-A conscientious instructor will invest time reading and understanding the draft work, to be able to provide quality feedback.
-But there is a more insidious aspect about getting feedback from an AI...who, actually, is providing the feedback?
-What data were used to train the machine that provides the feedback?
-Who fine-tuned the machine who provides the feedback?
-Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided?^[For instance, [@verma_chatgpt_2023]] (More on this below).
-
-- Personalized
+This is an utterly unrealistic expectation (also see automated grading below).
+Ideally, students should have the opportunity to discuss their drafted work with instructors before final submission (or again, could ask members of their study groups to proof-read their work, among many other ways to get feedback before submission).
+A conscientious instructor will invest time reading and understanding the draft work, so as to be able to provide quality feedback.
+But there is a more insidious aspect about getting feedback from an AI...who, actually, provides the feedback in this case?
+Whose data are used to train the machine that provides the feedback?
+Who fine-tunes the machine that provides the feedback?
+Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided?^[[Automation bias](https://en.wikipedia.org/wiki/Automation_bias), the human inclination to favor suggestions made by automated decision-making systems, means that even when a human takes responsibility for decisions, AI companies remain unaccountable for their role in influencing outcomes.
+A poignant example involving AI-based face recognition and online proctoring is related by @hill_accused_2022.].
+
+- Personalized learning
+
+Is learning really personalized if there is no person on the teaching side of things?
+This sounds like a gimmick to put a little lipstick on the underfunded massification of higher education.
+A competent instructor who is not tasked with talking at hundreds of anonymous faces in an auditorium can provide personalized learning opportunities to students, in ways that a plausible-sounding machine that regresses to the mean cannot do.
+
+- Creation of interactive and adaptive learning materials
 
 @mollick_instructors_2024^[Drs.
-Eric Mollick and Lilach Mollick are members of Wharton Interactive, in the business school of the University of Pennsylvania.
-Wharton Interactive offers to "democratize education" so to level up your learning!
-[@kelner_level_2022]] envision a future in which educators are innovators thanks to their adroit use of prompts for AIs.
+Eric Mollick and Lilach Mollick are members of Wharton Interactive in the business school of the University of Pennsylvania.
+Wharton Interactive offers to "democratize education" to help you level up your learning!
+(Discounts of up to 50\% are available to alumni) [@kelner_level_2022].] envision a future in which educators are innovators thanks to their adroit use of prompts for AIs.
 According to these authors, GenAI stands alone among educational technologies in that it can be "'programmed' through prompts alone", which should allow even instructors "without extensive technology or coding experience" to more easily create classroom applications.
 
-## What do GenAI tools do to students? {#what-can-ais-do-to-students}
+- Automated grading
+
+There is nothing new about automated grading: Scantron sheets for multiple choice examinations have existed for over four decades.
+They work well because the algorithm for grading multiple choices is simple, and importantly, it is deterministic (i.e., there is no randomness involved).
+There is no reason to use a neural network to grade multiple choice examinations.
+As well, there is no evidence at all that a neural network can do better at grading anything more complex than a multiple choice examination, especially since 1) it will tend to consider things that are closer to the mean as "better"; and 2) the amount of training required for a classification task of this nature is massive^[Labeling data is an essential, yet extremely expensive aspect of training GenAIs.
+**NOTE** expand on this.].
+
+- Writing reference letters
+
+Using a chatbot to write reference letters must be either one of the most pointless or more dishonest uses of a chatbots in higher education.
+Imagine that the reference letter honestly includes a disclosure that ChatGPT was used to write it^[As it should, according to McMaster's guidelines for acknowledging the use of AI.].
+If the objective of a reference letter is for someone to attest from the vantage point of some level of expertise as to the skills and qualities of a person, a chatbot-written letter is probably worse than useless: what message does the disclaimer send to the recipient of the letter?
+And how much stock will anyone place on the veracity of the letter?
+To avoid this catch, one would have to occult or obscure the AI assist, and in doing so would end up passing text written by countless others and filtered through a machine, as if it was their own.
+
+- qualitative and quantitative data analysis
+
+- AI-powered systems can provide smart tutoring.  
+- AI can enable virtual classrooms.  
+- AI can provide data insights for informed decisions.  
+- Natural Language Processing (NLP) can enhance language learning.  
+- AI can create interactive and adaptive learning materials.  
+- AI can identify early signs of learning difficulties.  
+- AI can aid in providing personalised professional development opportunities for teachers.
+- AI can streamline administrative tasks.
+
+## What do GenAIs do to students? {#what-can-ais-do-to-students}
 
@@
 It can help someone to accomplish a task more easily, but it will not help if used constantly.
-However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle that the end user cannot hope to learn^[Reports of people getting their fingers crushed by Tesla's cybertruck could be a result of 1) people trusting the technology more than they should; and 2) not understanding the technology: apparently the algorithm uses increasing force to close the trunk if it fails to latch on back-to-back cycles.
-Hence, the trunk will not slash a banana on its first attemt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature (in the sense that neural networks are blackboxes, and the development of AIs itself is a blackbox where things change based on the interests^[Sometimes the interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that those interests will always be align with the public good, and in fact, they almost never will be as long as there are only a handful of American companies that offer an "essential" service.] of their real developers, not those who fancy themselves as able to create apps without the need to learn how to code.).
+However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle that the end user cannot hope to master^[Reports of people getting their fingers crushed by Tesla's cybertruck could be a result of 1) people trusting the technology more than they should; and 2) not understanding the technology: apparently the algorithm uses increasing force to close the trunk if it fails to latch on back-to-back cycles.
+Hence, the trunk will not slash a banana on its first attemt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature (in the sense that neural networks are blackboxes, and the development of AIs itself is a blackbox where things change based on the interests of their corporate developers^[Sometimes said interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that corporate interests will always align with those of the public, and in fact, they seldom will as long as there are only a handful of American companies that offer an "essential" service.], and not the interests of their relatively small customers^[And this refers not only to individuals playing with ChatGPT and MidJourney.
+In 2022, McMaster University operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD).
+Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022.
+To put things in perspective (and using the average exchange rate of 2022), if McMaster used all of its 2022 revenue to pay Microsoft, that would leave McMaster with zero money and still account for less than _one third of one percent point_ of Microsoft's revenue.
+A mid-sized Canadian university is puny compared to the economic might of Microsoft.] who fancy themselves creators but never learned (or forgot) how to code.).
+
+<!--
+McMaster: 842,899,000 CAD
+Microsoft: 198,000,000,000 USD == 257,776,200,000 CAD
+
+https://www.exchange-rates.org/exchange-rate-history/usd-cad-2022
+Avg. exchange rate in 2022: 1.3019 CAD/US
+
+This is how much bigger Microsoft is compared to McMaster. 
+(198000000000 * 1.3019)/842899000
+
+Put another way: If McMaster gave ALL OF ITS REVENUE to Microsoft, it would only account for 842899000 / (198000000000 * 1.3019) or less than one third of a percentage point of Microsoft's revenue.
+--->
 
@@
 
-Mind you, these, and numerous other hapless results^[For instance, ; or returning misinformation about depresion [@tenbarge_glue_2024]] (as of May, 2024) are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
+Mind you, these, and numerous other hapless results^[Including AI overviews with misinformation about depression [@tenbarge_glue_2024].] (as of May, 2024) are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
 To what extent can more money improve AI?
@@
 
-## What do GenAIs do to everyone else? {#what-do-ais-do-to-everyone-else}
+@berber_peter_2024: "venture-capital firm Sequoia estimated [in March 2024] that the AI industry spent \$50 billion [in 2023] on the Nvidia chips used to train advanced AI models...but brought in only $3 billion in revenue."
+
+Charlie Pierce [writes](https://www.antipope.org/charlie/blog-static/2024/04/the-radiant-future-of-1995.html#more) that "the industry is running at a loss on power consumption alone, never mind labour costs (which are quite high: those generative LLMs require extensive human curation of the input data they require for training)".
+
+"The common feature of these bubbles is a shitload of hype and promises from hucksters who fail to deliver a viable product but suck up as much investment capital as they can.
+A handful of them survive: from dot-com 1.0, the stand-outs are Amazon and Google (Facebook, Twitter, Reddit, et al came along much later—social media was a later, smaller bubble).
+Other survivors include Paypal, eBay, and Doubleclick (the latter being merged with Google to form a monstrous global advertising monopoly).
+The survivors tend to leave behind infrastructure: the failures leave behind t-shirts, second hand Aeron chairs, and motivational posters."
+
+https://futurism.com/the-byte/google-bard-ai-ready NOT READY, SO WHY RELEASE IT?
+
+## What do GenAIs do to everyone and everything else? {#what-do-ais-do-to-everyone-else}
 
@@
 
+I struggle a bit with this one.
+For example, I fly more often than the [typical Canadian](https://ourworldindata.org/grapher/air-trips-per-capita?tab=table&time=earliest..latest&country=~CAN) and a little bit less often than the [typical world traveller](https://www.news24.com/news24/around-the-world-in-50-hours-average-person-takes-65-flights-per-year-20180828) (but I have not owned a car in decades).
+Who am I to dictate how much damage _other people_ have a right to inflict (or not) to the environment.
+
 Microsoft: Here's how burning the planet will help to save it.
@@
 "Environmental costs of AI are soaring" [@crawford_generative_2024]
+
+https://www.wired.com/story/the-generative-ai-search-race-has-a-dirty-secret/
 
@@
 Or when trying to obtain accurate voting information in the context of voter supression?
-[@angwin_seeking_2024]
+[@angwin_seeking_2024].
 
@@
 When using a Generative AI we must ask whose perspectives the tool will tend to reproduce and amplify.
-
-Training data, much like technology, are not politically neutral, and several processes contribute to reproduce malignant biases, A way to explain this is by means of the concept of _platformed racism_ [@matamoros-fernandez_platformed_2017].
-Platformed racism is the process whereby the functionalities of platforms are exploited to create and spread racist ideology (liking and sharing), often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has had a wide reach [@treyger_russian_2022] and furthermore is known to seed distrust between ethnic communities [@svedkauskas_analysis_2020].
-As the U.S.
-Department of State Reports, Russian intelligence is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S.
+In this sense, numerous observers [e.g., @ahuja_we_2023; @stross_tech_2023; @gebru_tescreal_2024] have noted the ideological leanings of various actors in the AI scene, an elite world of tech investors^[A group that author Charles Stross colorfully [characterizes](https://www.antipope.org/charlie/blog-static/2024/04/the-radiant-future-of-1995.html) as "fascist-adjacent straight white males with an unadmitted political agenda".
+This characterization is somewhat unfair: both [Sam Altman](https://www.advocate.com/news/sam-altman-openai-exclusive) and [Peter Thiel](https://www.npr.org/2016/07/21/486966882/i-am-proud-to-be-gay-tech-investor-peter-thiel-tells-gop-convention) are openly gay.] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these believes is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars, but not before super-intelligent AIs destroy humanity.], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the wellbeing of trillions of humans in the future is well worth the sacrifices of a few billions of humans today.
+Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that wealthy and well connected longtermists are willing to make.]).
+
+One does not need to posit a conspiracy theory^[Accusing the critics of being conspiracy theorists, as @sennesh_conspiracy_2023 do, distracts from the fact that there is no dearth of leftist ideas for a more just future; intellectual poverty is not what is at stake, instead a key difference is that these ideas seldom have the power of money behind them, in contrast to, say, effective altruism [@matthews_how_2022].] to worry about the influence (whether coordinated or not) of a small group of men (they are almost all of them men^[After Meta disbanded its Responsible AI Division, the company formed an AI advisory council composed of four very wealthy white men [@davis_metas_2024;@goel_meta_2024].]) who collectively can muster more than _half a trillion dollars_ as they try to implement their ideas [@stross_tech_2023].
+
+But leaving aside the question of a potential "unadmitted political agenda" by an ultra-wealthy elite, there is the much more immediate problem that training data, much like technology, are not politically neutral.
+Several processes have been identified that contribute to reproduce malignant biases, including the concept of _platformed racism_ [@matamoros-fernandez_platformed_2017].
+Platformed racism is a process whereby the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has had a wide reach [@treyger_russian_2022] and furthermore is known to seed distrust between ethnic communities [@svedkauskas_analysis_2020].
+A U.S.
+Department of State report notes that Russian intelligence is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S.
 politics [@us_department_of_state_gec_2024, p.
@@
 
-Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
-In addition to being misleading to the unadvised, AI is also a stupendous tool ripe for abuse by those who are advised _and_ have an agenda.
-Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than using AI to identify misinformation [@maiberg__google_2024].].
+Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
+In addition to being misleading to the unadvised, AI is also a fantastic tool ripe for abuse by those who are advised _and_ have an agenda.
+Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat misinformation [@maiberg__google_2024].].
 In another wretched incident, Google's AI overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Obama [@tenbarge_glue_2024].
@@
 
-ChatGPT was an awesome feat of marketing [@cowen_chatgpt_2023], but much coverage of the relevant issues in the media has been so credulous it has stretched credulity past Pluto [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.].
-There are now calls to press pause on the hype [@angwin_press_2024] but these are still drown by the credulous coverage.
+Neural networks and machine learning are not new, and they have their uses. But those uses have been drown, and to some extent discredited, by the current hype around AI.
+
+ChatGPT was a colossal feat of marketing [@cowen_chatgpt_2023], but much coverage of AI in the media has been stretched beyond the breaking point of credulity [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.].
+There are now calls to press pause on the hype [@angwin_press_2024] but these are still drown by the uncritical coverage.
+
+> An end to “batch” teaching.
+Instead of working with cohorts of learners in grades or classes, individual learning agendas are developed based on an assessment of the skills, competencies and capabilities the learner already has as well as ongoing conversations about the purpose of the learning.
+Where does the student want the learning to take them?
+AI systems, coupled with coaching, mentoring and some instruction and peer support networks then ensure the learner is supported on the carefully mapped learning journey.
+Teaching can still take place: Students challenged with a set of related issues can be brought together for a boot camp or focused period of instruction before returning to their individualized study activities.
+
+>An end to “time” as a learning metric.
+Currently, undergraduate courses are described in terms of credit hours.
+Learning and assessment on demand, fast-tracking or slow-tracking using adaptive assessment, and modular learning when the student is ready liberates learners from the time-based metric.
+
+> An end to exams as we know them.
+Exams are an efficient way to judge the output of batch teaching but are not good indicators of learner knowledge, skills and capabilities.
+With AI-enabled assessment generation, automated marking and assessment on demand, exams are no longer needed.
+More authentic assessments that are project-based, team challenges, work-based learning activities, and student self-directed projects can become the norm.
+
+>Increased feedback and advising.
+AI systems do not have “office hours.” They are available 24x7 and never tire of finding new ways of responding to student queries and needs.Subject focused chatbots which have been “fed” all necessary information about a subject will do an excellent job of subject-matter advising and teaching.
+Analytics suggest that some students need to be supported by teachers or others with special skills, such as writing coaches, statistical support workers and numeracy advisors.
+The aim will be to ensure every learner follows and is successful in the learning journey they have contracted to complete.
+
+> Greater accessibility.
+Higher education systems are organized around blocks of time: semesters or terms.
+There is no longer good reason for this.
+Students should be able to begin their learning at any time and be assessed when they are ready.
+A four-year undergraduate degree could easily be completed in 16 months by a diligent student, for example, or in even less time by someone arriving in the program with substantial life and work experience.
+AI systems can also leverage technologies to ensure the resources and supports for learners reflect who they are (culture, language, experience and skill) and what challenges they face (disabilities, learning styles, etc.).
+
+It has now been 55 years since Howard S.
+Barrows and Robyn M.
+Tamblyn [pioneered](https://en.wikipedia.org/wiki/Problem-based_learning#History) problem-based learning at McMaster, an accomplishment that the university [still touts today](https://provost.mcmaster.ca/teaching-learning/).
+Problem-based learning was not and is still not high tech, and yet remains a great pedagogical approach, influential to the date.
+Developing problem-based learning took human ingenuity and insight, not getting hooked on an expensive technology of questionable value today, of dubious longevity, and that in the here and now is unarguably extractive, abusive, and exploitative, and that advances an agenda that can only be squared with "integrity, quality, and excellence" via fabulous contortions of the imagination.
+
+Does the university imagine that it will be best known in half a century as "McMaster, the institution that pioneered the use of ChatGPT"?
+Is that what it really wants?
+[Regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean), after all, is also called a "reversion to mediocrity" [@barnett_regression_2017], and that is what an AI campus has to offer.
 
 I suggest that universities should avoid the mistakes of the media, an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) (also see the "pivot-to-video" and how tech companies suckered many media companies, including ).
+Tech corporations (or least ways some of them), can still hope to recoup some of their investment in AI.
+But a university gambles its most important asset when betting on AI: its reputation as a place were knowledge is created, and as a custodian of knowledge.
+AIs already have a reputation for producing mis- and disinformation--from the very titans of the field^[Google could not care less about its reputation.
+It is [too big to care](https://pluralistic.net/2024/04/04/teach-me-how-to-shruggie/#kagi).] to the unaware^[], and beyond, to bad-faith actors.
+In my view, there are no efficiencies, no promises of future excellence, that can compensate the risk of ruining an institutions reputation.
 
@@
 Universities main source of power is their reputation as places that preserve and expand knowledge in a principled way.
-"This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an institution of higher-education is not needed when there are cheaper alternatives for such an education ("act now before you are laid off!"):
-
-```{r ligemy-ai-bundle, out.width="50%", fig.cap="\\label{fig:ai-success-bundle}An AI success bundle from the \"The Real People of Ligency PR and Marketing Team\""}
-bundle_plot <- ggplot() +
-  labs(caption = "https://www.udemy.com/user/ligency-team/") +
-  theme_void()
-
-ggdraw() + 
-  draw_plot(bundle_plot) +
-  draw_image(image = "images/trail-blazing-ai-bundle.png")
-```
-
-"McMaster is best known for our pioneering work in problem-based learning, which began in our medical school in 1969 and has been adopted across the world." https://provost.mcmaster.ca/teaching-learning/
-
-Do we really want to be best known in 40 years as "McMaster, the institution that pioneered the use of ChatGPT"?
-
-Recall that [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean) is also called "reversion to mediocrity" [@barnett_regression_2017].
+"This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an institution of higher-education is not needed when there are cheaper alternatives for such an education ("act in the next four days to proof yourself against layoffs!"):
+
+![](images/trail-blazing-ai-bundle.png)
 
@@
 (see https://www.defenseone.com/technology/2024/05/ai-and-other-tech-creating-special-operations-renaissance/396388/)
+
+Peter Thiel on "AI is a military technology" https://www.businessinsider.com/peter-thiel-claim-ai-primarily-military-technology-experts-disagree-2019-8
+
+Crypto-currency to AI: https://en.wikipedia.org/wiki/Cognition_AI
+
+## Bonus
+
+```{r googly-eyes, out.width="50%", fig.cap="\\label{fig:googly-ai-eyes}Figure XX. In conclusion"}
+google_plot <- ggplot() +
+  labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
+  theme_void()
+
+ggdraw() + 
+  draw_plot(google_plot) +
+  draw_image(image = "images/ai-googly-eyes.png", width = 0.8)
+```
 
@@
 
-### Panel tasks{#ai-panel-tasks}
-
-The task of the Artificial Intelligence Advisory Committee Expert Panel on Teaching and Learning was initially defined as follows by the Vice-Provost Teaching and Learning in an email dated April 16:
-
-<!--https://app.fastmail.com/mail/search:erin/T6fae26856703ef40.Mfe5dff646c6ae452b84ef4bf-->
-
-> [T]he first task of this Expert Panel will be to provide recommendations to the AI Advisory Committee on updates to the Provisional Guidelines on the Use of Generative AI in Teaching and Learning.
-
-<!--This does not say anything about what the Committee will do with the recommendations; I cannot find anything about this.-->
+### AI Teaching and Learning Panel priority initiatives {#ai-panel-tasks}
+
+The [Terms of Reference](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/ai-advisory-committee/) of the Artificial Intelligence Advisory Committee identify the following priority initiatives for the Expert Panel on Teaching and Learning:
+
+- Provide recommendations to the AI Advisory Committee on updates to the Provisional Guidelines on the Use of Generative AI in Teaching and Learning.  
+- Review and recommend training, resources and professional development on AI literacy for both educators and students (e.g. grants, microcredentials, guidebooks).  
+- Engage in dialogue with both students and educators to understand their perspectives on AI in learning.  
+- Endorse ethical standards and guidelines for AI usage in teaching environments.  
+- Review and provide insights into assessments or tools that leverage AI for student evaluation.  
+- Engage in cross-institutional dialogues to share insights and learn from global best practices.  
+- Advocate for student involvement in the design and critique of AI-enabled teaching tools.
 

### 2024-06-01 README.Rmd
More writing

@@
 However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle that the end user cannot hope to master^[Reports of people getting their fingers crushed by Tesla's cybertruck could be a result of 1) people trusting the technology more than they should; and 2) not understanding the technology: apparently the algorithm uses increasing force to close the trunk if it fails to latch on back-to-back cycles.
-Hence, the trunk will not slash a banana on its first attemt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature (in the sense that neural networks are blackboxes, and the development of AIs itself is a blackbox where things change based on the interests of their corporate developers^[Sometimes said interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that corporate interests will always align with those of the public, and in fact, they seldom will as long as there are only a handful of American companies that offer an "essential" service.], and not the interests of their relatively small customers^[And this refers not only to individuals playing with ChatGPT and MidJourney.
+Hence, the trunk will not slash a banana on its first attemt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature (in the sense that neural networks are blackboxes, and the development of AIs itself is a blackbox where things change based on the interests of their corporate developers^[Sometimes said interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that corporate interests will always align with those of the public, and in fact, they seldom will as long as there are only a handful of American companies that offer an "essential" service.], and not the interests of their relatively small customers^[And this refers not only to individuals whose skills are yoked to ChatGPT and MidJourney.
 In 2022, McMaster University operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD).
 Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022.
-To put things in perspective (and using the average exchange rate of 2022), if McMaster used all of its 2022 revenue to pay Microsoft, that would leave McMaster with zero money and still account for less than _one third of one percent point_ of Microsoft's revenue.
-A mid-sized Canadian university is puny compared to the economic might of Microsoft.] who fancy themselves creators but never learned (or forgot) how to code.).
+To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero money and still account for less than _one third of one percent point_ of Microsoft's revenue.
+A mid-sized Canadian university is like an ant compared to the economic boot of Microsoft.] who fancy themselves creators but never learned (or forgot) how to code.).
 
@@
 ```
+
+Tools to posion AI models:
+
+@salman_raising_2023
+
+@shan_nightshade_2024
+
+@xu_shadowcast_2024
 
@@
 
+https://www.theguardian.com/commentisfree/article/2024/may/30/ugly-truth-ai-chatgpt-guzzling-resources-environment
+
 "Environmental costs of AI are soaring" [@crawford_generative_2024]
@@
 In this sense, numerous observers [e.g., @ahuja_we_2023; @stross_tech_2023; @gebru_tescreal_2024] have noted the ideological leanings of various actors in the AI scene, an elite world of tech investors^[A group that author Charles Stross colorfully [characterizes](https://www.antipope.org/charlie/blog-static/2024/04/the-radiant-future-of-1995.html) as "fascist-adjacent straight white males with an unadmitted political agenda".
-This characterization is somewhat unfair: both [Sam Altman](https://www.advocate.com/news/sam-altman-openai-exclusive) and [Peter Thiel](https://www.npr.org/2016/07/21/486966882/i-am-proud-to-be-gay-tech-investor-peter-thiel-tells-gop-convention) are openly gay.] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these believes is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars, but not before super-intelligent AIs destroy humanity.], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the wellbeing of trillions of humans in the future is well worth the sacrifices of a few billions of humans today.
+This characterization is somewhat unfair: both [Sam Altman](https://www.advocate.com/news/sam-altman-openai-exclusive) and [Peter Thiel](https://www.npr.org/2016/07/21/486966882/i-am-proud-to-be-gay-tech-investor-peter-thiel-tells-gop-convention) are openly gay.
+On the other hand, their political agenda hides in plain sight.
+@thiel_education_2009 has written about how he "no longer believe[s] that freedom and democracy are compatible".
+Elon Musk, well, Elon Musk once shared a meme comparing Canada's PM Justin Trudeau to Hitler, when defending the vandals who [terrorized Ottawa](https://fortune.com/2022/01/27/elon-musk-twitter-freedom-convoy-canadian-truckers-rule-covid-vaccine/) in 2022.] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars...just not before super-intelligent AIs destroy humanity.], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the wellbeing of trillions of humans in the future is well worth the sacrifices of a few billions of humans today.
 Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that wealthy and well connected longtermists are willing to make.]).
@@
 This dismissal, though, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do).
-Responses like fiddling with the model (fine tuning it) to provide more palatable outputs (like in the case of **FIND EXAMPLE**) are not more reassuring.
+Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not more reassuring.
 Who is to say what a handful of unaccountable corporations will find palatable tomorrow, let alone in the face of powerful authoritarian regimes?
@@
 But a university gambles its most important asset when betting on AI: its reputation as a place were knowledge is created, and as a custodian of knowledge.
-AIs already have a reputation for producing mis- and disinformation--from the very titans of the field^[Google could not care less about its reputation.
-It is [too big to care](https://pluralistic.net/2024/04/04/teach-me-how-to-shruggie/#kagi).] to the unaware^[], and beyond, to bad-faith actors.
+AIs already have a reputation for producing mis- and disinformation--from the very titans of the field^[Entities like Google and Amazon could not care less about their reputation.
+They are [too big](https://www.businessinsider.com/ex-amazon-ghaderi-exec-suing-ai-race-copyright-allegations-2024) [to care](https://pluralistic.net/2024/04/04/teach-me-how-to-shruggie/).] to the unaware^[Like a B.C.
+lawyer who was [reprimanded](https://www.cbc.ca/news/canada/british-columbia/lawyer-chatgpt-fake-precedent-1.7126393) for using AI hallucinations in a legal filing.
+This was not even the first time that supposedly highly skilled professionals were castigated for [using](https://www.washingtonpost.com/technology/2023/11/16/chatgpt-lawyer-fired-ai/) [chatbots](https://www.reuters.com/legal/new-york-lawyers-sanctioned-using-fake-chatgpt-cases-legal-brief-2023-06-22/).], and beyond, to bad-faith actors^[Like rival geopolitical powers that are relatively impervious to propaganda themselves for reason of being [authoritarian countries](https://www.theatlantic.com/magazine/archive/2024/06/china-russia-republican-party-relations/678271/).].
 In my view, there are no efficiencies, no promises of future excellence, that can compensate the risk of ruining an institutions reputation.

### 2024-06-15 README.Rmd
More work

@@
 - [AI in teaching and learning](#ai-in-teaching-and-learning)  
-- [What do GenAIs do for students?](#what-can-ais-do-for-students)  
+- [What do GenAIs do for students?](#what-do-ais-do-for-students)  
 - [What do GenAIs do to students?](#what-can-ais-do-to-students)  
@@
 ```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 1. An example of Atlassian AI writing code"}
-knitr::include_graphics("images/AI-bug-fix.png")
+knitr::include_graphics("images/ai-bug-fix.png")
 ```
@@
 The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves, all the way down to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023].
-
-Higher-education has not been spared the onslaught of promises.
-
-@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.] for example, presents three innovative use cases for AI in higher education.
+Not even the [bowel movements of our pets](https://finance.yahoo.com/news/next-generation-ai-powered-cat-140000795.html) are beyond the revolutionary all-seeing eye of AI.
+
+```{r ai-litter-box, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 2. Revolutionizing pet health"}
+knitr::include_graphics("images/ai-litter-box.png")
+```
+
+The forecast that AI will improve everything (from the natural environment of the globe to the dumps of our cats) was certain to, sooner or later, reach higher-education, a target-rich environment for selling hype--and indeed, colleges and universities have not been spared the onslaught of promises.
+Examples of this abound.
+
+@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.], for one, presents three innovative use cases for AI in higher education.
 First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies".].
 AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants.
-AI could also personalize the experience of consumers (also called students) by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.
+AI could also personalize the experience of consumers (occasionally also called students) by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.
 
@@
 Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9).
-On February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it.
-Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.
-
-Similarly hyperbolic claims aoput AI in higher education can be found peppered throughout the landscape, including by @johnson_ai_2023^[[Gridserve](https://www.gridserve.co.uk/about-us/) is a UK company "dedicated to enhancing the technological capabilities of the education sector through customised information and communication technology (ICT) services".] and @contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students".
+It did not take long for this prediction to turn sourer than beans in August: on February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it.
+Shortly after that, Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.
+
+Similarly hyperbolic claims about AI in higher education can be found peppered throughout the landscape, including by @johnson_ai_2023^[[Gridserve](https://www.gridserve.co.uk/about-us/) is a UK company "dedicated to enhancing the technological capabilities of the education sector through customised information and communication technology (ICT) services".] and @contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students".
 Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].].
@@
 
-- AI can provide personalised learning.  
+- AI can provide personalized learning.  
 - AI-powered systems can provide smart tutoring.  
@@
 
-Is it possible that AIs can really do all this?
+Is it really possible that AIs can do all this?
 
@@
 Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot hope to counteract [@maiberg__google_2024].].
-Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they do?
+Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they really do do?
 
@@
 
-In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 65 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone.
+In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 70 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone.
 As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems.
@@
 Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems".
-This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highligths the importance of institutions to provide the moral and ethical clarity that may elude individual instructors.].
+This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the importance of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].
 
@@
 
-GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning many, many hidden layers.
+GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning many, many, _many_ hidden layers.
 As an example, Large Language Models (such as those behind OpenAI's ChatGPT, Microsoft's Copilot, Google's Gemini, and facebook's LLaMA) are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].
@@
 In more conventional terms, neural networks are regression models--much more sophisticated and powerful than linear regression to be sure, but regression models nonetheless [@ripley_neural_1994].
-Here, it is important to note that the term regression refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing a prompt to a "statistically probable" outcome.
+Here, it is important to note that the term regression refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing an input (e.g., a prompt) to a "statistically probable" outcome.
 
@@
 
-```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 2. Simple Example of a Regression Model"}
+```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 3. Simple Example of a Regression Model"}
 ggplot(df,
@@
 
-```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 3. Training data and model output compared"}
+```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 4. Training data and model output compared"}
 
@@
 In comparison to the linear regression and its two parameters, this neural network uses seven parameters.
-Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the results of the model contigent on the starting conditions of the training process.
+Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the results of the model contingent on the starting conditions of the training process.
 ```{r}
@@
 
-```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 4. Simple Example of a Shallow Neural Network"}
+```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 5. Simple Example of a Shallow Neural Network"}
 knitr::include_graphics("images/nn1.png")
@@
 
-The plot below illustrates the predictions of this neural network, where we can see that the model is not limited to a linear response; neural networks are more flexible than linear regression, and they achieve this by being less parsimonious (they use more parameters).
-
-```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 5. Training data and model outputs compared: linear regression and shallow neural network"}
+The plot below illustrates the predictions of this neural network, where we can see that the model is not limited to a linear response.
+Neural networks are more flexible than linear regression, and they achieve this by being less parsimonious (they use more parameters).
+
+```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 6. Training data and model outputs compared: linear regression and shallow neural network"}
 ggplot() + 
@@
 
-```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 6. Example of a Deeper Neural Network"}
+```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 7. Example of a Deeper Neural Network"}
 knitr::include_graphics("images/nn2.png")
@@
 
-The increased complexity leads to greater flexibility, and we can see that the model makes predictions that resemble more closely the actual data.
-
-```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 7. Training data and model outputs compared: shallow neural network and deeper neural network"}
+The increased complexity leads to greater flexibility, and we can see that the model's predictions resemble more closely the actual data.
+
+```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 8. Training data and model outputs compared: shallow neural network and deeper neural network"}
 ggplot() + 
@@
 Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than the number of data points that exist in the dataset used to train the model.
-Roughly, every additional parameter needs a data input to train on, and when there are as many parameters as data inputs the model is overfit and will predict each value of the independent variable in the training dataset perfectly.
+Roughly, every additional parameter needs a data input to train on, and when there are as many parameters as data inputs the model becomes overfit and will predict each value of the independent variable in the training dataset perfectly.
 The model will fail to train if there are more parameters than data points.
 
-In a neural network trained to work with natural language, the "predictors" are words that come before and after a certain word.
-Several approaches exist to _embed_ a bag of words, using large text corpora the co-occurrence of words is used to convert words to numerical vectors.
-These vectors are decontextualized, that is, they only reflect the position of a given word with respect to other words in the training corpora; however, the numeric values are static even if the surrounding words in a specific prompt change.
-A recent breakthrough in this field is the development of _transformers_, a network architecture architecture that updates the numeric values of a word depending on the words that come before and after.
+In a neural network trained to work with natural language, the "predictors" are the words that come before and after a certain word.
+Several approaches exist to _embed_ a bag of words, which means using large text corpora to calculate the frequency of co-occurrence of words.
+These frequencies are then converted to numerical vectors that represent the position of a word in multidimensional space.
+These "coordinates" in the space of the vocabulary can then be used to calculate the "distances" between words.
+Alas, these vectors are decontextualized, that is, they only reflect the position of a given word with respect to other words in the training corpora, and the numeric values are static, even if the surrounding words in a specific prompt change.
+A recent breakthrough in the field of natural language processing is the development of _transformers_, a network architecture architecture that updates the numeric values of a word depending on the words that come before and after in a prompt.
 
@@
 - Their flexibility comes from using "deep" architectures, which means many layers with many, many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters.
-OpenAI no longer reports the number of parameters, but in the latest version this is estimated to be in the trillions.].
+OpenAI no longer reports the number of parameters of its newer models, but in the latest version this is estimated to be in the trillions.].
+The large number of parameters makes these models _extremely_ data hungry.
 - AIs are trained using an initial set of random values as parameters. The same training dataset can result in different models due to the randomness in the initial conditions.
@@
 The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data were used to train AIs.
-The increasingly large datasets  essentially expanded the space where the model could interpolate.].
+The increasingly large datasets essentially expand the space where the model could interpolate.].
 
@@
 Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it makes sense to use this technology.
-To begin, we must note the sheer number of examples of the use of AI by students.
-A report sponsored by Turnitin [@tyton_partners_generative_2023] found that 50% out of 1,600 post-secondary students in the study were users of GenAI (but only 22% of 1,000 faculty).
-Use of AI by students is a commonly cited reason for universities to embrace AI [e.g., @power_3_2024].
-McMaster's own internal research indicates that at least 61% of over 1,300 students polled, used AI during the Fall Term in 2023, 75% of them for school work [@yachouh_student_2024].
-
-But embracing AI just because the students use it is like putting the cart before the horse.
+To begin, we must note the number of examples of the use of AI by students.
+A Turnitin-sponsored  report [@tyton_partners_generative_2023] of 1,600 post secondary students and 1,000 faculty, found that 50% of students used GenAI, compared to only 22% faculty.
+Use of AI by students is commonly cited as a reason for universities to embrace AI [e.g., @power_3_2024].
+McMaster's own internal research (based on a sample of over 1,300 students) indicates that at least 61% of students used AI during the Fall Term in 2023, 75% of them for school work [@yachouh_student_2024].
+
+But embracing AI because the students use it is putting the cart before the horse.
 After all, smoking is banned on campus, and I do not recall the university asking the students if they smoked^["A tobacco and smoke-free campus is the next important step towards fulfilling our responsibilities as educators, healthcare professionals and to the communities we serve" said then-president Patrick Deane [@carter_mcmaster_2017].].
@@
 
-What does McMaster hope to achieve with AI? From the Overview on Generative AI document:
+So, what does McMaster hope to achieve with AI? From the Overview on Generative AI document:
 
@@
 
-## What do GenAIs do for students?{#what-can-ais-do-for-students}
+## What do GenAIs do for students?{#what-do-ais-do-for-students}
 
@@
 
-```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 8. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
+```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 9. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
 knitr::include_graphics("images/mac-library-econometrics.png")
@@
 
-```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 9. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
+```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 10. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
 knitr::include_graphics("images/mac-library-cell-biology.png")
@@
 
-```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 10. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
+```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 11. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
 knitr::include_graphics("images/mac-library-euphemism.png")
@@
 
-First, posing interesting questions for students to work on is an essential part of an instructor's job, but this part of the job becomes compromised when instructors operate in massive classrooms situations^[Average class sizes at McMaster (with a full time under-graduate student population of 30,400) are 127.8 in first/second year, and 52.3 in third/fourth year; this places McMaster dead last among all Canadian institutions in class sizes in first/second year, and second to last after UBC in third/fourth year [@macleans_canadas_2022].].
+First, posing interesting questions for students to work on is an essential part of an instructor's job, but this part of the job becomes compromised when instructors are pushed to operate in massive classrooms with sometimes hundreds of students^[Average class sizes at McMaster (with a full time under-graduate student population of 30,400) are 127.8 in first/second year, and 52.3 in third/fourth year; this places McMaster dead last among all Canadian institutions in terms of class sizes in first/second year, and second to last after UBC in third/fourth year [@macleans_canadas_2022].].
 In this case, AI is a substitute, and demonstrably not a good one, for a qualified instructor who can intellectually challenge their students.
@@
 Research by @vahed_continuous_2023 suggests that instructors need to develop assessment literacy, that is, an understanding of "the fundamental assessment concepts and procedures, and to engage in the ongoing process of adjusting instructional strategies to enrich students learning experiences".
-There is a need for innovative approaches in assessment, but it is unlikely that AI is the tool to develop the literacy that they require.
+There is a indeed a dire need for innovative approaches in assessment, but it is unlikely that AI is the tool to develop them.
 
@@
 Who fine-tunes the machine that provides the feedback?
-Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided?^[[Automation bias](https://en.wikipedia.org/wiki/Automation_bias), the human inclination to favor suggestions made by automated decision-making systems, means that even when a human takes responsibility for decisions, AI companies remain unaccountable for their role in influencing outcomes.
-A poignant example involving AI-based face recognition and online proctoring is related by @hill_accused_2022.].
+Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided^[[Automation bias](https://en.wikipedia.org/wiki/Automation_bias), the human inclination to favor suggestions made by automated decision-making systems, means that even when a human takes responsibility for decisions, AI companies remain unaccountable for their role in influencing outcomes.
+A poignant example involving AI-based face recognition and online proctoring is related by @hill_accused_2022.]?
 
@@
 Is learning really personalized if there is no person on the teaching side of things?
-This sounds like a gimmick to put a little lipstick on the underfunded massification of higher education.
-A competent instructor who is not tasked with talking at hundreds of anonymous faces in an auditorium can provide personalized learning opportunities to students, in ways that a plausible-sounding machine that regresses to the mean cannot do.
+An AI cannot possibly understand the needs of a person, let alone help to satisfy them.
+The idea of using AI to somehow personalize learning is less a plan to provide excellent learning opportunities, and more a gimmick to put a little lipstick on the underfunded massification of higher education: a competent instructor, freed from the task of talking at hundreds of anonymous faces in an auditorium, can provide excellent personalized learning opportunities to students in ways that a plausible-sounding machine that regresses to the mean cannot do.
 
@@
 
+Leaving the technology fetish about classroom applications aside, Mollick and Mollick are well aware of the ethical concerns around using AI (see p.
+6), including that "[m]odels are trained in ways that may violate copyright", exploiting the "efforts of low-wage workers in precarious conditions for part of their training", and the fact that "[m]odels are trained on biased data and can produce either subtly or overtly biased results".
+And yet, they choose to ignore every single one of these concerns for the sake of being innovators^[The title of the paper is "Instructors as Innovators: A future-focused approach to new AI learning opportunities, with prompts".].
+Machine-washing plagiarism while exploiting low-wage workers is innovative only in a way that offends human decency.
+
+But even if the datasets and training processes used to develop GenAIs were unimpeachable, would the practice of generating learning materials using AIs be "innovative"?
+Alas, no.
+As should be clear by now, any teaching materials generated in this way can only aspire to be average given the prompts used to get the machine going.
+As a shortcut to save time (and especially money), GenAIs might be of interest, but no one should buy the delusion that the use of this technology makes them innovators.
+In any case, the list of remedial measures required to verify that the materials produced by the AI algorithm are not nonsense is so long, that it is hard to see what is being gained [see @mollick_instructors_2024, pp.
+4-5].
+
 - Automated grading
 
-There is nothing new about automated grading: Scantron sheets for multiple choice examinations have existed for over four decades.
-They work well because the algorithm for grading multiple choices is simple, and importantly, it is deterministic (i.e., there is no randomness involved).
+There is nothing new about automated grading: [Scantron](https://www.scantron.com/company/) sheets for multiple choice examinations have existed for over four decades.
+They work well because the algorithm needed for grading multiple choices is simple, and importantly, it is deterministic (i.e., there is no randomness involved).
 There is no reason to use a neural network to grade multiple choice examinations.
-As well, there is no evidence at all that a neural network can do better at grading anything more complex than a multiple choice examination, especially since 1) it will tend to consider things that are closer to the mean as "better"; and 2) the amount of training required for a classification task of this nature is massive^[Labeling data is an essential, yet extremely expensive aspect of training GenAIs.
-**NOTE** expand on this.].
+But on the flip side, there is also no evidence that a neural network can perform well at grading anything more complex than a multiple choice examination, especially since 1) it will tend to consider things that are closer to the mean as "better"^[My own unpublished experiments with grading written reflections were underwhelming, missing the human-assigned grade over 30\% of the time]; and 2) the amount of training required for a classification task of this nature is massive^[Labeling data is an essential, yet extremely expensive aspect of training GenAIs.
+The algorithm needs a "dependent" variable for training, in this case a grade assigned to a piece of work: this is called a label.
+Grading assignments manually is the process of labeling.
+Given all the possible variations of assignments (by year, by subject, by instructor), a general grading machine would require vast volumes of _labelled_ training data, and even so there is no guarantee that the results will be adequate, let alone appropriate.].
 
@@
 
-Using a chatbot to write reference letters must be either one of the most pointless or more dishonest uses of a chatbots in higher education.
-Imagine that the reference letter honestly includes a disclosure that ChatGPT was used to write it^[As it should, according to McMaster's guidelines for acknowledging the use of AI.].
+Using a chatbot to write reference letters must be one of the most pointless uses of chatbots in higher education, or one of the most dishonest.
+Imagine that the reference letter honestly included a disclosure that ChatGPT was used to write it^[As it should, according to McMaster's guidelines for acknowledging the use of AI.].
 If the objective of a reference letter is for someone to attest from the vantage point of some level of expertise as to the skills and qualities of a person, a chatbot-written letter is probably worse than useless: what message does the disclaimer send to the recipient of the letter?
 And how much stock will anyone place on the veracity of the letter?
-To avoid this catch, one would have to occult or obscure the AI assist, and in doing so would end up passing text written by countless others and filtered through a machine, as if it was their own.
-
-- qualitative and quantitative data analysis
-
+To avoid this catch, one would have to occult or obscure the use of AI, and in doing so would end up passing text written by countless others and filtered through a machine, as if it was their own.
+
+Equally troublesome is the need to input some personal data, so as to avoid getting a generic letter from the machine.
+Does the author of the letter need consent to share personal information about the subject of the letter with OpenAI et al.?
+
+- Qualitative and quantitative data analysis
+
+Neural networks have been used for decades to do data analysis.
+There is nothing new here.
+But what is being promised is that we will be able to do data analysis without having to learn how to do data analysis.
+That somehow a chatbot will do it for us.
+The idea that a chatbot could be used to do data analysis betrays a profound misunderstanding of what chatbots do, and almost absolute ignorance of what data analysis entails.
+Ask a chatbot to do some arithmetic: it will produce something that is right on average.
+But average is not always right.
+And seeing how AIs are now being trained on the dregs of the internet (see below) it should not surprise anyone when an AI Overview responds to the prompt "2 + 2" with "potato".
+
+```{r flickr-potato, out.width="50%", fig.cap="Figure 12. Data analysis using a chatbot? Maybe not such a good idea"}
+knitr::include_graphics("images/potato.png")
+```
+
+An error like this is easy to spot.
+But we seldom use computers to do calculations that can be done by hand, and checking that an AI-generated response is not wrong is likely to be as time consuming as not using an AI to do data analysis in the first place [see @yuan_how_2023].
+
+From a different perspective, even the average output of an AI should be treated with caution.
+As noted [above](#really-what-do-they-do), the results of a neural network depend on the weights selected at random at the start of a training process.
+Change those weights, the results will change.
+As well, even small variations in the prompt can be impactful in models with millions upon millions of parameters.
+In statistical terms, the output is but one _realization_ of a random process.
+For the results to be useful, we should calculate the average of a large number of responses.
+Given the cost of training and querying a LLM, this is prohibitively expensive.
+
+- Identify early signs of learning difficulties.
+
+An AI should not be used to identify signs of learning difficulties for the same reasons that AIs cannot be trusted to identify text written by an AI.
+The false positive rates (blaming the innocent) are out of proportion compared to the benefits the practice might offer.
+A vigilant instructor will be aware of learning difficulties among their students (but not if the students are an ocean of anonymous faces).
+But were the process to become automated, who would decide that an intervention is needed, and who qualified for it?
+Would the intervention be offered to everyone flagged by the system, including the false positives?
+And, could those denied an intervention complain?
+Importantly, who will fund any intervention?
+And who will deliver it?
+Another AI?
+
+<!--
 - AI-powered systems can provide smart tutoring.  
@@
 - Natural Language Processing (NLP) can enhance language learning.  
-- AI can create interactive and adaptive learning materials.  
-- AI can identify early signs of learning difficulties.  
-- AI can aid in providing personalised professional development opportunities for teachers.
-- AI can streamline administrative tasks.
+-->
 
@@
 
+The preceding section illustrates how numerous tasks that AIs can notionally do in higher education are in fact poor replacements for a human-centered education--and that is even before considering the moral, ethical, and possibly legal compromises that the use of corporate-developed AIs demand (non-corporate developed AIs can be ignored: as @vandehei_behind_2024 remind us, only a handful of corporations have the resources to make AIs give the impression of competence).
+But what about the things that GenAIs can do _to_ students?
+
 ### Distract from learning
@@
 Reliance on AI can short-circuit the learning process.
-Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] convey the idea that prompt engineering will become an essential skill; perhaps the skill to end all skills.
-
-However, this skill is tied to a singular technology.
-How is prompt engineering a transferable skill?
-As @mollick_instructors_2024 note, prompt engineering is not even transferable _between_ different LLMs (see p.
+Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] convey the idea that prompt engineering will become an essential skill; perhaps even the skill to end all skills^[Prompt engineering is a pompous name for a task that amounts to trying to read the entrails of a Large Language Model: something that cannot possibly be performed in a systematic way, and that offers no guarantees that it will work in the same way again, especially after a model is updated.
+No genuine engineer would do their work in such fashion, and calling the activity of writing prompts "engineering" is a blatant attempt to steal the competence and reputation of actual engineering.].
+
+It is important to note that, even if prompt engineering offered some value to students, the skill is inextricable tied to a singular technology, with little transferability potential.
+As @mollick_instructors_2024 demonstrate, prompt engineering is not even transferable _between_ different LLMs (see p.
 5 and p.
 36).
-Good research skills transfer between contexts.
-Learning how to form an effective query to find bibliographical resources would work at McMaster, at the University of Toronto, and at the New York Public Library.
+Good skills transfer between contexts.
+Learning how to form an effective bibliographical query would work at retrieving relevant resources at McMaster, at the University of Toronto, and at the New York Public Library.
 Learning to identify relevant content would work while reading a novel, a book of history, or a text of econometrics.
-Writing eventually helps to create a style.
+Writing, the slow and painful process of converting thoughts into words, is essential to create a style of communication that is unique to every person.
+It might be argued, in fact, that writing _is_ learning [@klein_trends_2016].
 
@@
 To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero money and still account for less than _one third of one percent point_ of Microsoft's revenue.
-A mid-sized Canadian university is like an ant compared to the economic boot of Microsoft.] who fancy themselves creators but never learned (or forgot) how to code.).
+A mid-sized Canadian university is like an ant compared to the economic boot of Microsoft.
+In truth, as the case of Solar Winds shows, Microsoft can refuse to fix a security flaw that leaves even the US Government vulnerable(for instance, to a Russian hack), and do so with little fear of ever facing accountability [@burke_microsoft_2024].] who fancy themselves creators but never learned (or forgot) how to code.).
 
@@
 
-My perspective is that it is better to do things more slowly at the beginning, to really exercise the brain, so that the brain can do more exceptional things later (perhaps even more efficiently).
-
-But let us assume for a moment that AI in its current form really becomes essential. What is the expected longevity of skills like prompt engineering?
+One could ask whether there is value in doing things more slowly, less expertly at the beginning, in order to make the delicate process of thought work, so that the brain can accomplish genuinely exceptional things later, and perhaps even more efficiently.
+
+### Learning the wrong stuff and deskilling
+
+But let us assume for a moment that using AIs in their current form does becomes an in-demand tool in the short term. What is the expected longevity of skills like prompt engineering?
 
 Apart from a few methodological developments (like transformers), improvement in the performance of AIs (say, from ChatGPT-2 to the latest version), mostly comes from using bigger datasets.
-There has been a rush for data, as developers try to improve their models by training them using more and more data.
+There has been a rush for data^[This rush is perhaps more frenetic than gold rushes in the past.
+And, while the gold rushes of the 19th century were relatively democratic due to the low cost of immigration, the rush for data is largely dominated by a few, already ultra-wealthy players.], as developers try to improve their models by training them using more and more data.
 But, despite the large number of texts available in digital form, there is still only a finite amount of data to be had.
 
-Big corporations are mostly done grabbing all data that they could lay their hands on, sometimes by morally dubious, and at times by possibly illegal means.
+As of this writing, the big corporations that can afford to grab data, are mostly done with all the data grabbing, sometimes by morally dubious, and at times by possibly illegal means [@clark_ex-amazon_2024].
 
@@
 
-As the sources of human-generated natural language data dry up, the corporations developing AIs have turned to what might be the last juicy nuggets left on the internet.
+As the sources of human-generated natural language data dry up, the few corporations developing AIs have turned to what might be the last juicy nuggets of text left on the internet.
 
 OpenAI turned to Stack Overflow [@stack_overflow_stack_2024] for more data.
-Stack Overflow [bills itself](https://stackoverflow.com) as "A community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world." The community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) with the way Stack Overflow dealt with OpenAI, and believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution.
+Stack Overflow [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world." The community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) with the way Stack Overflow dealt with OpenAI, and believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution.
 Some members of the community have protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/).
@@
 
-Google, for its part, turned to Reddit for training data [@roth_google_2024] to the tune of \$60 million per year.
-The results of this deal have been less than stellar, as Reddit content begins to creep into Google's AI-assisted search.
-Reports of this include a suggestion by Google search to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024]^[The headline of the 404 Media report was "Google Is Paying Reddit $60 Million for Fucksmith to Tell Its Users to Eat Glue".].
-
-```{r google-reddit, out.width="50%", fig.cap="Figure 10. Google's AI overview result was traced back to a post by redditor Fucksmith"}
+Google, for its part, turned to Reddit for training data to the tune of \$60 million per year [@roth_google_2024].
+As Reddit content begins to creep into Google's AI-assisted search, the results of this deal have turned grotesque, with reports of AI Overview failures including a suggestion to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024]^[The headline of the 404 Media report was "Google Is Paying Reddit $60 Million for Fucksmith to Tell Its Users to Eat Glue".].
+
+```{r google-reddit, out.width="50%", fig.cap="Figure 13. Google's AI Overview result was traced back to a post by redditor Fucksmith"}
 knitr::include_graphics("images/google-cheese-glue.png")
@@
 
-Other AI Overview results include a recommendation by geologists to eat at least one small rock a day, a piece of advice tracked back to an article in the well-known satirical periodical The Onion.
-
-```{r google-the-onion, out.width="50%", fig.cap="Figure 10. Google's AI overview result was traced back to an article in The Onion"}
+Other AI Overview results included a recommendation by geologists to eat at least one small rock a day, a piece of advice tracked back to an article published by the well-known satirical periodical [The Onion](https://www.theonion.com/geologists-recommend-eating-at-least-one-small-rock-per-1846655112).
+
+```{r google-the-onion, out.width="50%", fig.cap="Figure 14. Google's AI Overview result was traced back to an article in The Onion"}
 knitr::include_graphics("images/google-the-rock.jpg")
@@
 To what extent can more money improve AI?
-Probably not by much, since the key constraint is the amount of data available for training, and as numerous examples
-
-a search for “how to rotate text in ms paint,” referring to the Microsoft Paint application.
-The top Google search result, viewed by NBC News, directs the reader to a sarcastic Reddit comment that says to press the “Flubblegorp” key on your keyboard.
-This key does not exist.
-
-NEXT:
+Probably not by much, since the key constraint is the amount of data available for training the deep neural networks behind GenAI.
+
+Model poisoning.
 
@@
 
-Model poisoning.
-
-Develop a garbage skill https://arstechnica.com/tech-policy/2023/05/lawyer-cited-6-fake-cases-made-up-by-chatgpt-judge-calls-it-unprecedented/
-
-### Learning the wrong stuff
+Develop a garbage skill
 
@@
 It is likely that there is a similar authority bias in which students tend to be more positive
-
-### Deskilling
-
-Words go here.
 
@@
 
-```{r puzzle-web-search, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}A search using Duck Duck Go"}
+```{r puzzle-web-search, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}Figure 15. A search using Duck Duck Go"}
 ggdraw() + 
@@
 
-```{r puzzle-chatgpt, out.width="50%", fig.cap="\\label{fig:puzzle-chatgpt}A conversation with ChatGPT-4"}
+```{r puzzle-chatgpt, out.width="50%", fig.cap="\\label{fig:puzzle-chatgpt}Figure 16. A conversation with ChatGPT-4"}
 puzzle_plot <- ggplot() +
@@
 @thiel_education_2009 has written about how he "no longer believe[s] that freedom and democracy are compatible".
-Elon Musk, well, Elon Musk once shared a meme comparing Canada's PM Justin Trudeau to Hitler, when defending the vandals who [terrorized Ottawa](https://fortune.com/2022/01/27/elon-musk-twitter-freedom-convoy-canadian-truckers-rule-covid-vaccine/) in 2022.] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars...just not before super-intelligent AIs destroy humanity.], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the wellbeing of trillions of humans in the future is well worth the sacrifices of a few billions of humans today.
-Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that wealthy and well connected longtermists are willing to make.]).
-
-One does not need to posit a conspiracy theory^[Accusing the critics of being conspiracy theorists, as @sennesh_conspiracy_2023 do, distracts from the fact that there is no dearth of leftist ideas for a more just future; intellectual poverty is not what is at stake, instead a key difference is that these ideas seldom have the power of money behind them, in contrast to, say, effective altruism [@matthews_how_2022].] to worry about the influence (whether coordinated or not) of a small group of men (they are almost all of them men^[After Meta disbanded its Responsible AI Division, the company formed an AI advisory council composed of four very wealthy white men [@davis_metas_2024;@goel_meta_2024].]) who collectively can muster more than _half a trillion dollars_ as they try to implement their ideas [@stross_tech_2023].
+Elon Musk, well, Elon Musk once shared a meme comparing Canada's PM Justin Trudeau to Hitler, while defending the vandals who [terrorized Ottawa](https://fortune.com/2022/01/27/elon-musk-twitter-freedom-convoy-canadian-truckers-rule-covid-vaccine/) in 2022.] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars...but maybe not before super-intelligent AIs destroy humanity.
+Just in case, wealthy tech barons like Peter Thiel and Sam Altman, have made suitable [escape plans](https://www.theguardian.com/news/2018/feb/15/why-silicon-valley-billionaires-are-prepping-for-the-apocalypse-in-new-zealand).], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the well-being of trillions of humans in the future is well worth the sacrifices of a few billions of humans today.
+Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that those wealthy and well connected longtermists are willing to make.]).
+
+One does not need to posit a conspiracy theory^[Accusing the critics of being conspiracy theorists, as @sennesh_conspiracy_2023 do, distracts from the fact that there is no dearth of leftist ideas for a more just future; intellectual poverty is not what is at stake, instead a key difference is that progressive and humanistic ideas seldom have the power of money behind them.
+This is in contrast to, say, effective altruism [@matthews_how_2022].] to worry about the influence (whether coordinated or not) of a small group of men (they are almost all of them men^[After Meta disbanded its Responsible AI Division, the company formed an AI advisory council composed of four very wealthy white men [@davis_metas_2024;@goel_meta_2024].]) who collectively can muster more than _half a trillion dollars_ as they try to turn their individual convictions into a reality for all [@stross_tech_2023].
 
 But leaving aside the question of a potential "unadmitted political agenda" by an ultra-wealthy elite, there is the much more immediate problem that training data, much like technology, are not politically neutral.
-Several processes have been identified that contribute to reproduce malignant biases, including the concept of _platformed racism_ [@matamoros-fernandez_platformed_2017].
-Platformed racism is a process whereby the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has had a wide reach [@treyger_russian_2022] and furthermore is known to seed distrust between ethnic communities [@svedkauskas_analysis_2020].
+Several processes have been identified that contribute to reproduce malignant biases, including that of _platformed racism_ [@matamoros-fernandez_platformed_2017].
+Platformed racism results when the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has already had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020].
 A U.S.
@@
 51].].
-Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs (this is part of the Unfathomable Data problem discussed by @bender_dangers_2021).
+Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs^[this is part of the Unfathomable Data problem discussed by @bender_dangers_2021.
 In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].].
@@
 
-Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
+Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
 In addition to being misleading to the unadvised, AI is also a fantastic tool ripe for abuse by those who are advised _and_ have an agenda.
-Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat misinformation [@maiberg__google_2024].].
-In another wretched incident, Google's AI overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Obama [@tenbarge_glue_2024].
-Mishaps like this, and others, are dismissed by Google as attempts to "to trip up the technology with uncommon questions" [@tenbarge_glue_2024].
-This dismissal, though, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do).
-Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not more reassuring.
-Who is to say what a handful of unaccountable corporations will find palatable tomorrow, let alone in the face of powerful authoritarian regimes?
-
-```{r google-obama, out.width="50%", fig.cap="\\label{fig:google-ai-overview-of-obama}Google's AI overview of Muslim US presidents"}
+Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].].
+In another wretched incident, Google's AI Overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Mr.
+Obama [@tenbarge_glue_2024].
+Mishaps like this, and others, are dismissed by Google as attempts "to trip up the technology with uncommon questions" [@tenbarge_glue_2024].
+This dismissal, arrogant as it is, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do).
+Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring.
+Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow, let alone in the face of powerful authoritarian regimes that might threaten their profits^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China.
+These authors, whose one true expertise is access journalism, never stop to consider why an authoritarian regime might want to have powerful pattern recognition machines.
+To compound things, they neglect to disclose the stake that Axios, the website they founded, has on AI.
+VandeHei has not been timid to express himself otherwise, and is on record stating that AI "will eviscerate the weak, the ordinary, the unprepared in media" [@robertson_axios_2024].
+Who should be concerned about maintaining an edge?]?
+
+```{r google-obama, out.width="50%", fig.cap="\\label{fig:google-ai-overview-of-obama}Figure 17. Google's AI Overview of Muslim US presidents"}
 google_plot <- ggplot() +
@@
 
-```{r googly-eyes, out.width="50%", fig.cap="\\label{fig:googly-ai-eyes}Figure XX. In conclusion"}
+```{r googly-eyes, out.width="50%", fig.cap="\\label{fig:googly-ai-eyes}Figure 18. In conclusion"}
 google_plot <- ggplot() +

### 2024-06-15 _README.Rmd
More work

--- 2024-06-01 README.Rmd
@@
 
-# An AI is haunting campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the shade of the spectre of unregulated surveillance capitalism.]: Some considerations about the use of AI for teaching and learning in post-secondary education
+# An AI is haunting campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: Some considerations about the use of AI for teaching and learning in post-secondary education
 
@@
 Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.].
-It seemed inevitable that universities _had to do something_ with it, at the risk of being let behind.
+It seemed inevitable that universities _had to do something_ with it, at the risk of being left behind.
 
@@
 Notice that this statement (like the definition of the work before the panels above), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption.
-This attitude is almost fatalistic^[In the sense of "fated to".] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn of the bandwagon, is just the most recent of several cycles of hype [see @siegel_ai_2023].] [@angwin_press_2024].
+This attitude is almost fatalistic^[In the sense of "fated to".] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn at the very top of the bandwagon, is just the most recent of several cycles of hype [see @siegel_ai_2023].] [@angwin_press_2024].
 
@@
 Or, what happens when "integrity", "quality", and/or "excellence" clash with "the economic needs of our society" (say, a mandate by democratically elected governments to spend as little as possible in post-secondary education^[Between 2000-2020, Ontario had the lowest spending on post-secondary education of all provinces in Canada, relative to its share of spending [@caut_provincial].
-Public funding of universities in Ontario as percent of revenue fell from 45.6% in 2001/2 to 35.5% in 2019/20 [@statistics_canada_college_2022].
+Public funding of universities in Ontario as percent of revenue fell from 45.6% in 2001/2 to 35.5% in 2019/20 [@statistics_canada_college_2022], and yet, looking at the state of universities in Ontario, greater reliance on the private sector has not done much to improve the outlook.
 As noted by @usher_state_2023, Ontario is "a province where underfunding [higher-education] institutions is a pan-partisan affair".
@@
 
-@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.] for example, presents three innovative use cases for AI in higher education.
+@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.] writing for Forbes, presents three innovative use cases for AI in higher education.
 First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies".].
@@
 
-In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school).
+In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia Business School](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school) in France, an institution that Financial Times ranks among the [top 100 in Europe](https://en.wikipedia.org/wiki/Excelia_Group#Rankings).
 Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence", whatever that means.] argue that AIs will likely be the primary way to access information in the future.
@@
 
-Similarly hyperbolic claims aoput AI in higher education can be found peppered throughout the landscape, including by @johnson_ai_2023^[[Gridserve](https://www.gridserve.co.uk/about-us/) is a UK company "dedicated to enhancing the technological capabilities of the education sector through customised information and communication technology (ICT) services".] and @contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students".
+Similarly hyperbolic claims about AI in higher education can be found peppered throughout the landscape, including by @johnson_ai_2023^[[Gridserve](https://www.gridserve.co.uk/about-us/) is a UK company "dedicated to enhancing the technological capabilities of the education sector through customised information and communication technology (ICT) services".] and @contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students".
 Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].].
-A fairly comprehensive list [from @johnson_ai_2023] is as follows (notice the use of "can" in the original, not "could", or "maybe will"):
+A fairly comprehensive list of things that AI can do [from @johnson_ai_2023] is as follows (notice the use of "can" in the original, not "potentially could", or "maybe will, at some indeterminate future"):
 
@@
 
-Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it makes sense to use this technology.
+Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it might make sense to use this technology.
 To begin, we must note the sheer number of examples of the use of AI by students.
@@
 
+Mollick and Mollick are well aware of the ethical concerns around using AI (see p.
+6), including that "[m]odels are trained in ways that may violate copyright", exploiting the "efforts of low-wage workers in precarious conditions for part of their training", and the fact that "[m]odels are trained on biased data and can produce either subtly or overtly biased results".
+And yet, they choose to ignore every single one of these concerns for the sake of being innovators^[The title of the paper is "Instructors as Innovators: A future-focused approach to new AI learning opportunities, with prompts".].
+Machine-washing plagiarism while exploiting low-wage workers is innovative only in a truly amoral way.
+
 - Automated grading
 
-There is nothing new about automated grading: Scantron sheets for multiple choice examinations have existed for over four decades.
+There is nothing new about automated grading: [Scantron](https://www.scantron.com/company/) sheets for multiple choice examinations have existed for over four decades.
 They work well because the algorithm for grading multiple choices is simple, and importantly, it is deterministic (i.e., there is no randomness involved).
 There is no reason to use a neural network to grade multiple choice examinations.
-As well, there is no evidence at all that a neural network can do better at grading anything more complex than a multiple choice examination, especially since 1) it will tend to consider things that are closer to the mean as "better"; and 2) the amount of training required for a classification task of this nature is massive^[Labeling data is an essential, yet extremely expensive aspect of training GenAIs.
+But on the flip side, there is no evidence at all that a neural network can do better at grading anything more complex than a multiple choice examination, especially since 1) it will tend to consider things that are closer to the mean as "better"; and 2) the amount of training required for a classification task of this nature is massive^[Labeling data is an essential, yet extremely expensive aspect of training GenAIs.
 **NOTE** expand on this.].
@@
 
-Using a chatbot to write reference letters must be either one of the most pointless or more dishonest uses of a chatbots in higher education.
-Imagine that the reference letter honestly includes a disclosure that ChatGPT was used to write it^[As it should, according to McMaster's guidelines for acknowledging the use of AI.].
+Using a chatbot to write reference letters must be either one of the most pointless or one of the more dishonest uses of chatbots in higher education.
+Imagine that the reference letter honestly included a disclosure that ChatGPT was used to write it^[As it should, according to McMaster's guidelines for acknowledging the use of AI.].
 If the objective of a reference letter is for someone to attest from the vantage point of some level of expertise as to the skills and qualities of a person, a chatbot-written letter is probably worse than useless: what message does the disclaimer send to the recipient of the letter?
@@
 
-- qualitative and quantitative data analysis
-
+- Qualitative and quantitative data analysis
+
+Neural networks have been used for decades to do data analysis.
+There is nothing new here.
+But what we are being promised is that we will be able to do data analysis without having to learn how to do data analysis.
+That somehow a chatbot will do it for us.
+The idea that a chatbot could be used to do data analysis betrays a profound misunderstanding of what chatbots do, and almost absolute ignorance of what data analysis entails.
+Ask a chatbot to do some arithmetic: it will produce something that is right on average.
+But average is not always right.
+And seeing how AIs are now being trained on the dregs of the internet (see below) it should not surprise anyone when an AI overview responds to the prompt "2 + 2" with "potato".
+
+```{r flickr-potato, out.width="50%", fig.cap="Figure 10. Data analysis using a chatbot? Maybe not such a good idea"}
+knitr::include_graphics("images/potato.png")
+```
+
+An error like this is easy to spot.
+But we seldom use computers to do calculations that can be done by hand, and usually checking that an AI-generated response is not wrong will be as time consuming as not using an AI to do data analysis in the first place [see @yuan_how_2023].
+
+From a different perspective, even the average output of an AI should be treated with caution.
+As noted [above](#really-what-do-they-do), the results of a neural network depend on the weights selected at random at the start of a training process.
+Change those weights, the results will change.
+As well, even small variations in the prompt can be impactful in models with millions upon millions of parameters.
+In statistical terms, the output is but one _realization_ of a random process.
+For the results to be useful, we should calculate the average of a large number of responses.
+Given the cost of training and querying a LLM, this is prohibitively expensive.
+
+<!--
 - AI-powered systems can provide smart tutoring.  
@@
 - Natural Language Processing (NLP) can enhance language learning.  
-- AI can create interactive and adaptive learning materials.  
-- AI can identify early signs of learning difficulties.  
-- AI can aid in providing personalised professional development opportunities for teachers.
-- AI can streamline administrative tasks.
+-->
+
+- Identify early signs of learning difficulties.
+
+An AI should not be used to identify signs of learning difficulties for the same reasons that AIs cannot be trusted to identify text written by an AI.
+The false positive rates (blaming the innocent) are out of proportion compared to the benefits.
+A vigilant instructor will be aware of learning difficulties among their students (but not if the students are an ocean of anonymous faces).
+But were the process to become automated, who would decide that an intervention is needed, and who qualified for it?
+Would the intervention be offered to everyone flagged by the system, including the false positives?
+And, could those denied an intervention complain?
+Importantly, who will fund any intervention?
+And who will deliver it?
+Another AI?
 
@@
 Reliance on AI can short-circuit the learning process.
-Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] convey the idea that prompt engineering will become an essential skill; perhaps the skill to end all skills.
+Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] convey the idea that prompt engineering will become an essential skill; perhaps the skill to end all skills^[Prompt engineering is a pompous name for a task that amounts to trying to read the entrails of a Large Language Model: something that cannot possibly be performed in a systematic way, and that offers no guarantees that it will work in the same way, especially after a model is updated.
+No genuine engineer would do their work in such fashion, and calling the activity of writing prompts "engineering" is an attempt to steal the competence and reputation of actual engineering.].
 
@@
 Learning to identify relevant content would work while reading a novel, a book of history, or a text of econometrics.
-Writing eventually helps to create a style.
+Writing is essential to eventually develop one's own communication style.
 
@@
 
-My perspective is that it is better to do things more slowly at the beginning, to really exercise the brain, so that the brain can do more exceptional things later (perhaps even more efficiently).
+My perspective is that it is better to do things more slowly at the beginning, to exercise the brain, so that the brain can do more exceptional things later (perhaps even more efficiently).
+
+### Learning the wrong stuff and deskilling
 
@@
 
-```{r google-reddit, out.width="50%", fig.cap="Figure 10. Google's AI overview result was traced back to a post by redditor Fucksmith"}
+```{r google-reddit, out.width="50%", fig.cap="Figure 11. Google's AI overview result was traced back to a post by redditor Fucksmith"}
 knitr::include_graphics("images/google-cheese-glue.png")
@@
 
-```{r google-the-onion, out.width="50%", fig.cap="Figure 10. Google's AI overview result was traced back to an article in The Onion"}
+```{r google-the-onion, out.width="50%", fig.cap="Figure 12. Google's AI overview result was traced back to an article in The Onion"}
 knitr::include_graphics("images/google-the-rock.jpg")
@@
 To what extent can more money improve AI?
-Probably not by much, since the key constraint is the amount of data available for training, and as numerous examples
-
-a search for “how to rotate text in ms paint,” referring to the Microsoft Paint application.
-The top Google search result, viewed by NBC News, directs the reader to a sarcastic Reddit comment that says to press the “Flubblegorp” key on your keyboard.
-This key does not exist.
-
-NEXT:
+Probably not by much, since the key constraint is the amount of data available for training.
+
+Model poisoning.
 
@@
 
-### Extractive
+### Extractive and Exploitative
 
@@
 @norem_intels_2023
-
-### Exploitative
 
@@
 
-Organization ranging from Free Open Source projects to elements of the federal government in the United States have limited/restricted the use of AI.
+Organizations ranging from Free Open Source projects to elements of the federal government in the United States have limited/restricted the use of AI.
 
@@
 I struggle a bit with this one.
-For example, I fly more often than the [typical Canadian](https://ourworldindata.org/grapher/air-trips-per-capita?tab=table&time=earliest..latest&country=~CAN) and a little bit less often than the [typical world traveller](https://www.news24.com/news24/around-the-world-in-50-hours-average-person-takes-65-flights-per-year-20180828) (but I have not owned a car in decades).
-Who am I to dictate how much damage _other people_ have a right to inflict (or not) to the environment.
+For example, I fly more often than the [typical Canadian](https://ourworldindata.org/grapher/air-trips-per-capita?tab=table&time=earliest..latest&country=~CAN) and a little bit less often than the [typical world traveller](https://www.news24.com/news24/around-the-world-in-50-hours-average-person-takes-65-flights-per-year-20180828) but I have not owned a car in decades.
+Who am I to dictate how much damage _other people_ have a right to inflict (or not) to the environment?
 
@@
 @thiel_education_2009 has written about how he "no longer believe[s] that freedom and democracy are compatible".
-Elon Musk, well, Elon Musk once shared a meme comparing Canada's PM Justin Trudeau to Hitler, when defending the vandals who [terrorized Ottawa](https://fortune.com/2022/01/27/elon-musk-twitter-freedom-convoy-canadian-truckers-rule-covid-vaccine/) in 2022.] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars...just not before super-intelligent AIs destroy humanity.], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the wellbeing of trillions of humans in the future is well worth the sacrifices of a few billions of humans today.
+Elon Musk, well, Elon Musk once shared a meme comparing Canada's PM Justin Trudeau to Hitler, when defending the vandals who [terrorized Ottawa](https://fortune.com/2022/01/27/elon-musk-twitter-freedom-convoy-canadian-truckers-rule-covid-vaccine/) in 2022.] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars...just not before super-intelligent AIs destroy humanity.
+Just in case, wealthy tech barons like Peter Thiel and Sam Altman, have made suitable [escape plans](https://www.theguardian.com/news/2018/feb/15/why-silicon-valley-billionaires-are-prepping-for-the-apocalypse-in-new-zealand).], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the wellbeing of trillions of humans in the future is well worth the sacrifices of a few billions of humans today.
 Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that wealthy and well connected longtermists are willing to make.]).
@@
 Several processes have been identified that contribute to reproduce malignant biases, including the concept of _platformed racism_ [@matamoros-fernandez_platformed_2017].
-Platformed racism is a process whereby the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has had a wide reach [@treyger_russian_2022] and furthermore is known to seed distrust between ethnic communities [@svedkauskas_analysis_2020].
+Platformed racism is a process whereby the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020].
 A U.S.
@@
 In addition to being misleading to the unadvised, AI is also a fantastic tool ripe for abuse by those who are advised _and_ have an agenda.
-Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat misinformation [@maiberg__google_2024].].
+Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].].
 In another wretched incident, Google's AI overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Obama [@tenbarge_glue_2024].

### 2024-06-17 README.Rmd
More work

@@
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(knitr)
 library(magick) # Advanced Graphics and Image-Processing in R
@@
 
-In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 70 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone.
+In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 75 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone.
 As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems.
@@
 Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems".
-This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the importance of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].
+This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].
 
@@
 
-### Distract from learning
+### Distract from learning and deskilling
 
@@
 However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle that the end user cannot hope to master^[Reports of people getting their fingers crushed by Tesla's cybertruck could be a result of 1) people trusting the technology more than they should; and 2) not understanding the technology: apparently the algorithm uses increasing force to close the trunk if it fails to latch on back-to-back cycles.
-Hence, the trunk will not slash a banana on its first attemt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature (in the sense that neural networks are blackboxes, and the development of AIs itself is a blackbox where things change based on the interests of their corporate developers^[Sometimes said interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that corporate interests will always align with those of the public, and in fact, they seldom will as long as there are only a handful of American companies that offer an "essential" service.], and not the interests of their relatively small customers^[And this refers not only to individuals whose skills are yoked to ChatGPT and MidJourney.
+Hence, the trunk will not slash a banana on its first attempt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature--in the sense that 1) neural networks are blackboxes; and 2) the development of AIs itself is a blackbox where things change based on the interests of their corporate developers^[Sometimes said interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that corporate interests will always align with those of the public, and in fact, they seldom will as long as there are only a handful of American companies that offer an "essential" service.] instead of the interests of their relatively small customers^[And this refers not only to individuals whose skills are yoked to ChatGPT and MidJourney.
 In 2022, McMaster University operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD).
 Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022.
-To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero money and still account for less than _one third of one percent point_ of Microsoft's revenue.
-A mid-sized Canadian university is like an ant compared to the economic boot of Microsoft.
-In truth, as the case of Solar Winds shows, Microsoft can refuse to fix a security flaw that leaves even the US Government vulnerable(for instance, to a Russian hack), and do so with little fear of ever facing accountability [@burke_microsoft_2024].] who fancy themselves creators but never learned (or forgot) how to code.).
+To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero dollars and still account for less than _one third of one percent point_ of Microsoft's revenue.
+A mid-sized Canadian university is like an ant scuttling under the economic boot of Microsoft.
+In truth, Microsoft feels that it can stand up to even the US Government.
+As seen in the case of Solar Winds, Microsoft refused to fix a known security flaw that left the US Government vulnerable to a Russian hack, and did so with little fear of ever facing accountability [@burke_microsoft_2024].].
 
@@
 
-One could ask whether there is value in doing things more slowly, less expertly at the beginning, in order to make the delicate process of thought work, so that the brain can accomplish genuinely exceptional things later, and perhaps even more efficiently.
-
-### Learning the wrong stuff and deskilling
+One could ask whether there is not more value in doing things slowly, less expertly at the beginning, in order to make the delicate process of thought work, so that the brain can accomplish genuinely exceptional things later--perhaps even more efficiently.
 
@@
 
-Apart from a few methodological developments (like transformers), improvement in the performance of AIs (say, from ChatGPT-2 to the latest version), mostly comes from using bigger datasets.
-There has been a rush for data^[This rush is perhaps more frenetic than gold rushes in the past.
-And, while the gold rushes of the 19th century were relatively democratic due to the low cost of immigration, the rush for data is largely dominated by a few, already ultra-wealthy players.], as developers try to improve their models by training them using more and more data.
-But, despite the large number of texts available in digital form, there is still only a finite amount of data to be had.
-
-As of this writing, the big corporations that can afford to grab data, are mostly done with all the data grabbing, sometimes by morally dubious, and at times by possibly illegal means [@clark_ex-amazon_2024].
-
+Apart from a few methodological developments (like transformers), improvement in the performance of AIs (say, from ChatGPT-2 to the latest version), mostly are the result of using bigger datasets.
+This has led to a rush, as developers try to improve their models by training them on more and more data^[A rush that feels at least as frenetic as any gold rush in the past.
+And, while the gold rushes of the 19th century were relatively democratic due to their low entry costs--both in terms of immigration and tools--the current data rush has enormous entry costs and is largely dominated by Venture Capital (VC) and large corporations.].
+But, despite the large number of text items and images available in digital form, there is still only a finite amount of data to be had.
+
+As of this writing, the few big players that can afford to grab the data, are mostly done with all the data grabbing, sometimes by morally dubious, and at times by possibly illegal means [@clark_ex-amazon_2024].
 IBM, for example, uses data drawn from the Project CodeNet Dataset.
 The data are "sourced" from two online judge web sites, AIZU Online Judge and AtCoder [@ibm_automatic_2021].
-According to IBM "[the] end goal of CodeNet is to enable developers to create systems that can modernize existing codebases, as well as fix errors and security vulnerabilities in code." When we inspect the Terms of Use of AIZU (a Japanese academic project), "Commercial uses of the registered codes are prohibited".
+According to IBM "[the] end goal of CodeNet is to enable developers to create systems that can modernize existing codebases, as well as fix errors and security vulnerabilities in code".
+When we inspect the Terms of Use of AIZU (a Japanese academic project), "Commercial uses of the registered codes are prohibited".
 IBM studiously avoids the word "commercial" in its posts about LLM, foundational models, and CodeNet.
@@
 
-As the sources of human-generated natural language data dry up, the few corporations developing AIs have turned to what might be the last juicy nuggets of text left on the internet.
-
-OpenAI turned to Stack Overflow [@stack_overflow_stack_2024] for more data.
-Stack Overflow [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world." The community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) with the way Stack Overflow dealt with OpenAI, and believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution.
-Some members of the community have protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/).
-It is an open question how long Stack Overflow will remain a community-based space, when it treats the members of the community as unwitting providers of data for OpenAI.
+As the fountain of human-generated content, generously shared by countless people over a period of decades, becomes exhausted by a handful of corporations developing AIs, there has been a turn towards what might be the last juicy nuggets of information left on the internet.
+
+OpenAI, as a leader in data grabbing, turned to Stack Overflow for more data [@stack_overflow_stack_2024].
+Stack Overflow [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world".
+Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) (to say the least) with the way Stack Overflow dealt with OpenAI, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not_ do.
+Some members of the community have protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica unhelpfully termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/).
+It is an open question how long Stack Overflow will remain a community-based space, after it started treating its members as unwitting, and even unwilling providers of data for OpenAI.
 
 Google, for its part, turned to Reddit for training data to the tune of \$60 million per year [@roth_google_2024].
-As Reddit content begins to creep into Google's AI-assisted search, the results of this deal have turned grotesque, with reports of AI Overview failures including a suggestion to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024]^[The headline of the 404 Media report was "Google Is Paying Reddit $60 Million for Fucksmith to Tell Its Users to Eat Glue".].
+As Reddit content begins to creep into Google's AI-assisted search, the results of this deal have turned grotesque, with reports of AI Overview failures that include a suggestion to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024]^[The headline of the 404 Media report was "Google Is Paying Reddit $60 Million for Fucksmith to Tell Its Users to Eat Glue".].
 
@@
 
-Mind you, these, and numerous other hapless results^[Including AI overviews with misinformation about depression [@tenbarge_glue_2024].] (as of May, 2024) are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
-To what extent can more money improve AI?
-Probably not by much, since the key constraint is the amount of data available for training the deep neural networks behind GenAI.
-
-Model poisoning.
-
-Peak ChatGPT.
-
-Develop a garbage skill
-
-Words go here.
-
-Automation bias: https://en.wikipedia.org/wiki/Automation_bias
-
-It is likely that there is a similar authority bias in which students tend to be more positive
-
-### Persistence of skills (how long will learning to use a chatbot be useful?)
-
-How useful will be AIs in the future? Think of model poisoning, and the possibility that we are going to see (or may already have seen) Peak ChatGPT.
-
-De-duplication of training data improves LLMs [@lee2022deduplicating]. No wonder: it amounts to having more different information.
-
-But there is the following implication: once human-generated content is exhausted and/or has been overwhelmed by chatbot generated content, de-duplication becomes increasingly challenging.
-
-The variance of the model's output is substantially reduced: the output will, of necessity, be less varied, less diverse, and more norm-conforming:
+Mind you, these, and numerous other hapless results^[Including AI overviews with misinformation about depression [@tenbarge_glue_2024].] are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
+To what extent can more money improve GenAIs?
+Probably not by much, since a key constraint is the amount of data available for training the deep neural networks behind GenAI [@villalobos_position_2024].
+And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data^[Disturbing as this is, given the appetite for risky behaviors by corporate actors, and the potential for abuse of nominally private and confidential information.
+It would be too naive to expose even a morsel of juicy data to transmission over the internet, and expect Google et al., not to see it as fair game.].
+As GenAIs become more widely used, they can produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites^[A blog post on a website called _Ancient Rome_ by a certain Ellen Hunter (whose profile picture is likely the output of a generative AI and who does not seem to exist elsewhere) discusses the topic of windows in classical Rome and Greece [@hunter_did_2023].
+Under subheading 4.1 (_What Existed Before Windows?_), the blog explains that "MS-DOS was a disk operating system that was developed by Microsoft.
+It was the predecessor to the Windows operating system and was used on IBM PC compatible computers.
+MS-DOS was first released in 1981 and was the most popular operating system for personal computers during the 1980s".
+Clearly, the text is the output of a chatbot, and was posted to the internet without regard to its accuracy; but why waste time on even a cursory check, when the objective of the blog is not to inform, but to attract clicks for monetizing ads?] and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the most thoroughly vetted form of writing in the world [@maiberg_ai_2024]^[And let us not fool ourselves with the idea that this will be limited to smaller, scammy paper mills, when examples begin to crop up in publications by IEEE, Elsevier, and Springer, as evidenced by the telltale use of "tortured phrases" [@cabanac_tortured_2021] and chatbot stock expressions like "[a]s of my last knowledge update" [e.g., @swain_deep_2023; @al-worafi_pharmacy_2023; @sumit_dynamic_2024; @zhang_retracted_2024].].
+
+What can we expect as, in what probably is inevitable by now, future GenAI models begin to be trained using data that was generated by previous GenAIs?
+Revisiting the examples in a preceding section ([But really, what do AIs do?](#really-what-do-they-do)) we can see that the predictions of the models are more uniform than the original data used for training.
+Technically, the variance of a model's output is substantially reduced compared to the variance of the training data: the outputs of a model will necessarily (as they regress towards the mean) be less varied, less diverse, and more norm-conforming (see Table 1).
+
 ```{r data-vs-prediction-variance}
@@
   group_by(type) |> 
-  summarize(variance = var(y))
-```
-
-Tools to posion AI models:
-
-@salman_raising_2023
-
-@shan_nightshade_2024
-
-@xu_shadowcast_2024
+  summarize(variance = var(y)) |>
+  transmute(Model = factor(type, 
+                       levels = c("training data", "regression output", "nn1 output", "nn2 output"),
+                       labels = c("Training data", "Linear regression", "Neural network 1", "Neural network 2"),
+                       ordered = TRUE),
+         Variance = round(variance, 2)) |>
+  kable(caption = "\\ref{tab:data-vs-prediction-variance}Table 1. Variance reduction in model outputs")
+```
+
+The implication should be clear: training models with the output of models is a recipe for less varied outputs.
+We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to a bland version of their own previous outputs.
+And this is before we even consider the possibility of AIs being deliberately "poisoned".
+As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways, too subtle for the changes to be detectable by a human, but deleterious to the training process of a GenAI [@salman_raising_2023].
+One such tool is @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts.
+Another tool is @xu_shadowcast_2024's _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.
+
+All in all, the evidence suggests that, as fresh data for training future GenAIs dries up, the performance of these machines may not improve drastically from what we have seen so far, that is, flashy results that bear no relationship to the truth, being in fact amoral bullshit-producing machines^[American philosopher Harry G.
+Frankfurt defines _bullshit_ as speech intended to persuade without regards to the truth [@frankfurt_bullshit_2005].
+AIs, lacking intent, do not lie--but they do bullshit.].
+In fact, the performance may begin to deteriorate as the output of GenAIs begins to be used to train other GenAIs.
+This effect could easily be compounded if, as is likely the case, human-generated content is exhausted, and/or is overwhelmed by GenAI content, and/or is corrupted by purposefully bad data supplied by actors ranging from outraged creators to adversarial regimes.
+All of which begs the following questions.
+How useful will AIs be in the future?
+How persistent will skills like prompt "engineering" be?
+Will learning to prompt an AI be worth sacrificing the opportunity to develop non-AI-powered skills like coding, writing, and thinking?
+Who, after all, stands to benefit the most by this form of deskilling, in exchange for working to develop an ability that may turn out to be of ephemeral value at best and a dead-end at worst?
+
+<!--
+Develop a garbage skill
+
+Automation bias: https://en.wikipedia.org/wiki/Automation_bias
+
+It is likely that there is a similar authority bias in which students tend to be more positive
+
+### Persistence of skills (how long will learning to use a chatbot be useful?)
+
+How useful will be AIs in the future? Think of model poisoning, and the possibility that we are going to see (or may already have seen) Peak ChatGPT.
+
+De-duplication of training data improves LLMs [@lee2022deduplicating]. No wonder: it amounts to having more different information.
+
+But there is the following implication: , de-duplication becomes increasingly challenging.
+-->
 
 ## What do GenAIs do for the mega-corporations pushing them? {#what-do-ais-to-corporations}
+
+Deskilling leads to...
 
@@
 @thiel_education_2009 has written about how he "no longer believe[s] that freedom and democracy are compatible".
-Elon Musk, well, Elon Musk once shared a meme comparing Canada's PM Justin Trudeau to Hitler, while defending the vandals who [terrorized Ottawa](https://fortune.com/2022/01/27/elon-musk-twitter-freedom-convoy-canadian-truckers-rule-covid-vaccine/) in 2022.] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars...but maybe not before super-intelligent AIs destroy humanity.
-Just in case, wealthy tech barons like Peter Thiel and Sam Altman, have made suitable [escape plans](https://www.theguardian.com/news/2018/feb/15/why-silicon-valley-billionaires-are-prepping-for-the-apocalypse-in-new-zealand).], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the well-being of trillions of humans in the future is well worth the sacrifices of a few billions of humans today.
-Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that those wealthy and well connected longtermists are willing to make.]).
+Sam Bankman-Fried's FTX donated millions to a group with racist ties [@wilson_sam_2024].
+And Elon Musk?
+Well, Elon Musk once shared a meme comparing Canada's PM Justin Trudeau to Hitler while defending the vandals who terrorized Ottawa in 2022 [@henderson_and_elon_2022].] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars...but maybe not before super-intelligent AIs destroy humanity.
+Just in case, wealthy tech barons like Peter Thiel and Sam Altman, have made suitable [escape plans](https://www.theguardian.com/news/2018/feb/15/why-silicon-valley-billionaires-are-prepping-for-the-apocalypse-in-new-zealand).], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the well-being of trillions of humans in an imagined distant, star-faring future, is well worth the sacrifices of a few billions of humans today.
+Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that wealthy and well connected longtermists are willing to make.]).
 
@@
 51].].
-Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs^[this is part of the Unfathomable Data problem discussed by @bender_dangers_2021.
+Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs^[This is part of the Unfathomable Data problem discussed by @bender_dangers_2021.
 In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].].
@@
 Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring.
-Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow, let alone in the face of powerful authoritarian regimes that might threaten their profits^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China.
+Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China.
 These authors, whose one true expertise is access journalism, never stop to consider why an authoritarian regime might want to have powerful pattern recognition machines.

### 2024-06-17 README.Rmd
Working on corporate side of things

@@
 Notice that this statement (like the definition of the work before the panels above), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption.
-This attitude is almost fatalistic^[In the sense of "fated to".] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn of the bandwagon, is just the most recent of several cycles of hype [see @siegel_ai_2023].] [@angwin_press_2024].
+This attitude is almost fatalistic^[In the sense of "fated to".] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent of several cycles of hype [see @siegel_ai_2023]--fundamentally a vastly overblown tech bubble [@vinsel_dont_2023].] [@angwin_press_2024].
 
@@
 We also need to understand the goals of the organization that considers using the technology.
-When does it make sense for a university like McMaster to "integrate" AIs as part of the core mission of the organization, accounting for what the technology does for some people and what it does to others?
+When does it make sense for a university like McMaster to "integrate" AIs as part of its core mission, accounting for what the technology does for some people and what it does to others?
 
@@
 
+And so, what do AIs do that can advance the mission of the university?
+
 ## What do AIs do?{#what-do-ais-do}
@@
 A key reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they have been presented quite deliberately as _everything tools_^[Almost as flexible as, well, a naturally intelligent human, but faster and supposedly cheaper.].
-While hammers are understood to have a limited range of uses that rely on concentrating force, and word processors are mostly used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
+While hammers are understood to have a limited range of uses that mostly have to do with concentrating force, and word processors are generally only used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
 
@@
 According to Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.], AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024].
-But there is no reason to stop at self-driving vehicles: McMaster University's Canada Excellence Research Chair Laureate Ali Emadi "hopes to help revolutionize and influence everything from personal travel to city planning" by putting AI directly into our transportation systems [@dorey_driving_2023].
+But why stop at self-driving vehicles?
+McMaster University's Canada Excellence Research Chair Laureate Ali Emadi "hopes to help revolutionize and influence everything from personal travel to city planning" by putting AI directly into our transportation systems [@dorey_driving_2023].
 
@@
 
-AI will even write code, perhaps helping itself to some self-programming along the way: AIs that write themselves, or AI self-help?
+AIs will even write code, perhaps helping themselves to some self-programming along the way: AIs that write themselves, or AI self-help?
 
@@
 
-The forecast that AI will improve everything (from the natural environment of the globe to the dumps of our cats) was certain to, sooner or later, reach higher-education, a target-rich environment for selling hype--and indeed, colleges and universities have not been spared the onslaught of promises.
+The forecast that AI will improve everything (from global climate to the dumps of our cats) was certain to, sooner or later, reach higher-education, a target-rich environment for selling hype--and indeed, colleges and universities have not been spared the onslaught of promises.
+
 Examples of this abound.
 
-@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.], for one, presents three innovative use cases for AI in higher education.
+@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.], writing for Forbes, presents three innovative use cases for AI in higher education.
 First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies".].
@@
 
-In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school).
+In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school), a top business school in France.
 Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence", whatever that means.] argue that AIs will likely be the primary way to access information in the future.
@@
 Universities will use AIs to improve learning and challenge the critical thinking of students.
-Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience.], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been essential to understanding my own courses, what I teach and how I teach it; automating the task of writing syllabi is another way of degrading the quality of instruction that students can expect to receive.].
+Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience.], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been an activity essential to understand my own courses, what I teach and how I teach it; automating the task of writing syllabi is another way of degrading the quality of instruction that students can expect to receive.].
 
@@
 A benefit of portraying AI as an _everything tool_ (for the proponents of the technology, at least), is that conversations around it quickly turn into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength.
-To be sure, AIs _can_ do some things well, but critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed ["flooding the zone with shit"](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation).
+To be sure, AIs can do _some_ things well, but critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed ["flooding the zone with shit"](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation).
 Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot hope to counteract [@maiberg__google_2024].].
@@
 Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems".
-This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].
+This disclaimer, consuming fully one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].
 
@@
 Below is an example of the simplest regression model with only one independent variable (i.e., a bivariate model).
-The plot shows the pairs of $y$ and $x$ values, and the line is the model, the conditional mean $\hat{y}$.
+The plot shows the pairs of $y$ and $x$ values, and the line is the model, i.e., the conditional mean $\hat{y}$.
 This model needs only two parameters: an intercept (the conditianl mean $\hat{y}$ when $x = 0$), and the slope of the line, in other words, the rate of change of $\hat{y}$ with respect to $x$.
@@
 
-Indeed, the first neural network is "better" than the linear regression in that it reduces the typical error by `r paste0(prettyNum(delta_mad_nn1, digits = 2), "%")`, whereas the deeper network reduces the typical error by `r paste0(prettyNum(delta_mad_nn2, digits = 2), "%")`, also with respect to the linear regression^[The "typical error" here is the mean absolute deviation of the predictions.].
+Indeed, the first neural network is "better" than the linear regression in that it reduces the typical error by `r paste0(prettyNum(abs(delta_mad_nn1), digits = 2), "%")`, whereas the deeper network reduces the typical error by `r paste0(prettyNum(abs(delta_mad_nn2), digits = 2), "%")`, also with respect to the linear regression^[The "typical error" here is the mean absolute deviation of the predictions.].
 
@@
 Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than the number of data points that exist in the dataset used to train the model.
-Roughly, every additional parameter needs a data input to train on, and when there are as many parameters as data inputs the model becomes overfit and will predict each value of the independent variable in the training dataset perfectly.
+Roughly, every additional parameter needs one additional datum to train on, and when there are as many parameters as data inputs the model becomes overfit and will predict each value of the independent variable in the training dataset perfectly.
 The model will fail to train if there are more parameters than data points.
@@
 The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data were used to train AIs.
-The increasingly large datasets essentially expand the space where the model could interpolate.].
+The increasingly large datasets essentially expand the space where the model can interpolate.].
 
@@
 
-Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it makes sense to use this technology.
+Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it might make sense to use this technology.
 To begin, we must note the number of examples of the use of AI by students.
@@
 An emphasis on practice questions, however, signals an underlying weakness of the process, namely, a focus on learning for the examination, as opposed to learning for its own sake.
-A growing body of evidence points at the effectiveness of alternative assessment schemes, such as continuous assessment through course work, and even seen examinations instead of the traditional unseen examinations [@turner_see_2018].
+A growing body of evidence points towards the effectiveness of alternative assessment schemes, such as continuous assessment through course work, and even seen examinations instead of the traditional unseen examinations [@turner_see_2018].
 Alas, unseen examinations continue to dominate the evaluation landscape [@buckley_are_2023].
@@
 There is nothing new about automated grading: [Scantron](https://www.scantron.com/company/) sheets for multiple choice examinations have existed for over four decades.
-They work well because the algorithm needed for grading multiple choices is simple, and importantly, it is deterministic (i.e., there is no randomness involved).
+They work well because the algorithm needed for grading multiple choices is simple, and importantly, deterministic (i.e., there is no randomness involved).
 There is no reason to use a neural network to grade multiple choice examinations.
@@
 Grading assignments manually is the process of labeling.
-Given all the possible variations of assignments (by year, by subject, by instructor), a general grading machine would require vast volumes of _labelled_ training data, and even so there is no guarantee that the results will be adequate, let alone appropriate.].
+Given all the possible variations of assignments (by year, by subject, by instructor), a general grading machine would require vast volumes of _labelled_ training data, and even so there is no guarantee that the results would be adequate, let alone appropriate.].
 
@@
 A vigilant instructor will be aware of learning difficulties among their students (but not if the students are an ocean of anonymous faces).
-But were the process to become automated, who would decide that an intervention is needed, and who qualified for it?
+But were the process to become automated, who would decide when an intervention was warranted, and who qualified for it?
 Would the intervention be offered to everyone flagged by the system, including the false positives?
 And, could those denied an intervention complain?
-Importantly, who will fund any intervention?
-And who will deliver it?
+Importantly, who would fund any intervention?
+And who would deliver it?
 Another AI?
@@
 Reliance on AI can short-circuit the learning process.
-Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] convey the idea that prompt engineering will become an essential skill; perhaps even the skill to end all skills^[Prompt engineering is a pompous name for a task that amounts to trying to read the entrails of a Large Language Model: something that cannot possibly be performed in a systematic way, and that offers no guarantees that it will work in the same way again, especially after a model is updated.
-No genuine engineer would do their work in such fashion, and calling the activity of writing prompts "engineering" is a blatant attempt to steal the competence and reputation of actual engineering.].
-
-It is important to note that, even if prompt engineering offered some value to students, the skill is inextricable tied to a singular technology, with little transferability potential.
-As @mollick_instructors_2024 demonstrate, prompt engineering is not even transferable _between_ different LLMs (see p.
+Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] convey the idea that prompt engineering will become an essential skill; perhaps even the skill to end all skills^[Prompt engineering is a pompous name for a task that more closely resembles reading the entrails of a Large Language Model than actual engineering.
+It is a task that cannot possibly be performed in a systematic way, and that offers no guarantees that it will work in the same way again, especially after a model is updated.
+No genuine engineer would do their work in such fashion--and calling the activity of writing prompts "engineering" is a blatant attempt to steal the competence and reputation of actual engineering.].
+
+It is important to note that, even if prompt "engineering" offered some value to students, the skill is inextricable tied to a singular technology, with little transferability potential.
+As @mollick_instructors_2024 demonstrate, prompt "engineering" is not even transferable _between_ different LLMs (see p.
 5 and p.
@@
 Learning to identify relevant content would work while reading a novel, a book of history, or a text of econometrics.
-Writing, the slow and painful process of converting thoughts into words, is essential to create a style of communication that is unique to every person.
+Writing, the slow and demanding process of converting thoughts into words, is essential to create a style of communication that is unique to each person.
 It might be argued, in fact, that writing _is_ learning [@klein_trends_2016].
@@
 It can help someone to accomplish a task more easily, but it will not help if used constantly.
-However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle that the end user cannot hope to master^[Reports of people getting their fingers crushed by Tesla's cybertruck could be a result of 1) people trusting the technology more than they should; and 2) not understanding the technology: apparently the algorithm uses increasing force to close the trunk if it fails to latch on back-to-back cycles.
-Hence, the trunk will not slash a banana on its first attempt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature--in the sense that 1) neural networks are blackboxes; and 2) the development of AIs itself is a blackbox where things change based on the interests of their corporate developers^[Sometimes said interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that corporate interests will always align with those of the public, and in fact, they seldom will as long as there are only a handful of American companies that offer an "essential" service.] instead of the interests of their relatively small customers^[And this refers not only to individuals whose skills are yoked to ChatGPT and MidJourney.
+However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle that the end user cannot hope to fully master^[Reports of people getting their fingers crushed by Tesla's cybertruck could be a result of 1) people trusting the technology more than they should; and 2) not understanding the technology: apparently the algorithm uses increasing force to close the trunk if it fails to latch on back-to-back cycles.
+Hence, the trunk will not slash a banana on its first attempt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature--in the sense that 1) neural networks are blackboxes; and 2) the development of AIs itself is a blackbox wherein things can more easily change based on the interests of their corporate owners^[Sometimes said interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that corporate interests will always align with those of the public, and in fact, they seldom will as long as there are only a handful of American companies that offer an "essential" service.] than in response to the interests of their relatively small customers^[And this refers not only to individuals whose skills become yoked to ChatGPT and MidJourney.
 In 2022, McMaster University operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD).
@@
 A mid-sized Canadian university is like an ant scuttling under the economic boot of Microsoft.
-In truth, Microsoft feels that it can stand up to even the US Government.
-As seen in the case of Solar Winds, Microsoft refused to fix a known security flaw that left the US Government vulnerable to a Russian hack, and did so with little fear of ever facing accountability [@burke_microsoft_2024].].
+In truth, Microsoft is so incredibly large, that it feels like it can stand up to even the US Government.
+As revealed by the case of Solar Winds, Microsoft refused to fix a known security flaw that left the US Government vulnerable to a Russian hack, and probably did so with little fear of ever facing accountability [@burke_microsoft_2024].].
 
@@
 
-But let us assume for a moment that using AIs in their current form does becomes an in-demand tool in the short term. What is the expected longevity of skills like prompt engineering?
+But let us assume for a moment that using AIs in their current form does becomes an in-demand tool in the short term. What is the expected longevity of skills like prompt "engineering"?
 
@@
 This has led to a rush, as developers try to improve their models by training them on more and more data^[A rush that feels at least as frenetic as any gold rush in the past.
-And, while the gold rushes of the 19th century were relatively democratic due to their low entry costs--both in terms of immigration and tools--the current data rush has enormous entry costs and is largely dominated by Venture Capital (VC) and large corporations.].
+And, while the gold rushes of the 19th century were relatively democratic due to their low entry costs--both in terms of immigration and tools--the current data rush has enormous entry costs and is largely dominated by venture capital (that is, equity investments that support pre-launch, launch, and early stages of a business) and established large corporations.].
 But, despite the large number of text items and images available in digital form, there is still only a finite amount of data to be had.
@@
 One might ask if IBM is doing all this research as a non-commercial effort?
-Will those developers creating systems, modernizing databases, fixing errors, etc., be able to access this for free, and use the systems for education and research only?
+Will those developers creating systems, modernizing databases, fixing errors, etc., be able to access CodeNet for free, and use the systems for education and research only?
 What does "commercial uses" even mean?
@@
 Stack Overflow [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world".
-Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) (to say the least) with the way Stack Overflow dealt with OpenAI, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not_ do.
-Some members of the community have protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica unhelpfully termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/).
-It is an open question how long Stack Overflow will remain a community-based space, after it started treating its members as unwitting, and even unwilling providers of data for OpenAI.
+Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) (to say the least) with the way Stack Overflow dealt with OpenAI, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not do_.
+Some members of the community protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica unhelpfully termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/).
+It is an open question how long Stack Overflow will remain a community-based space, especially after it started treating its members as unwitting, and even unwilling providers of data for OpenAI.
 
@@
 To what extent can more money improve GenAIs?
-Probably not by much, since a key constraint is the amount of data available for training the deep neural networks behind GenAI [@villalobos_position_2024].
-And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data^[Disturbing as this is, given the appetite for risky behaviors by corporate actors, and the potential for abuse of nominally private and confidential information.
+Probably not by much, since a key constraint is the amount of data available for training the deep neural networks behind the technology [@villalobos_position_2024].
+And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data^[A disturbing idea in and of itself, given the appetite for risky behaviors by corporate actors, and the potential for abuse of nominally private and confidential information.
 It would be too naive to expose even a morsel of juicy data to transmission over the internet, and expect Google et al., not to see it as fair game.].
@@
 And this is before we even consider the possibility of AIs being deliberately "poisoned".
-As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways, too subtle for the changes to be detectable by a human, but deleterious to the training process of a GenAI [@salman_raising_2023].
+As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023].
 One such tool is @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts.
@@
 
-All in all, the evidence suggests that, as fresh data for training future GenAIs dries up, the performance of these machines may not improve drastically from what we have seen so far, that is, flashy results that bear no relationship to the truth, being in fact amoral bullshit-producing machines^[American philosopher Harry G.
+All in all, the evidence suggests that, as fresh data for training future GenAIs dries up, the performance of these machines may not improve drastically from what we have seen so far, that is, flashy results that bear no relationship to the truth, being in fact the output of bullshit-producing machines^[American philosopher Harry G.
 Frankfurt defines _bullshit_ as speech intended to persuade without regards to the truth [@frankfurt_bullshit_2005].
@@
 Will learning to prompt an AI be worth sacrificing the opportunity to develop non-AI-powered skills like coding, writing, and thinking?
-Who, after all, stands to benefit the most by this form of deskilling, in exchange for working to develop an ability that may turn out to be of ephemeral value at best and a dead-end at worst?
+Who, after all, stands to benefit the most by this form of deskilling, in exchange for the effort required to develop an ability that may turn out to be of ephemeral value at best and a dead-end at worst?
 
@@
 ## What do GenAIs do for the mega-corporations pushing them? {#what-do-ais-to-corporations}
+
+As I argue above, AIs do not do much for students or instructors, other than distract them from learning and developing genuinely transferable skills.
+
+But AIs must do something for someone, otherwise why would they be everywhere?
+What do AIs do for the likes of OpenAI, Microsoft, and Google?
+So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money.
+Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].
+
+<!--
+Ministry of Colleges and Universities: 163,482,281 CAD
+Venture capitalists: 75,000,000,000 USD == 257,776,200,000 CAD
+
+https://www.exchange-rates.org/exchange-rate-history/usd-cad-2022
+Avg. exchange rate in 2022: 1.3019 CAD/US
+
+This is how much bigger venture capital in AI is compared to research in Ontario. 
+163482281/(75000000000 * 1.3019)
+--->
+
+But the industry has not necessarily made money for investors yet.
+For example, @berber_peter_2024 reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue.
+This has not prevented start-ups from pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs of the scammy environment in which GenAI tech thrives were needed, @berber_peter_2024 also reports how Cognition, a startup that pivoted to AI from cryptocurrency, was receiving billionaire infusions of speculative money.].
+
+@sriram_ai_2023
+
+By some accounts   @amdur_venture_2023
+
+Distractions:
+@vinsel_dont_2023 @siegel_ai_2023
 
@@
 To read:
-
-@sriram_ai_2023
-
-@berber_peter_2024: "venture-capital firm Sequoia estimated [in March 2024] that the AI industry spent \$50 billion [in 2023] on the Nvidia chips used to train advanced AI models...but brought in only $3 billion in revenue."
 

### 2024-06-18 README.Rmd
Moving on to what AI does to everyone and everything

@@
 
-> Mission Statement: Our organization seeks to level up on all matters related.  
+> _Mission Statement: Our organization seeks to level up on all matters related._  
 > -- An Organization Run by its Marketing Department
@@
 
-In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 75 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone.
+In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 80 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone.
 As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems.
@@
 
+> _ChatGPT doesn't give you information. It gives you information-shaped sentences._
+> --Neil Gaiman
+
 According to McMaster's internal research:
@@
 But average is not always right.
-And seeing how AIs are now being trained on the dregs of the internet (see below) it should not surprise anyone when an AI Overview responds to the prompt "2 + 2" with "potato".
+And seeing how AIs are now being trained on the dregs of the internet, it should surprise approximately no one when an AI Overview responds to the prompt "2 + 2" with "potato".
 
@@
 An error like this is easy to spot.
-But we seldom use computers to do calculations that can be done by hand, and checking that an AI-generated response is not wrong is likely to be as time consuming as not using an AI to do data analysis in the first place [see @yuan_how_2023].
+But we seldom use computers to do calculations that can be done by hand, and checking that an AI-generated response is not wrong is likely to be as time consuming as not using an AI to do quantitative data analysis in the first place [see @yuan_how_2023].
 
@@
 Given the cost of training and querying a LLM, this is prohibitively expensive.
+
+But what about a task that involves logic instead of arithmetic? LLMs are good for it only if the goat knows how to swim.
+
+```{r puzzle-web-search, eval=FALSE, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}Figure 15. A search using Duck Duck Go"}
+ggdraw() + 
+  draw_image(image = "images/logical-puzzle-web-search.png")
+```
+
+```{r puzzle-chatgpt, out.width="50%", fig.cap="\\label{fig:puzzle-chatgpt}Figure 16. ChatGPT-4 makes the goat swim"}
+puzzle_plot <- ggplot() +
+  labs(caption = "https://assemblag.es/@hugo/112450401657084750") +
+  theme_void()
+
+ggdraw() + 
+  draw_plot(puzzle_plot) +
+  draw_image(image = "images/logical-puzzle-chat-gpt-4.png")
+
+```
+
+As research by @nezhurina_alice_2024 demonstrates, relatively simple natural language puzzles that are easily solved by humans, often result in confidently expressed nonsense when used to prompt an AI.
+Wrong answers, furthermore, are frequently backed up by confabulations, that is, reasoning-like or plausible sounding explanations that are illogical.
+If anything, mathematics should be easy: after all, computers have been doing mathematics since at least the [Antikythera mechanism](https://en.wikipedia.org/wiki/Antikythera_mechanism) (ca.
+178 BC), which preceded other mechanical machines by centuries--including Babbage's [Analytical Engine](https://en.wikipedia.org/wiki/Analytical_engine), the machine that [Ada Lovelace](https://en.wikipedia.org/wiki/Ada_Lovelace) used to code an algorithm to write Bernoulli numbers.
+Electronic computers have been doing mathematics reliably for decades now.
+Computers can do mathematics well because they are implemented using a crisp, unambiguous language.
+Natural language is full of nuances that are not easily captured using the medium of text alone.
+The stochastic connections between words that LLMs produce are simply probabilistic word salads that only sound plausible thanks to the billions of data points used to train them and the billions of parameters that the models use.
+But if the objective of qualitative research is to identify interesting aspects of the human experience, it is certainly not the case that they will be found in a regression to the mean.
 
@@
 
-The preceding section illustrates how numerous tasks that AIs can notionally do in higher education are in fact poor replacements for a human-centered education--and that is even before considering the moral, ethical, and possibly legal compromises that the use of corporate-developed AIs demand (non-corporate developed AIs can be ignored: as @vandehei_behind_2024 remind us, only a handful of corporations have the resources to make AIs give the impression of competence).
+The preceding section illustrates how numerous tasks that AIs can notionally do in higher education are in fact poor replacements for a human-centered education--and that is even before considering the moral, ethical, and possibly legal compromises that the use of corporate-developed AIs demand (non-corporate developed AIs can be ignored: as @vandehei_behind_2024 remind us, only a handful of corporations have the resources needed to create the illusion of compentence in an AI).
+
 But what about the things that GenAIs can do _to_ students?
@@
 
-But let us assume for a moment that using AIs in their current form does becomes an in-demand tool in the short term. What is the expected longevity of skills like prompt "engineering"?
+However, let us assume for a moment that using AIs in their current form does becomes an in-demand tool in the short term. What is the expected longevity of skills like prompt "engineering"?
 
@@
 This has led to a rush, as developers try to improve their models by training them on more and more data^[A rush that feels at least as frenetic as any gold rush in the past.
-And, while the gold rushes of the 19th century were relatively democratic due to their low entry costs--both in terms of immigration and tools--the current data rush has enormous entry costs and is largely dominated by venture capital (that is, equity investments that support pre-launch, launch, and early stages of a business) and established large corporations.].
+And, while the gold rushes of the 19th century were relatively democratic due to their low entry costs--both in terms of immigration and tools--the current data rush has enormous entry costs and is largely dominated by large corporations and venture capital (that is, equity investments that support pre-launch, launch, and early stages of a business).].
 But, despite the large number of text items and images available in digital form, there is still only a finite amount of data to be had.
@@
 
-As the fountain of human-generated content, generously shared by countless people over a period of decades, becomes exhausted by a handful of corporations developing AIs, there has been a turn towards what might be the last juicy nuggets of information left on the internet.
-
-OpenAI, as a leader in data grabbing, turned to Stack Overflow for more data [@stack_overflow_stack_2024].
+As the fountain of human-generated content--generously (one might even say naively) shared by untold numbers of people over a period of decades--becomes exhausted by a handful of corporations developing AIs, there has been a turn towards what might be the last juicy nuggets of information left on the internet.
+
+OpenAI, a global leader in data grabbing, turned to Stack Overflow for more data [@stack_overflow_stack_2024].
 Stack Overflow [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world".
@@
 Some members of the community protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica unhelpfully termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/).
-It is an open question how long Stack Overflow will remain a community-based space, especially after it started treating its members as unwitting, and even unwilling providers of data for OpenAI.
+It is an open question how long Stack Overflow will remain a community-based space, especially after it started treating its members as unwitting, and even unwilling providers of data for OpenAI [@grimm_stack_2024].
 
@@
 MS-DOS was first released in 1981 and was the most popular operating system for personal computers during the 1980s".
-Clearly, the text is the output of a chatbot, and was posted to the internet without regard to its accuracy; but why waste time on even a cursory check, when the objective of the blog is not to inform, but to attract clicks for monetizing ads?] and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the most thoroughly vetted form of writing in the world [@maiberg_ai_2024]^[And let us not fool ourselves with the idea that this will be limited to smaller, scammy paper mills, when examples begin to crop up in publications by IEEE, Elsevier, and Springer, as evidenced by the telltale use of "tortured phrases" [@cabanac_tortured_2021] and chatbot stock expressions like "[a]s of my last knowledge update" [e.g., @swain_deep_2023; @al-worafi_pharmacy_2023; @sumit_dynamic_2024; @zhang_retracted_2024].].
+Clearly, the text is the output of a chatbot, and was posted to the internet without regard to its accuracy; but why waste time on even a cursory check, when the objective of the blog is not to inform, but to attract clicks for monetizing ads?] and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the most thoroughly vetted form of writing in the world [@maiberg_ai_2024]^[And let us not fool ourselves with the idea that this will be limited to smaller, scammy paper mills, when examples begin to crop up in publications by IEEE, Elsevier, and Springer, as evidenced by the telltale use of "tortured phrases" [@cabanac_tortured_2021] and chatbot stock expressions like "[a]s of my last knowledge update...", "[a]s of my knowledge cutoff...", "[c]ertainly!
+Here are...", and "[r]egenerate response" [e.g., @al-worafi_pharmacy_2023; @farag_exploring_2023; @li_revitalizing_2023; @shafiq_escrao_2023; @swain_deep_2023; @alsaif_nd2o3-_2024; @gatell_impact_2024; @remzan_ensemble_2024; @sumit_dynamic_2024; @zhang_retracted_2024].
+One particularly egregious example straight out includes the phrase "[p]lease note that as an AI language model, I am unable to generate specific tables or conduct tests, so the actual results should be included in the table" [@yang_utilization_2023].].
 
@@
 
-The implication should be clear: training models with the output of models is a recipe for less varied outputs.
-We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to a bland version of their own previous outputs.
+The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs.
+We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to duller versions of their own previous outputs.
 And this is before we even consider the possibility of AIs being deliberately "poisoned".
@@
 
+> _Teach a person to use ChatGPT, and they will be in thrall to a corporation for as long as that is their skill. Teach them to think, and they will be free for life._  
+> -- Anonymous
+
 As I argue above, AIs do not do much for students or instructors, other than distract them from learning and developing genuinely transferable skills.
@@
 What do AIs do for the likes of OpenAI, Microsoft, and Google?
-So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money.
+So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as the resources needed to have a University Committee to advise on AI^[Some observers have even wondered whether the AI fever is not actually becoming detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].].
 Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].
@@
 
-But the industry has not necessarily made money for investors yet.
+Alas, the industry has not necessarily made money for investors yet.
 For example, @berber_peter_2024 reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue.
-This has not prevented start-ups from pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs of the scammy environment in which GenAI tech thrives were needed, @berber_peter_2024 also reports how Cognition, a startup that pivoted to AI from cryptocurrency, was receiving billionaire infusions of speculative money.].
-
-@sriram_ai_2023
-
-By some accounts   @amdur_venture_2023
-
-Distractions:
-@vinsel_dont_2023 @siegel_ai_2023
-
-Deskilling leads to...
-
-So far, not money.
-But since corporations are not in the development of AI for charitable motives, at some point they will want to make money out of them.
-Who will pay?
-And how much will they need to pay?
-Or have to pay, if their skills are inexorable dependent on this one technology?
-
-To read:
-
-Charlie Pierce [writes](https://www.antipope.org/charlie/blog-static/2024/04/the-radiant-future-of-1995.html#more) that "the industry is running at a loss on power consumption alone, never mind labour costs (which are quite high: those generative LLMs require extensive human curation of the input data they require for training)".
-
-"The common feature of these bubbles is a shitload of hype and promises from hucksters who fail to deliver a viable product but suck up as much investment capital as they can.
-A handful of them survive: from dot-com 1.0, the stand-outs are Amazon and Google (Facebook, Twitter, Reddit, et al came along much later—social media was a later, smaller bubble).
-Other survivors include Paypal, eBay, and Doubleclick (the latter being merged with Google to form a monstrous global advertising monopoly).
-The survivors tend to leave behind infrastructure: the failures leave behind t-shirts, second hand Aeron chairs, and motivational posters."
-
-https://futurism.com/the-byte/google-bard-ai-ready NOT READY, SO WHY RELEASE IT?
+@patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google.
+At a guess, the industry is running at a loss: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer.
+And a large number of users are not even paying subscribers: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023].
+The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing the most intensive, disturbing, degrading, and damaging aspects of training AIs to the Global South [@nix_analysis_2024].
+The fact that the industry is not profitable has not prevented start-ups from pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted to AI from cryptocurrency, after cryptocurrency went sour with the frauds at FTX.].
+
+Venture capital is not known for its charitable aims, though, and at some point investors will want their money back with a heap of profit on top [@sriram_ai_2023].
+Who will pay to make investors whole?
+And how much will those who end up paying need to pay?
+Or, in the worst case, _have_ to pay, if their skills are inexorable tied to this one technology?
+Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020].
+From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service^[Universities, instead of being innovators, have increasingly become renters in the land of the platform economy [@komljenovic_rise_2021].].
 
@@
 Compare the results:
-
-```{r puzzle-web-search, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}Figure 15. A search using Duck Duck Go"}
-ggdraw() + 
-  draw_image(image = "images/logical-puzzle-web-search.png")
-```
-
-```{r puzzle-chatgpt, out.width="50%", fig.cap="\\label{fig:puzzle-chatgpt}Figure 16. A conversation with ChatGPT-4"}
-puzzle_plot <- ggplot() +
-  labs(caption = "https://assemblag.es/@hugo/112450401657084750") +
-  theme_void()
-
-ggdraw() + 
-  draw_plot(puzzle_plot) +
-  draw_image(image = "images/logical-puzzle-chat-gpt-4.png")
-
-```
 

### 2024-06-18 README.Rmd
More work

@@
 Natural language is full of nuances that are not easily captured using the medium of text alone.
-The stochastic connections between words that LLMs produce are simply probabilistic word salads that only sound plausible thanks to the billions of data points used to train them and the billions of parameters that the models use.
+The stochastic connections between words that LLMs produce are simply probabilistic word salads^[@xu_hallucination_2024 use a formal approach to show that LLMs are inherently incapable of mapping outputs to a ground truth, and thus their results will only ever resemble a ground truth purely by chance.
+Recent reporting strongly hints that industry leaders like  Sundar Pichai and Tim Cook understand this well [@patel_google_2024; @tyrangiel_opinion_2024]] that only sound plausible thanks to the billions of data points used to train them and the billions of parameters that the models use.
 But if the objective of qualitative research is to identify interesting aspects of the human experience, it is certainly not the case that they will be found in a regression to the mean.
@@
 What do AIs do for the likes of OpenAI, Microsoft, and Google?
-So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as the resources needed to have a University Committee to advise on AI^[Some observers have even wondered whether the AI fever is not actually becoming detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].].
+So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as those spent by universities to support an AI-dedicated committee^[Some observers have even wondered whether the AI fever is not actually becoming detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].].
 Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].
@@
 Alas, the industry has not necessarily made money for investors yet.
-For example, @berber_peter_2024 reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue.
+
+@berber_peter_2024, for instance, reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue.
 @patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google.
-At a guess, the industry is running at a loss: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer.
+At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer.
 And a large number of users are not even paying subscribers: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023].
 The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing the most intensive, disturbing, degrading, and damaging aspects of training AIs to the Global South [@nix_analysis_2024].
-The fact that the industry is not profitable has not prevented start-ups from pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted to AI from cryptocurrency, after cryptocurrency went sour with the frauds at FTX.].
+The fact that the industry is not profitable has not prevented start-ups from hurriedly pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted from cryptocurrencyto AI, coincidentally as cryptocurrency was going sour after the frauds at FTX were revealed.
+Barely disguised grift has grown abundant in the vicinity of AI at almost every level [@cook_5_2024].].
 
@@
 Who will pay to make investors whole?
-And how much will those who end up paying need to pay?
-Or, in the worst case, _have_ to pay, if their skills are inexorable tied to this one technology?
-Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020].
-From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service^[Universities, instead of being innovators, have increasingly become renters in the land of the platform economy [@komljenovic_rise_2021].].
+And how much those who end up paying will need to pay?
+Or, in the worst case, _have_ to pay, if their skills become inexorably tied to this one technology?
+Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability^[And, with profitability becoming a more distant possibility, there is always the option for large investors to cosmically inflate the hype [@salmon_what_2024] to convince bigger suckers to give them money [@reuters_openais_2024] and/or to fleece retail investors [@temkin_vcs_2024].]: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020].
+From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service^[Universities, rather than being innovators, have increasingly become renters in the land of the platform economy [@komljenovic_rise_2021].].
 
@@
 
-### Extractive
-
-In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )
-
-@germain_google_2023
-@norem_intels_2023
-
-### Exploitative
-
-ChatGPT 4.o was announced on May 13, 2024, with a flashy display of the model's new capabilities, which included the ability to respond vocally to prompts, and to identify images via a mobile device's camera.
-4.o's voice, in particular, caused some titillation in the pants of even its creators [@roose_is_2024].
+### AI is an eminently extractive and exploitative industry
+
+ChatGPT 4.o was announced on May 13, 2024, with fanfare and a flashy display of the model's new capabilities, which included the ability to respond vocally to prompts, and to identify images via a mobile device's camera.
+4.o's voice, in particular, caused some titillation in the nether parts of even its creators [@roose_is_2024].
 The voice was modeled after the digital personal assistant of the movie _Her_, played by Scarlett Johannson.
@@
 
-Alas, very few have the social standing of Scarlett Johansson to challenge OpenAI. Most people in the world lack that power.
-
-The ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted. But AI companies do not spare the rich either.
+Alas, very few have the social standing of Scarlett Johansson to challenge OpenAI.
+Most people in the world lack that power.
+In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )
+
+@germain_google_2023
+@norem_intels_2023
+
+The ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted.
 
@@
 You'll get more confusing sentences that don't make sense, and you will potentially get more of this content that they worked so hard to filter out because they were only filtering it in English.
-
-Think about linguistic hegemony, and how the models are only as good (or as bad) as their inputs.
 
@@
 
-## As an aside {#as-an-aside}
-
-On antropomorphizing GenAI tools.
-This happens mostly in ways that are barely liminal.
-For example, in the document "Overview of AI at McMaster" we read that GenAI's somehow have an "ability to engage in nuanced conversation".
-This phrasing could be simply a literary device, but even so it is indicative of a tendency to ascribe human attributes to a statistical model.
-Doing this risks amplifying existing biases.
-But in other cases, the effort is more blatant.
-Wrapping a statistical model in a human-like skin may actually be more devious, as in the new, sexy ChatGPT interface:
-
-<video width="320" height="240" controls>
-<source src="video/open-ai-commercial.mp4" type="video/mp4">
-</video>
-
-**NOTES:** Need to research more the argument that the interface is so important in the case of these tools because what they do well is trivial and what they do poorly is what is being hyped.
-Found the article [@farrugia_known_2024]; see here: https://fasterandworse.com/known-purpose-and-trusted-potential/
-
 ## Concluding remarks
@@
 Neural networks and machine learning are not new, and they have their uses. But those uses have been drown, and to some extent discredited, by the current hype around AI.
+
+@morrone_signals_2024 Carefully manufactured and well funded hype.
 
@@
 
-## Loose thoughts
+## As an aside {#as-an-aside}
+
+On antropomorphizing GenAI tools.
+This happens mostly in ways that are barely liminal.
+For example, in the document "Overview of AI at McMaster" we read that GenAI's somehow have an "ability to engage in nuanced conversation".
+This phrasing could be simply a literary device, but even so it is indicative of a tendency to ascribe human attributes to a statistical model.
+Doing this risks amplifying existing biases.
+But in other cases, the effort is more blatant.
+Wrapping a statistical model in a human-like skin may actually be more devious, as in the new, sexy ChatGPT interface:
+
+<video width="320" height="240" controls>
+<source src="video/open-ai-commercial.mp4" type="video/mp4">
+</video>
+
+**NOTES:** Need to research more the argument that the interface is so important in the case of these tools because what they do well is trivial and what they do poorly is what is being hyped.
+Found the article [@farrugia_known_2024]; see here: https://fasterandworse.com/known-purpose-and-trusted-potential/
 
@@
 
-Crypto-currency to AI: https://en.wikipedia.org/wiki/Cognition_AI
+AI industry and the military: @haskins_im_2024
 

### 2024-06-18 ai-haunting-campus/ai-haunting-campus.qmd
Created quarto file to render report to pdf

--- 2024-06-18 README.Rmd
@@
 ---
-output: github_document
-bibliography: bibliography/bibliography.bib
+title: "An AI is haunting campus"
+subtitle: "Some considerations about the use of AI for teaching and learning in post-secondary education"
+author: Antonio Páez (McMaster University)
+date: "`r Sys.Date()`"
+format:
+    pdf: 
+      fontsize: 7pt
+    html: 
+      self-contained: true
+      grid: 
+        margin-width: 350px
+execute: 
+  echo: fenced
+  freeze: auto
+reference-location: margin
+#citation-location: margin
+bibliography:
+  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
 ---
-
-<!-- README.md is generated from README.Rmd. Please edit that file -->
 
@@
 
-# An AI is haunting campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the shade of the spectre of unregulated surveillance capitalism.]: Some considerations about the use of AI for teaching and learning in post-secondary education
-
-<!-- badges: start -->
-
-<!-- badges: end -->
-
-By: Antonio Páez (paezha@mcmaster.ca)
+::: {.column-margin}
+Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.
+:::
 
@@
 
----
-
-## Contents
-
-- [Background](#background)  
-- [The university's mission](#the-mission)  
-- [What do AIs do?](#what-do-ais-do)  
-- [A note on the terms of the conversation](#terms-of-debate)  
-- [But really, what do AIs do?](#really-what-do-they-do)  
-- [AI in teaching and learning](#ai-in-teaching-and-learning)  
-- [What do GenAIs do for students?](#what-do-ais-do-for-students)  
-- [What do GenAIs do to students?](#what-can-ais-do-to-students)  
-- [What do GenAIs do for the mega-corporations pushing these tools?](#what-do-ais-to-corporations)  
-- [What do GenAIs do to everyone else?](#what-do-ais-do-to-everyone-else)
-
----
-
-## Background {#background}
+Note: _This essay was written as an open research project, and it can be found in the following repository:_ https://github.com/paezha/University-AI-Panel
+
+## Background
 
@@
 
-## The university's mission {#the-mission}
+# The university's mission {#the-mission}
 
@@
 
-```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 1. An example of Atlassian AI writing code"}
-knitr::include_graphics("images/ai-bug-fix.png")
+```{r ai-writing-code, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}An example of Atlassian AI writing code"}
+knitr::include_graphics(glue::glue(here::here(), "/images/ai-bug-fix.png"))
 ```
@@
 
-```{r ai-litter-box, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 2. Revolutionizing pet health"}
-knitr::include_graphics("images/ai-litter-box.png")
+```{r ai-litter-box, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}Revolutionizing pet health"}
+knitr::include_graphics(glue::glue(here::here(), "/images/ai-litter-box.png"))
 ```
@@
 Shortly after that, Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.
+
+\newpage
 
@@
 
-```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 3. Simple Example of a Regression Model"}
+```{r plot-simulated-data, message=FALSE, out.width="70%", fig.cap="Simple Example of a Regression Model"}
 ggplot(df,
@@
 
-```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 4. Training data and model output compared"}
+```{r data-vs-predictions-comparison, out.width="70%", fig.cap="Training data and model output compared"}
 
@@
 
-```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 5. Simple Example of a Shallow Neural Network"}
-knitr::include_graphics("images/nn1.png")
+```{r plot-shallow-neural-network, out.width="70%", fig.cap="Simple Example of a Shallow Neural Network"}
+knitr::include_graphics(glue::glue(here::here(), "/images/nn1.png"))
 ```
@@
 
-```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 6. Training data and model outputs compared: linear regression and shallow neural network"}
+```{r data-vs-predictions-comparison-2, out.width="70%", fig.cap="Training data and model outputs compared: linear regression and shallow neural network"}
 ggplot() + 
@@
 
-```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 7. Example of a Deeper Neural Network"}
-knitr::include_graphics("images/nn2.png")
+```{r plot-deeper-neural-network, out.width="70%", fig.cap="Example of a Deeper Neural Network"}
+knitr::include_graphics(glue::glue(here::here(), "/images/nn2.png"))
 ```
@@
 
-```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 8. Training data and model outputs compared: shallow neural network and deeper neural network"}
+```{r data-vs-predictions-comparison-3, out.width="70%", fig.cap="Training data and model outputs compared: shallow neural network and deeper neural network"}
 ggplot() + 
@@
 
-```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 9. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
-knitr::include_graphics("images/mac-library-econometrics.png")
-```
-
-```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 10. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
-knitr::include_graphics("images/mac-library-cell-biology.png")
-```
-
-```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 11. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
-knitr::include_graphics("images/mac-library-euphemism.png")
+```{r mac-library-econometrics, out.width="70%", fig.cap="Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
+knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-econometrics.png"))
+```
+
+```{r mac-library-cell-biology, out.width="70%", fig.cap="Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
+knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-cell-biology.png"))
+```
+
+```{r mac-library-euphemism, out.width="70%", fig.cap="Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
+knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-euphemism.png"))
 ```
@@
 
-```{r flickr-potato, out.width="50%", fig.cap="Figure 12. Data analysis using a chatbot? Maybe not such a good idea"}
-knitr::include_graphics("images/potato.png")
+```{r flickr-potato, out.width="70%", fig.cap="Data analysis using a chatbot? Maybe not such a good idea"}
+knitr::include_graphics(glue::glue(here::here(), "/images/potato.png"))
 ```
@@
 
-```{r puzzle-web-search, eval=FALSE, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}Figure 15. A search using Duck Duck Go"}
+```{r puzzle-web-search, eval=FALSE, out.width="70%", fig.cap="A search using Duck Duck Go"}
 ggdraw() + 
-  draw_image(image = "images/logical-puzzle-web-search.png")
-```
-
-```{r puzzle-chatgpt, out.width="50%", fig.cap="\\label{fig:puzzle-chatgpt}Figure 16. ChatGPT-4 makes the goat swim"}
+  draw_image(image = glue::glue(here::here(), "/images/logical-puzzle-web-search.png"))
+```
+
+```{r puzzle-chatgpt, out.width="70%", fig.cap="ChatGPT-4 makes the goat swim"}
 puzzle_plot <- ggplot() +
   labs(caption = "https://assemblag.es/@hugo/112450401657084750") +
-  theme_void()
+  theme_void() +
+  theme(text = element_text(size = 8))
 
@@
   draw_plot(puzzle_plot) +
-  draw_image(image = "images/logical-puzzle-chat-gpt-4.png")
+  draw_image(image = glue::glue(here::here(), "/images/logical-puzzle-chat-gpt-4.png"))
 
@@
 
+\newpage
+
 In a sense, an AI is like a crutch.
@@
 
-```{r google-reddit, out.width="50%", fig.cap="Figure 13. Google's AI Overview result was traced back to a post by redditor Fucksmith"}
-knitr::include_graphics("images/google-cheese-glue.png")
+```{r google-reddit, out.width="70%", fig.cap="Google's AI Overview result was traced back to a post by redditor Fucksmith"}
+knitr::include_graphics(glue::glue(here::here(), "/images/google-cheese-glue.png"))
 ```
@@
 
-```{r google-the-onion, out.width="50%", fig.cap="Figure 14. Google's AI Overview result was traced back to an article in The Onion"}
-knitr::include_graphics("images/google-the-rock.jpg")
-```
+```{r google-the-onion, out.width="70%", fig.cap="Google's AI Overview result was traced back to an article in The Onion"}
+knitr::include_graphics(glue::glue(here::here(), "/images/google-the-rock.jpg"))
+```
+
+\newpage
 
@@
          Variance = round(variance, 2)) |>
-  kable(caption = "\\ref{tab:data-vs-prediction-variance}Table 1. Variance reduction in model outputs")
+  kable(caption = "Variance reduction in model outputs")
 ```
@@
 As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023].
-One such tool is @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts.
-Another tool is @xu_shadowcast_2024's _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.
+One such tool is _PhotoGuard_, developed by a team at MIT to send signals to an AI that prevents it from manipulating images [@heikkila_this_2023].
+Other examples include @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts.
+And more recently, @xu_shadowcast_2024 presented _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.
 
@@
 
+## What do GenAIs do for the Province?
+
+As I argue above, AIs do not seem to do much for students or instructors, other than distract them from learning and developing genuinely transferable skills.
+On the other hand, it is hard not to suspect that the provincial government is hoping to use GenAIs to further reduce the need to fund higher education.
+But AIs must do something for someone, otherwise why would they be everywhere?
+For Ontario, the main patron and effectively the boss of McMaster, it provides a rationale (an excuse even) to force the institution to find "efficiencies".
+Tom Goldstein, a computer science professor at the University of Maryland, thinks that the lure of GenAI will be hard to resist for many, irrespective of their flaws...after all, "[e]ven though [GenAI is] expensive...it's still far less expensive than human labor" [@oremus_chatgpt_2023].
+Without adjudicating the merits of the argument that human labor is expensive^[When the alternative is substandard quality of life for many of our fellow citizens, and/or higher unemployment.], it is hard to argue against the lure of GenAI for governments that do not seem to believe in higher education.
+
 ## What do GenAIs do for the mega-corporations pushing them? {#what-do-ais-to-corporations}
@@
 
-As I argue above, AIs do not do much for students or instructors, other than distract them from learning and developing genuinely transferable skills.
-
-But AIs must do something for someone, otherwise why would they be everywhere?
-What do AIs do for the likes of OpenAI, Microsoft, and Google?
-So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as those spent by universities to support an AI-dedicated committee^[Some observers have even wondered whether the AI fever is not actually becoming detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].].
+But neither the province nor the university are developing the AIs that they hope will make them cheaper to operate.
+So what do AIs do for the the entities that develop them, namely OpenAI et al.?
+So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as those spent by universities to support an AI-dedicated committee^[Some observers have even wondered whether the AI fever has not actually become detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].].
 Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].
@@
 @patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google.
-At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer.
+At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 monthly queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer.
 And a large number of users are not even paying subscribers: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023].
 The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing the most intensive, disturbing, degrading, and damaging aspects of training AIs to the Global South [@nix_analysis_2024].
-The fact that the industry is not profitable has not prevented start-ups from hurriedly pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted from cryptocurrencyto AI, coincidentally as cryptocurrency was going sour after the frauds at FTX were revealed.
+The fact that the industry is not profitable has not prevented start-ups from hurriedly pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted from cryptocurrency to AI, coincidentally as cryptocurrency was going sour after the frauds at FTX were revealed.
 Barely disguised grift has grown abundant in the vicinity of AI at almost every level [@cook_5_2024].].
@@
 Or, in the worst case, _have_ to pay, if their skills become inexorably tied to this one technology?
-Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability^[And, with profitability becoming a more distant possibility, there is always the option for large investors to cosmically inflate the hype [@salmon_what_2024] to convince bigger suckers to give them money [@reuters_openais_2024] and/or to fleece retail investors [@temkin_vcs_2024].]: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020].
+Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability^[And, with profitability increasingly looking like a distant possibility, there is always the option for large investors to cosmically inflate the hype [@salmon_what_2024] in order to convince bigger suckers to give them money [@reuters_openais_2024] and/or to fleece retail investors [@temkin_vcs_2024].]: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020].
 From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service^[Universities, rather than being innovators, have increasingly become renters in the land of the platform economy [@komljenovic_rise_2021].].
@@
 
-The key here is "labeled examples".
+The key here is "labelled examples".
 What are those labels?
@@
 Or bits of text that have been labelled as...well...questionable.
-Labelling data is extremely time consuming and labor intensive.
+Labeling data is extremely time consuming and labor intensive.
 What's a corporation to do?
@@
 
+```{r google-obama, out.width="90%", fig.cap="\\label{fig:google-ai-overview-of-obama}Google's AI Overview of Muslim US presidents"}
+google_plot <- ggplot() +
+  labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
+  theme_void() +
+  theme(text = element_text(size = 8))
+
+ggdraw() + 
+  draw_plot(google_plot) +
+  draw_image(image = glue::glue(here::here(), "/images/google-obama.png"))
+```
+
 One does not need to posit a conspiracy theory^[Accusing the critics of being conspiracy theorists, as @sennesh_conspiracy_2023 do, distracts from the fact that there is no dearth of leftist ideas for a more just future; intellectual poverty is not what is at stake, instead a key difference is that progressive and humanistic ideas seldom have the power of money behind them.
@@
 In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].].
-The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].
+The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021].
+This is not a bug of the system, but rather a feature, as Meta's supression of information about Gaza in 2024 shows [@paul_former_2024].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].
 
@@
 Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring.
+
+\newpage
+
 Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China.
@@
 VandeHei has not been timid to express himself otherwise, and is on record stating that AI "will eviscerate the weak, the ordinary, the unprepared in media" [@robertson_axios_2024].
-Who should be concerned about maintaining an edge?]?
-
-```{r google-obama, out.width="50%", fig.cap="\\label{fig:google-ai-overview-of-obama}Figure 17. Google's AI Overview of Muslim US presidents"}
-google_plot <- ggplot() +
-  labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
-  theme_void()
-
-ggdraw() + 
-  draw_plot(google_plot) +
-  draw_image(image = "images/google-obama.png")
-```
+Who is concerned about maintaining an edge?]?
 
@@
 
-Neural networks and machine learning are not new, and they have their uses. But those uses have been drown, and to some extent discredited, by the current hype around AI.
-
-@morrone_signals_2024 Carefully manufactured and well funded hype.
-
-ChatGPT was a colossal feat of marketing [@cowen_chatgpt_2023], but much coverage of AI in the media has been stretched beyond the breaking point of credulity [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.].
-There are now calls to press pause on the hype [@angwin_press_2024] but these are still drown by the uncritical coverage.
-
+Neural networks and machine learning are nothing new, and they have their uses.
+But those uses have been suffocated, and to some extent discredited, by the current hype around Generative AI.
+This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools that have been more than willing to believe in magic beans [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.].
+There are now calls to press pause on the hype [@angwin_press_2024] but they are barely audible above the reverberating echos inside the AI hype chamber.
+
+Universities and colleges would be wise to heed the advice to press the pause button on the AI hype, as there is a real risk of them falling prey to a confidence trick.
+Higher education is often driven by a deeply ingrained belief in innovation (even, or perhaps particularly, for its own sake), and this impulse is compounded by the way AI adoption has been drawn as a false dilemma, thus triggering a fear of missing out.
+There are other pressures at work, including a decades-long history of manufactured crises in education, in Canada as elsewhere [@cizek_give_1999; @usher_state_2023].
+
+In this essay I have tried to describe the many problems with AI as an _everything tool_.
+Tall in my mind are the moral and ethical concerns around an extractive and exploitative technology that promises democracy but delivers further concentration of wealth and power.
+
+<!--
 > An end to “batch” teaching.
@@
 AI systems can also leverage technologies to ensure the resources and supports for learners reflect who they are (culture, language, experience and skill) and what challenges they face (disabilities, learning styles, etc.).
+-->
 
@@
 
-I suggest that universities should avoid the mistakes of the media, an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) (also see the "pivot-to-video" and how tech companies suckered many media companies, including ).
+I suggest that universities should avoid the mistakes of the media, an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) [also see the "pivot-to-video" and how tech companies suckered many media companies into giving them money in exchange for fake engagement statistics [@hazard_owen_facebooks_2021].].
 Tech corporations (or least ways some of them), can still hope to recoup some of their investment in AI.
@@
 "This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an institution of higher-education is not needed when there are cheaper alternatives for such an education ("act in the next four days to proof yourself against layoffs!"):
-
+<!--
 ![](images/trail-blazing-ai-bundle.png)
-
-What do AIs do? For the province, the main patron and effectively the boss of McMaster, it provides a rationale (an excuse even) to force the institution to find "efficiencies".
+-->
 
@@
 
-```{r googly-eyes, out.width="50%", fig.cap="\\label{fig:googly-ai-eyes}Figure 18. In conclusion"}
+```{r googly-eyes, out.width="70%", fig.cap="\\label{fig:googly-ai-eyes}In conclusion"}
 google_plot <- ggplot() +
   labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
-  theme_void()
+  theme_void()  +
+  theme(text = element_text(size = 8))
 
@@
   draw_plot(google_plot) +
-  draw_image(image = "images/ai-googly-eyes.png", width = 0.8)
-```
-
-## Notes
+  draw_image(image = glue::glue(here::here(), "/images/ai-googly-eyes.png", width = 0.8))
+```
+
+## Appendices
 

### 2024-06-19 README.Rmd
First full draft completed

@@
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(glue)
+library(here)
 library(knitr)
@@
 
-By: Antonio Páez (paezha@mcmaster.ca)
+Antonio Páez (paezha@mcmaster.ca)
 
@@
 _No generative AI tool was used to research or write this essay, only good old-fashioned library and internet search skills_.
+
+Note: _This essay was written as an open research project, and it can be found in the following repository:_ https://github.com/paezha/University-AI-Panel
 
@@
 ```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 1. An example of Atlassian AI writing code"}
-knitr::include_graphics("images/ai-bug-fix.png")
+knitr::include_graphics(glue::glue(here::here(), "/images/ai-bug-fix.png"))
 ```
@@
 ```{r ai-litter-box, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 2. Revolutionizing pet health"}
-knitr::include_graphics("images/ai-litter-box.png")
+knitr::include_graphics(glue::glue(here::here(), "/images/ai-litter-box.png"))
 ```
@@
 
-In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 80 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone.
+In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 90 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone.
 As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems.
@@
 ```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 5. Simple Example of a Shallow Neural Network"}
-knitr::include_graphics("images/nn1.png")
+knitr::include_graphics(glue::glue(here::here(), "/images/nn1.png"))
 ```
@@
 ```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 7. Example of a Deeper Neural Network"}
-knitr::include_graphics("images/nn2.png")
+knitr::include_graphics(glue::glue(here::here(), "/images/nn2.png"))
 ```
@@
 ```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 9. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
-knitr::include_graphics("images/mac-library-econometrics.png")
+knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-econometrics.png"))
 ```
@@
 ```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 10. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
-knitr::include_graphics("images/mac-library-cell-biology.png")
+knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-cell-biology.png"))
 ```
@@
 ```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 11. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
-knitr::include_graphics("images/mac-library-euphemism.png")
+knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-euphemism.png"))
 ```
@@
 Who fine-tunes the machine that provides the feedback?
-Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided^[[Automation bias](https://en.wikipedia.org/wiki/Automation_bias), the human inclination to favor suggestions made by automated decision-making systems, means that even when a human takes responsibility for decisions, AI companies remain unaccountable for their role in influencing outcomes.
+Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided^[[Automation bias](https://en.wikipedia.org/wiki/Automation_bias), the human inclination to favor suggestions made by automated decision-making systems, means that even when a human takes responsibility for decisions, AI companies remain unaccountable for their role in influencing outcomes^[GenAI is not a toy, as Air Canada discovered when the company was found liable for "negligent misrepresentation", after a court denied claims by the airline that the chatbot it deployed was somehow responsible for its own actions as an effectively separate legal entity [@garcia_what_2024].].
 A poignant example involving AI-based face recognition and online proctoring is related by @hill_accused_2022.]?
@@
 ```{r flickr-potato, out.width="50%", fig.cap="Figure 12. Data analysis using a chatbot? Maybe not such a good idea"}
-knitr::include_graphics("images/potato.png")
+knitr::include_graphics(glue::glue(here::here(), "/images/potato.png"))
 ```
@@
 ggdraw() + 
-  draw_image(image = "images/logical-puzzle-web-search.png")
+  draw_image(image = glue::glue(here::here(), "/images/logical-puzzle-web-search.png"))
 ```
@@
 ```{r google-reddit, out.width="50%", fig.cap="Figure 13. Google's AI Overview result was traced back to a post by redditor Fucksmith"}
-knitr::include_graphics("images/google-cheese-glue.png")
+knitr::include_graphics(glue::glue(here::here(), "/images/google-cheese-glue.png"))
 ```
@@
 ```{r google-the-onion, out.width="50%", fig.cap="Figure 14. Google's AI Overview result was traced back to an article in The Onion"}
-knitr::include_graphics("images/google-the-rock.jpg")
+knitr::include_graphics(glue::glue(here::here(), "/images/google-the-rock.jpg"))
 ```
@@
 
+## What do GenAIs do for the Province?
+
+As I argue above, AIs do not seem to do much for students or instructors, other than distract them from learning and developing genuinely transferable skills.
+
+But AIs must do something for someone, otherwise why would they be everywhere?
+Given history, one could be forgiven to think that the provincial government is hoping to use GenAIs to further reduce the need to fund higher education.
+GenAIs appear on the horizon as an  innovation-sounding rationale (a mandate even!) for the province to force institutions of higher education to find "efficiencies".
+Tom Goldstein, a computer science professor at the University of Maryland, thinks that the lure of GenAI will be hard to resist for many, irrespective of the technology's flaws...after all, "[e]ven though [GenAI is] expensive...it's still far less expensive than human labor" [@oremus_chatgpt_2023].
+Without adjudicating the merits of the argument that human labor is expensive^[When the alternative is substandard quality of life for many of our fellow citizens, and/or higher unemployment.], one would hav to be blind not to see the glamour that GenAI will cast on governments that do not seem to believe in higher education in the first place.
+
+```{r googly-eyes, out.width="70%", fig.cap="\\label{fig:googly-ai-eyes}'I like the look of his cuts'"}
+google_plot <- ggplot() +
+  labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
+  theme_void()  +
+  theme(text = element_text(size = 8))
+
+ggdraw() + 
+  draw_plot(google_plot) +
+  draw_image(image = glue::glue(here::here(), "/images/ai-googly-eyes.png", width = 0.8))
+```
+
 ## What do GenAIs do for the mega-corporations pushing them? {#what-do-ais-to-corporations}
@@
 
-ChatGPT 4.o was announced on May 13, 2024, with fanfare and a flashy display of the model's new capabilities, which included the ability to respond vocally to prompts, and to identify images via a mobile device's camera.
+ChatGPT 4.o was announced on May 13, 2024, with great fanfare and a flashy display of the model's new capabilities.
+These included the chatbot's ability to respond vocally to prompts, and to identify images via a mobile device's camera.
 4.o's voice, in particular, caused some titillation in the nether parts of even its creators [@roose_is_2024].
 The voice was modeled after the digital personal assistant of the movie _Her_, played by Scarlett Johannson.
-This film (to approximately no one's surprise) is, by his own account, one of OpenAI's Altman's favorites.
+By Sam Altman's own account, this is one of his favorite movies.
 The background story of 4.o's voice is somewhat sordid: Altman reportedly tried to hire Ms.
-Johannson to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when Ms.
-Johannson declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways.
+Johannson to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when she declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways.
 This voice was retired (probably to the profound sadness of many an AI engineer) after Ms.
@@
 
-Alas, very few have the social standing of Scarlett Johansson to challenge OpenAI.
-Most people in the world lack that power.
-In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )
-
-@germain_google_2023
-@norem_intels_2023
-
-The ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted.
-
-@williams_exploited_2022 (Summary needed)
-
-@perrigo_2_2023 (Follows summary)
-
+Alas, very few have the social standing and wealth of Scarlett Johansson to challenge OpenAI.
+Most people in the world lack that power and the AI industry has not been shy to exploit that imbalance [@roberts_commercial_2016], with Google's privacy policy for using the whole internet as the source of data for its AI efforts [@germain_google_2023], and Adobe landing in hot water by adopting terms of use widely thought to be confiscatory [@weatherbed_adobes_2024].
+
+Besides being extractive, the AI industry is exploitative in darker ways.
 ChatGPT was trained on vast amounts of data scooped from the internet, where many societal problems are amplified.
 Trained on horrible content, it was hard to sell ChatGPT when it was prone to racist, sexist, obscene outbursts.
-But companies like facebook, who did not want to pay for moderation, had hit on a solution of sorts <!--: train an AI to detect content that was unpalatable.-->
-
-The solution?
-
-> [F]eed an AI with labeled examples of violence, hate speech, and sexual abuse, and that tool could learn to detect those forms of toxicity in the wild.
-That detector would be built into ChatGPT to check whether it was echoing the toxicity of its training data, and filter it out before it ever reached the user.
-It could also help scrub toxic text from the training datasets of future AI models.
-
-The key here is "labeled examples".
-What are those labels?
-Many machine algorithms do not learn by themselves...they learn from humans, and need what is called "labelled" data: a images that humans have labelled as "dog", "cats", etc.
-Or bits of text that have been labelled as...well...questionable.
-Labelling data is extremely time consuming and labor intensive.
-What's a corporation to do?
-
-> To get those labels, OpenAI sent tens of thousands of snippets of text to an outsourcing firm in Kenya, beginning in November 2021.
-Much of that text appeared to have been pulled from the darkest recesses of the internet.
-Some of it described situations in graphic detail like child sexual abuse, bestiality, murder, suicide, torture, self harm, and incest.
-"
-
-What did this mean to the people doing all this labeling in one of the poorest regions in the world?
+Fortunately (or unfortunately, depending on perspective) companies like facebook, who did not want to pay for moderation, had already hit on a solution of sorts, which was basically to train AIs to detect content that was unpalatable.
+Indeed, they had found that feeding an AI with labeled examples of all the worst that exists on the internet (e.g., violence, hate speech, child sexual abuse) could produce tools that learned to detect various manifestations of such toxicity in the wild, to filter it out before it reached users.
+However, to obtain the essential labeled inputs, OpenAI sent tens of thousands of samples to Kenya, with text that described murder, suicide, bestiality, self-harm, incest, torture, and child sexual abuse situations in graphic detail.
+
+What did this mean to the people doing all the labeling in one of the poorest regions in the world?
 First of all, an income: "data labelers employed...on behalf of OpenAI were paid a take-home wage of between around $1.32 and $2 per hour depending on seniority and performance." But at what cost?
-
-> One Sama worker tasked with reading and labeling text for OpenAI told TIME he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child.
-“That was torture,” he said.
-“You will read a number of statements like that all through the week.
-By the time it gets to Friday, you are disturbed from thinking through that picture."
+As one worker tasked with reading and labeling text for OpenAI told TIME, "he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child.
+'That was torture,' he said.
+'You will read a number of statements like that all through the week.
+By the time it gets to Friday, you are disturbed from thinking through that picture'" [@perrigo_2_2023].
+
+<!--
+
+In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )
+
+@norem_intels_2023
+
+The ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted.
+
+@williams_exploited_2022 (Summary needed)
 
@@
 You'll get more confusing sentences that don't make sense, and you will potentially get more of this content that they worked so hard to filter out because they were only filtering it in English.
-
-@roberts_commercial_2016 Commercial {Content} {Moderation}: {Digital} {Laborers}' {Dirty} {Work}
+-->
 
 ### Copyright (and privacy) issues
+
+### Privacy issues
 
@@
 
-Where will new data come from? [@murgia_why_2023]
-
-Note how several organizations have responded to the AIfication of all the things: by banning it.
+This means that data cannot be presumed to be private once that it has been used to train an AI.
+In response organizations such as the National Archives and Records Administration in the US [@koebler_national_2024], and the US Congress [@singh_after_2024] have banned the use of ChatGPT and Copilot.
+@singh_after_2024 also notes how companies like Samsung and Apple have restricted their use of chatbots.
+
+NetBSD (a project that produces a free, Unix-like Open Source operating system) states in its commit guidelines (contributions to code) that "[c]ode generated by a large language model or similar technology...is presumed to be tainted code, and must not be committed without prior written approval by core." [@netbsd_netbsd_nodate].
 The concerns cited are seldom ethical or moral, but mostly about the risk of sensitive information leaking.
-As LLMs begin to be increasingly trained on model-produce data, overfitting (or memorization) can become a larger risk.
-
-And this is just a result of using an overfitted model. What about data leaks? Some stories regarding Microsoft's security track record:
-
-https://www.theregister.com/2024/04/12/microsoft_cisa_order/
-
-https://techreport.com/news/microsoft-azure-hit-with-the-largest-data-breach-in-its-history-hundreds-of-executive-accounts-compromised/
-
-Also, how confident can the public be about assurances that a big tech corporation will not data mine and commercialize data by the companies that sell the services of LLMs?
-Microsoft is so big that it can stand up to the US government:
-
-https://www.theregister.com/2024/04/05/microsoft_government_contracts/
-
-<!-- "Nice operation you have there...be a shame if your systems stopped working all of a sudden."-->
-
-Organization ranging from Free Open Source projects to elements of the federal government in the United States have limited/restricted the use of AI.
-
-NetBSD (a project that produces a free, Unix-like Open Source operating system) states in its commit guidelines (contributions to code) that "[c]ode generated by a large language model or similar technology...is presumed to be tainted code, and must not be committed without prior written approval by core." [@netbsd_netbsd_nodate].
-
-The National Archives and Records Administration (USA) bans use of ChatGPT [@koebler_national_2024]
-
-US Congress bans use of ChatGPT and Copilot for staff [@singh_after_2024].
-This article also note how companies like Samsung and Apple have restricted the use of chatbots citing concerns with data security.
+As LLMs begin to be trained on model-produced data [@murgia_why_2023], overfitting (or memorization) will likely become a larger risk.
+
+And this is just a result of using an overfitted model.
+What about data leaks?
+Microsoft's data security [track](https://www.theregister.com/2024/04/12/microsoft_cisa_order/) [record](https://techreport.com/news/microsoft-azure-hit-with-the-largest-data-breach-in-its-history-hundreds-of-executive-accounts-compromised/) does not inspire confidence.
 
@@
 
+<!--
 I struggle a bit with this one.
@@
 Who am I to dictate how much damage _other people_ have a right to inflict (or not) to the environment.
-
-Microsoft: Here's how burning the planet will help to save it.
-
-https://blogs.microsoft.com/on-the-issues/2024/05/15/microsoft-environmental-sustainability-report-2024/ [@smith_our_2024]
-
-https://blogs.microsoft.com/on-the-issues/2023/11/16/accelerating-sustainability-ai-playbook/ [@nakagawa_accelerating_2023]
-
+-->
+
+Every human activity has an environmental impact, and at a time when we face unprecedented challenges with climate change, we must be discerning in how we consume resources.
+The choice is so clear that Microsoft has basically told us: we need to burn the planet [@smith_our_2024] to create the AIs that will help us save the planet [@nakagawa_accelerating_2023].
+
+@crawford_generative_2024 has already sounded the alarm: the environmental costs of AI are soaring, and doing so mostly in secret [@luccioni_power_2024].
+It has been estimated that ChatGPT consumes substantially more energy than a regular web search, up to 25 times more, and according to some estimates, the whole of AI may consume twice as much energy as the whole of France by 2030 [@brussels_chatgpt_2024].
+As the examples in the preceding sections illustrate, this is in exchange for results that are not better than a simple search, and often are worse.
+Widespread AI deployment may in fact make the climate crisis worse, not only by guzzling resources, but also by spreading misinformation [@gordon_multiple_2024].
+
+<!--
 https://www.theguardian.com/commentisfree/article/2024/may/30/ugly-truth-ai-chatgpt-guzzling-resources-environment
 
-"Environmental costs of AI are soaring" [@crawford_generative_2024]
-
 https://www.wired.com/story/the-generative-ai-search-race-has-a-dirty-secret/
 
-ChatGPT consumes substantially more energy than a regular web search, up to 25 times more, and according to some estimates, the whole of AI may consume twice as much energy as the whole of France by 2030 [@brussels_chatgpt_2024].
-
 ...are the results better?
@@
 
-The example above may seem like an amusement, but a GenAI is not a toy, as Air Canada discover when the company was found liable for "negligent misrepresentation", after a court denied claims by the airline that the chatbot was somehow responsible for its own actions as an effectively separate legal entity [@garcia_what_2024].
+The example above may seem like an amusement, but a .
 
@@
 [@angwin_seeking_2024].
+-->
 
@@
 VandeHei has not been timid to express himself otherwise, and is on record stating that AI "will eviscerate the weak, the ordinary, the unprepared in media" [@robertson_axios_2024].
-Who should be concerned about maintaining an edge?]?
+Who is concerned about maintaining an edge?]?
 
@@
   labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
-  theme_void()
+  theme_void() +
+  theme(text = element_text(size = 8))
 
@@
   draw_plot(google_plot) +
-  draw_image(image = "images/google-obama.png")
+  draw_image(image = glue::glue(here::here(), "/images/google-obama.png"))
 ```
@@
 
-Neural networks and machine learning are not new, and they have their uses. But those uses have been drown, and to some extent discredited, by the current hype around AI.
-
-@morrone_signals_2024 Carefully manufactured and well funded hype.
-
-ChatGPT was a colossal feat of marketing [@cowen_chatgpt_2023], but much coverage of AI in the media has been stretched beyond the breaking point of credulity [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.].
-There are now calls to press pause on the hype [@angwin_press_2024] but these are still drown by the uncritical coverage.
-
-> An end to “batch” teaching.
-Instead of working with cohorts of learners in grades or classes, individual learning agendas are developed based on an assessment of the skills, competencies and capabilities the learner already has as well as ongoing conversations about the purpose of the learning.
+Neural networks and machine learning are nothing new, and they have their uses.
+But those uses have been suffocated, and to some extent discredited, by the current hype around Generative AI.
+This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools that have been more than willing to believe in magic beans [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.].
+There are now calls to press pause on the hype [@angwin_press_2024] but these voices are barely audible above the thundering echos inside the AI hype chamber.
+
+Universities and colleges would be wise to heed the advice to press the pause button on the AI hype, as there is a real risk of them falling prey to a confidence trick.
+Higher education is often driven by a deeply ingrained belief in innovation (even, or perhaps particularly, for its own sake), and this impulse is compounded by the way AI adoption has been drawn as a false dilemma, thus triggering a fear of missing out.
+There are other pressures at work, including a decades-long history of manufactured crises in education, in Canada as elsewhere [@cizek_give_1999; @usher_state_2023].
+
+In this essay I have tried to describe the many problems with AI as an _everything tool_.
+Tall in my mind are the moral and ethical concerns around an extractive and exploitative technology that promises democracy but delivers further concentration of wealth and power.
+GenAI is environmentally damaging for what it delivers: confidently stated confabulations.
+I have little illusion that these arguments will prove persuasive, since the university is already aware of them because they usually appear in the last slides of every presentation about--and yet here we are, asking not whether, but where it makes sense to adopt AI.
+
+I would still appeal, though, to the university's self-regard and sense of self-preservation.
+
+It has now been 55 years since Howard S.
+Barrows and Robyn M.
+Tamblyn pioneered [problem-based learning](https://en.wikipedia.org/wiki/Problem-based_learning#History) at McMaster, an accomplishment that the university [still touts today](https://provost.mcmaster.ca/teaching-learning/).
+Problem-based learning was not and is still not high tech, and yet remains a remarkable pedagogical breakthrough, an approach that has stayed influential for decades.
+Developing problem-based learning took human ingenuity and insight.
+What the university is considering today is which of our core missions can we hitch to an expensive technology of questionable value today, of dubious longevity, and that in the here and now advances an agenda in education that can only be squared with "integrity, quality, and excellence" via fabulous contortions of the imagination.
+
+Does the university imagine that it will be best known in half a century as "McMaster, the institution that pioneered the use of ChatGPT"?
+Is that what it really wants?
+All that AIs offer, after all, is a very sophisticated [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean), which not by coincidence is also called a "reversion to mediocrity" [@barnett_regression_2017].
+We would do our students an unforgivable disservice by offering a skill that reversed them to mediocrity.
+"This university uses ChatGPT to teach" is simply a way of saying that we are too cheap to offer a human-centered education.
+But in truth, a university is too expensive an institution to be attractive under such conditions, when there are far cheaper alternatives for an AI-based education ("for only $9.99 proof yourself against layoffs").
+
+```{r ai-bundle, out.width="110%", fig.cap="\\label{fig:trail-blazing-bundle}A trail-blazing bundle: ChatGPT for data science and data analysis in Python"}
+ggdraw() + 
+  draw_image(image = glue::glue(here::here(), "/images/trail-blazing-ai-bundle.png", width = 0.8))
+```
+
+Going down this path would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors, a university's main source of authority is its reputation as a place that preserves and expands knowledge in a principled way.
+
+I suggest that universities would do well to avoid the mistakes of other sectors, including those in the media that have contributed to hype AI.
+The media is an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) [also see the "pivot-to-video" and how tech companies suckered many media companies into giving them money in exchange for fake engagement statistics [@hazard_owen_facebooks_2021].].
+Tech corporations (or at least some of them), can still hope to recoup some of the fabulous ammounts they have poured into AI.
+But a university gambles its most important asset when betting on AI: its reputation as a place were knowledge is created, and as a custodian of knowledge.
+AIs already have a reputation for producing mis- and disinformation--from the very titans of the field^[Entities like Google and Amazon could not care less about their reputation.
+They are [too big](https://www.businessinsider.com/ex-amazon-ghaderi-exec-suing-ai-race-copyright-allegations-2024) [to care](https://pluralistic.net/2024/04/04/teach-me-how-to-shruggie/).] to the unaware^[Like a B.C.
+lawyer who was [reprimanded](https://www.cbc.ca/news/canada/british-columbia/lawyer-chatgpt-fake-precedent-1.7126393) for using AI hallucinations in a legal filing.
+This was not even the first time that supposedly highly skilled professionals were castigated for [using](https://www.washingtonpost.com/technology/2023/11/16/chatgpt-lawyer-fired-ai/) [chatbots](https://www.reuters.com/legal/new-york-lawyers-sanctioned-using-fake-chatgpt-cases-legal-brief-2023-06-22/).], and beyond, to bad-faith actors^[Like rival geopolitical powers that are relatively impervious to propaganda themselves for reason of being [authoritarian countries](https://www.theatlantic.com/magazine/archive/2024/06/china-russia-republican-party-relations/678271/).].
+
+In my view, there are no efficiencies, no questionable promises of future excellence, that can compensate the one thing that makes a university stand tall among other institutions--its pursuit of innovation, yes, but above all of the truth, because what else is worth discovering, preserving, and communicating?
+
+<!--
+> Instead of working with cohorts of learners in grades or classes, individual learning agendas are developed based on an assessment of the skills, competencies and capabilities the learner already has as well as ongoing conversations about the purpose of the learning.
 Where does the student want the learning to take them?
@@
 AI systems can also leverage technologies to ensure the resources and supports for learners reflect who they are (culture, language, experience and skill) and what challenges they face (disabilities, learning styles, etc.).
-
-It has now been 55 years since Howard S.
-Barrows and Robyn M.
-Tamblyn [pioneered](https://en.wikipedia.org/wiki/Problem-based_learning#History) problem-based learning at McMaster, an accomplishment that the university [still touts today](https://provost.mcmaster.ca/teaching-learning/).
-Problem-based learning was not and is still not high tech, and yet remains a great pedagogical approach, influential to the date.
-Developing problem-based learning took human ingenuity and insight, not getting hooked on an expensive technology of questionable value today, of dubious longevity, and that in the here and now is unarguably extractive, abusive, and exploitative, and that advances an agenda that can only be squared with "integrity, quality, and excellence" via fabulous contortions of the imagination.
-
-Does the university imagine that it will be best known in half a century as "McMaster, the institution that pioneered the use of ChatGPT"?
-Is that what it really wants?
-[Regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean), after all, is also called a "reversion to mediocrity" [@barnett_regression_2017], and that is what an AI campus has to offer.
-
-I suggest that universities should avoid the mistakes of the media, an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) (also see the "pivot-to-video" and how tech companies suckered many media companies, including ).
-Tech corporations (or least ways some of them), can still hope to recoup some of their investment in AI.
-But a university gambles its most important asset when betting on AI: its reputation as a place were knowledge is created, and as a custodian of knowledge.
-AIs already have a reputation for producing mis- and disinformation--from the very titans of the field^[Entities like Google and Amazon could not care less about their reputation.
-They are [too big](https://www.businessinsider.com/ex-amazon-ghaderi-exec-suing-ai-race-copyright-allegations-2024) [to care](https://pluralistic.net/2024/04/04/teach-me-how-to-shruggie/).] to the unaware^[Like a B.C.
-lawyer who was [reprimanded](https://www.cbc.ca/news/canada/british-columbia/lawyer-chatgpt-fake-precedent-1.7126393) for using AI hallucinations in a legal filing.
-This was not even the first time that supposedly highly skilled professionals were castigated for [using](https://www.washingtonpost.com/technology/2023/11/16/chatgpt-lawyer-fired-ai/) [chatbots](https://www.reuters.com/legal/new-york-lawyers-sanctioned-using-fake-chatgpt-cases-legal-brief-2023-06-22/).], and beyond, to bad-faith actors^[Like rival geopolitical powers that are relatively impervious to propaganda themselves for reason of being [authoritarian countries](https://www.theatlantic.com/magazine/archive/2024/06/china-russia-republican-party-relations/678271/).].
-In my view, there are no efficiencies, no promises of future excellence, that can compensate the risk of ruining an institutions reputation.
-
-Why?
-The reputational risk...universities are under siege from unsympathetic or actively hostile actors.
-Universities main source of power is their reputation as places that preserve and expand knowledge in a principled way.
-"This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an institution of higher-education is not needed when there are cheaper alternatives for such an education ("act in the next four days to proof yourself against layoffs!"):
-
-![](images/trail-blazing-ai-bundle.png)
-
-What do AIs do? For the province, the main patron and effectively the boss of McMaster, it provides a rationale (an excuse even) to force the institution to find "efficiencies".
 
@@
 AI industry and the military: @haskins_im_2024
-
-## Bonus
-
-```{r googly-eyes, out.width="50%", fig.cap="\\label{fig:googly-ai-eyes}Figure 18. In conclusion"}
-google_plot <- ggplot() +
-  labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
-  theme_void()
-
-ggdraw() + 
-  draw_plot(google_plot) +
-  draw_image(image = "images/ai-googly-eyes.png", width = 0.8)
-```
-
-## Notes
+-->
+
+## Appendices
 

### 2024-06-19 ai-haunting-campus/ai-haunting-campus.qmd
First full draft completed

@@
 
+License: [CC-BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/)
+
+![](by-sa.png){width=50}
+
 ## Background
@@
 As I argue above, AIs do not seem to do much for students or instructors, other than distract them from learning and developing genuinely transferable skills.
-On the other hand, it is hard not to suspect that the provincial government is hoping to use GenAIs to further reduce the need to fund higher education.
+
 But AIs must do something for someone, otherwise why would they be everywhere?
-For Ontario, the main patron and effectively the boss of McMaster, it provides a rationale (an excuse even) to force the institution to find "efficiencies".
-Tom Goldstein, a computer science professor at the University of Maryland, thinks that the lure of GenAI will be hard to resist for many, irrespective of their flaws...after all, "[e]ven though [GenAI is] expensive...it's still far less expensive than human labor" [@oremus_chatgpt_2023].
-Without adjudicating the merits of the argument that human labor is expensive^[When the alternative is substandard quality of life for many of our fellow citizens, and/or higher unemployment.], it is hard to argue against the lure of GenAI for governments that do not seem to believe in higher education.
+Given history, one could be forgiven to think that the provincial government is hoping to use GenAIs to further reduce the need to fund higher education.
+GenAIs appear on the horizon as an  innovation-sounding rationale (a mandate even!) for the province to force institutions of higher education to find "efficiencies".
+Tom Goldstein, a computer science professor at the University of Maryland, thinks that the lure of GenAI will be hard to resist for many, irrespective of the technology's flaws...after all, "[e]ven though [GenAI is] expensive...it's still far less expensive than human labor" [@oremus_chatgpt_2023].
+Without adjudicating the merits of the argument that human labor is expensive^[When the alternative is substandard quality of life for many of our fellow citizens, and/or higher unemployment.], one would hav to be blind not to see the glamour that GenAI will cast on governments that do not seem to believe in higher education in the first place.
+
+```{r googly-eyes, out.width="70%", fig.cap="\\label{fig:googly-ai-eyes}'I like the look of his cuts'"}
+google_plot <- ggplot() +
+  labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
+  theme_void()  +
+  theme(text = element_text(size = 8))
+
+ggdraw() + 
+  draw_plot(google_plot) +
+  draw_image(image = glue::glue(here::here(), "/images/ai-googly-eyes.png", width = 0.8))
+```
 
@@
 
-Alas, very few have the social standing of Scarlett Johansson to challenge OpenAI.
-Most people in the world lack that power.
-In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )
-
-@germain_google_2023
-@norem_intels_2023
-
-The ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted.
-
-@williams_exploited_2022 (Summary needed)
-
-@perrigo_2_2023 (Follows summary)
-
+Alas, very few have the social standing and wealth of Scarlett Johansson to challenge OpenAI.
+Most people in the world lack that power and the AI industry has not been shy to exploit that imbalance [@roberts_commercial_2016], with Google's privacy policy for using the whole internet as the source of data for its AI efforts [@germain_google_2023], and Adobe landing in hot water by adopting terms of use widely thought to be confiscatory [@weatherbed_adobes_2024].
+
+Besides being extractive, the AI industry is exploitative in darker ways.
 ChatGPT was trained on vast amounts of data scooped from the internet, where many societal problems are amplified.
 Trained on horrible content, it was hard to sell ChatGPT when it was prone to racist, sexist, obscene outbursts.
-But companies like facebook, who did not want to pay for moderation, had hit on a solution of sorts <!--: train an AI to detect content that was unpalatable.-->
-
-The solution?
-
-> [F]eed an AI with labeled examples of violence, hate speech, and sexual abuse, and that tool could learn to detect those forms of toxicity in the wild.
-That detector would be built into ChatGPT to check whether it was echoing the toxicity of its training data, and filter it out before it ever reached the user.
-It could also help scrub toxic text from the training datasets of future AI models.
-
-The key here is "labelled examples".
-What are those labels?
-Many machine algorithms do not learn by themselves...they learn from humans, and need what is called "labelled" data: a images that humans have labelled as "dog", "cats", etc.
-Or bits of text that have been labelled as...well...questionable.
-Labeling data is extremely time consuming and labor intensive.
-What's a corporation to do?
-
-> To get those labels, OpenAI sent tens of thousands of snippets of text to an outsourcing firm in Kenya, beginning in November 2021.
-Much of that text appeared to have been pulled from the darkest recesses of the internet.
-Some of it described situations in graphic detail like child sexual abuse, bestiality, murder, suicide, torture, self harm, and incest.
-"
-
-What did this mean to the people doing all this labeling in one of the poorest regions in the world?
+Fortunately (or unfortunately, depending on perspective) companies like facebook, who did not want to pay for moderation, had already hit on a solution of sorts, which was basically to train AIs to detect content that was unpalatable.
+Indeed, they had found that feeding an AI with labeled examples of all the worst that exists on the internet (e.g., violence, hate speech, child sexual abuse) could produce tools that learned to detect various manifestations of such toxicity in the wild, to filter it out before it reached users.
+However, to obtain the essential labeled inputs, OpenAI sent tens of thousands of samples to Kenya, with text that described murder, suicide, bestiality, self-harm, incest, torture, and child sexual abuse situations in graphic detail.
+
+What did this mean to the people doing all the labeling in one of the poorest regions in the world?
 First of all, an income: "data labelers employed...on behalf of OpenAI were paid a take-home wage of between around $1.32 and $2 per hour depending on seniority and performance." But at what cost?
-
-> One Sama worker tasked with reading and labeling text for OpenAI told TIME he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child.
-“That was torture,” he said.
-“You will read a number of statements like that all through the week.
-By the time it gets to Friday, you are disturbed from thinking through that picture."
+As one worker tasked with reading and labeling text for OpenAI told TIME, "he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child.
+'That was torture,' he said.
+'You will read a number of statements like that all through the week.
+By the time it gets to Friday, you are disturbed from thinking through that picture'" [@perrigo_2_2023].
+
+<!--
+
+In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )
+
+@norem_intels_2023
+
+The ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted.
+
+@williams_exploited_2022 (Summary needed)
 
@@
 You'll get more confusing sentences that don't make sense, and you will potentially get more of this content that they worked so hard to filter out because they were only filtering it in English.
-
-@roberts_commercial_2016 Commercial {Content} {Moderation}: {Digital} {Laborers}' {Dirty} {Work}
-
-### Copyright (and privacy) issues
+-->
+
+### Privacy issues
 
@@
 [@angwin_seeking_2024].
+-->
+
+\newpage
 
@@
 Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that wealthy and well connected longtermists are willing to make.]).
-
-```{r google-obama, out.width="90%", fig.cap="\\label{fig:google-ai-overview-of-obama}Google's AI Overview of Muslim US presidents"}
-google_plot <- ggplot() +
-  labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
-  theme_void() +
-  theme(text = element_text(size = 8))
-
-ggdraw() + 
-  draw_plot(google_plot) +
-  draw_image(image = glue::glue(here::here(), "/images/google-obama.png"))
-```
 
@@
 
+```{r google-obama, out.width="90%", fig.cap="\\label{fig:google-ai-overview-of-obama}Google's AI Overview of Muslim US presidents"}
+google_plot <- ggplot() +
+  labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
+  theme_void() +
+  theme(text = element_text(size = 8))
+
+ggdraw() + 
+  draw_plot(google_plot) +
+  draw_image(image = glue::glue(here::here(), "/images/google-obama.png"))
+```
+
+\newpage
+
 Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
@@
 
-\newpage
-
 Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China.
@@
 This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools that have been more than willing to believe in magic beans [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.].
-There are now calls to press pause on the hype [@angwin_press_2024] but they are barely audible above the reverberating echos inside the AI hype chamber.
+There are now calls to press pause on the hype [@angwin_press_2024] but these voices are barely audible above the thundering echos inside the AI hype chamber.
 
@@
 Tall in my mind are the moral and ethical concerns around an extractive and exploitative technology that promises democracy but delivers further concentration of wealth and power.
+GenAI is environmentally damaging for what it delivers: confidently stated confabulations.
+I have little illusion that these arguments will prove persuasive, since the university is already aware of them because they usually appear in the last slides of every presentation about--and yet here we are, asking not whether, but where it makes sense to adopt AI.
+
+I would still appeal, though, to the university's self-regard and sense of self-preservation.
+
+It has now been 55 years since Howard S.
+Barrows and Robyn M.
+Tamblyn pioneered [problem-based learning](https://en.wikipedia.org/wiki/Problem-based_learning#History) at McMaster, an accomplishment that the university [still touts today](https://provost.mcmaster.ca/teaching-learning/).
+Problem-based learning was not and is still not high tech, and yet remains a remarkable pedagogical breakthrough, an approach that has stayed influential for decades.
+Developing problem-based learning took human ingenuity and insight.
+What the university is considering today is which of our core missions can we hitch to an expensive technology of questionable value today, of dubious longevity, and that in the here and now advances an agenda in education that can only be squared with "integrity, quality, and excellence" via fabulous contortions of the imagination.
+
+Does the university imagine that it will be best known in half a century as "McMaster, the institution that pioneered the use of ChatGPT"?
+Is that what it really wants?
+All that AIs offer, after all, is a very sophisticated [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean), which not by coincidence is also called a "reversion to mediocrity" [@barnett_regression_2017].
+We would do our students an unforgivable disservice by offering a skill that reversed them to mediocrity.
+"This university uses ChatGPT to teach" is simply a way of saying that we are too cheap to offer a human-centered education.
+But in truth, a university is too expensive an institution to be attractive under such conditions, when there are far cheaper alternatives for an AI-based education ("for only $9.99 proof yourself against layoffs").
+
+```{r ai-bundle, out.width="110%", fig.cap="\\label{fig:trail-blazing-bundle}A trail-blazing bundle: ChatGPT for data science and data analysis in Python"}
+ggdraw() + 
+  draw_image(image = glue::glue(here::here(), "/images/trail-blazing-ai-bundle.png", width = 0.8))
+```
+
+Going down this path would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors, a university's main source of authority is its reputation as a place that preserves and expands knowledge in a principled way.
+
+I suggest that universities would do well to avoid the mistakes of other sectors, including those in the media that have contributed to hype AI.
+The media is an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) [also see the "pivot-to-video" and how tech companies suckered many media companies into giving them money in exchange for fake engagement statistics [@hazard_owen_facebooks_2021].].
+Tech corporations (or at least some of them), can still hope to recoup some of the fabulous ammounts they have poured into AI.
+But a university gambles its most important asset when betting on AI: its reputation as a place were knowledge is created, and as a custodian of knowledge.
+AIs already have a reputation for producing mis- and disinformation--from the very titans of the field^[Entities like Google and Amazon could not care less about their reputation.
+They are [too big](https://www.businessinsider.com/ex-amazon-ghaderi-exec-suing-ai-race-copyright-allegations-2024) [to care](https://pluralistic.net/2024/04/04/teach-me-how-to-shruggie/).] to the unaware^[Like a B.C.
+lawyer who was [reprimanded](https://www.cbc.ca/news/canada/british-columbia/lawyer-chatgpt-fake-precedent-1.7126393) for using AI hallucinations in a legal filing.
+This was not even the first time that supposedly highly skilled professionals were castigated for [using](https://www.washingtonpost.com/technology/2023/11/16/chatgpt-lawyer-fired-ai/) [chatbots](https://www.reuters.com/legal/new-york-lawyers-sanctioned-using-fake-chatgpt-cases-legal-brief-2023-06-22/).], and beyond, to bad-faith actors^[Like rival geopolitical powers that are relatively impervious to propaganda themselves for reason of being [authoritarian countries](https://www.theatlantic.com/magazine/archive/2024/06/china-russia-republican-party-relations/678271/).].
+
+In my view, there are no efficiencies, no questionable promises of future excellence, that can compensate the one thing that makes a university stand tall among other institutions--its pursuit of innovation, yes, but above all of the truth, because what else is worth discovering, preserving, and communicating?
 
 <!--
-> An end to “batch” teaching.
-Instead of working with cohorts of learners in grades or classes, individual learning agendas are developed based on an assessment of the skills, competencies and capabilities the learner already has as well as ongoing conversations about the purpose of the learning.
+> Instead of working with cohorts of learners in grades or classes, individual learning agendas are developed based on an assessment of the skills, competencies and capabilities the learner already has as well as ongoing conversations about the purpose of the learning.
 Where does the student want the learning to take them?
@@
 AI systems can also leverage technologies to ensure the resources and supports for learners reflect who they are (culture, language, experience and skill) and what challenges they face (disabilities, learning styles, etc.).
--->
-
-It has now been 55 years since Howard S.
-Barrows and Robyn M.
-Tamblyn [pioneered](https://en.wikipedia.org/wiki/Problem-based_learning#History) problem-based learning at McMaster, an accomplishment that the university [still touts today](https://provost.mcmaster.ca/teaching-learning/).
-Problem-based learning was not and is still not high tech, and yet remains a great pedagogical approach, influential to the date.
-Developing problem-based learning took human ingenuity and insight, not getting hooked on an expensive technology of questionable value today, of dubious longevity, and that in the here and now is unarguably extractive, abusive, and exploitative, and that advances an agenda that can only be squared with "integrity, quality, and excellence" via fabulous contortions of the imagination.
-
-Does the university imagine that it will be best known in half a century as "McMaster, the institution that pioneered the use of ChatGPT"?
-Is that what it really wants?
-[Regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean), after all, is also called a "reversion to mediocrity" [@barnett_regression_2017], and that is what an AI campus has to offer.
-
-I suggest that universities should avoid the mistakes of the media, an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) [also see the "pivot-to-video" and how tech companies suckered many media companies into giving them money in exchange for fake engagement statistics [@hazard_owen_facebooks_2021].].
-Tech corporations (or least ways some of them), can still hope to recoup some of their investment in AI.
-But a university gambles its most important asset when betting on AI: its reputation as a place were knowledge is created, and as a custodian of knowledge.
-AIs already have a reputation for producing mis- and disinformation--from the very titans of the field^[Entities like Google and Amazon could not care less about their reputation.
-They are [too big](https://www.businessinsider.com/ex-amazon-ghaderi-exec-suing-ai-race-copyright-allegations-2024) [to care](https://pluralistic.net/2024/04/04/teach-me-how-to-shruggie/).] to the unaware^[Like a B.C.
-lawyer who was [reprimanded](https://www.cbc.ca/news/canada/british-columbia/lawyer-chatgpt-fake-precedent-1.7126393) for using AI hallucinations in a legal filing.
-This was not even the first time that supposedly highly skilled professionals were castigated for [using](https://www.washingtonpost.com/technology/2023/11/16/chatgpt-lawyer-fired-ai/) [chatbots](https://www.reuters.com/legal/new-york-lawyers-sanctioned-using-fake-chatgpt-cases-legal-brief-2023-06-22/).], and beyond, to bad-faith actors^[Like rival geopolitical powers that are relatively impervious to propaganda themselves for reason of being [authoritarian countries](https://www.theatlantic.com/magazine/archive/2024/06/china-russia-republican-party-relations/678271/).].
-In my view, there are no efficiencies, no promises of future excellence, that can compensate the risk of ruining an institutions reputation.
-
-Why?
-The reputational risk...universities are under siege from unsympathetic or actively hostile actors.
-Universities main source of power is their reputation as places that preserve and expand knowledge in a principled way.
-"This university uses ChatGPT" is a way to say that the university is redundant...an organization as expensive as an institution of higher-education is not needed when there are cheaper alternatives for such an education ("act in the next four days to proof yourself against layoffs!"):
-<!--
-![](images/trail-blazing-ai-bundle.png)
--->
 
@@
 AI industry and the military: @haskins_im_2024
-
-## Bonus
-
-```{r googly-eyes, out.width="70%", fig.cap="\\label{fig:googly-ai-eyes}In conclusion"}
-google_plot <- ggplot() +
-  labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
-  theme_void()  +
-  theme(text = element_text(size = 8))
-
-ggdraw() + 
-  draw_plot(google_plot) +
-  draw_image(image = glue::glue(here::here(), "/images/ai-googly-eyes.png", width = 0.8))
-```
+-->
 

### 2024-06-29 ai-haunting-campus/ai-haunting-campus.qmd
Fixed a troublesome footnote and created a word document to count words.

@@
 
-A key reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they have been presented quite deliberately as _everything tools_^[Almost as flexible as, well, a naturally intelligent human, but faster and supposedly cheaper.].
+A key reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they have been presented quite deliberately as _everything tools_^[Almost as flexible as, well, a naturally intelligent human, but faster and supposedly cheaper.
+[Here](https://www.youtube.com/watch?v=yUoj9B8OpR8) is OpenAi's Mira Murati talking about "AI Everywhere: Transforming Our World, Empowering Humanity"--just maybe not all of humanity: "some creative jobs maybe will go away, but maybe they shouldn't have been there in the first place".
+One can be sure she means the jobs that produced the materials that OpenAI stole to train its AIs.].
 While hammers are understood to have a limited range of uses that mostly have to do with concentrating force, and word processors are generally only used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
@@
 Who fine-tunes the machine that provides the feedback?
-Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided^[[Automation bias](https://en.wikipedia.org/wiki/Automation_bias), the human inclination to favor suggestions made by automated decision-making systems, means that even when a human takes responsibility for decisions, AI companies remain unaccountable for their role in influencing outcomes^[GenAI is not a toy, as Air Canada discovered when the company was found liable for "negligent misrepresentation", after a court denied claims by the airline that the chatbot it deployed was somehow responsible for its own actions as an effectively separate legal entity [@garcia_what_2024].].
-A poignant example involving AI-based face recognition and online proctoring is related by @hill_accused_2022.]?
+Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided^[[Automation bias](https://en.wikipedia.org/wiki/Automation_bias), the human inclination to favor suggestions made by automated decision-making systems, means that even when a human takes responsibility for decisions, AI companies remain unaccountable for their role in influencing outcomes.
+GenAI is not a toy, as Air Canada discovered when the company was found liable for "negligent misrepresentation", after a court denied claims by the airline that the chatbot it deployed was somehow responsible for its own actions as an effectively separate legal entity [@garcia_what_2024].]?
+A poignant example involving AI-based face recognition and online proctoring is related by @hill_accused_2022.
 

### 2024-06-29 paper/paper.qmd
Fixed a troublesome footnote and created a word document to count words.

--- 2024-06-29 ai-haunting-campus/ai-haunting-campus.qmd
@@
 ---
-title: "An AI is haunting campus"
-subtitle: "Some considerations about the use of AI for teaching and learning in post-secondary education"
-author: Antonio Páez (McMaster University)
-date: "`r Sys.Date()`"
-format:
-    pdf: 
-      fontsize: 7pt
-    html: 
-      self-contained: true
-      grid: 
-        margin-width: 350px
-execute: 
-  echo: fenced
-  freeze: auto
-reference-location: margin
-#citation-location: margin
+title: "An AI is Haunting Campus"
+format: docx
 bibliography:
@@
 ```
-
-::: {.column-margin}
-Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.
-:::
-
-Disclaimer: _All opinions expressed here are mine and any resemblance to those of my employer(s) is purely coincidental_.
-_No generative AI tool was used to research or write this essay, only good old-fashioned library and internet search skills_.
-
-Note: _This essay was written as an open research project, and it can be found in the following repository:_ https://github.com/paezha/University-AI-Panel
-
-License: [CC-BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/)
-
-![](by-sa.png){width=50}
 
@@
 
-<!--
-> Instead of working with cohorts of learners in grades or classes, individual learning agendas are developed based on an assessment of the skills, competencies and capabilities the learner already has as well as ongoing conversations about the purpose of the learning.
-Where does the student want the learning to take them?
-AI systems, coupled with coaching, mentoring and some instruction and peer support networks then ensure the learner is supported on the carefully mapped learning journey.
-Teaching can still take place: Students challenged with a set of related issues can be brought together for a boot camp or focused period of instruction before returning to their individualized study activities.
-
->An end to “time” as a learning metric.
-Currently, undergraduate courses are described in terms of credit hours.
-Learning and assessment on demand, fast-tracking or slow-tracking using adaptive assessment, and modular learning when the student is ready liberates learners from the time-based metric.
-
-> An end to exams as we know them.
-Exams are an efficient way to judge the output of batch teaching but are not good indicators of learner knowledge, skills and capabilities.
-With AI-enabled assessment generation, automated marking and assessment on demand, exams are no longer needed.
-More authentic assessments that are project-based, team challenges, work-based learning activities, and student self-directed projects can become the norm.
-
->Increased feedback and advising.
-AI systems do not have “office hours.” They are available 24x7 and never tire of finding new ways of responding to student queries and needs.Subject focused chatbots which have been “fed” all necessary information about a subject will do an excellent job of subject-matter advising and teaching.
-Analytics suggest that some students need to be supported by teachers or others with special skills, such as writing coaches, statistical support workers and numeracy advisors.
-The aim will be to ensure every learner follows and is successful in the learning journey they have contracted to complete.
-
-> Greater accessibility.
-Higher education systems are organized around blocks of time: semesters or terms.
-There is no longer good reason for this.
-Students should be able to begin their learning at any time and be assessed when they are ready.
-A four-year undergraduate degree could easily be completed in 16 months by a diligent student, for example, or in even less time by someone arriving in the program with substantial life and work experience.
-AI systems can also leverage technologies to ensure the resources and supports for learners reflect who they are (culture, language, experience and skill) and what challenges they face (disabilities, learning styles, etc.).
-
-## As an aside {#as-an-aside}
-
-On antropomorphizing GenAI tools.
-This happens mostly in ways that are barely liminal.
-For example, in the document "Overview of AI at McMaster" we read that GenAI's somehow have an "ability to engage in nuanced conversation".
-This phrasing could be simply a literary device, but even so it is indicative of a tendency to ascribe human attributes to a statistical model.
-Doing this risks amplifying existing biases.
-But in other cases, the effort is more blatant.
-Wrapping a statistical model in a human-like skin may actually be more devious, as in the new, sexy ChatGPT interface:
-
-<video width="320" height="240" controls>
-<source src="video/open-ai-commercial.mp4" type="video/mp4">
-</video>
-
-**NOTES:** Need to research more the argument that the interface is so important in the case of these tools because what they do well is trivial and what they do poorly is what is being hyped.
-Found the article [@farrugia_known_2024]; see here: https://fasterandworse.com/known-purpose-and-trusted-potential/
-
-Are GenAI reality-warping machines?
-Think about the small and large ways in which these tools can be used to rewrite history, for relatively banal commercial objectives (monetizing clicks) or worse, for malicious propaganda campaigns.
-
-**NOTE** Russia copied millions of articles from wikipedia wholesale, to replace the open encyclopedia with a home-grown version without all the bits that the current regime finds objectionable [@corfield_russia_2023; @jankowicz_russia_2023].
-
->The first step in liquidating a people,' said Hubl, 'is to erase its memory.
-Destroy its books, its culture, its history.
-Then have somebody write new books, manufacture a new culture, invent a new history.
-Before long the nation will begin to forget what it is and what it was.
-The world around it will forget even faster.
->-― Milan Kundera, The Book of Laughter and Forgetting
-
-Potential for weaponization of GenAI.
-**Farfetched (?) idea** Could it be that GenAI will be at some point declared a military-grade technology with restricted applications?
-(see https://www.defenseone.com/technology/2024/05/ai-and-other-tech-creating-special-operations-renaissance/396388/)
-
-Peter Thiel on "AI is a military technology" https://www.businessinsider.com/peter-thiel-claim-ai-primarily-military-technology-experts-disagree-2019-8
-
-AI industry and the military: @haskins_im_2024
--->
-
-## Appendices
-
-### AI Committee announcement {#ai-committee-announced}
-
-This is the announcement of the launch of the AI Committee (see email from the Office of the Provost, dated March 20, 2024):
-
-<!--https://app.fastmail.com/mail/search:expert+panels/T2d9b9367655ae218.M6205d28844daa67dfd3432ff-->
-
-> "McMaster launches AI committee
->
-> Whether you are a researcher, faculty member, student, administrator, or have another role on campus, you are likely thinking about Artificial Intelligence (AI) and its challenges and potential.
->
-> AI is a powerful tool and I believe that universities have an important role to play in its adoption.
-At McMaster, we are well placed to prioritize an ethical approach to shaping policies and developing new ways of using AI to ensure a responsible and beneficial integration where it makes sense.
->
-> I’m pleased to announce that McMaster has launched an AI Advisory Committee to explore the use of AI across campus.
-Three subcommittees have also been started to examine the use of AI in teaching and learning, research and operational excellence in our work.
-It is a strategy that makes McMaster one of the few institutions taking a pan-university approach to AI.
->
-> Our earlier work to develop provisional guidelines for the use of generative AI in teaching and learning drew interest from other post-secondary institutions, which adopted McMaster's guidelines after we made them available through a Creative Commons licence.
->
-> The committee will be led by three co-chairs, each bringing different expertise.
-They are Matheus Grasselli, deputy provost, Gayleen Gray, associate vice-president and chief technology officer and Gianni Parise, acting deputy vice-president, Research.
-They will be supported by Erin Aspenlieder, special advisor to the provost on generative AI.
->
-> The three subcommittees will each by led by a subject matter expert.
-Kim Dej, vice-provost, Teaching and Learning, will lead the exploration of AI in teaching and learning, Maggie Pooran, executive director, Health, Safety, Well-Being & Labour Relations, will explore its use in operational excellence, and Martin Horn, associate dean, Faculty of Humanities, will look at implications for research.
->
-> Each expert subcommittee will have a pool of contributors made up of members of the McMaster community.
-Depending on the initiative, some or all members will be invited to complete the specific work based on their expertise, interest and availability.
->
-> If you would like to be considered for the expert panels, please complete this expression of interest form."
-
-### AI Teaching and Learning Panel priority initiatives {#ai-panel-tasks}
-
-The [Terms of Reference](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/ai-advisory-committee/) of the Artificial Intelligence Advisory Committee identify the following priority initiatives for the Expert Panel on Teaching and Learning:
-
-- Provide recommendations to the AI Advisory Committee on updates to the Provisional Guidelines on the Use of Generative AI in Teaching and Learning.  
-- Review and recommend training, resources and professional development on AI literacy for both educators and students (e.g. grants, microcredentials, guidebooks).  
-- Engage in dialogue with both students and educators to understand their perspectives on AI in learning.  
-- Endorse ethical standards and guidelines for AI usage in teaching environments.  
-- Review and provide insights into assessments or tools that leverage AI for student evaluation.  
-- Engage in cross-institutional dialogues to share insights and learn from global best practices.  
-- Advocate for student involvement in the design and critique of AI-enabled teaching tools.
-
 ## References

### 2024-07-02 paper/highlights.qmd
Paper submitted to Computers & Education

---
title: "An AI is Haunting Campus: A Call for Human-Centered Higher Education"
subtitle: "Highlights"
format: docx
---

- AI has been presented as an _everything tool_ in higher education
- AIs are regression models: they use inputs to regress to a probable output, i.e., a mean 
- AIs cannot achieve the variety of tasks that are promised without compromising the values of universities
- Students are poorly served by all the attention paid to AIs
- AIs are poor complements to a human-centered education

### 2024-07-02 paper/paper.qmd
Paper submitted to Computers & Education

@@
 ---
-title: "An AI is Haunting Campus"
+title: "An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: A Call for Human-Centered Higher Education"
 format: docx
@@
 
-## Background
-
-> _Mission Statement: Our organization seeks to level up on all matters related._  
-> -- An Organization Run by its Marketing Department
+## Introduction
 
 On November 30, 2022, OpenAI released what became its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language.
-The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels: by early December, ChatGPT was being panned as "the best artificial intelligence chatbot ever released to the general public" [@roose_brilliance_2023], and a tool able to generate "impressively detailed" and "human-like" text, leading some to wonder whether it could "replace humans" [@lock_what_2022].
+The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels [@roose_brilliance_2023; @lock_what_2022].
+Soon after, seemingly everyone wanted in on the Generative AI (GenAI) craze^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent cycle of hype [see @siegel_ai_2023].], and no surprise, given how the technology was hailed as nothing short of revolutionary.
 
-Universities, much like every other organization dealing with information, had to respond in a compressed span of time to the rapid release of generative AI (GenAI) tools like ChatGPT, most of them created and unleashed in the wild by a handful of large American corporations including [Meta](https://en.wikipedia.org/wiki/Meta_Platforms) (or were we to deadname it, facebook), [OpenAI](https://en.wikipedia.org/wiki/OpenAI), [Google](https://en.wikipedia.org/wiki/Alphabet_Inc.), and [Microsoft](https://en.wikipedia.org/wiki/Microsoft).
+According to some, AI technologies in education "offer new tools and applications that have the potential to transform traditional teaching and learning methods" [@adiguzel_revolutionizing_2023, p.
+1], and have "the potential to fundamentally change the classroom experience and the knowledge and skills outcomes of our students" [@dwivedi_opinion_2023, p.20].
+Furthermore, the technology "is here to stay...[t]herefore, teachers and students must develop the specific digital competencies needed to use such tools in ways that are pedagogically beneficial and ethical" [@kohnke_chatgpt_2023, p.
+546].
 
-GenAI was hailed as nothing short of revolutionary.
-McMaster, for example, [stated](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) that:
-
-> [u]nderstanding how generative AI works, and how it can support work within organizations represents an ongoing area of exploration and innovation.
-The opportunities of generative AI are exciting: creating workplace efficiencies to allow for different kinds of work, adding capabilities for individuals and teams, and offering personalized uses.
-
-AI was quickly painted as an _innovative_ and _exciting_ technology, that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p.
+AI was quickly painted as an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, and that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p.
 64].
 Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.].
-It seemed inevitable that universities _had to do something_ with it, at the risk of being let behind.
+Institutions of higher education were compelled to get in on the act, forming high-level advisory committees and drafting guidelines to guide the university community leverage AI in principled ways [e.g., @office_of_teaching_and_learning_provisional_2023; @mcmaster_university_provisional_2024; @trent_university_generative_2024; @university_of_toronto_ai_2024].
 
-The fear of missing out, combined with a modicum of concern for the ethical implications of GenAIs^[Ethical concerns for the most part have had to do with policing academic dishonesty among students.], has led to institutional responses that have been less than coherent.
-A case in point is Emory University, where the university gave a $10,000 prize to a group of students who proposed developing an AI tool for an Entrepreneurship Summit, only for the students to be subsequently suspended by the school's Honor Council for the potential of their tool for cheating^[There was no evidence that cheating had happened.] [@koebler__university_2024].
+For perspective, hammers are powerful tools, and so are thousands of other instruments found at universities.
+And yet not every single one commands a high-level committee to advice on its use.
+Just what is it about AIs that requires pan-university consultations on how to use them?
+To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, examine what it actually does.
 
-For its part, McMaster was a leader in responding to the rapid, and disruptive, release of these _revolutionary_ tools, and its provisional [guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) (released under a Creative Commons Licence) have since inspired those of other post-secondary institutions in Canada, including the University of Guelph [-@office_of_teaching_and_learning_provisional_2023] and Trent University [-@trent_university_generative_2024].
-
-The aim of the provisional guidelines was to inform the university community about the use of AI in "operational excellence".
-As we approach the Fall Term of 2024, with almost two years of cohabitation experience with AI for the masses at our backs, the time has come to update these guidelines.
-Pertinently, the institution, via its Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus.
-This committee is composed of three sub-committees tasked with examining the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work".
-
-The sub-committees, in turn, convened panels to [provide recommendations](#ai-panel-tasks) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/).
-The [priority initiatives](#ai-panel-tasks) for the members of the AI Teaching and Learning Panel are to help shape the guidelines for teaching and learning at McMaster, a consequential task "for how educators take up generative AI in their course and assessment design and how students approach the use of this technology".
-
-According to the announcement of this AI Committee at McMaster:
-
-> AI is a powerful tool and I believe that universities have an important role to play in its adoption.  
-> -- Susan Tighe, Provost
-
-Notice that this statement (like the definition of the work before the panels above), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption.
-This attitude is almost fatalistic^[In the sense of "fated to".] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent of several cycles of hype [see @siegel_ai_2023]--fundamentally a vastly overblown tech bubble [@vinsel_dont_2023].] [@angwin_press_2024].
-
-To put things in perspective, hammers are powerful tools.
-Word processors are powerful tools.
-Lasers are powerful tools.
-And yet we do not have high-level university committees for each and every tool that is used at the institution.
-What is it about AIs that, unlike many other tools, requires a pan-university consultation on an "_ethical approach to shaping policies and developing new ways of using AI to ensure a responsible and beneficial integration where it makes sense_"?
-
-(The last part of the statement above carries quite a bit of weight here: where does adoption of AI make sense? Presumably this is what the Committee on AI will try to answer.)
-
-To understand why many organizations, including McMaster University, have decided to spend considerable resources studying the adoption and responsible use of AIs, it is essential to understand what the technology claims to do that sets it apart from other tools, and what it in actuality does.
-
-In doing this, it is essential that we remain alert to the fact that technologies are _never_ politically neutral, and therefore it is important to ask two questions when considering their adoption:
+In doing this, we must remain alert to the fact that technologies are never politically neutral, and two questions must be asked when considering their adoption:
 
@@
 
-We also need to understand the goals of the organization that considers using the technology.
-When does it make sense for a university like McMaster to "integrate" AIs as part of its core mission, accounting for what the technology does for some people and what it does to others?
+As well, we need to understand the character of the organization that considers the use of the technology.
 
-Let us begin with the mission of McMaster as an institution of higher education.
+This essay aims to provide some clarification around 1) the things that GenAIs claim to do in education; and 2) what the technology actually does; within 3) the context of a principled organization.
+I conclude that the promise of generative AI has been overstated, that is is unlikely to deliver, and that in any case the technology, as sold by the tech industry, is antithetical to many of the values that universities claim to hold dear.
 
@@
 
-McMaster's [mission](https://president.mcmaster.ca/mission-vision/) is "the discovery, communication and preservation of knowledge," accomplished with a commitment to "creativity, innovation and excellence".
-These aspirations must respond to some underlying values, and indeed, the university explicitly values "integrity, quality, inclusiveness and teamwork".
-Ultimately, the university aims to "inspire critical thinking, personal growth, and a passion for lifelong learning" and to serve "the social, cultural, and economic needs of our community and our society."
+Universities present themselves as institutions that serve "the social, cultural, and economic needs of [their] community and [their] society" [@office_of_the_president_mcmaster_university_mission_2024]; that strive "to make valued and socially responsible contributions to our local communities, to Canada, and to the world" [@trent_university_vision_2024]; and that "serve society and...enhance the quality of life through scholarship" [@university_of_guelph_mission_2024].
 
-A mission statement is like an onion, and one needs to peel multiple layers to try to understand the core principles enunciated there.
-Why are things like "integrity" and "critical thought" valued?
-Or, what happens when "integrity", "quality", and/or "excellence" clash with "the economic needs of our society" (say, a mandate by democratically elected governments to spend as little as possible in post-secondary education^[Between 2000-2020, Ontario had the lowest spending on post-secondary education of all provinces in Canada, relative to its share of spending [@caut_provincial].
-Public funding of universities in Ontario as percent of revenue fell from 45.6% in 2001/2 to 35.5% in 2019/20 [@statistics_canada_college_2022].
-As noted by @usher_state_2023, Ontario is "a province where underfunding [higher-education] institutions is a pan-partisan affair".
-Is it any wonder that universities in the province are pushed to find "efficiencies"?])?
+This overarching goal is achieved through "the discovery, communication and preservation of knowledge" [@office_of_the_president_mcmaster_university_mission_2024]; a dedication "to fostering an academic community in which the learning and scholarship of every member may flourish" [@the_office_of_the_governing_council_secretariat_institutional_1992]; by "[encouraging] and [celebrating] excellence and innovation in teaching, learning, research and student development" [@trent_university_vision_2024]; with a determination "to put the learner at the centre of all [the university] does" [@university_of_guelph_mission_2024].
 
-Clearly, there is some tension implicit in the mission of a university.
-But being generous, we can take the mission statement at face value and assume that the highest priorities are those related to knowledge obtained, preserved, and communicated with creativity, integrity, excellence, and quality.
+Unprincipled organizations do not need to question what the adoption of a technology does to and for whom--as long as it advances whatever its goal is at any particular moment.
+But universities aim to achieve their mission while adhering to "integrity, quality, inclusiveness and teamwork" [@office_of_the_president_mcmaster_university_mission_2024]; with "vigilant protection for individual human rights, and a resolute commitment to the principles of equal opportunity, equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992]; and  as they "foster sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].
 
-And so, what do AIs do that can advance the mission of the university?
+Universities must strike a careful balance between what they aim to achieve and how they achieve it.
+And so, what do AIs do that can advance the mission of universities without compromising their core values?
 
-## What do AIs do?{#what-do-ais-do}
+## What do GenAIs do?{#what-do-ais-do}
 
@@
 
-A key reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they have been presented quite deliberately as _everything tools_^[Almost as flexible as, well, a naturally intelligent human, but faster and supposedly cheaper.
-[Here](https://www.youtube.com/watch?v=yUoj9B8OpR8) is OpenAi's Mira Murati talking about "AI Everywhere: Transforming Our World, Empowering Humanity"--just maybe not all of humanity: "some creative jobs maybe will go away, but maybe they shouldn't have been there in the first place".
-One can be sure she means the jobs that produced the materials that OpenAI stole to train its AIs.].
-While hammers are understood to have a limited range of uses that mostly have to do with concentrating force, and word processors are generally only used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
+A reason why many organizations have spent substantial resources considering the adoption of AIs is that they have been presented, quite deliberately, as _everything tools_: tools capable of doing many things a human can do, but faster and supposedly cheaper.
+While hammers are understood to have a limited range of uses that mostly have to do with blunt force, the release of GenAIs to the masses was accompanied by innumerable promises--quite a few of them made by entities and individuals that had a stake in the adoption of the technology^[Ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
 
@@
 
-Also, that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" by "automating customer self-service, accelerating response times on an increasing number of tasks as well as providing greater accuracy, enhanced routing and intelligent context gathering" [@ibm_what_2023].
+Also, that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" [@ibm_what_2023].
+AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024].
+In mental health care, chatbots could complement human therapists by using patient data to analyze behavioral patterns to then mimic practitioner questions [@silva_4_2023].
+AIs would revolutionize dating by helping "people connect and find potential partners" [@khalatian_matchmaking_2023].
 
-According to Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.], AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024].
-But why stop at self-driving vehicles?
-McMaster University's Canada Excellence Research Chair Laureate Ali Emadi "hopes to help revolutionize and influence everything from personal travel to city planning" by putting AI directly into our transportation systems [@dorey_driving_2023].
-
-In mental health care, chatbots could complement human therapists "via comprehensive data access and [analysis of] behavioral patterns" to "mimic practitioner questions and subsequently make recommendations based on a user’s inputs" [@silva_4_2023].
-
-AIs will even write code, perhaps helping themselves to some self-programming along the way: AIs that write themselves, or AI self-help?
-
-```{r ai-writing-code, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}An example of Atlassian AI writing code"}
-knitr::include_graphics(glue::glue(here::here(), "/images/ai-bug-fix.png"))
-```
-
-The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves, all the way down to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023].
-Not even the [bowel movements of our pets](https://finance.yahoo.com/news/next-generation-ai-powered-cat-140000795.html) are beyond the revolutionary all-seeing eye of AI.
+The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves to micromanaging the [bowel movements of our pets](https://finance.yahoo.com/news/next-generation-ai-powered-cat-140000795.html).
 
 ```{r ai-litter-box, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}Revolutionizing pet health"}
+#| label: fig-revolutionizing-pet-care
+
 knitr::include_graphics(glue::glue(here::here(), "/images/ai-litter-box.png"))
@@
 
-The forecast that AI will improve everything (from global climate to the dumps of our cats) was certain to, sooner or later, reach higher-education, a target-rich environment for selling hype--and indeed, colleges and universities have not been spared the onslaught of promises.
+Colleges and universities have not been spared the onslaught of promises.
 
-Examples of this abound.
+@power_3_2024, presents three innovative uses for AI in higher education.
+First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably from working at the university].
+AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants.
+AI could also personalize the experience of consumers by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.
 
-@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.], writing for Forbes, presents three innovative use cases for AI in higher education.
-First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies".].
-AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants.
-AI could also personalize the experience of consumers (occasionally also called students) by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.
-
-Mr. Power's message to the higher-education sector? It is time for institutions to "level up": jump on the bandwagon, fellas, don't let it leave you behind!
-
-In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school), a top business school in France.
-Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence", whatever that means.] argue that AIs will likely be the primary way to access information in the future.
-For this reason, institutions must equip students with the skills needed to use AI competently, in particular train them to perfect the art of _prompt engineering_ (the ability to craft effective queries that prompt a GenAI to provide the most useful outputs).
+In another instance, @hie_how_2023 argue that AIs will likely be the primary way to access information in the future.
+For this reason, institutions must equip students with the skills like _prompt engineering_, the ability to craft effective queries that prompt a GenAI to provide the most useful outputs.
 Universities will use AIs to improve learning and challenge the critical thinking of students.
-Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience.], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been an activity essential to understand my own courses, what I teach and how I teach it; automating the task of writing syllabi is another way of degrading the quality of instruction that students can expect to receive.].
+Faculty will also benefit from AI: it will be used to evaluate their students, and will handle repetitive tasks like writing course syllabi.
 
@@
 Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9).
-It did not take long for this prediction to turn sourer than beans in August: on February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it.
-Shortly after that, Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.
+This prediction aged poorly: on February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it.
+Shortly after that, Meta [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.
 
-\newpage
-
-Similarly hyperbolic claims about AI in higher education can be found peppered throughout the landscape, including by @johnson_ai_2023^[[Gridserve](https://www.gridserve.co.uk/about-us/) is a UK company "dedicated to enhancing the technological capabilities of the education sector through customised information and communication technology (ICT) services".] and @contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students".
-Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].].
-A fairly comprehensive list [from @johnson_ai_2023] is as follows (notice the use of "can" in the original, not "could", or "maybe will"):
+Similarly hyperbolic claims about AI in higher education can be found peppered throughout the landscape [e.g., @adiguzel_revolutionizing_2023; @contact_north_five_2024; @dwivedi_opinion_2023; @johnson_ai_2023; ].
+A fairly comprehensive list [from @johnson_ai_2023] is as follows (notice the use of "can" in the original):
 
@@
 
-A benefit of portraying AI as an _everything tool_ (for the proponents of the technology, at least), is that conversations around it quickly turn into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength.
-To be sure, AIs can do _some_ things well, but critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed ["flooding the zone with shit"](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation).
-Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot hope to counteract [@maiberg__google_2024].].
-Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they really do do?
-
+Discussions of AI as an _everything tool_ quickly devolve into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength.
+To be sure, AIs can do _some_ things well, but critics are forced to respond to every single propounded use with evidence, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself.
+Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they really do?
 A second discursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.
 
-In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 90 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone.
-As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems.
-For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training.
-Models are trained on biased data and can produce either subtly or overtly biased results.
-And because these biases seem to come from an objective machine, they can be especially pernicious.
-Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems".
-This disclaimer, consuming fully one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].
+In what follows, it is not my intention to confront every single possible use of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop.
+As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the compromises demanded by AI.
 
-## But really, what do AIs do? {#really-what-do-they-do}
+## But what do GenAIs really do? {#really-what-do-they-do}
 
-GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning many, many, _many_ hidden layers.
-As an example, Large Language Models (such as those behind OpenAI's ChatGPT, Microsoft's Copilot, Google's Gemini, and facebook's LLaMA) are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].
+GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning _many_ hidden layers.
+As an example, Large Language Models are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].
 
-But, what are [neural](https://en.wikipedia.org/wiki/Large_language_model#Further_reading) [networks](https://medium.com/@asifurrahmanaust/lesson-3-neural-network-is-nothing-but-a-linear-regression-e05a328a0f23)?
-
-Behind all the jargon, the neural networks that underpin GenAI are models designed to produce "statistically probable outputs" when prompted [@ibm_what_2021].
-In more conventional terms, neural networks are regression models--much more sophisticated and powerful than linear regression to be sure, but regression models nonetheless [@ripley_neural_1994].
+Behind all the jargon, though, the neural networks that underpin GenAI are models designed to produce "statistically probable outputs" when prompted [@ibm_what_2021].
+In more conventional terms, neural networks are regression models--much more sophisticated and powerful than linear regression to be sure--but regression models nonetheless [@ripley_neural_1994].
 Here, it is important to note that the term regression refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing an input (e.g., a prompt) to a "statistically probable" outcome.
@@
 
-Below is an example of the simplest regression model with only one independent variable (i.e., a bivariate model).
+@fig-bivariate-model is an example of the simplest regression model with only one independent variable (i.e., a bivariate model).
 The plot shows the pairs of $y$ and $x$ values, and the line is the model, i.e., the conditional mean $\hat{y}$.
@@
 Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.
+
 ```{r simulate-simple-example}
@@
 
-```{r plot-simulated-data, message=FALSE, out.width="70%", fig.cap="Simple Example of a Regression Model"}
+```{r plot-simulated-data, message=FALSE, out.width="70%", fig.cap="Bivariate Regression Model"}
+#| label: fig-bivariate-model
+
 ggplot(df,
@@
 ```{r data-vs-predictions-comparison, out.width="70%", fig.cap="Training data and model output compared"}
+#| label: fig-data-vs-preds-01
 
@@
 
-Neural networks do something similar but using a net-like architecture that is more flexible at identifying the characteristics of the data.
-For instance, the following figure shows a relatively simple neural network trained using using the same data.
+Neural networks do something similar but using a more flexible net-like architecture.
+For instance, @fig-nn1 shows a relatively simple neural network trained using using the same data.
 This network is not "deep", as it has only one "hidden layer" comprised of two neurons between the input and the output layers.
@@
 ```{r plot-shallow-neural-network, out.width="70%", fig.cap="Simple Example of a Shallow Neural Network"}
+#| label: fig-nn1
+
 knitr::include_graphics(glue::glue(here::here(), "/images/nn1.png"))
@@
 
-The plot below illustrates the predictions of this neural network, where we can see that the model is not limited to a linear response.
-Neural networks are more flexible than linear regression, and they achieve this by being less parsimonious (they use more parameters).
+@fig-data-vs-preds-02 illustrates the predictions of the neural network in @fig-nn1, where we can see that the model is not limited to a linear response.
+This flexibility by the neural network is achieved by being less parsimonious (it uses more parameters).
 
 ```{r data-vs-predictions-comparison-2, out.width="70%", fig.cap="Training data and model outputs compared: linear regression and shallow neural network"}
+#| label: fig-data-vs-preds-02
+
 ggplot() + 
@@
 
-More complex neural network can be designed.
-The following plot is of a network deeper than the previous one, with two hidden layers, each consisting of five neurons.
-This model uses forty-six parameters.
+More complex neural networks can be designed. @fig-nn2 shows a network deeper than the previous one, with two hidden layers, each consisting of five neurons. This model uses forty-six parameters. 
 ```{r train-deeper-neural-network}
@@
 ```{r plot-deeper-neural-network, out.width="70%", fig.cap="Example of a Deeper Neural Network"}
+#| label: fig-nn2
+
 knitr::include_graphics(glue::glue(here::here(), "/images/nn2.png"))
@@
 
-The increased complexity leads to greater flexibility, and we can see that the model's predictions resemble more closely the actual data.
+The increased complexity leads to greater flexibility, and we can see that the model's predictions resemble more closely the actual data (@fig-data-vs-preds-03).
 
 ```{r data-vs-predictions-comparison-3, out.width="70%", fig.cap="Training data and model outputs compared: shallow neural network and deeper neural network"}
+#| label: fig-data-vs-preds-03
+
 ggplot() + 
@@
 These "coordinates" in the space of the vocabulary can then be used to calculate the "distances" between words.
-Alas, these vectors are decontextualized, that is, they only reflect the position of a given word with respect to other words in the training corpora, and the numeric values are static, even if the surrounding words in a specific prompt change.
 A recent breakthrough in the field of natural language processing is the development of _transformers_, a network architecture architecture that updates the numeric values of a word depending on the words that come before and after in a prompt.
@@
 - AIs are _very_ flexible non-linear regression models.
-- Their flexibility comes from using "deep" architectures, which means many layers with many, many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters.
+- Their flexibility comes from using "deep" architectures, which means many layers with many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters.
 OpenAI no longer reports the number of parameters of its newer models, but in the latest version this is estimated to be in the trillions.].
@@
 - Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think).
-The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data were used to train AIs.
-The increasingly large datasets essentially expand the space where the model can interpolate.].
+The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became more subtle as bigger datasets were used to train AIs.
+The increasingly large datasets essentially expanded the space where the model can interpolate.].
 
-## AI in teaching and learning {#ai-in-teaching-and-learning}
+## GenAIs in teaching and learning {#ai-in-teaching-and-learning}
 
-Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it might make sense to use this technology.
-To begin, we must note the number of examples of the use of AI by students.
-A Turnitin-sponsored  report [@tyton_partners_generative_2023] of 1,600 post secondary students and 1,000 faculty, found that 50% of students used GenAI, compared to only 22% faculty.
-Use of AI by students is commonly cited as a reason for universities to embrace AI [e.g., @power_3_2024].
-McMaster's own internal research (based on a sample of over 1,300 students) indicates that at least 61% of students used AI during the Fall Term in 2023, 75% of them for school work [@yachouh_student_2024].
-
-But embracing AI because the students use it is putting the cart before the horse.
-After all, smoking is banned on campus, and I do not recall the university asking the students if they smoked^["A tobacco and smoke-free campus is the next important step towards fulfilling our responsibilities as educators, healthcare professionals and to the communities we serve" said then-president Patrick Deane [@carter_mcmaster_2017].].
-A university is supposed to be a place of education (higher, at that), and not a fast-food locale where customers get what they want.
-
-So, what does McMaster hope to achieve with AI? From the Overview on Generative AI document:
-
-> Some of [the uses of AI] in our context could include qualitative and quantitative data analysis, writing text (e.g.
-reference letters, grant applications, job descriptions, report summaries), translation, captioning, multiple formats for learning and engagement, and personalized support.
+Some of the things that universities hope to achieve with AI include writing text (e.g.
+reference letters, grant applications, job descriptions, report summaries), conducting qualitative and quantitative data analysis, translation, captioning, providing multiple formats for learning and engagement, and for personalized support.
+Anecdotally, students report wanting to use generative AI not to complete assignments for them, but to help enhance learning in their courses by having concepts explained differently, by working on practice questions or by getting immediate feedback on drafted work.
+If we take a student-centered perspective, we might begin by asking about some of the things that AIs purportedly do for students.
 
@@
 
-> _ChatGPT doesn't give you information. It gives you information-shaped sentences._
-> --Neil Gaiman
-
-According to McMaster's internal research:
-
-> Anecdotally students at McMaster report wanting to use generative AI not to complete assignments for them, but to help enhance learning in their courses by having concepts explained differently, by working on practice questions or by getting immediate feedback on drafted work.
-
-<!-- From Review Report in private folder: May 6 Review Report_AI Expert Panel on Teaching and Learning.docx -->
-
 - Students want concepts explained to them differently.
 
-There are many traditional ways for students to get different perspectives on concepts.
+There are many ways for students to get different perspectives on concepts.
 They can form study groups, attend office hours to engage with their instructors, ask their Teaching Assistants, ask relatives and friends, and so on.
-But even for a student who for whatever reason cannot or will not do any of these things, the fact remains that relatively common concepts have already been explained by humans in dozens if not hundreds of different ways, often in writing that is available through an excellent university library.
-A search of McMaster's library catalog, looking only for resources available online (for convenience), reveals that there are hundreds, thousands, and sometimes tens of thousands of resources available to explore the same concept from multiple perspectives.
-
-```{r mac-library-econometrics, out.width="70%", fig.cap="Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
-knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-econometrics.png"))
-```
-
-```{r mac-library-cell-biology, out.width="70%", fig.cap="Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
-knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-cell-biology.png"))
-```
-
-```{r mac-library-euphemism, out.width="70%", fig.cap="Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
-knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-euphemism.png"))
-```
-
+But even for a student who for whatever reason cannot or will not do any of these things, the fact remains that relatively common concepts have already been explained by humans in dozens if not hundreds of different ways, often in writing that is available through their university library.
 Is an AI tool essential to find alternative explanations for common concepts?
@@
 for reasons of lying far from the mean).
-This strongly suggests that any outputs about highly specialized queries should be treated with caution^[The Overview on Generative AI notes that "if you ask [a chatbot] about yourself or a more obscure topic, [it] will almost certainly make up nonsense." This is true even when asking about a person with an extensive online presence (the internet being the likely source of the training datasets).
-As recently as 2023, a famous law professor discovered that he was being named as a sexual harasser by ChatGPT, when in reality he had never been accused of such conduct [@verma_chatgpt_2023].].
-
-In summary: there is no dearth of resources to learn about relatively common concepts, and AIs should not be used to try to learn more specialized concepts.
+This strongly suggests that any outputs about specialized queries should be treated with caution.
 
@@
 An emphasis on practice questions, however, signals an underlying weakness of the process, namely, a focus on learning for the examination, as opposed to learning for its own sake.
+Posing interesting questions is an essential part of an instructor's job, but this part of the job becomes compromised when instructors are pushed to operate in massive classrooms.
+In this case, AI is a substitute, and demonstrably not a good one, for a qualified instructor.
+At issue, at well, is traditional evaluation approaches.
 A growing body of evidence points towards the effectiveness of alternative assessment schemes, such as continuous assessment through course work, and even seen examinations instead of the traditional unseen examinations [@turner_see_2018].
@@
 Ungrading, in particular, is increasingly recognized for its potential to help learners internalize the motivation to learn, and for fostering adaptability, creative thinking, and self-management [@gorichanaz_it_2024].
-
-Two issues appear to be relevant.
-
-First, posing interesting questions for students to work on is an essential part of an instructor's job, but this part of the job becomes compromised when instructors are pushed to operate in massive classrooms with sometimes hundreds of students^[Average class sizes at McMaster (with a full time under-graduate student population of 30,400) are 127.8 in first/second year, and 52.3 in third/fourth year; this places McMaster dead last among all Canadian institutions in terms of class sizes in first/second year, and second to last after UBC in third/fourth year [@macleans_canadas_2022].].
-In this case, AI is a substitute, and demonstrably not a good one, for a qualified instructor who can intellectually challenge their students.
-
-The second issue is traditional evaluation.
-As noted, studying for the examination often detracts from genuine learning.
-Why are other assessment approaches seldom considered?
+If studying for the examination often detracts from genuine learning, why are other assessment approaches seldom considered?
 Research by @vahed_continuous_2023 suggests that instructors need to develop assessment literacy, that is, an understanding of "the fundamental assessment concepts and procedures, and to engage in the ongoing process of adjusting instructional strategies to enrich students learning experiences".
@@
 
-This is an utterly unrealistic expectation (also see automated grading below).
-Ideally, students should have the opportunity to discuss their drafted work with instructors before final submission (or again, could ask members of their study groups to proof-read their work, among many other ways to get feedback before submission).
-A conscientious instructor will invest time reading and understanding the draft work, so as to be able to provide quality feedback.
-But there is a more insidious aspect about getting feedback from an AI...who, actually, provides the feedback in this case?
+This is an utterly unrealistic expectation.
+Ideally, students should have the opportunity to discuss their drafted work with instructors before final submission (or could ask members of their study groups to proof-read their work, among many other ways to get feedback before submission).
+A conscientious instructor will invest time reading the drafted work to provide quality feedback.
+There is, however, a more insidious aspect about getting feedback from an AI...who, actually, provides the feedback in this case?
 Whose data are used to train the machine that provides the feedback?
 Who fine-tunes the machine that provides the feedback?
-Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided^[[Automation bias](https://en.wikipedia.org/wiki/Automation_bias), the human inclination to favor suggestions made by automated decision-making systems, means that even when a human takes responsibility for decisions, AI companies remain unaccountable for their role in influencing outcomes.
-GenAI is not a toy, as Air Canada discovered when the company was found liable for "negligent misrepresentation", after a court denied claims by the airline that the chatbot it deployed was somehow responsible for its own actions as an effectively separate legal entity [@garcia_what_2024].]?
-A poignant example involving AI-based face recognition and online proctoring is related by @hill_accused_2022.
+Whose perspectives and views are used to provide the feedback, and can they ever be accountable for the feedback provided?
+A poignant example of this dilemma involves AI-based face recognition and online proctoring, as related by @hill_accused_2022.
 
-- Personalized learning
+- Creation of innovative learning materials
 
-Is learning really personalized if there is no person on the teaching side of things?
-An AI cannot possibly understand the needs of a person, let alone help to satisfy them.
-The idea of using AI to somehow personalize learning is less a plan to provide excellent learning opportunities, and more a gimmick to put a little lipstick on the underfunded massification of higher education: a competent instructor, freed from the task of talking at hundreds of anonymous faces in an auditorium, can provide excellent personalized learning opportunities to students in ways that a plausible-sounding machine that regresses to the mean cannot do.
-
-- Creation of interactive and adaptive learning materials
-
-@mollick_instructors_2024^[Drs.
-Eric Mollick and Lilach Mollick are members of Wharton Interactive in the business school of the University of Pennsylvania.
-Wharton Interactive offers to "democratize education" to help you level up your learning!
-(Discounts of up to 50\% are available to alumni) [@kelner_level_2022].] envision a future in which educators are innovators thanks to their adroit use of prompts for AIs.
+@mollick_instructors_2024 envision a future in which educators are innovators thanks to their adroit use of prompts for AIs.
 According to these authors, GenAI stands alone among educational technologies in that it can be "'programmed' through prompts alone", which should allow even instructors "without extensive technology or coding experience" to more easily create classroom applications.
@@
 Leaving the technology fetish about classroom applications aside, Mollick and Mollick are well aware of the ethical concerns around using AI (see p.
-6), including that "[m]odels are trained in ways that may violate copyright", exploiting the "efforts of low-wage workers in precarious conditions for part of their training", and the fact that "[m]odels are trained on biased data and can produce either subtly or overtly biased results".
-And yet, they choose to ignore every single one of these concerns for the sake of being innovators^[The title of the paper is "Instructors as Innovators: A future-focused approach to new AI learning opportunities, with prompts".].
-Machine-washing plagiarism while exploiting low-wage workers is innovative only in a way that offends human decency.
+6), including that "[m]odels are trained in ways that may violate copyright", that exploit the "efforts of low-wage workers in precarious conditions for part of their training", and the fact that "[m]odels are trained on biased data and can produce either subtly or overtly biased results".
+And yet, they choose to ignore every single one of these concerns for the sake of innovation.
+Machine-washed plagiarism that exploits low-wage workers while using inordinate amounts of energy is innovative only in a way that offends "integrity" [@office_of_the_president_mcmaster_university_mission_2024], "individual human rights...equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992], and "sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].
 
-But even if the datasets and training processes used to develop GenAIs were unimpeachable, would the practice of generating learning materials using AIs be "innovative"?
+But what if the datasets and training processes used to develop GenAIs were unimpeachable, would the practice of generating learning materials using AIs then be "innovative"?
 Alas, no.
-As should be clear by now, any teaching materials generated in this way can only aspire to be average given the prompts used to get the machine going.
-As a shortcut to save time (and especially money), GenAIs might be of interest, but no one should buy the delusion that the use of this technology makes them innovators.
-In any case, the list of remedial measures required to verify that the materials produced by the AI algorithm are not nonsense is so long, that it is hard to see what is being gained [see @mollick_instructors_2024, pp.
+As should be clear by now, any teaching materials generated in this way can only aspire to be average given the prompts used to get the machine going, and no one should buy the delusion that the use of this technology makes them innovators.
+In any case, the list of remedial measures required to verify that the materials generated are not nonsense is so long, that it is hard to see what is being gained [see @mollick_instructors_2024, pp.
 4-5].
@@
 There is nothing new about automated grading: [Scantron](https://www.scantron.com/company/) sheets for multiple choice examinations have existed for over four decades.
-They work well because the algorithm needed for grading multiple choices is simple, and importantly, deterministic (i.e., there is no randomness involved).
+They work well because the algorithm needed for grading multiple choices is simple, and importantly, non-random.
 There is no reason to use a neural network to grade multiple choice examinations.
-But on the flip side, there is also no evidence that a neural network can perform well at grading anything more complex than a multiple choice examination, especially since 1) it will tend to consider things that are closer to the mean as "better"^[My own unpublished experiments with grading written reflections were underwhelming, missing the human-assigned grade over 30\% of the time]; and 2) the amount of training required for a classification task of this nature is massive^[Labeling data is an essential, yet extremely expensive aspect of training GenAIs.
-The algorithm needs a "dependent" variable for training, in this case a grade assigned to a piece of work: this is called a label.
-Grading assignments manually is the process of labeling.
-Given all the possible variations of assignments (by year, by subject, by instructor), a general grading machine would require vast volumes of _labelled_ training data, and even so there is no guarantee that the results would be adequate, let alone appropriate.].
+But on the flip side, there is also no evidence that a neural network can perform well at grading anything more complex than a multiple choice examination, since 1) it will tend to consider things that are closer to the mean as "better"; and 2) the amount of training data required for a classification task of this nature is massive.
+Labeling data is an essential, yet extremely expensive aspect of training GenAIs.
+The algorithm needs a "dependent" variable for training, a "label", which in this case a grade that a human assigned to a piece of work.
+Given all the possible variations of assignments (by year, by subject, by instructor), a general grading machine would require vast volumes of _labelled_ training data, and even so there is no guarantee that the results would be adequate, let alone appropriate.
 
@@
 
-Using a chatbot to write reference letters must be one of the most pointless uses of chatbots in higher education, or one of the most dishonest.
-Imagine that the reference letter honestly included a disclosure that ChatGPT was used to write it^[As it should, according to McMaster's guidelines for acknowledging the use of AI.].
+Using a chatbot to write reference letters must be one of the most pointless uses of chatbots, or one of the most dishonest.
+Imagine that the reference letter honestly included a disclosure that ChatGPT was used to write it.
 If the objective of a reference letter is for someone to attest from the vantage point of some level of expertise as to the skills and qualities of a person, a chatbot-written letter is probably worse than useless: what message does the disclaimer send to the recipient of the letter?
 And how much stock will anyone place on the veracity of the letter?
-To avoid this catch, one would have to occult or obscure the use of AI, and in doing so would end up passing text written by countless others and filtered through a machine, as if it was their own.
-
-Equally troublesome is the need to input some personal data, so as to avoid getting a generic letter from the machine.
-Does the author of the letter need consent to share personal information about the subject of the letter with OpenAI et al.?
+To avoid this catch, one would have to occult or obscure the use of AI, an act that lays bare the dishonesty of the activity.
 
@@
 
-Neural networks have been used for decades to do data analysis.
-There is nothing new here.
-But what is being promised is that we will be able to do data analysis without having to learn how to do data analysis.
-That somehow a chatbot will do it for us.
-The idea that a chatbot could be used to do data analysis betrays a profound misunderstanding of what chatbots do, and almost absolute ignorance of what data analysis entails.
+Computers have been doing mathematics reliably since at least the [Antikythera mechanism](https://en.wikipedia.org/wiki/Antikythera_mechanism) (ca.
+178 BC),.
+Electronic computers have been doing mathematics for decades now, and neural networks have been used for decades to do data analysis.
+There is nothing new about that.
+But the promise rather is that we will be able to do data analysis without having to _learn_ how to do data analysis.
+The idea that a chatbot could be used to do data analysis betrays a profound misunderstanding of both what chatbots do and what data analysis entails.
 Ask a chatbot to do some arithmetic: it will produce something that is right on average.
 But average is not always right.
-And seeing how AIs are now being trained on the dregs of the internet, it should surprise approximately no one when an AI Overview responds to the prompt "2 + 2" with "potato".
+And seeing how AIs are now being trained on the dregs of the internet, it should be no surprise when an AI Overview responds to the prompt "2 + 2" with "potato".
 
@@
 An error like this is easy to spot.
-But we seldom use computers to do calculations that can be done by hand, and checking that an AI-generated response is not wrong is likely to be as time consuming as not using an AI to do quantitative data analysis in the first place [see @yuan_how_2023].
+But we seldom use computers to do calculations that can be done by hand, and checking that an AI-generated response is not wrong is as time consuming as doing the analysis without an AI in the first place [see @yuan_how_2023].
 
@@
 Change those weights, the results will change.
-As well, even small variations in the prompt can be impactful in models with millions upon millions of parameters.
-In statistical terms, the output is but one _realization_ of a random process.
+As well, even small variations in the prompt can be impactful in models with billions or trillions of parameters.
+In statistical terms, any output is but one _realization_ of a random process.
 For the results to be useful, we should calculate the average of a large number of responses.
@@
 
-But what about a task that involves logic instead of arithmetic? LLMs are good for it only if the goat knows how to swim.
-
-```{r puzzle-web-search, eval=FALSE, out.width="70%", fig.cap="A search using Duck Duck Go"}
-ggdraw() + 
-  draw_image(image = glue::glue(here::here(), "/images/logical-puzzle-web-search.png"))
-```
-
-```{r puzzle-chatgpt, out.width="70%", fig.cap="ChatGPT-4 makes the goat swim"}
-puzzle_plot <- ggplot() +
-  labs(caption = "https://assemblag.es/@hugo/112450401657084750") +
-  theme_void() +
-  theme(text = element_text(size = 8))
-
-ggdraw() + 
-  draw_plot(puzzle_plot) +
-  draw_image(image = glue::glue(here::here(), "/images/logical-puzzle-chat-gpt-4.png"))
-
-```
-
-As research by @nezhurina_alice_2024 demonstrates, relatively simple natural language puzzles that are easily solved by humans, often result in confidently expressed nonsense when used to prompt an AI.
-Wrong answers, furthermore, are frequently backed up by confabulations, that is, reasoning-like or plausible sounding explanations that are illogical.
-If anything, mathematics should be easy: after all, computers have been doing mathematics since at least the [Antikythera mechanism](https://en.wikipedia.org/wiki/Antikythera_mechanism) (ca.
-178 BC), which preceded other mechanical machines by centuries--including Babbage's [Analytical Engine](https://en.wikipedia.org/wiki/Analytical_engine), the machine that [Ada Lovelace](https://en.wikipedia.org/wiki/Ada_Lovelace) used to code an algorithm to write Bernoulli numbers.
-Electronic computers have been doing mathematics reliably for decades now.
-Computers can do mathematics well because they are implemented using a crisp, unambiguous language.
-Natural language is full of nuances that are not easily captured using the medium of text alone.
-The stochastic connections between words that LLMs produce are simply probabilistic word salads^[@xu_hallucination_2024 use a formal approach to show that LLMs are inherently incapable of mapping outputs to a ground truth, and thus their results will only ever resemble a ground truth purely by chance.
-Recent reporting strongly hints that industry leaders like  Sundar Pichai and Tim Cook understand this well [@patel_google_2024; @tyrangiel_opinion_2024]] that only sound plausible thanks to the billions of data points used to train them and the billions of parameters that the models use.
+But what about a task that involves logic instead of arithmetic?
+Research by @nezhurina_alice_2024 demonstrates that relatively simple natural language puzzles that are easily solved by humans, often result in confidently expressed errors when used to prompt an AI.
+Wrong answers, furthermore, are frequently backed up by confabulations, that is, nonsense expressed in reasoning-like terms.
+The onnections between words that LLMs deliver are but probabilistic word salads: @xu_hallucination_2024 use a formal approach to show that LLMs are inherently incapable of mapping outputs to a ground truth, and thus their results will only ever resemble a ground truth purely by chance.
+The outputs of a chatbot only sound plausible thanks to the vast amount of data points and parameters that the models use.
 But if the objective of qualitative research is to identify interesting aspects of the human experience, it is certainly not the case that they will be found in a regression to the mean.
@@
 An AI should not be used to identify signs of learning difficulties for the same reasons that AIs cannot be trusted to identify text written by an AI.
-The false positive rates (blaming the innocent) are out of proportion compared to the benefits the practice might offer.
-A vigilant instructor will be aware of learning difficulties among their students (but not if the students are an ocean of anonymous faces).
+The false positive and negative rates are out of proportion compared to the benefits the practice might offer.
+A vigilant instructor will be aware of learning difficulties among their students.
 But were the process to become automated, who would decide when an intervention was warranted, and who qualified for it?
@@
 And, could those denied an intervention complain?
-Importantly, who would fund any intervention?
-And who would deliver it?
+Importantly, who would fund an intervention, and who would deliver it?
 Another AI?
 
-<!--
-- AI-powered systems can provide smart tutoring.  
-- AI can enable virtual classrooms.  
-- AI can provide data insights for informed decisions.  
-- Natural Language Processing (NLP) can enhance language learning.  
--->
+- Personalized learning
+
+Anyone receiving "personalized learning" from an AI can be sure that they will be treated as the machine-generated average of their class, and not as a person.
 
@@
 
-The preceding section illustrates how numerous tasks that AIs can notionally do in higher education are in fact poor replacements for a human-centered education--and that is even before considering the moral, ethical, and possibly legal compromises that the use of corporate-developed AIs demand (non-corporate developed AIs can be ignored: as @vandehei_behind_2024 remind us, only a handful of corporations have the resources needed to create the illusion of compentence in an AI).
-
+The preceding section illustrates how numerous tasks that AIs can notionally do in higher education are in fact poor replacements for a human-centered education.
 But what about the things that GenAIs can do _to_ students?
@@
 It is a task that cannot possibly be performed in a systematic way, and that offers no guarantees that it will work in the same way again, especially after a model is updated.
-No genuine engineer would do their work in such fashion--and calling the activity of writing prompts "engineering" is a blatant attempt to steal the competence and reputation of actual engineering.].
+Calling the activity of writing prompts "engineering" is a blatant attempt to steal the competence and reputation of actual engineering.].
 
-It is important to note that, even if prompt "engineering" offered some value to students, the skill is inextricable tied to a singular technology, with little transferability potential.
-As @mollick_instructors_2024 demonstrate, prompt "engineering" is not even transferable _between_ different LLMs (see p.
+It is important to note that, even if prompt "engineering" offered some value, the skill is tied to a singular technology and offers little transferability potential.
+@mollick_instructors_2024 demonstrate how prompt "engineering" is not even transferable _between_ different LLMs (see p.
 5 and p.
@@
 Good skills transfer between contexts.
-Learning how to form an effective bibliographical query would work at retrieving relevant resources at McMaster, at the University of Toronto, and at the New York Public Library.
-Learning to identify relevant content would work while reading a novel, a book of history, or a text of econometrics.
+Learning how to do an effective bibliographical query transfers between libraries.
+Learning to identify relevant content works while reading a novel, a book of history, or a text of econometrics.
 Writing, the slow and demanding process of converting thoughts into words, is essential to create a style of communication that is unique to each person.
@@
 
-\newpage
-
 In a sense, an AI is like a crutch.
-It can help someone to accomplish a task more easily, but it will not help if used constantly.
-However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle that the end user cannot hope to fully master^[Reports of people getting their fingers crushed by Tesla's cybertruck could be a result of 1) people trusting the technology more than they should; and 2) not understanding the technology: apparently the algorithm uses increasing force to close the trunk if it fails to latch on back-to-back cycles.
-Hence, the trunk will not slash a banana on its first attempt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature--in the sense that 1) neural networks are blackboxes; and 2) the development of AIs itself is a blackbox wherein things can more easily change based on the interests of their corporate owners^[Sometimes said interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that corporate interests will always align with those of the public, and in fact, they seldom will as long as there are only a handful of American companies that offer an "essential" service.] than in response to the interests of their relatively small customers^[And this refers not only to individuals whose skills become yoked to ChatGPT and MidJourney.
-In 2022, McMaster University operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD).
+However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle, but one that the user cannot hope to fully master due to its blackbox nature--in the sense that 1) neural networks are blackboxes; and 2) the development of AIs itself is a blackbox wherein things can more easily change based on the interests of their corporate owners than in response to the interests of their relatively small customers^[And this refers not only to individuals whose skills become yoked to ChatGPT and MidJourney.
+As an example, in 2022 McMaster University operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD).
 Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022.
 To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero dollars and still account for less than _one third of one percent point_ of Microsoft's revenue.
-A mid-sized Canadian university is like an ant scuttling under the economic boot of Microsoft.
-In truth, Microsoft is so incredibly large, that it feels like it can stand up to even the US Government.
-As revealed by the case of Solar Winds, Microsoft refused to fix a known security flaw that left the US Government vulnerable to a Russian hack, and probably did so with little fear of ever facing accountability [@burke_microsoft_2024].].
+A mid-sized Canadian university is like an ant scuttling under the economic boot of Microsoft.].
 
@@
 
-One could ask whether there is not more value in doing things slowly, less expertly at the beginning, in order to make the delicate process of thought work, so that the brain can accomplish genuinely exceptional things later--perhaps even more efficiently.
+However, let us assume for a moment that using AIs in their current form does becomes an in-demand tool in the short term. What is the expected longevity of a skill like prompt "engineering"?
 
-However, let us assume for a moment that using AIs in their current form does becomes an in-demand tool in the short term. What is the expected longevity of skills like prompt "engineering"?
-
-Apart from a few methodological developments (like transformers), improvement in the performance of AIs (say, from ChatGPT-2 to the latest version), mostly are the result of using bigger datasets.
-This has led to a rush, as developers try to improve their models by training them on more and more data^[A rush that feels at least as frenetic as any gold rush in the past.
-And, while the gold rushes of the 19th century were relatively democratic due to their low entry costs--both in terms of immigration and tools--the current data rush has enormous entry costs and is largely dominated by large corporations and venture capital (that is, equity investments that support pre-launch, launch, and early stages of a business).].
+Besides a few methodological developments, improvement in the performance of AIs has mostly relied on using bigger datasets.
+This has led to a rush, as developers try to train their models with bigger datasets^[A rush that feels as frenetic as any gold rush in the past.
+However, while the gold rushes of the 19th century were relatively democratic due to their low entry costs, the current data rush has enormous entry costs and is largely dominated by large corporations and venture capital.].
 But, despite the large number of text items and images available in digital form, there is still only a finite amount of data to be had.
@@
 As of this writing, the few big players that can afford to grab the data, are mostly done with all the data grabbing, sometimes by morally dubious, and at times by possibly illegal means [@clark_ex-amazon_2024].
-IBM, for example, uses data drawn from the Project CodeNet Dataset.
-The data are "sourced" from two online judge web sites, AIZU Online Judge and AtCoder [@ibm_automatic_2021].
-According to IBM "[the] end goal of CodeNet is to enable developers to create systems that can modernize existing codebases, as well as fix errors and security vulnerabilities in code".
-When we inspect the Terms of Use of AIZU (a Japanese academic project), "Commercial uses of the registered codes are prohibited".
-IBM studiously avoids the word "commercial" in its posts about LLM, foundational models, and CodeNet.
-One might ask if IBM is doing all this research as a non-commercial effort?
-Will those developers creating systems, modernizing databases, fixing errors, etc., be able to access CodeNet for free, and use the systems for education and research only?
-What does "commercial uses" even mean?
+And so, as the fountain of human-generated content has been reduced to a drip by a handful of corporations developing AIs, there has been a turn towards what might be the last untapped drops of information left on the internet.
 
-As the fountain of human-generated content--generously (one might even say naively) shared by untold numbers of people over a period of decades--becomes exhausted by a handful of corporations developing AIs, there has been a turn towards what might be the last juicy nuggets of information left on the internet.
+OpenAI, a global leader in data grabbing, resorted to data from Stack Overflow [@stack_overflow_stack_2024], a website that [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges".
+Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) (to say the least) with the way Stack Overflow dealt with OpenAI, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not do_.
+It is an open question how long Stack Overflow will remain a community-based space after it started treating its members as unwitting, and even unwilling providers of data for OpenAI [@grimm_stack_2024].
 
-OpenAI, a global leader in data grabbing, turned to Stack Overflow for more data [@stack_overflow_stack_2024].
-Stack Overflow [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world".
-Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) (to say the least) with the way Stack Overflow dealt with OpenAI, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not do_.
-Some members of the community protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica unhelpfully termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/).
-It is an open question how long Stack Overflow will remain a community-based space, especially after it started treating its members as unwitting, and even unwilling providers of data for OpenAI [@grimm_stack_2024].
-
-Google, for its part, turned to Reddit for training data to the tune of \$60 million per year [@roth_google_2024].
-As Reddit content begins to creep into Google's AI-assisted search, the results of this deal have turned grotesque, with reports of AI Overview failures that include a suggestion to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024]^[The headline of the 404 Media report was "Google Is Paying Reddit $60 Million for Fucksmith to Tell Its Users to Eat Glue".].
+Google, for its part, appealed to Reddit for training data to the tune of \$60 million per year [@roth_google_2024].
+As Reddit content begins to creep into Google's AI-assisted search, the results of this deal have turned grotesque, with reports of AI Overview failures that include a suggestion to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024].
 
@@
 
-\newpage
+Mind you, these, and numerous other hapless results, are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
+To what extent can more money improve GenAIs?
+Probably not by much, since a key constraint is the amount of data available for model training [@villalobos_position_2024].
+And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data.
+As GenAIs become more widely used, they can produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the most thoroughly vetted form of writing in the world [@cabanac_tortured_2021; @maiberg_ai_2024].
 
-Mind you, these, and numerous other hapless results^[Including AI overviews with misinformation about depression [@tenbarge_glue_2024].] are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
-To what extent can more money improve GenAIs?
-Probably not by much, since a key constraint is the amount of data available for training the deep neural networks behind the technology [@villalobos_position_2024].
-And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data^[A disturbing idea in and of itself, given the appetite for risky behaviors by corporate actors, and the potential for abuse of nominally private and confidential information.
-It would be too naive to expose even a morsel of juicy data to transmission over the internet, and expect Google et al., not to see it as fair game.].
-As GenAIs become more widely used, they can produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites^[A blog post on a website called _Ancient Rome_ by a certain Ellen Hunter (whose profile picture is likely the output of a generative AI and who does not seem to exist elsewhere) discusses the topic of windows in classical Rome and Greece [@hunter_did_2023].
-Under subheading 4.1 (_What Existed Before Windows?_), the blog explains that "MS-DOS was a disk operating system that was developed by Microsoft.
-It was the predecessor to the Windows operating system and was used on IBM PC compatible computers.
-MS-DOS was first released in 1981 and was the most popular operating system for personal computers during the 1980s".
-Clearly, the text is the output of a chatbot, and was posted to the internet without regard to its accuracy; but why waste time on even a cursory check, when the objective of the blog is not to inform, but to attract clicks for monetizing ads?] and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the most thoroughly vetted form of writing in the world [@maiberg_ai_2024]^[And let us not fool ourselves with the idea that this will be limited to smaller, scammy paper mills, when examples begin to crop up in publications by IEEE, Elsevier, and Springer, as evidenced by the telltale use of "tortured phrases" [@cabanac_tortured_2021] and chatbot stock expressions like "[a]s of my last knowledge update...", "[a]s of my knowledge cutoff...", "[c]ertainly!
-Here are...", and "[r]egenerate response" [e.g., @al-worafi_pharmacy_2023; @farag_exploring_2023; @li_revitalizing_2023; @shafiq_escrao_2023; @swain_deep_2023; @alsaif_nd2o3-_2024; @gatell_impact_2024; @remzan_ensemble_2024; @sumit_dynamic_2024; @zhang_retracted_2024].
-One particularly egregious example straight out includes the phrase "[p]lease note that as an AI language model, I am unable to generate specific tables or conduct tests, so the actual results should be included in the table" [@yang_utilization_2023].].
+What can we expect as future GenAI models begin to be trained using data that was generated by previous GenAIs?
+Revisiting the examples in a preceding section ([But really, what do AIs do?](#really-what-do-they-do)) we can see that the predictions of a model are more uniform than the original data used for training.
+Technically, the variance of a model's output is substantially reduced compared to the variance of the training data: the outputs of a model will be less varied, less diverse, and more norm-conforming.
+The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs.
+We may be witnessing _Peak ChatGPT_, and future model versions, instead of improving, may begin to regress to duller versions of their own previous outputs.
+To compound things, as backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023], including _PhotoGuard_ [@heikkila_this_2023], _Nightshade_ [@shan_nightshade_2024], _Shadowcast_ [@xu_shadowcast_2024].
 
-What can we expect as, in what probably is inevitable by now, future GenAI models begin to be trained using data that was generated by previous GenAIs?
-Revisiting the examples in a preceding section ([But really, what do AIs do?](#really-what-do-they-do)) we can see that the predictions of the models are more uniform than the original data used for training.
-Technically, the variance of a model's output is substantially reduced compared to the variance of the training data: the outputs of a model will necessarily (as they regress towards the mean) be less varied, less diverse, and more norm-conforming (see Table 1).
-
-```{r data-vs-prediction-variance}
-df_pred_all |> 
-  group_by(type) |> 
-  summarize(variance = var(y)) |>
-  transmute(Model = factor(type, 
-                       levels = c("training data", "regression output", "nn1 output", "nn2 output"),
-                       labels = c("Training data", "Linear regression", "Neural network 1", "Neural network 2"),
-                       ordered = TRUE),
-         Variance = round(variance, 2)) |>
-  kable(caption = "Variance reduction in model outputs")
-```
-
-The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs.
-We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to duller versions of their own previous outputs.
-And this is before we even consider the possibility of AIs being deliberately "poisoned".
-As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023].
-One such tool is _PhotoGuard_, developed by a team at MIT to send signals to an AI that prevents it from manipulating images [@heikkila_this_2023].
-Other examples include @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts.
-And more recently, @xu_shadowcast_2024 presented _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.
-
-All in all, the evidence suggests that, as fresh data for training future GenAIs dries up, the performance of these machines may not improve drastically from what we have seen so far, that is, flashy results that bear no relationship to the truth, being in fact the output of bullshit-producing machines^[American philosopher Harry G.
-Frankfurt defines _bullshit_ as speech intended to persuade without regards to the truth [@frankfurt_bullshit_2005].
-AIs, lacking intent, do not lie--but they do bullshit.].
-In fact, the performance may begin to deteriorate as the output of GenAIs begins to be used to train other GenAIs.
-This effect could easily be compounded if, as is likely the case, human-generated content is exhausted, and/or is overwhelmed by GenAI content, and/or is corrupted by purposefully bad data supplied by actors ranging from outraged creators to adversarial regimes.
-All of which begs the following questions.
+All in all, the evidence suggests that the scarcity of fresh data, paired with more abundant model-generated and/or poisoned data, will make the performance of these machines stagnate or deteriorate.
+Which begs the following questions.
 How useful will AIs be in the future?
 How persistent will skills like prompt "engineering" be?
-Will learning to prompt an AI be worth sacrificing the opportunity to develop non-AI-powered skills like coding, writing, and thinking?
-Who, after all, stands to benefit the most by this form of deskilling, in exchange for the effort required to develop an ability that may turn out to be of ephemeral value at best and a dead-end at worst?
-
-<!--
-Develop a garbage skill
-
-Automation bias: https://en.wikipedia.org/wiki/Automation_bias
-
-It is likely that there is a similar authority bias in which students tend to be more positive
-
-### Persistence of skills (how long will learning to use a chatbot be useful?)
-
-How useful will be AIs in the future? Think of model poisoning, and the possibility that we are going to see (or may already have seen) Peak ChatGPT.
-
-De-duplication of training data improves LLMs [@lee2022deduplicating]. No wonder: it amounts to having more different information.
-
-But there is the following implication: , de-duplication becomes increasingly challenging.
--->
-
-## What do GenAIs do for the Province?
-
-As I argue above, AIs do not seem to do much for students or instructors, other than distract them from learning and developing genuinely transferable skills.
-
-But AIs must do something for someone, otherwise why would they be everywhere?
-Given history, one could be forgiven to think that the provincial government is hoping to use GenAIs to further reduce the need to fund higher education.
-GenAIs appear on the horizon as an  innovation-sounding rationale (a mandate even!) for the province to force institutions of higher education to find "efficiencies".
-Tom Goldstein, a computer science professor at the University of Maryland, thinks that the lure of GenAI will be hard to resist for many, irrespective of the technology's flaws...after all, "[e]ven though [GenAI is] expensive...it's still far less expensive than human labor" [@oremus_chatgpt_2023].
-Without adjudicating the merits of the argument that human labor is expensive^[When the alternative is substandard quality of life for many of our fellow citizens, and/or higher unemployment.], one would hav to be blind not to see the glamour that GenAI will cast on governments that do not seem to believe in higher education in the first place.
-
-```{r googly-eyes, out.width="70%", fig.cap="\\label{fig:googly-ai-eyes}'I like the look of his cuts'"}
-google_plot <- ggplot() +
-  labs(caption = "Source: https://godforsaken.website/@alex/112541393417513370") +
-  theme_void()  +
-  theme(text = element_text(size = 8))
-
-ggdraw() + 
-  draw_plot(google_plot) +
-  draw_image(image = glue::glue(here::here(), "/images/ai-googly-eyes.png", width = 0.8))
-```
+Will learning how to prompt an AI be worth sacrificing the opportunity to develop transferable skills like coding, writing, and thinking?
+Who, after all, stands to benefit the most by the form of deskilling that AIs encourage?
 
@@
 
-> _Teach a person to use ChatGPT, and they will be in thrall to a corporation for as long as that is their skill. Teach them to think, and they will be free for life._  
-> -- Anonymous
-
-But neither the province nor the university are developing the AIs that they hope will make them cheaper to operate.
-So what do AIs do for the the entities that develop them, namely OpenAI et al.?
-So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as those spent by universities to support an AI-dedicated committee^[Some observers have even wondered whether the AI fever has not actually become detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].].
-Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].
-
-<!--
-Ministry of Colleges and Universities: 163,482,281 CAD
-Venture capitalists: 75,000,000,000 USD == 257,776,200,000 CAD
-
-https://www.exchange-rates.org/exchange-rate-history/usd-cad-2022
-Avg. exchange rate in 2022: 1.3019 CAD/US
-
-This is how much bigger venture capital in AI is compared to research in Ontario. 
-163482281/(75000000000 * 1.3019)
---->
+AIs do not seem to do much of value for students, as they can be wasteful at best and harmful at worst.
+But the technology must do something for someone, otherwise it would not be ubiquitous.
+So what do AIs do for the the organizations that develop them?
+So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities^[Some observers have even wondered whether the AI fever has not actually become detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].].
+Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.] [@tricot_venture_2021].
 
@@
 The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing the most intensive, disturbing, degrading, and damaging aspects of training AIs to the Global South [@nix_analysis_2024].
-The fact that the industry is not profitable has not prevented start-ups from hurriedly pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted from cryptocurrency to AI, coincidentally as cryptocurrency was going sour after the frauds at FTX were revealed.
-Barely disguised grift has grown abundant in the vicinity of AI at almost every level [@cook_5_2024].].
 
@@
 And how much those who end up paying will need to pay?
-Or, in the worst case, _have_ to pay, if their skills become inexorably tied to this one technology?
-Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability^[And, with profitability increasingly looking like a distant possibility, there is always the option for large investors to cosmically inflate the hype [@salmon_what_2024] in order to convince bigger suckers to give them money [@reuters_openais_2024] and/or to fleece retail investors [@temkin_vcs_2024].]: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020].
-From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service^[Universities, rather than being innovators, have increasingly become renters in the land of the platform economy [@komljenovic_rise_2021].].
+Or, in the worst case, will _have_ to pay, if their skills become tied to this one technology?
+Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020; @komljenovic_rise_2021].
+From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service.
 
@@
 
-ChatGPT 4.o was announced on May 13, 2024, with great fanfare and a flashy display of the model's new capabilities.
-These included the chatbot's ability to respond vocally to prompts, and to identify images via a mobile device's camera.
-4.o's voice, in particular, caused some titillation in the nether parts of even its creators [@roose_is_2024].
-The voice was modeled after the digital personal assistant of the movie _Her_, played by Scarlett Johannson.
-By Sam Altman's own account, this is one of his favorite movies.
-The background story of 4.o's voice is somewhat sordid: Altman reportedly tried to hire Ms.
-Johannson to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when she declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways.
-This voice was retired (probably to the profound sadness of many an AI engineer) after Ms.
-Johansson threatened legal action^[Altman's denials that OpenAI had not used Ms.
-Johansson's voice sounded hollow, after he [posted](https://x.com/sama/status/1790075827666796666) on social media the word "her" at the time the new version of ChatGPT was announced.] [@allyn_scarlett_2024].
+ChatGPT 4.o was announced on May 13, 2024, with great fanfare and a flashy display of the model's new capabilities, which included functions to respond vocally to prompts, and to identify images via a mobile device's camera.
+4.o's voice, in particular, caused considerable titillation [@roose_is_2024].
+The voice was modeled after the digital personal assistant of the movie _Her_, Sam Altman's favorite movie by his own account.
+The background story of 4.o's voice is more sordid: Altman reportedly tried to hire Scarlett Johannson, who played the AI assistant in the movie, to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when she declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways.
+This voice was retired after Ms.
+Johansson threatened legal action [@allyn_scarlett_2024].
 
-Alas, very few have the social standing and wealth of Scarlett Johansson to challenge OpenAI.
-Most people in the world lack that power and the AI industry has not been shy to exploit that imbalance [@roberts_commercial_2016], with Google's privacy policy for using the whole internet as the source of data for its AI efforts [@germain_google_2023], and Adobe landing in hot water by adopting terms of use widely thought to be confiscatory [@weatherbed_adobes_2024].
+Alas, very few have the wealth and social standing of Scarlett Johansson to challenge OpenAI.
+Most people in the world lack that power and the AI industry has not been shy to exploit the imbalance [@roberts_commercial_2016; @germain_google_2023; @weatherbed_adobes_2024].
 
@@
 ChatGPT was trained on vast amounts of data scooped from the internet, where many societal problems are amplified.
-Trained on horrible content, it was hard to sell ChatGPT when it was prone to racist, sexist, obscene outbursts.
-Fortunately (or unfortunately, depending on perspective) companies like facebook, who did not want to pay for moderation, had already hit on a solution of sorts, which was basically to train AIs to detect content that was unpalatable.
+Trained on horrible content, it was hard to sell a ChatGPT prone to racist, sexist, obscene outbursts.
+Fortunately for OpenAI, companies like facebook had already hit on a solution of sorts to reduce the cost of moderation.
+The idea was basically to train AIs to detect content that was unpalatable.
 Indeed, they had found that feeding an AI with labeled examples of all the worst that exists on the internet (e.g., violence, hate speech, child sexual abuse) could produce tools that learned to detect various manifestations of such toxicity in the wild, to filter it out before it reached users.
@@
 
-What did this mean to the people doing all the labeling in one of the poorest regions in the world?
-First of all, an income: "data labelers employed...on behalf of OpenAI were paid a take-home wage of between around $1.32 and $2 per hour depending on seniority and performance." But at what cost?
+What did this do for and to the people doing all the labeling in one of the poorest regions in the world?
+First of all, it provided an income: "data labelers employed...on behalf of OpenAI were paid a take-home wage of between around $1.32 and $2 per hour depending on seniority and performance." But at what cost?
 As one worker tasked with reading and labeling text for OpenAI told TIME, "he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child.
@@
 
-<!--
-
-In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )
-
-@norem_intels_2023
-
-The ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted.
-
-@williams_exploited_2022 (Summary needed)
-
-@hao_hidden_2023
-
-Quotes from Hao's podcast:
-
-> Karen Hao: One of the reasons why ChatGPT was able to become so virally popular and continued to sustain popularity is because it is largely not spewing really awful things.
-People feel comfortable using the product knowing that it's not going to do that.
-
-> Annie Minoff: At least it won't do it in English. If Alex wanted to use ChatGPT in his native language, Swahili, would he be able to do that?
-
-> Karen Hao: You can interact with ChatGPT in Swahili, but ChatGPT was developed primarily to work in English.
-So a lot of the scrubbing, the content moderation, the important safety measures within the chatbot were done in English.
-So when you prompt it in Swahili, you'll get more misinformation.
-You'll get more confusing sentences that don't make sense, and you will potentially get more of this content that they worked so hard to filter out because they were only filtering it in English.
--->
-
 ### Privacy issues
@@
 LLM are known to regurgitate verbatim chunks of the training datasets.
-This is called _memorization_, a word that contributes to antropomorphize what is essentially a statistical model.
-Memorization, in fact, is just a fancy way of saying "model overfitting"--in other words, an extreme regression to the mean that excels at predicting data in the training set, but is poor at projecting new data points.
-Overfitting happens when a model has too many parameters or too few data inputs (Models have [billions of parameters](https://en.wikipedia.org/wiki/Large_language_model#Training_cost) and are trained on [billions of pages of text](https://www.ibm.com/topics/large-language-models])).
-Data memorization--the verbatim or quasi-verbatim output of training data--can be as high as 7% in the case of some LLMs [@peng_near_duplicate_2023, p.
+This is called _memorization_, or in more conventional terms, model overfitting--an extreme regression to the mean that excels at predicting data in the training set.
+Overfitting happens when a model has too many parameters or too few data inputs.
+Data "memorization"--the verbatim or quasi-verbatim output of training data--can be as high as 7% in the case of some LLMs [@peng_near_duplicate_2023, p.
 8].
-
-This means that data cannot be presumed to be private once that it has been used to train an AI.
-In response organizations such as the National Archives and Records Administration in the US [@koebler_national_2024], and the US Congress [@singh_after_2024] have banned the use of ChatGPT and Copilot.
-@singh_after_2024 also notes how companies like Samsung and Apple have restricted their use of chatbots.
-
-NetBSD (a project that produces a free, Unix-like Open Source operating system) states in its commit guidelines (contributions to code) that "[c]ode generated by a large language model or similar technology...is presumed to be tainted code, and must not be committed without prior written approval by core." [@netbsd_netbsd_nodate].
-The concerns cited are seldom ethical or moral, but mostly about the risk of sensitive information leaking.
-As LLMs begin to be trained on model-produced data [@murgia_why_2023], overfitting (or memorization) will likely become a larger risk.
-
-And this is just a result of using an overfitted model.
-What about data leaks?
-Microsoft's data security [track](https://www.theregister.com/2024/04/12/microsoft_cisa_order/) [record](https://techreport.com/news/microsoft-azure-hit-with-the-largest-data-breach-in-its-history-hundreds-of-executive-accounts-compromised/) does not inspire confidence.
+This means that data cannot be presumed to be private once that they have been used to train an AI.
+And this issue is just a possible result of using overfitted models.
+But what about data leaks?
+So far, organizations such as the National Archives and Records Administration in the US, the US Congress, and companies like Samsung and Apple have restricted or banned the use of ChatGPT and Copilot for data security reasons [@koebler_national_2024; @singh_after_2024]: no data transferred in the form of a prompt can be presumed to be safe.
 
@@
 
-<!--
-I struggle a bit with this one.
-For example, I fly more often than the [typical Canadian](https://ourworldindata.org/grapher/air-trips-per-capita?tab=table&time=earliest..latest&country=~CAN) and a little bit less often than the [typical world traveller](https://www.news24.com/news24/around-the-world-in-50-hours-average-person-takes-65-flights-per-year-20180828) (but I have not owned a car in decades).
-Who am I to dictate how much damage _other people_ have a right to inflict (or not) to the environment.
--->
+Every human activity has an environmental impact, and at a time when we face unprecedented challenges with climate change, we must be discerning in how we consume resources.
+The choice is clear and Microsoft has told us: we need to burn the planet [@smith_our_2024] to create the AIs that will help us save the planet [@nakagawa_accelerating_2023].
 
-Every human activity has an environmental impact, and at a time when we face unprecedented challenges with climate change, we must be discerning in how we consume resources.
-The choice is so clear that Microsoft has basically told us: we need to burn the planet [@smith_our_2024] to create the AIs that will help us save the planet [@nakagawa_accelerating_2023].
-
-@crawford_generative_2024 has already sounded the alarm: the environmental costs of AI are soaring, and doing so mostly in secret [@luccioni_power_2024].
-It has been estimated that ChatGPT consumes substantially more energy than a regular web search, up to 25 times more, and according to some estimates, the whole of AI may consume twice as much energy as the whole of France by 2030 [@brussels_chatgpt_2024].
-As the examples in the preceding sections illustrate, this is in exchange for results that are not better than a simple search, and often are worse.
-Widespread AI deployment may in fact make the climate crisis worse, not only by guzzling resources, but also by spreading misinformation [@gordon_multiple_2024].
-
-<!--
-https://www.theguardian.com/commentisfree/article/2024/may/30/ugly-truth-ai-chatgpt-guzzling-resources-environment
-
-https://www.wired.com/story/the-generative-ai-search-race-has-a-dirty-secret/
-
-...are the results better?
-
-Compare the results:
-
-The example above may seem like an amusement, but a .
-
-Similar GenAI-generated mishaps abound.
-
-But what when an AI replicates the both-sides perspective of much of the media when trying to learn about an existential crisis like climate changes?
-[@gordon_multiple_2024].
-Or when trying to obtain accurate voting information in the context of voter supression?
-[@angwin_seeking_2024].
--->
-
-\newpage
+Researchers like @crawford_generative_2024 have already sounded the alarm: the environmental costs of AI are soaring, and doing so mostly in secret [@luccioni_power_2024].
+It has been estimated that ChatGPT consumes up to 25 times more energy than a regular web search,, and according to some estimates, the whole of AI may consume twice as much energy as the whole of France by 2030 [@brussels_chatgpt_2024].
+As the examples in the preceding sections illustrate, this is in exchange for results that are not better than a simple search, and often are considerably worse.
 
@@
 Sam Bankman-Fried's FTX donated millions to a group with racist ties [@wilson_sam_2024].
-And Elon Musk?
-Well, Elon Musk once shared a meme comparing Canada's PM Justin Trudeau to Hitler while defending the vandals who terrorized Ottawa in 2022 [@henderson_and_elon_2022].] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars...but maybe not before super-intelligent AIs destroy humanity.
-Just in case, wealthy tech barons like Peter Thiel and Sam Altman, have made suitable [escape plans](https://www.theguardian.com/news/2018/feb/15/why-silicon-valley-billionaires-are-prepping-for-the-apocalypse-in-new-zealand).], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the well-being of trillions of humans in an imagined distant, star-faring future, is well worth the sacrifices of a few billions of humans today.
-Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that wealthy and well connected longtermists are willing to make.]).
+Elon Musk once shared a meme comparing Canada PM Justin Trudeau to Hitler while defending the vandals who terrorized Ottawa in 2022 [@henderson_and_elon_2022].] known for fringe ideologies that combine uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that those who can afford it, are destined to use technology to indefinitely extend life and conquer the stars...but maybe not before super-intelligent AIs destroy humanity.] with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the well-being of trillions of humans in an imagined distant, star-faring future, is well worth the sacrifices of a few billions of humans today.]).
 
-One does not need to posit a conspiracy theory^[Accusing the critics of being conspiracy theorists, as @sennesh_conspiracy_2023 do, distracts from the fact that there is no dearth of leftist ideas for a more just future; intellectual poverty is not what is at stake, instead a key difference is that progressive and humanistic ideas seldom have the power of money behind them.
-This is in contrast to, say, effective altruism [@matthews_how_2022].] to worry about the influence (whether coordinated or not) of a small group of men (they are almost all of them men^[After Meta disbanded its Responsible AI Division, the company formed an AI advisory council composed of four very wealthy white men [@davis_metas_2024;@goel_meta_2024].]) who collectively can muster more than _half a trillion dollars_ as they try to turn their individual convictions into a reality for all [@stross_tech_2023].
-
-But leaving aside the question of a potential "unadmitted political agenda" by an ultra-wealthy elite, there is the much more immediate problem that training data, much like technology, are not politically neutral.
-Several processes have been identified that contribute to reproduce malignant biases, including that of _platformed racism_ [@matamoros-fernandez_platformed_2017].
-Platformed racism results when the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has already had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020].
+One does not need to posit a conspiracy theory to worry about the influence (coordinated or not) of a small group of uber-wealthy people who collectively can muster more than _half a trillion dollars_ as they try to turn their beliefs into a reality for all [@stross_tech_2023].
+So leaving aside the question of a potential "unadmitted political agenda", there is still the immediate problem that the datasets used to train AIs, much like the technology itself, are not politically neutral.
+_Platformed racism_ [@matamoros-fernandez_platformed_2017], for instance, is known to reproduce malignant biases.
+Platformed racism results when the platforms are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has already had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020].
 A U.S.
@@
 51].].
-Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs^[This is part of the Unfathomable Data problem discussed by @bender_dangers_2021.
-In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].].
-The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021].
-This is not a bug of the system, but rather a feature, as Meta's supression of information about Gaza in 2024 shows [@paul_former_2024].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].
+Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has already been passed to their AIs^[This is part of the Unfathomable Data problem discussed by @bender_dangers_2021.].
+The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].
 
-```{r google-obama, out.width="90%", fig.cap="\\label{fig:google-ai-overview-of-obama}Google's AI Overview of Muslim US presidents"}
-google_plot <- ggplot() +
-  labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
-  theme_void() +
-  theme(text = element_text(size = 8))
+Search engines already tended to reinforce racism [@noble_algorithms_2018]; given the above, no one should expect AI-augmented search engines to do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
+In addition, AI is also a tool ripe for abuse.
+Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].].
+In another wretched incident, Google's AI Overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that follows on a years-long Russian-assisted right-wing propaganda campaign to otherize Mr.
+Obama [@tenbarge_glue_2024].
 
-ggdraw() + 
-  draw_plot(google_plot) +
-  draw_image(image = glue::glue(here::here(), "/images/google-obama.png"))
-```
-
-\newpage
-
-Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
-In addition to being misleading to the unadvised, AI is also a fantastic tool ripe for abuse by those who are advised _and_ have an agenda.
-Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].].
-In another wretched incident, Google's AI Overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Mr.
-Obama [@tenbarge_glue_2024].
 Mishaps like this, and others, are dismissed by Google as attempts "to trip up the technology with uncommon questions" [@tenbarge_glue_2024].
-This dismissal, arrogant as it is, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do).
+This dismissal only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do).
 Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring.
-
-Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China.
-These authors, whose one true expertise is access journalism, never stop to consider why an authoritarian regime might want to have powerful pattern recognition machines.
-To compound things, they neglect to disclose the stake that Axios, the website they founded, has on AI.
-VandeHei has not been timid to express himself otherwise, and is on record stating that AI "will eviscerate the weak, the ordinary, the unprepared in media" [@robertson_axios_2024].
-Who is concerned about maintaining an edge?]?
+Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes?
 
@@
 
-<!-- AI is a way to cheapen university education; it compromises the ideals of quality, integrity, and excellence. -->
-
 Neural networks and machine learning are nothing new, and they have their uses.
 But those uses have been suffocated, and to some extent discredited, by the current hype around Generative AI.
-This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools that have been more than willing to believe in magic beans [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.].
+This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools that have been more than willing to believe in magic beans [@burneko_if_2024].
 There are now calls to press pause on the hype [@angwin_press_2024] but these voices are barely audible above the thundering echos inside the AI hype chamber.
 
-Universities and colleges would be wise to heed the advice to press the pause button on the AI hype, as there is a real risk of them falling prey to a confidence trick.
+Universities and colleges would be wise to heed the advice to press the pause button on the AI hype to avoid falling prey to a confidence trick.
 Higher education is often driven by a deeply ingrained belief in innovation (even, or perhaps particularly, for its own sake), and this impulse is compounded by the way AI adoption has been drawn as a false dilemma, thus triggering a fear of missing out.
@@
 
-In this essay I have tried to describe the many problems with AI as an _everything tool_.
-Tall in my mind are the moral and ethical concerns around an extractive and exploitative technology that promises democracy but delivers further concentration of wealth and power.
-GenAI is environmentally damaging for what it delivers: confidently stated confabulations.
-I have little illusion that these arguments will prove persuasive, since the university is already aware of them because they usually appear in the last slides of every presentation about--and yet here we are, asking not whether, but where it makes sense to adopt AI.
+In this essay I have tried to describe the numerous issues with AI as an _everything tool_.
+
+I hope to have conveyed the sense that AI as marketed by the likes of OpenAI and Microsoft, cannot be used in a way that is consistent with values like integrity, respect for human rights, and environmental sustainability.
+This is well-known: most discussions of AI in education acknowledge that their use is rife with problems.
+For instance, @mollick_instructors_2024 note issues with copyright, exploitation of low-wage workers, reproduction of biases, the risk of automation bias, and privacy loss.
+These are issues that "educators may want to consider before deciding whether to use these systems".
+This disclaimer, consuming fully one third of a page in a 76-page document, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual educators working under desperately under-resourced conditions, or simply willing to win a race to the bottom.
+And yet, I have little illusion that these arguments will prove persuasive.
 
@@
 
-It has now been 55 years since Howard S.
-Barrows and Robyn M.
-Tamblyn pioneered [problem-based learning](https://en.wikipedia.org/wiki/Problem-based_learning#History) at McMaster, an accomplishment that the university [still touts today](https://provost.mcmaster.ca/teaching-learning/).
-Problem-based learning was not and is still not high tech, and yet remains a remarkable pedagogical breakthrough, an approach that has stayed influential for decades.
-Developing problem-based learning took human ingenuity and insight.
-What the university is considering today is which of our core missions can we hitch to an expensive technology of questionable value today, of dubious longevity, and that in the here and now advances an agenda in education that can only be squared with "integrity, quality, and excellence" via fabulous contortions of the imagination.
-
-Does the university imagine that it will be best known in half a century as "McMaster, the institution that pioneered the use of ChatGPT"?
-Is that what it really wants?
-All that AIs offer, after all, is a very sophisticated [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean), which not by coincidence is also called a "reversion to mediocrity" [@barnett_regression_2017].
-We would do our students an unforgivable disservice by offering a skill that reversed them to mediocrity.
+We would do our students an unforgivable disservice by offering them a skill that reversed them to mediocrity and left them at the mercy of predatory corporations.
 "This university uses ChatGPT to teach" is simply a way of saying that we are too cheap to offer a human-centered education.
-But in truth, a university is too expensive an institution to be attractive under such conditions, when there are far cheaper alternatives for an AI-based education ("for only $9.99 proof yourself against layoffs").
-
-```{r ai-bundle, out.width="110%", fig.cap="\\label{fig:trail-blazing-bundle}A trail-blazing bundle: ChatGPT for data science and data analysis in Python"}
-ggdraw() + 
-  draw_image(image = glue::glue(here::here(), "/images/trail-blazing-ai-bundle.png", width = 0.8))
-```
+But in truth, a university is too expensive an institution to be attractive under such conditions, when there are far less expensive alternatives for an AI-based education ("for only $9.99 proof yourself against layoffs").
 
 Going down this path would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors, a university's main source of authority is its reputation as a place that preserves and expands knowledge in a principled way.
-
-I suggest that universities would do well to avoid the mistakes of other sectors, including those in the media that have contributed to hype AI.
-The media is an industry that has innovated itself [out of the trust of the public](https://reutersinstitute.politics.ox.ac.uk/news/julia-angwin-fears-public-sphere-about-get-worse-ai-makes-it-easier-flood-zone-misinformation) [also see the "pivot-to-video" and how tech companies suckered many media companies into giving them money in exchange for fake engagement statistics [@hazard_owen_facebooks_2021].].
-Tech corporations (or at least some of them), can still hope to recoup some of the fabulous ammounts they have poured into AI.
-But a university gambles its most important asset when betting on AI: its reputation as a place were knowledge is created, and as a custodian of knowledge.
-AIs already have a reputation for producing mis- and disinformation--from the very titans of the field^[Entities like Google and Amazon could not care less about their reputation.
-They are [too big](https://www.businessinsider.com/ex-amazon-ghaderi-exec-suing-ai-race-copyright-allegations-2024) [to care](https://pluralistic.net/2024/04/04/teach-me-how-to-shruggie/).] to the unaware^[Like a B.C.
-lawyer who was [reprimanded](https://www.cbc.ca/news/canada/british-columbia/lawyer-chatgpt-fake-precedent-1.7126393) for using AI hallucinations in a legal filing.
-This was not even the first time that supposedly highly skilled professionals were castigated for [using](https://www.washingtonpost.com/technology/2023/11/16/chatgpt-lawyer-fired-ai/) [chatbots](https://www.reuters.com/legal/new-york-lawyers-sanctioned-using-fake-chatgpt-cases-legal-brief-2023-06-22/).], and beyond, to bad-faith actors^[Like rival geopolitical powers that are relatively impervious to propaganda themselves for reason of being [authoritarian countries](https://www.theatlantic.com/magazine/archive/2024/06/china-russia-republican-party-relations/678271/).].
-
-In my view, there are no efficiencies, no questionable promises of future excellence, that can compensate the one thing that makes a university stand tall among other institutions--its pursuit of innovation, yes, but above all of the truth, because what else is worth discovering, preserving, and communicating?
+Give the principle away, what is left to sustain universities as institutions that add value to society?
 

### 2024-07-02 paper/title-page.qmd
Paper submitted to Computers & Education

---
title: "An AI is Haunting Campus: A Call for Human-Centered Higher Education"
author: Antonio Páez
format: docx
---

- orcid: 0000-0001-6912-9919
- email: paezha@mcmaster.ca
- department: School of Earth, Environment and Society
- university: McMaster University
- address: 1280 Main St W, Hamilton, ON, L8S 4K1 Canada

**Abstract**

AIs have been deliberately presented as everything tools that can do a fabulous array of tasks in higher education. This presentation is misleading, and there is a need to examine the claims about what AIs can do and what they actually do in the context of teaching and learning.  In this paper I show how discussions about AI hinge on making a large number of unverified claims (Gish Gallop) while presenting the technology as inevitable (a False Dilemma). However, once we examine what the technology does, and particularly what it does to whom and for whom, widespread adoption of AI in teaching and learning is inimical to the principles that most institutions of higher education claim to hold dear.

### 2024-08-10 README.Rmd
Small edits to text and compedu rendered version of submitted paper

@@
 On November 30, 2022, OpenAI released what became its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language.
-The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels: by early December, ChatGPT was being panned as "the best artificial intelligence chatbot ever released to the general public" [@roose_brilliance_2023], and a tool able to generate "impressively detailed" and "human-like" text, leading some to wonder whether it could "replace humans" [@lock_what_2022].
+The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023].
+It took less than two weeks for the hype to reach stratospheric levels, and by early December it had opinion-makers raving: not only was ChatGPT "the best artificial intelligence chatbot ever released to the general public" [@roose_brilliance_2023], some wondered whether it could even "replace humans" with its "impressively detailed" and "human-like" text "replace humans" [@lock_what_2022].
 

### 2024-08-10 README.Rmd
Fixed copy-edit error

@@
 The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023].
-It took less than two weeks for the hype to reach stratospheric levels, and by early December it had opinion-makers raving: not only was ChatGPT "the best artificial intelligence chatbot ever released to the general public" [@roose_brilliance_2023], some wondered whether it could even "replace humans" with its "impressively detailed" and "human-like" text "replace humans" [@lock_what_2022].
+It took less than two weeks for the hype to reach stratospheric levels, and by early December it had opinion-makers raving: not only was ChatGPT "the best artificial intelligence chatbot ever released to the general public" [@roose_brilliance_2023], some even wondered whether it could "replace humans" with its "impressively detailed" and "human-like" text [@lock_what_2022].
 

### 2024-10-28 paper/paper.qmd
Submission to Higher Education

@@
 ```
+
+## Abstract
+
+The use of Generative Artificial Intelligence (AI) in higher education has been a sensational topic since the launch of ChatGPT and similar tools.
+AI has been presented as an everything tool that can do a fabulous array of tasks, from creating content to grading to providing feedback to students.
+This presentation of AI is misleading, and there is a need to examine the claims about what the technology can do and what it actually does in the context of teaching and learning.
+In this paper I scan the academic and gray literature, along with media reports regarding AI, to show how discussions about AI hinge on making a large number of unverified claims (Gish Gallop) while presenting the technology as inevitable (a False Dilemma).
+However, once we examine what AI does, and particularly what it does for whom and to whom, a picture emerges of a technology that is intrinsically inimical to the principles that most institutions of higher education claim to hold dear, including the best interest of their students.
 

### 2024-11-02 paper/paper.qmd
Submission to Critical Studies in Education

@@
 ---
-title: "An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: A Call for Human-Centered Higher Education"
+title: "An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: What Role for the Human in Higher Education?"
 format: docx

### 2025-01-24 paper/paper.qmd
Revisions for possible submission to Critical AI

@@
 This presentation of AI is misleading, and there is a need to examine the claims about what the technology can do and what it actually does in the context of teaching and learning.
-In this paper I scan the academic and gray literature, along with media reports regarding AI, to show how discussions about AI hinge on making a large number of unverified claims (Gish Gallop) while presenting the technology as inevitable (a False Dilemma).
+In this essay I scan the academic and gray literature, along with media reports regarding AI, to show how discussions about AI hinge on making a large number of unverified claims (Gish Gallop) while presenting the technology as inevitable (a False Dilemma).
 However, once we examine what AI does, and particularly what it does for whom and to whom, a picture emerges of a technology that is intrinsically inimical to the principles that most institutions of higher education claim to hold dear, including the best interest of their students.
@@
 The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels [@roose_brilliance_2023; @lock_what_2022].
-Soon after, seemingly everyone wanted in on the Generative AI (GenAI) craze^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent cycle of hype [see @siegel_ai_2023].], and no surprise, given how the technology was hailed as nothing short of revolutionary.
+Soon after, seemingly everyone wanted in on the Generative AI (GenAI) craze^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent cycle of hype [see @siegel_ai_2023].].
 
@@
 
-AI was quickly painted as an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, and that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p.
+AI was quickly painted as nothing short of revolutionary: an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, and that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p.
 64].
@@
 For perspective, hammers are powerful tools, and so are thousands of other instruments found at universities.
-And yet not every single one commands a high-level committee to advice on its use.
+And yet not every single one commands a high-level university committee to advice on their use.
 Just what is it about AIs that requires pan-university consultations on how to use them?
@@
 
-This essay aims to provide some clarification around 1) the things that GenAIs claim to do in education; and 2) what the technology actually does; within 3) the context of a principled organization.
+This essay aims to provide some clarification around 1) claims about the potential of GenAIs in education; and 2) what GenAIs actually do--within 3) the context of a principled organization.
 I conclude that the promise of generative AI has been overstated, that is is unlikely to deliver, and that in any case the technology, as sold by the tech industry, is antithetical to many of the values that universities claim to hold dear.
@@
 
-This overarching goal is achieved through "the discovery, communication and preservation of knowledge" [@office_of_the_president_mcmaster_university_mission_2024]; a dedication "to fostering an academic community in which the learning and scholarship of every member may flourish" [@the_office_of_the_governing_council_secretariat_institutional_1992]; by "[encouraging] and [celebrating] excellence and innovation in teaching, learning, research and student development" [@trent_university_vision_2024]; with a determination "to put the learner at the centre of all [the university] does" [@university_of_guelph_mission_2024].
+This goal is achieved through "the discovery, communication and preservation of knowledge" [@office_of_the_president_mcmaster_university_mission_2024]; a dedication "to fostering an academic community in which the learning and scholarship of every member may flourish" [@the_office_of_the_governing_council_secretariat_institutional_1992]; by "[encouraging] and [celebrating] excellence and innovation in teaching, learning, research and student development" [@trent_university_vision_2024]; with a determination "to put the learner at the centre of all [the university] does" [@university_of_guelph_mission_2024].
 
-Unprincipled organizations do not need to question what the adoption of a technology does to and for whom--as long as it advances whatever its goal is at any particular moment.
+Unprincipled organizations do not need to question their choices of technology--as long as the technology advances whatever its goal is at any particular moment.
 But universities aim to achieve their mission while adhering to "integrity, quality, inclusiveness and teamwork" [@office_of_the_president_mcmaster_university_mission_2024]; with "vigilant protection for individual human rights, and a resolute commitment to the principles of equal opportunity, equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992]; and  as they "foster sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].
@@
 But those uses have been suffocated, and to some extent discredited, by the current hype around Generative AI.
-This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools that have been more than willing to believe in magic beans [@burneko_if_2024].
+This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools willing to believe in magic mirrors [@burneko_if_2024].
 There are now calls to press pause on the hype [@angwin_press_2024] but these voices are barely audible above the thundering echos inside the AI hype chamber.
@@
 
-I hope to have conveyed the sense that AI as marketed by the likes of OpenAI and Microsoft, cannot be used in a way that is consistent with values like integrity, respect for human rights, and environmental sustainability.
+AI, as marketed by the likes of OpenAI, Googla, and Microsoft, cannot be used in a way that is consistent with values like integrity, respect for human rights, and environmental sustainability.
 This is well-known: most discussions of AI in education acknowledge that their use is rife with problems.
@@
 These are issues that "educators may want to consider before deciding whether to use these systems".
-This disclaimer, consuming fully one third of a page in a 76-page document, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual educators working under desperately under-resourced conditions, or simply willing to win a race to the bottom.
+This disclaimer highlights the important role of institutions to provide the moral and ethical clarity that may elude individual educators working under desperately under-resourced conditions, or simply willing to win a race to the bottom.
 And yet, I have little illusion that these arguments will prove persuasive.
@@
 I would still appeal, though, to the university's self-regard and sense of self-preservation. 
-
 We would do our students an unforgivable disservice by offering them a skill that reversed them to mediocrity and left them at the mercy of predatory corporations.
@@
 
-Going down this path would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors, a university's main source of authority is its reputation as a place that preserves and expands knowledge in a principled way.
-Give the principle away, what is left to sustain universities as institutions that add value to society?
+Going down this path would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors.
+A university’s main source of authority is its reputation as a place that preserves and expands knowledge in a principled way.
+If we give this principle away, what would be left to sustain universities as institutions that add value to society?
 

### 2025-01-25 paper/paper.qmd
Revisions for submission to Critical AI

@@
 The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels [@roose_brilliance_2023; @lock_what_2022].
-Soon after, seemingly everyone wanted in on the Generative AI (GenAI) craze^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent cycle of hype [see @siegel_ai_2023].].
+Soon after, everyone wanted in on the Generative AI (GenAI) craze^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots leading the parade, is just the most recent cycle of hype [see @siegel_ai_2023].].
 
-According to some, AI technologies in education "offer new tools and applications that have the potential to transform traditional teaching and learning methods" [@adiguzel_revolutionizing_2023, p.
+AI technologies in education "offer new tools and applications that have the potential to transform traditional teaching and learning methods" [@adiguzel_revolutionizing_2023, p.
 1], and have "the potential to fundamentally change the classroom experience and the knowledge and skills outcomes of our students" [@dwivedi_opinion_2023, p.20].
@@
 
-AI was quickly painted as nothing short of revolutionary: an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, and that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p.
-64].
-Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.].
-Institutions of higher education were compelled to get in on the act, forming high-level advisory committees and drafting guidelines to guide the university community leverage AI in principled ways [e.g., @office_of_teaching_and_learning_provisional_2023; @mcmaster_university_provisional_2024; @trent_university_generative_2024; @university_of_toronto_ai_2024].
+AI is painted as revolutionary: an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, while leading to _efficiencies_^["Efficiencies" being an euphemism for cutting costs [@oconnell_let_2012, p.
+64], usually by finding "redundancies" rather than better things for people to do.].
+Higher education institutions were compelled to get in on the act, forming high-level advisory committees and drafting guidelines for university communities to leverage AI [e.g., @office_of_teaching_and_learning_provisional_2023; @mcmaster_university_provisional_2024; @trent_university_generative_2024; @university_of_toronto_ai_2024].
 
-For perspective, hammers are powerful tools, and so are thousands of other instruments found at universities.
-And yet not every single one commands a high-level university committee to advice on their use.
+But hammers are powerful tools, and so are thousands of other instruments found at universities.
+And yet not all of them get high-level university committees to advice on their use.
 Just what is it about AIs that requires pan-university consultations on how to use them?
-To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, examine what it actually does.
+To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, also examine what it actually does.
 
-In doing this, we must remain alert to the fact that technologies are never politically neutral, and two questions must be asked when considering their adoption:
+In doing this, we must remain alert to the fact that technologies are never politically neutral, and two questions must be asked:
 
@@
 
-This essay aims to provide some clarification around 1) claims about the potential of GenAIs in education; and 2) what GenAIs actually do--within 3) the context of a principled organization.
-I conclude that the promise of generative AI has been overstated, that is is unlikely to deliver, and that in any case the technology, as sold by the tech industry, is antithetical to many of the values that universities claim to hold dear.
+This essay examines claims about the uses of GenAIs in education, a sober assessment of what they actually do, while keeping in mind the mission of nominally principled organizations.
+I conclude that promises about GenAI are overstated and unlikely to be met, and that current AIs are antithetical to the values that universities claim to hold dear.
 
-# The university's mission {#the-mission}
+# The mission {#the-mission}
 
-Universities present themselves as institutions that serve "the social, cultural, and economic needs of [their] community and [their] society" [@office_of_the_president_mcmaster_university_mission_2024]; that strive "to make valued and socially responsible contributions to our local communities, to Canada, and to the world" [@trent_university_vision_2024]; and that "serve society and...enhance the quality of life through scholarship" [@university_of_guelph_mission_2024].
+Universities pride themselves of serving "the social, cultural, and economic needs of [their] community and [their] society" [@office_of_the_president_mcmaster_university_mission_2024]; of striving "to make valued and socially responsible contributions to our local communities, to Canada, and to the world" [@trent_university_vision_2024]; and of "[serving] society and...[enhancing] the quality of life through scholarship" [@university_of_guelph_mission_2024].
 
-This goal is achieved through "the discovery, communication and preservation of knowledge" [@office_of_the_president_mcmaster_university_mission_2024]; a dedication "to fostering an academic community in which the learning and scholarship of every member may flourish" [@the_office_of_the_governing_council_secretariat_institutional_1992]; by "[encouraging] and [celebrating] excellence and innovation in teaching, learning, research and student development" [@trent_university_vision_2024]; with a determination "to put the learner at the centre of all [the university] does" [@university_of_guelph_mission_2024].
+These goals are achieved through "the discovery, communication and preservation of knowledge" [@office_of_the_president_mcmaster_university_mission_2024]; a dedication "to fostering an academic community in which the learning and scholarship of every member may flourish" [@the_office_of_the_governing_council_secretariat_institutional_1992]; by "[encouraging] and [celebrating] excellence and innovation in teaching, learning, research and student development" [@trent_university_vision_2024]; and with a determination "to put the learner at the centre of all [the university] does" [@university_of_guelph_mission_2024].
 
-Unprincipled organizations do not need to question their choices of technology--as long as the technology advances whatever its goal is at any particular moment.
+Unprincipled organizations need not question their choices of technology--as long as the technology advances whatever their goal might be at any moment.
 But universities aim to achieve their mission while adhering to "integrity, quality, inclusiveness and teamwork" [@office_of_the_president_mcmaster_university_mission_2024]; with "vigilant protection for individual human rights, and a resolute commitment to the principles of equal opportunity, equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992]; and  as they "foster sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].
@@
 
-A reason why many organizations have spent substantial resources considering the adoption of AIs is that they have been presented, quite deliberately, as _everything tools_: tools capable of doing many things a human can do, but faster and supposedly cheaper.
-While hammers are understood to have a limited range of uses that mostly have to do with blunt force, the release of GenAIs to the masses was accompanied by innumerable promises--quite a few of them made by entities and individuals that had a stake in the adoption of the technology^[Ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
+One reason many organizations have spent substantial resources on AIs is that they have been presented as _everything tools_: a technology capable of doing many things humans do, but faster and supposedly cheaper.
+While most tools are understood to have a limited range of uses, the release of GenAIs was accompanied by innumerable promises--most of them made by entities and individuals with a stake in the adoption of the technology^[Ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].
 
-The public was promised that GenAI would "accelerate sustainability" [@nakagawa_accelerating_2023].
-
-Also, that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" [@ibm_what_2023].
-AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024].
-In mental health care, chatbots could complement human therapists by using patient data to analyze behavioral patterns to then mimic practitioner questions [@silva_4_2023].
-AIs would revolutionize dating by helping "people connect and find potential partners" [@khalatian_matchmaking_2023].
-
-The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves to micromanaging the [bowel movements of our pets](https://finance.yahoo.com/news/next-generation-ai-powered-cat-140000795.html).
-
-```{r ai-litter-box, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}Revolutionizing pet health"}
-#| label: fig-revolutionizing-pet-care
-
-knitr::include_graphics(glue::glue(here::here(), "/images/ai-litter-box.png"))
-```
+The list of promises have grown to encompass pretty much everything, from saving the planet [by "accelerating sustainability"; @nakagawa_accelerating_2023] to "doubling human lifespans" within the decade [@novak_anthropic_2025].
+Transforming health care, finance, human resources, insurance, "and beyond" [@ibm_what_2023] pales besides such momentous accomplishments.
 
@@
 
-@power_3_2024, presents three innovative uses for AI in higher education.
-First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably from working at the university].
-AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants.
-AI could also personalize the experience of consumers by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.
+For example, @power_3_2024, presents three "innovative" uses for AI in higher education.
+First, handing over manual tasks to AIs could improve the budget by "releasing" administrative staff.
+AI could assess traits like the "grit" and "empathy" in admission applications.
+AI could also personalize the experience of consumers [i.e., students] by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.
 
-In another instance, @hie_how_2023 argue that AIs will likely be the primary way to access information in the future.
-For this reason, institutions must equip students with the skills like _prompt engineering_, the ability to craft effective queries that prompt a GenAI to provide the most useful outputs.
-Universities will use AIs to improve learning and challenge the critical thinking of students.
-Faculty will also benefit from AI: it will be used to evaluate their students, and will handle repetitive tasks like writing course syllabi.
-
-@hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr.
+@hie_how_2023 argue that AIs will be the primary way to access information in the future.
+For this reason, institutions must equip students with skills like _prompt engineering_, the ability to craft effective queries for GenAIs.
+Universities will use AIs to improve learning and challenge students' critical thinking.
+Faculty will also benefit from AI by automating student evaluation and writing course syllabi.
+Moreover, @hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr.
 Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9).
 This prediction aged poorly: on February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it.
-Shortly after that, Meta [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.
+Shortly after that, Meta [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they should at least try to shape it as best they can.
 
@@
 
-Is it really possible that AIs can do all this?
+Is it really possible for AIs to do all this?
 
@@
 
-Discussions of AI as an _everything tool_ quickly devolve into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength.
-To be sure, AIs can do _some_ things well, but critics are forced to respond to every single propounded use with evidence, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself.
-Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they really do?
+Presenting AI as an _everything tool_ is purposeful.
+This device makes debates around the technology quickly devolve into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical style where a party makes an overwhelming number of arguments with little regard to their accuracy or strength.
+To be sure, AIs can do _some_ things well, but critics are forced to counter every suspect promise with evidence, a more time consuming task than making unsubstantiated and/or misleading claims.
+Proponents of AI can always admit that AIs may not actually do _that one thing_, but what about the myriad of other things that people say they really do?
 A second discursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.
 
-In what follows, it is not my intention to confront every single possible use of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop.
-As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the compromises demanded by AI.
+It is not my intention to confront every single possible use of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop.
+As well, I reject the false dilemma, and remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone submitting to the compromises demanded by AI.
 
@@
 GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning _many_ hidden layers.
-As an example, Large Language Models are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].
+Large Language Models are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].
 
@@
 
-A regression model takes data as inputs, typically a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable).
-With these inputs, the model finds a conditional mean that becomes the output of the model, or $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables).
-The process of finding that conditional mean is to satisfy some criterion (e.g., that the conditional mean minimizes the error of the model--the distance between $y$ and $\hat{y}$).
+A regression model uses as inputs a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable) to find a conditional mean that becomes the output of the model, or $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables).
+The process of finding that conditional mean is to satisfy some criterion (e.g., that the conditional mean minimizes the error of the model).
 
-@fig-bivariate-model is an example of the simplest regression model with only one independent variable (i.e., a bivariate model).
+@fig-bivariate-model is an example of the simplest regression model with only one independent variable.
 The plot shows the pairs of $y$ and $x$ values, and the line is the model, i.e., the conditional mean $\hat{y}$.
-This model needs only two parameters: an intercept (the conditianl mean $\hat{y}$ when $x = 0$), and the slope of the line, in other words, the rate of change of $\hat{y}$ with respect to $x$.
+This model needs only two parameters: an intercept (the conditional mean $\hat{y}$ when $x = 0$), and the slope of the line (the rate of change of $\hat{y}$ with respect to $x$).
 Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.
@@
 
-Neural networks do something similar but using a more flexible net-like architecture.
-For instance, @fig-nn1 shows a relatively simple neural network trained using using the same data.
-This network is not "deep", as it has only one "hidden layer" comprised of two neurons between the input and the output layers.
-In comparison to the linear regression and its two parameters, this neural network uses seven parameters.
-Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the results of the model contingent on the starting conditions of the training process.
+Neural networks do something similar but using a net-like architecture.
+For instance, @fig-nn1 shows a relatively simple neural network trained using using the same data as above.
+This network is not "deep", as it has only one "hidden layer" (the two neurons between the input and the output layers).
+In comparison to the linear regression and its two parameters, this neural network uses seven parameters that need to be pre-loaded with random values, which makes the results of the model contingent on the starting conditions of the training process.
 ```{r}
@@
 
-@fig-data-vs-preds-02 illustrates the predictions of the neural network in @fig-nn1, where we can see that the model is not limited to a linear response.
-This flexibility by the neural network is achieved by being less parsimonious (it uses more parameters).
+Unlike a linear regression model, the predictions of the neural network can be non-linear (@fig-data-vs-preds-02 ).
+This flexibility is achieved by being a less parsimonious model (it uses more parameters).
 
@@
 
-More complex neural networks can be designed. @fig-nn2 shows a network deeper than the previous one, with two hidden layers, each consisting of five neurons. This model uses forty-six parameters. 
+More complex neural networks can be designed. @fig-nn2 shows a deeper network, with two hidden layers of five neurons each. This model uses forty-six parameters. 
 ```{r train-deeper-neural-network}
@@
 
-There is a limit as to how flexible a model can be, and this is true of linear regression and of neural networks.
+There is a limit as to how flexible a regression model can be.
 Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than the number of data points that exist in the dataset used to train the model.
-Roughly, every additional parameter needs one additional datum to train on, and when there are as many parameters as data inputs the model becomes overfit and will predict each value of the independent variable in the training dataset perfectly.
+Roughly, every additional parameter needs one additional datum to train on, and when there are as many parameters as data inputs the model becomes overfit and predicts each value of the independent variable in the training dataset perfectly.
 The model will fail to train if there are more parameters than data points.
@@
 In a neural network trained to work with natural language, the "predictors" are the words that come before and after a certain word.
-Several approaches exist to _embed_ a bag of words, which means using large text corpora to calculate the frequency of co-occurrence of words.
-These frequencies are then converted to numerical vectors that represent the position of a word in multidimensional space.
-These "coordinates" in the space of the vocabulary can then be used to calculate the "distances" between words.
-A recent breakthrough in the field of natural language processing is the development of _transformers_, a network architecture architecture that updates the numeric values of a word depending on the words that come before and after in a prompt.
+Several approaches exist to _embed_ a "bag" of words, which means using massive amounts of text to calculate how frequently words go together.
+These frequencies are then converted to numerical vectors that represent the "position" of a word in multidimensional space, and then the "distances" between words.
 
-Here are some key ideas to keep in mind when thinking about AIs:
+Here are some key concepts to keep in mind when discussing AIs:
 
 - AIs are neural networks, which is to say, regression models.
-- A regression model generates outputs that are regressions to the mean (i.e., statistically probable outputs).
-- AIs are _very_ flexible non-linear regression models.
-- Their flexibility comes from using "deep" architectures, which means many layers with many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters.
+- Regression models generate outputs that are regressions to the mean.
+- AIs are _very_ flexible non-linear regression models thanks to the use of "deep" architectures, which means many layers with many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters.
 OpenAI no longer reports the number of parameters of its newer models, but in the latest version this is estimated to be in the trillions.].
-The large number of parameters makes these models _extremely_ data hungry.
-- AIs are trained using an initial set of random values as parameters. The same training dataset can result in different models due to the randomness in the initial conditions.
+- The large number of parameters makes these models _extremely_ data hungry. 
+- AIs are trained using an initial set of random values. The same training dataset can result in different models due to the randomness in the initial conditions.
 - The flexibility of AIs means that outputs will vary seemingly at random with relatively small changes in the prompt. 
 - Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think).
-The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became more subtle as bigger datasets were used to train AIs.
-The increasingly large datasets essentially expanded the space where the model can interpolate.].
+The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is why so-called hallucinations became more subtle as bigger datasets were used to train AIs.
+The increasingly large datasets essentially expanded the space where models could interpolate.].
+- However, when data around a topic are sparse, the regression model can output exactly the training data, because the correlation is perfect.
 
@@
 
-Some of the things that universities hope to achieve with AI include writing text (e.g.
+Uses of AI in universities include writing text (e.g.
 reference letters, grant applications, job descriptions, report summaries), conducting qualitative and quantitative data analysis, translation, captioning, providing multiple formats for learning and engagement, and for personalized support.
 Anecdotally, students report wanting to use generative AI not to complete assignments for them, but to help enhance learning in their courses by having concepts explained differently, by working on practice questions or by getting immediate feedback on drafted work.
-If we take a student-centered perspective, we might begin by asking about some of the things that AIs purportedly do for students.
+In short, AIs are supposed to help both teachers and students.
 
-## What do GenAIs do for students?{#what-do-ais-do-for-students}
-
-- Students want concepts explained to them differently.
+### Different explanations of concepts
 
@@
 
-- Practice questions
+### Practice questions
 
@@
 
-- Immediate feedback on drafted work
+### Immediate feedback on drafted work
 
@@
 
-- Creation of innovative learning materials
+### Creation of innovative learning materials
 
@@
 
-- Automated grading
+### Automated grading
 
@@
 
-- Writing reference letters
+### Writing reference letters
 
@@
 
-- Qualitative and quantitative data analysis
+### Qualitative and quantitative data analysis
 
@@
 An error like this is easy to spot.
-But we seldom use computers to do calculations that can be done by hand, and checking that an AI-generated response is not wrong is as time consuming as doing the analysis without an AI in the first place [see @yuan_how_2023].
+But we do not use computers to do calculations that can be done by hand, and checking the correctedness of an AI-generated response is as time consuming as doing the analysis without an AI [see @yuan_how_2023].
 
@@
 For the results to be useful, we should calculate the average of a large number of responses.
-Given the cost of training and querying a LLM, this is prohibitively expensive.
+Given the cost of training and querying LLMs, this is prohibitively expensive.
 
 But what about a task that involves logic instead of arithmetic?
-Research by @nezhurina_alice_2024 demonstrates that relatively simple natural language puzzles that are easily solved by humans, often result in confidently expressed errors when used to prompt an AI.
+Research by @nezhurina_alice_2024 demonstrates that relatively simple natural language puzzles that are easily solved by humans often result in confidently expressed errors when used to prompt an AI.
 Wrong answers, furthermore, are frequently backed up by confabulations, that is, nonsense expressed in reasoning-like terms.
-The onnections between words that LLMs deliver are but probabilistic word salads: @xu_hallucination_2024 use a formal approach to show that LLMs are inherently incapable of mapping outputs to a ground truth, and thus their results will only ever resemble a ground truth purely by chance.
+The connections between words that LLMs deliver are but probabilistic word salads: @xu_hallucination_2024 use a formal approach to show that LLMs are inherently incapable of mapping outputs to truths, and thus their results will only ever resemble truth purely by chance.
 The outputs of a chatbot only sound plausible thanks to the vast amount of data points and parameters that the models use.
@@
 
-- Identify early signs of learning difficulties.
+### Identify early signs of learning difficulties.
 
 An AI should not be used to identify signs of learning difficulties for the same reasons that AIs cannot be trusted to identify text written by an AI.
-The false positive and negative rates are out of proportion compared to the benefits the practice might offer.
+The false positive and negative rates are disproportionate compared to the benefits.
 A vigilant instructor will be aware of learning difficulties among their students.
@@
 
-- Personalized learning
+### Personalized learning
 
-Anyone receiving "personalized learning" from an AI can be sure that they will be treated as the machine-generated average of their class, and not as a person.
+Anyone receiving "personalized learning opportunities" from an AI can be sure that they will be treated as the machine-generated average of the dataset used to train the model, and not as a person.
 
-## What do GenAIs do to students? {#what-can-ais-do-to-students}
+## What do GenAIs do to students and teachers? {#what-can-ais-do-to-students}
 
-The preceding section illustrates how numerous tasks that AIs can notionally do in higher education are in fact poor replacements for a human-centered education.
-But what about the things that GenAIs can do _to_ students?
+The preceding section illustrates numerous AI tasks in higher education are in fact poor replacements for a human-centered education. But what about the things that GenAIs can do _to_ students?
 
@@
 
-Reliance on AI can short-circuit the learning process.
-Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] convey the idea that prompt engineering will become an essential skill; perhaps even the skill to end all skills^[Prompt engineering is a pompous name for a task that more closely resembles reading the entrails of a Large Language Model than actual engineering.
-It is a task that cannot possibly be performed in a systematic way, and that offers no guarantees that it will work in the same way again, especially after a model is updated.
-Calling the activity of writing prompts "engineering" is a blatant attempt to steal the competence and reputation of actual engineering.].
+Reliance on AI short-circuits the learning process.
+Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] imply that prompt engineering will become an essential skill, perhaps even the skill to end all skills^[Prompt engineering is a pompous name for a task that more closely resembles reading the entrails of a Large Language Model than actual engineering.
+It is a task that cannot be performed in a systematic way, and that offers no guarantees that it will work in the same way again, between models or versions.
+Calling the activity of writing prompts "engineering" is to steal the competence and reputation of actual engineering.].
 
-It is important to note that, even if prompt "engineering" offered some value, the skill is tied to a singular technology and offers little transferability potential.
+It is important to note that, even if prompt "engineering" offered some value, the skill is tied to a singular technology that offers little transferability potential.
 @mollick_instructors_2024 demonstrate how prompt "engineering" is not even transferable _between_ different LLMs (see p.
@@
 Good skills transfer between contexts.
-Learning how to do an effective bibliographical query transfers between libraries.
-Learning to identify relevant content works while reading a novel, a book of history, or a text of econometrics.
+Learning to do effective bibliographical queries transfers between libraries.
+Learning to identify relevant content works for reading a novel, a book of history, or a text of econometrics.
 Writing, the slow and demanding process of converting thoughts into words, is essential to create a style of communication that is unique to each person.
@@
 
-In a sense, an AI is like a crutch.
-However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle, but one that the user cannot hope to fully master due to its blackbox nature--in the sense that 1) neural networks are blackboxes; and 2) the development of AIs itself is a blackbox wherein things can more easily change based on the interests of their corporate owners than in response to the interests of their relatively small customers^[And this refers not only to individuals whose skills become yoked to ChatGPT and MidJourney.
-As an example, in 2022 McMaster University operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD).
+An AI is like a crutch, but what we are being promised is an All-Terrain Vehicle, however one that the user will never own, and cannot hope to fully master due to its blackbox nature (neural networks are blackboxes and their development is also a blackbox wherein things can more easily change based on the interests of their corporate owners than in response to the interests of their relatively small customers^[And this refers not only to individuals.
+As an example, in 2022 McMaster University, a major Canadian university, operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD).
 Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022.
-To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero dollars and still account for less than _one third of one percent point_ of Microsoft's revenue.
-A mid-sized Canadian university is like an ant scuttling under the economic boot of Microsoft.].
+To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero dollars and still account for less than _one third of one percent point_ of Microsoft's revenue.].
 
@@
 
-Besides a few methodological developments, improvement in the performance of AIs has mostly relied on using bigger datasets.
-This has led to a rush, as developers try to train their models with bigger datasets^[A rush that feels as frenetic as any gold rush in the past.
-However, while the gold rushes of the 19th century were relatively democratic due to their low entry costs, the current data rush has enormous entry costs and is largely dominated by large corporations and venture capital.].
-But, despite the large number of text items and images available in digital form, there is still only a finite amount of data to be had.
+Improvement in the performance of AIs has relied on using bigger datasets.
+This has led to a rush, as developers train their models using bigger datasets^[A rush that feels as frenetic as any gold rush in the past.
+However, while the gold rushes of the 19th century were relatively democratic due to their low entry costs, the current data rush has enormous entry costs and is dominated by large corporations and venture capital.].
+But, despite the large number of text items and images available in digital form, the supply is finite.
 
-As of this writing, the few big players that can afford to grab the data, are mostly done with all the data grabbing, sometimes by morally dubious, and at times by possibly illegal means [@clark_ex-amazon_2024].
-And so, as the fountain of human-generated content has been reduced to a drip by a handful of corporations developing AIs, there has been a turn towards what might be the last untapped drops of information left on the internet.
+As of this writing, the few big players that can afford to grab the data, are mostly done with it, using morally dubious and possibly illegal means [@clark_ex-amazon_2024; @milmo_zuckerberg_2025; @smalley_linkedin_2025].
+And so, as the fountain of human-generated content has been reduced to a drip, there has been a turn towards what might be the last untapped drops of information left on the internet.
 
 OpenAI, a global leader in data grabbing, resorted to data from Stack Overflow [@stack_overflow_stack_2024], a website that [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges".
-Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) (to say the least) with the way Stack Overflow dealt with OpenAI, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not do_.
+Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) with the deal, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not_ and _cannot_ do.
 It is an open question how long Stack Overflow will remain a community-based space after it started treating its members as unwitting, and even unwilling providers of data for OpenAI [@grimm_stack_2024].
@@
 
-Other AI Overview results included a recommendation by geologists to eat at least one small rock a day, a piece of advice tracked back to an article published by the well-known satirical periodical [The Onion](https://www.theonion.com/geologists-recommend-eating-at-least-one-small-rock-per-1846655112).
+Other AI Overview results included a recommendation by geologists to eat at least one small rock a day, a piece of advice tracked back to an article published by the satirical periodical [The Onion](https://www.theonion.com/geologists-recommend-eating-at-least-one-small-rock-per-1846655112).
 
@@
 
-Mind you, these, and numerous other hapless results, are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
+These, and numerous other hapless results, are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
 To what extent can more money improve GenAIs?
@@
 And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data.
-As GenAIs become more widely used, they can produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the most thoroughly vetted form of writing in the world [@cabanac_tortured_2021; @maiberg_ai_2024].
+As GenAIs become more widely used, they can produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the world's most thoroughly vetted form of writing [@cabanac_tortured_2021; @maiberg_ai_2024].
 
 What can we expect as future GenAI models begin to be trained using data that was generated by previous GenAIs?
-Revisiting the examples in a preceding section ([But really, what do AIs do?](#really-what-do-they-do)) we can see that the predictions of a model are more uniform than the original data used for training.
-Technically, the variance of a model's output is substantially reduced compared to the variance of the training data: the outputs of a model will be less varied, less diverse, and more norm-conforming.
+Revisiting the examples in a preceding section ([But really, what do AIs do?](#really-what-do-they-do)) we can see that the predictions of a model are more uniform than the original data used for training: the outputs of a model will be less varied, less diverse, and more norm-conforming.
 The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs.
 We may be witnessing _Peak ChatGPT_, and future model versions, instead of improving, may begin to regress to duller versions of their own previous outputs.
-To compound things, as backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023], including _PhotoGuard_ [@heikkila_this_2023], _Nightshade_ [@shan_nightshade_2024], _Shadowcast_ [@xu_shadowcast_2024].
+To compound things, as the backlash against an exploitative data rush is suuported by new tools to generate data that are deleterious to the training process of a GenAIs [@salman_raising_2023], including _PhotoGuard_ [@heikkila_this_2023], _Nightshade_ [@shan_nightshade_2024], and _Shadowcast_ [@xu_shadowcast_2024].
 
@@
 
-## What do GenAIs do for the mega-corporations pushing them? {#what-do-ais-to-corporations}
+## What do GenAIs do for the mega-corporations pushing them? {#what-do-ais-for-corporations}
 
-AIs do not seem to do much of value for students, as they can be wasteful at best and harmful at worst.
+AIs do not seem to do much of value for students and teachers, as they are wasteful at best and harmful at worst.
 But the technology must do something for someone, otherwise it would not be ubiquitous.
 So what do AIs do for the the organizations that develop them?
-So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities^[Some observers have even wondered whether the AI fever has not actually become detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].].
+So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities^[Some observers believe that the AI fever has become detrimental to actual innovation [@siegel_ai_2023; @vinsel_dont_2023].].
 Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.] [@tricot_venture_2021].
@@
 At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 monthly queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer.
-And a large number of users are not even paying subscribers: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023].
-The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing the most intensive, disturbing, degrading, and damaging aspects of training AIs to the Global South [@nix_analysis_2024].
+And a large number of users do not pay: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023].
+The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing  to the Global South the most disturbing, degrading, and damaging aspects of training AIs [@nix_analysis_2024].
 
-Venture capital is not known for its charitable aims, though, and at some point investors will want their money back with a heap of profit on top [@sriram_ai_2023].
+Venture capital is not known for its charitable nature, though, and at some point investors will want their money back with a heap of profit on top [@sriram_ai_2023].
 Who will pay to make investors whole?
@@
 
-ChatGPT 4.o was announced on May 13, 2024, with great fanfare and a flashy display of the model's new capabilities, which included functions to respond vocally to prompts, and to identify images via a mobile device's camera.
-4.o's voice, in particular, caused considerable titillation [@roose_is_2024].
-The voice was modeled after the digital personal assistant of the movie _Her_, Sam Altman's favorite movie by his own account.
+ChatGPT 4.o was announced on May 13, 2024, with great fanfare and a flashy display of the model's new capabilities, including functions to respond vocally to prompts, and to identify images via a mobile device's camera.
+4.o's voice, in particular, caused considerable titillation [@roose_is_2024], modeled as it was after the digital personal assistant of the movie _Her_ (Sam Altman's favorite movie by his own account).
 The background story of 4.o's voice is more sordid: Altman reportedly tried to hire Scarlett Johannson, who played the AI assistant in the movie, to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when she declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways.
@@
 The idea was basically to train AIs to detect content that was unpalatable.
-Indeed, they had found that feeding an AI with labeled examples of all the worst that exists on the internet (e.g., violence, hate speech, child sexual abuse) could produce tools that learned to detect various manifestations of such toxicity in the wild, to filter it out before it reached users.
-However, to obtain the essential labeled inputs, OpenAI sent tens of thousands of samples to Kenya, with text that described murder, suicide, bestiality, self-harm, incest, torture, and child sexual abuse situations in graphic detail.
-
-What did this do for and to the people doing all the labeling in one of the poorest regions in the world?
-First of all, it provided an income: "data labelers employed...on behalf of OpenAI were paid a take-home wage of between around $1.32 and $2 per hour depending on seniority and performance." But at what cost?
+Indeed, they had found that feeding an AI with labeled examples of all the worst that exists on the internet (e.g., violence, hate speech, child sexual abuse) could produce tools that learned to detect various manifestations of such toxicity, to filter it out before it reached users.
+To obtain the essential labeled inputs, OpenAI sent tens of thousands of samples to Kenya, with text that described murder, suicide, bestiality, self-harm, incest, torture, and child sexual abuse situations in graphic detail.
+A sanitized ChatGPT is only possible through the labor (with wages around $1.32 and $2 per hour) of workers who were exposed to the worst of the internet.
 As one worker tasked with reading and labeling text for OpenAI told TIME, "he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child.
@@
 This means that data cannot be presumed to be private once that they have been used to train an AI.
-And this issue is just a possible result of using overfitted models.
-But what about data leaks?
+And there is besides the issue of data leaks.
 So far, organizations such as the National Archives and Records Administration in the US, the US Congress, and companies like Samsung and Apple have restricted or banned the use of ChatGPT and Copilot for data security reasons [@koebler_national_2024; @singh_after_2024]: no data transferred in the form of a prompt can be presumed to be safe.
@@
 When using a Generative AI we must ask whose perspectives the tool will tend to reproduce and amplify.
-In this sense, numerous observers [e.g., @ahuja_we_2023; @stross_tech_2023; @gebru_tescreal_2024] have noted the ideological leanings of various actors in the AI scene, an elite world of tech investors^[A group that author Charles Stross colorfully [characterizes](https://www.antipope.org/charlie/blog-static/2024/04/the-radiant-future-of-1995.html) as "fascist-adjacent straight white males with an unadmitted political agenda".
-This characterization is somewhat unfair: both [Sam Altman](https://www.advocate.com/news/sam-altman-openai-exclusive) and [Peter Thiel](https://www.npr.org/2016/07/21/486966882/i-am-proud-to-be-gay-tech-investor-peter-thiel-tells-gop-convention) are openly gay.
-On the other hand, their political agenda hides in plain sight.
-@thiel_education_2009 has written about how he "no longer believe[s] that freedom and democracy are compatible".
+In this sense, numerous observers [e.g., @ahuja_we_2023; @stross_tech_2023; @gebru_tescreal_2024] have noted the ideological leanings of various actors in the AI scene, an elite world of tech investors with a political agenda hiding in plain sight^[@thiel_education_2009 has written about how he "no longer believe[s] that freedom and democracy are compatible".
 Sam Bankman-Fried's FTX donated millions to a group with racist ties [@wilson_sam_2024].
-Elon Musk once shared a meme comparing Canada PM Justin Trudeau to Hitler while defending the vandals who terrorized Ottawa in 2022 [@henderson_and_elon_2022].] known for fringe ideologies that combine uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that those who can afford it, are destined to use technology to indefinitely extend life and conquer the stars...but maybe not before super-intelligent AIs destroy humanity.] with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the well-being of trillions of humans in an imagined distant, star-faring future, is well worth the sacrifices of a few billions of humans today.]).
+Elon Musk once shared a meme comparing Canada PM Justin Trudeau to Hitler while defending the vandals who terrorized Ottawa in 2022 [@henderson_and_elon_2022].].
 
-One does not need to posit a conspiracy theory to worry about the influence (coordinated or not) of a small group of uber-wealthy people who collectively can muster more than _half a trillion dollars_ as they try to turn their beliefs into a reality for all [@stross_tech_2023].
-So leaving aside the question of a potential "unadmitted political agenda", there is still the immediate problem that the datasets used to train AIs, much like the technology itself, are not politically neutral.
-_Platformed racism_ [@matamoros-fernandez_platformed_2017], for instance, is known to reproduce malignant biases.
-Platformed racism results when the platforms are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has already had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020].
+Besides concern about the influence of a small group of uber-wealthy people who collectively can muster more than _half a trillion dollars_ to turn their beliefs into reality [@stross_tech_2023], there is still leaves immediate problem that the datasets used to train AIs are, like the technology itself, not politically neutral.
+_Platformed racism_ [@matamoros-fernandez_platformed_2017], for instance, is known to reproduce malignant biases, that can emerge in a grass-roots fashion, but are also assisted by bad-faith actors^[Russian disinformation is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020].
 A U.S.
-Department of State report notes that Russian intelligence is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S.
+Department of State report notes that Russian intelligence exploits "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S.
 politics [@us_department_of_state_gec_2024, p.
@@
 Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has already been passed to their AIs^[This is part of the Unfathomable Data problem discussed by @bender_dangers_2021.].
-The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].
 
@@
 
-Neural networks and machine learning are nothing new, and they have their uses.
-But those uses have been suffocated, and to some extent discredited, by the current hype around Generative AI.
-This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools willing to believe in magic mirrors [@burneko_if_2024].
-There are now calls to press pause on the hype [@angwin_press_2024] but these voices are barely audible above the thundering echos inside the AI hype chamber.
+Neural networks are nothing new and they have their uses.
+But those uses have been suffocated, and to some extent discredited, by the current, carefully manufactured hype around Generative AI [@morrone_signals_2024].
+There are now calls to press pause on the hype [e.g., @angwin_press_2024] and universities and colleges would be wise to heed this advice, lest they fall prey to a confidence trick.
+Higher education is often driven by a deeply ingrained belief in innovation (even, or perhaps particularly, for its own sake), and this impulse is compounded by how AI adoption has been drawn as a false dilemma,  triggering a fear of missing out.
+Other pressures at work include a decades-long history of manufactured crises in education [@cizek_give_1999; @usher_state_2023].
 
-Universities and colleges would be wise to heed the advice to press the pause button on the AI hype to avoid falling prey to a confidence trick.
-Higher education is often driven by a deeply ingrained belief in innovation (even, or perhaps particularly, for its own sake), and this impulse is compounded by the way AI adoption has been drawn as a false dilemma, thus triggering a fear of missing out.
-There are other pressures at work, including a decades-long history of manufactured crises in education, in Canada as elsewhere [@cizek_give_1999; @usher_state_2023].
+This essay described the numerous issues with AI as an _everything tool_.
 
-In this essay I have tried to describe the numerous issues with AI as an _everything tool_.
-
-AI, as marketed by the likes of OpenAI, Googla, and Microsoft, cannot be used in a way that is consistent with values like integrity, respect for human rights, and environmental sustainability.
+Current AI technologies cannot be used in a way that is consistent with values like integrity, respect for human rights, and environmental sustainability.
 This is well-known: most discussions of AI in education acknowledge that their use is rife with problems.
-For instance, @mollick_instructors_2024 note issues with copyright, exploitation of low-wage workers, reproduction of biases, the risk of automation bias, and privacy loss.
+For instance, @mollick_instructors_2024 note issues with copyright, exploitation, reproduction of biases, automation bias, and privacy loss.
 These are issues that "educators may want to consider before deciding whether to use these systems".
@@
 
-I would still appeal, though, to the university's self-regard and sense of self-preservation. 
-We would do our students an unforgivable disservice by offering them a skill that reversed them to mediocrity and left them at the mercy of predatory corporations.
+I would still appeal, though, to the university's sense of self-regard and self-preservation. 
+We would do our students an unforgivable disservice by offering them a skill that regressed them to mediocrity and left them at the mercy of predatory corporations.
 "This university uses ChatGPT to teach" is simply a way of saying that we are too cheap to offer a human-centered education.
 But in truth, a university is too expensive an institution to be attractive under such conditions, when there are far less expensive alternatives for an AI-based education ("for only $9.99 proof yourself against layoffs").
-
-Going down this path would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors.
+Following the hype would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors.
 A university’s main source of authority is its reputation as a place that preserves and expands knowledge in a principled way.

### 2025-01-26 paper/essay.qmd
Converted the paper into an essay for submission to Critical AI

--- 2025-01-25 paper/paper.qmd
@@
 ---
-title: "An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: What Role for the Human in Higher Education?"
+title: "An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]"
 format: docx
@@
 
-## Abstract
-
-The use of Generative Artificial Intelligence (AI) in higher education has been a sensational topic since the launch of ChatGPT and similar tools.
+<!--
+ABSTRACT
+The use of Generative Artificial Intelligence (AI) in higher education has become a sensational topic.
 AI has been presented as an everything tool that can do a fabulous array of tasks, from creating content to grading to providing feedback to students.
@@
 However, once we examine what AI does, and particularly what it does for whom and to whom, a picture emerges of a technology that is intrinsically inimical to the principles that most institutions of higher education claim to hold dear, including the best interest of their students.
+-->
 
@@
 
-AI technologies in education "offer new tools and applications that have the potential to transform traditional teaching and learning methods" [@adiguzel_revolutionizing_2023, p.
-1], and have "the potential to fundamentally change the classroom experience and the knowledge and skills outcomes of our students" [@dwivedi_opinion_2023, p.20].
-Furthermore, the technology "is here to stay...[t]herefore, teachers and students must develop the specific digital competencies needed to use such tools in ways that are pedagogically beneficial and ethical" [@kohnke_chatgpt_2023, p.
-546].
-
-AI is painted as revolutionary: an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, while leading to _efficiencies_^["Efficiencies" being an euphemism for cutting costs [@oconnell_let_2012, p.
+AI is painted as revolutionary in education: an _innovative_ and _exciting_ technology, that moreover can lead to juicy _efficiencies_^["Efficiencies" being an euphemism for cutting costs [@oconnell_let_2012, p.
 64], usually by finding "redundancies" rather than better things for people to do.].
@@
 
-But hammers are powerful tools, and so are thousands of other instruments found at universities.
-And yet not all of them get high-level university committees to advice on their use.
-Just what is it about AIs that requires pan-university consultations on how to use them?
+Among the thousands of other tools used in universities, just what is it about AIs that requires pan-university consultations on how to use them?
 To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, also examine what it actually does.
@@
 A regression model uses as inputs a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable) to find a conditional mean that becomes the output of the model, or $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables).
-The process of finding that conditional mean is to satisfy some criterion (e.g., that the conditional mean minimizes the error of the model).
+To regress the inputs to a conditional mean some criterion needs to be adopted, for example, that the conditional mean minimizes the error of the model).
 
-@fig-bivariate-model is an example of the simplest regression model with only one independent variable.
-The plot shows the pairs of $y$ and $x$ values, and the line is the model, i.e., the conditional mean $\hat{y}$.
-This model needs only two parameters: an intercept (the conditional mean $\hat{y}$ when $x = 0$), and the slope of the line (the rate of change of $\hat{y}$ with respect to $x$).
-Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.
-
-```{r simulate-simple-example}
-set.seed(13253)
-
-b0 <- 1.0
-b1 <- 1.5
-
-df <- data.frame(x = b0 + b1 * runif(100, 
-                                0, 
-                                10),
-                 e = rnorm(100),
-                 type = "training data") |>
-  mutate(y = x + e)
-
-x <- df |> pull(x)
-y <- df |> pull(y)
-
-mod <- lm(y ~ x, 
-          data = df)
-```
-
-```{r plot-simulated-data, message=FALSE, out.width="70%", fig.cap="Bivariate Regression Model"}
-#| label: fig-bivariate-model
-
-ggplot(df,
-       aes(x = x,
-           y = y)) +
-  geom_point() +
-  geom_smooth(method = "lm") + 
-  theme_minimal()
-```
-
-```{r data-vs-predictions-comparison, out.width="70%", fig.cap="Training data and model output compared"}
-#| label: fig-data-vs-preds-01
-
-df_pred <- data.frame(x = x)
-
-y_hat <- predict(mod, df_pred)
-
-df_pred_lr <- df_pred |> 
-  mutate(y = y_hat,
-         type = "regression output")
-
-df_pred_all <- rbind(df |>
-                   select(x, y, type),
-                 df_pred_lr |>
-                   select(x, y, type))
-
-ggplot() + 
-  geom_point(data = df_pred_all,
-  aes(x = x,
-      y = y,
-      color = type,
-      shape = type,
-      size = type)) +
-  scale_color_manual(values = c("training data" = "black",
-                                "regression output" = "blue",
-                                "nn1 output" = "chocolate1",
-                                "nn2 output" = "brown")) +
-  scale_shape_manual(values = c("training data" = 16,
-                                "regression output" = 17,
-                                "nn1 output" = 18,
-                                "nn2 output" = 19)) +
-  scale_size_manual(values = c("training data" = 1,
-                                "regression output" = 2,
-                                "nn1 output" = 2,
-                                "nn2 output" = 2)) +
-  theme_minimal() +
-  theme(legend.position = "bottom")
-```
-
-Neural networks do something similar but using a net-like architecture.
-For instance, @fig-nn1 shows a relatively simple neural network trained using using the same data as above.
-This network is not "deep", as it has only one "hidden layer" (the two neurons between the input and the output layers).
-In comparison to the linear regression and its two parameters, this neural network uses seven parameters that need to be pre-loaded with random values, which makes the results of the model contingent on the starting conditions of the training process.
-```{r}
-# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
-f <- as.formula("y ~ x")
-```
-
-```{r train-shallow-neural-network}
-# Train a neural network using only two neurons in one hidden layer.
-The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node.
-The default activation function is the logistic function.
-set.seed(33225433)
-
-nn1 <- neuralnet(f, 
-                data = df, 
-                hidden = c(2), 
-                linear.output = T) 
-```
-
-```{r save-shallow-neural-network}
-# For some reason, the figure cannot be saved directly, so I saved it manually using screen capture
-plot(nn1)
-```
-
-```{r plot-shallow-neural-network, out.width="70%", fig.cap="Simple Example of a Shallow Neural Network"}
-#| label: fig-nn1
-
-knitr::include_graphics(glue::glue(here::here(), "/images/nn1.png"))
-```
-
-```{r shallow-neural-network-predictions}
-y_hat <- predict(nn1,
-                 df_pred)
-
-df_pred_nn1 <- df_pred |> 
-  mutate(y = y_hat,
-         type = "nn1 output")
-
-df_pred_all <- rbind(df_pred_all |>
-                   select(x, y, type),
-                 df_pred_nn1 |>
-                   select(x, y, type))
-```
-
-Unlike a linear regression model, the predictions of the neural network can be non-linear (@fig-data-vs-preds-02 ).
-This flexibility is achieved by being a less parsimonious model (it uses more parameters).
-
-```{r data-vs-predictions-comparison-2, out.width="70%", fig.cap="Training data and model outputs compared: linear regression and shallow neural network"}
-#| label: fig-data-vs-preds-02
-
-ggplot() + 
-  geom_point(data = df_pred_all,
-  aes(x = x,
-      y = y,
-      color = type,
-      shape = type,
-      size = type)) +
-  scale_color_manual(values = c("training data" = "black",
-                                "regression output" = "blue",
-                                "nn1 output" = "chocolate1",
-                                "nn2 output" = "brown")) +
-  scale_shape_manual(values = c("training data" = 16,
-                                "regression output" = 17,
-                                "nn1 output" = 18,
-                                "nn2 output" = 19)) +
-  scale_size_manual(values = c("training data" = 1,
-                                "regression output" = 2,
-                                "nn1 output" = 2,
-                                "nn2 output" = 2)) +
-  theme_minimal() +
-  theme(legend.position = "bottom")
-```
-
-More complex neural networks can be designed. @fig-nn2 shows a deeper network, with two hidden layers of five neurons each. This model uses forty-six parameters. 
-```{r train-deeper-neural-network}
-# Train a neural network using only two neurons in one hidden layer.
-The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node.
-The default activation function is the logistic function.
-set.seed(242)
-nn2 <- neuralnet(f, 
-                data = df, 
-                hidden = c(5, 5), 
-                linear.output = T)
-```
-
-```{r save-plot-deeper-neural-network}
-# For some reason, the figure cannot be saved directly, so I saved it manually using screen capture
-plot(nn2)
-```
-
-```{r plot-deeper-neural-network, out.width="70%", fig.cap="Example of a Deeper Neural Network"}
-#| label: fig-nn2
-
-knitr::include_graphics(glue::glue(here::here(), "/images/nn2.png"))
-```
-
-```{r}
-y_hat <- predict(nn2,
-                 df_pred)
-
-df_pred_nn2 <- df_pred |> 
-  mutate(y = y_hat,
-         type = "nn2 output")
-
-df_pred_all <- rbind(df_pred_all |>
-                   select(x, y, type),
-                 df_pred_nn2 |>
-                   select(x, y, type))
-
-```
-
-The increased complexity leads to greater flexibility, and we can see that the model's predictions resemble more closely the actual data (@fig-data-vs-preds-03).
-
-```{r data-vs-predictions-comparison-3, out.width="70%", fig.cap="Training data and model outputs compared: shallow neural network and deeper neural network"}
-#| label: fig-data-vs-preds-03
-
-ggplot() + 
-  geom_point(data = df_pred_all |>
-               filter(type != "regression output"),
-  aes(x = x,
-      y = y,
-      color = type,
-      shape = type,
-      size = type)) +
-  scale_color_manual(values = c("training data" = "black",
-                                "regression output" = "blue",
-                                "nn1 output" = "chocolate1",
-                                "nn2 output" = "brown")) +
-  scale_shape_manual(values = c("training data" = 16,
-                                "regression output" = 17,
-                                "nn1 output" = 18,
-                                "nn2 output" = 19)) +
-  scale_size_manual(values = c("training data" = 1,
-                                "regression output" = 2,
-                                "nn1 output" = 2,
-                                "nn2 output" = 2)) +
-  theme_minimal() +
-  theme(legend.position = "bottom")
-```
-
-```{r}
-# How do these models perform?
-mad_lr <- data.frame(df |>
-             mutate(y_hat = df_pred_lr$y)) |>
-  summarize(mse = sum(abs(y - y_hat)/n()))
-
-mad_nn1 <- data.frame(df |>
-             mutate(y_hat = df_pred_nn1$y)) |>
-  summarize(mad = sum(abs(y - y_hat)/n()))
-
-mad_nn2 <- data.frame(df |>
-             mutate(y_hat = df_pred_nn2$y)) |>
-  summarize(mad = sum(abs(y - y_hat)/n()))
-
-delta_mad_nn1 <- (mad_nn1 - mad_lr)/mad_lr * 100
-
-delta_mad_nn2 <- (mad_nn2 - mad_lr)/mad_lr * 100
-```
-
-Indeed, the first neural network is "better" than the linear regression in that it reduces the typical error by `r paste0(prettyNum(abs(delta_mad_nn1), digits = 2), "%")`, whereas the deeper network reduces the typical error by `r paste0(prettyNum(abs(delta_mad_nn2), digits = 2), "%")`, also with respect to the linear regression^[The "typical error" here is the mean absolute deviation of the predictions.].
-
-There is a limit as to how flexible a regression model can be.
-Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than the number of data points that exist in the dataset used to train the model.
-Roughly, every additional parameter needs one additional datum to train on, and when there are as many parameters as data inputs the model becomes overfit and predicts each value of the independent variable in the training dataset perfectly.
-The model will fail to train if there are more parameters than data points.
+Neural networks do something similar but using a net-like architecture that allows the model to make non-linear predictions.
+This flexibility is achieved by using more parameters, but has limits: while each additional parameter makes the model more flexible, it requires one additional datum to train on, and when there are as many parameters as data inputs the model becomes overfit and predicts each value of the independent variable in the training dataset perfectly.
 
@@
 
-This is an utterly unrealistic expectation.
-Ideally, students should have the opportunity to discuss their drafted work with instructors before final submission (or could ask members of their study groups to proof-read their work, among many other ways to get feedback before submission).
-A conscientious instructor will invest time reading the drafted work to provide quality feedback.
+A conscientious instructor who is not overextended will invest time reading drafted work to provide quality feedback.
 There is, however, a more insidious aspect about getting feedback from an AI...who, actually, provides the feedback in this case?
@@
 Leaving the technology fetish about classroom applications aside, Mollick and Mollick are well aware of the ethical concerns around using AI (see p.
-6), including that "[m]odels are trained in ways that may violate copyright", that exploit the "efforts of low-wage workers in precarious conditions for part of their training", and the fact that "[m]odels are trained on biased data and can produce either subtly or overtly biased results".
+6), including copyright violations, exploitation of low-wage workers, as well as the use of biased data.
 And yet, they choose to ignore every single one of these concerns for the sake of innovation.
-Machine-washed plagiarism that exploits low-wage workers while using inordinate amounts of energy is innovative only in a way that offends "integrity" [@office_of_the_president_mcmaster_university_mission_2024], "individual human rights...equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992], and "sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].
+Exploitative, world-burning machine-washed plagiarism is innovative only in a way that offends "integrity" [@office_of_the_president_mcmaster_university_mission_2024], "individual human rights...equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992], and "sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].
 
@@
 Alas, no.
-As should be clear by now, any teaching materials generated in this way can only aspire to be average given the prompts used to get the machine going, and no one should buy the delusion that the use of this technology makes them innovators.
+As should be clear by now, AI-generated teaching materials can only aspire to be average given the prompts used, and no one should buy the delusion that the use of this technology makes them innovators.
 In any case, the list of remedial measures required to verify that the materials generated are not nonsense is so long, that it is hard to see what is being gained [see @mollick_instructors_2024, pp.
@@
 But average is not always right.
-And seeing how AIs are now being trained on the dregs of the internet, it should be no surprise when an AI Overview responds to the prompt "2 + 2" with "potato".
+And seeing how AIs are now being trained on the dregs of the internet, it should be no surprise when an AI Overview responds to the prompt "2 + 2" with "potato" (@fig-flickr-potate).
 
 ```{r flickr-potato, out.width="70%", fig.cap="Data analysis using a chatbot? Maybe not such a good idea"}
+#| label: fig-flickr-potato
+
 knitr::include_graphics(glue::glue(here::here(), "/images/potato.png"))
@@
 But we do not use computers to do calculations that can be done by hand, and checking the correctedness of an AI-generated response is as time consuming as doing the analysis without an AI [see @yuan_how_2023].
-
-From a different perspective, even the average output of an AI should be treated with caution.
-As noted [above](#really-what-do-they-do), the results of a neural network depend on the weights selected at random at the start of a training process.
-Change those weights, the results will change.
-As well, even small variations in the prompt can be impactful in models with billions or trillions of parameters.
-In statistical terms, any output is but one _realization_ of a random process.
-For the results to be useful, we should calculate the average of a large number of responses.
-Given the cost of training and querying LLMs, this is prohibitively expensive.
 
@@
 
-It is important to note that, even if prompt "engineering" offered some value, the skill is tied to a singular technology that offers little transferability potential.
-@mollick_instructors_2024 demonstrate how prompt "engineering" is not even transferable _between_ different LLMs (see p.
+It is important to note that, even if prompt "engineering" offered some value, it is a skill that offers little transferability potential.
+@mollick_instructors_2024 demonstrate how prompt "engineering" is not transferable _between_ different LLMs (see p.
 5 and p.
 36).
-Good skills transfer between contexts.
 Learning to do effective bibliographical queries transfers between libraries.
@@
 
-An AI is like a crutch, but what we are being promised is an All-Terrain Vehicle, however one that the user will never own, and cannot hope to fully master due to its blackbox nature (neural networks are blackboxes and their development is also a blackbox wherein things can more easily change based on the interests of their corporate owners than in response to the interests of their relatively small customers^[And this refers not only to individuals.
-As an example, in 2022 McMaster University, a major Canadian university, operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD).
-Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022.
-To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero dollars and still account for less than _one third of one percent point_ of Microsoft's revenue.].
-
-<!--
-McMaster: 842,899,000 CAD
-Microsoft: 198,000,000,000 USD == 257,776,200,000 CAD
-
-https://www.exchange-rates.org/exchange-rate-history/usd-cad-2022
-Avg. exchange rate in 2022: 1.3019 CAD/US
-
-This is how much bigger Microsoft is compared to McMaster. 
-(198000000000 * 1.3019)/842899000
-
-Put another way: If McMaster gave ALL OF ITS REVENUE to Microsoft, it would only account for 842899000 / (198000000000 * 1.3019) or less than one third of a percentage point of Microsoft's revenue.
---->
+AIs are presented as all-terrain vehicles, but are more like a crutch, however one that the user will never own, and cannot hope to fully master due to its black-box nature [@judelson_restaging_2024]
 
@@
 However, while the gold rushes of the 19th century were relatively democratic due to their low entry costs, the current data rush has enormous entry costs and is dominated by large corporations and venture capital.].
-But, despite the large number of text items and images available in digital form, the supply is finite.
+But, despite the large number of text items and images available in digital form, the supply is finite and, like an overexploited river^[As of this writing, the few big players that can afford to grab the data, are mostly done with it, using morally dubious and possibly illegal means [@clark_ex-amazon_2024; @milmo_zuckerberg_2025; @smalley_linkedin_2025].], human-generated content has been reduced to a drip [q.v., @delacroix_sustainable_2024].
+In response to this, there has been a turn towards what might be the last untapped drops of genuine human-generated information left on the internet, often to the disgust of humans there^[OpenAI resorted to data from Stack Overflow [@stack_overflow_stack_2024], a website that [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges".
+Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) with the deal, and has come to believe that the terms were abusive].
 
-As of this writing, the few big players that can afford to grab the data, are mostly done with it, using morally dubious and possibly illegal means [@clark_ex-amazon_2024; @milmo_zuckerberg_2025; @smalley_linkedin_2025].
-And so, as the fountain of human-generated content has been reduced to a drip, there has been a turn towards what might be the last untapped drops of information left on the internet.
-
-OpenAI, a global leader in data grabbing, resorted to data from Stack Overflow [@stack_overflow_stack_2024], a website that [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges".
-Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) with the deal, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not_ and _cannot_ do.
-It is an open question how long Stack Overflow will remain a community-based space after it started treating its members as unwitting, and even unwilling providers of data for OpenAI [@grimm_stack_2024].
-
-Google, for its part, appealed to Reddit for training data to the tune of \$60 million per year [@roth_google_2024].
+Google, for example, appealed to Reddit for training data to the tune of \$60 million per year [@roth_google_2024].
 As Reddit content begins to creep into Google's AI-assisted search, the results of this deal have turned grotesque, with reports of AI Overview failures that include a suggestion to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024].
@@
 ```{r google-reddit, out.width="70%", fig.cap="Google's AI Overview result was traced back to a post by redditor Fucksmith"}
+#| label: fig-google-reddit
+
 knitr::include_graphics(glue::glue(here::here(), "/images/google-cheese-glue.png"))
@@
 
-Other AI Overview results included a recommendation by geologists to eat at least one small rock a day, a piece of advice tracked back to an article published by the satirical periodical [The Onion](https://www.theonion.com/geologists-recommend-eating-at-least-one-small-rock-per-1846655112).
+Other AI Overview results included a recommendation by geologists to eat at least one small rock a day, a piece of advice tracked back to an article published by the satirical periodical [The Onion](https://www.theonion.com/geologists-recommend-eating-at-least-one-small-rock-per-1846655112) (@fig-google-the-onion).
 
 ```{r google-the-onion, out.width="70%", fig.cap="Google's AI Overview result was traced back to an article in The Onion"}
+#| label: fig-google-the-onion
+
 knitr::include_graphics(glue::glue(here::here(), "/images/google-the-rock.jpg"))
@@
 
-These, and numerous other hapless results, are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
+These results are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024].
 To what extent can more money improve GenAIs?
 Probably not by much, since a key constraint is the amount of data available for model training [@villalobos_position_2024].
-And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data.
-As GenAIs become more widely used, they can produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the world's most thoroughly vetted form of writing [@cabanac_tortured_2021; @maiberg_ai_2024].
-
-What can we expect as future GenAI models begin to be trained using data that was generated by previous GenAIs?
-Revisiting the examples in a preceding section ([But really, what do AIs do?](#really-what-do-they-do)) we can see that the predictions of a model are more uniform than the original data used for training: the outputs of a model will be less varied, less diverse, and more norm-conforming.
-The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs.
-We may be witnessing _Peak ChatGPT_, and future model versions, instead of improving, may begin to regress to duller versions of their own previous outputs.
+And dismal as this is, the outlook is even worse: as GenAIs become more widely used, they produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the world's most thoroughly vetted form of writing [@cabanac_tortured_2021; @maiberg_ai_2024].
+The result is predictable: future GenAI models will inevitably be trained using data that was generated by previous GenAIs, leading to outputs that will be less varied, less diverse, and more norm-conforming.
+In other words, we may be witnessing _Peak ChatGPT_, and future model versions, instead of improving, may begin to regress to duller versions of their own previous outputs.
 To compound things, as the backlash against an exploitative data rush is suuported by new tools to generate data that are deleterious to the training process of a GenAIs [@salman_raising_2023], including _PhotoGuard_ [@heikkila_this_2023], _Nightshade_ [@shan_nightshade_2024], and _Shadowcast_ [@xu_shadowcast_2024].
@@
 All in all, the evidence suggests that the scarcity of fresh data, paired with more abundant model-generated and/or poisoned data, will make the performance of these machines stagnate or deteriorate.
-Which begs the following questions.
-How useful will AIs be in the future?
-How persistent will skills like prompt "engineering" be?
-Will learning how to prompt an AI be worth sacrificing the opportunity to develop transferable skills like coding, writing, and thinking?
-Who, after all, stands to benefit the most by the form of deskilling that AIs encourage?
 
-## What do GenAIs do for the mega-corporations pushing them? {#what-do-ais-for-corporations}
+## What do GenAIs do for the corporations pushing them? {#what-do-ais-for-corporations}
 
-AIs do not seem to do much of value for students and teachers, as they are wasteful at best and harmful at worst.
+AIs do not offer much of value in teaching and learning: they are wasteful at best and harmful at worst.
 But the technology must do something for someone, otherwise it would not be ubiquitous.
-So what do AIs do for the the organizations that develop them?
 So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities^[Some observers believe that the AI fever has become detrimental to actual innovation [@siegel_ai_2023; @vinsel_dont_2023].].
-Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.] [@tricot_venture_2021].
-
-Alas, the industry has not necessarily made money for investors yet.
+Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD) [@tricot_venture_2021].
+But, the industry has not necessarily made money for investors yet.
 
@@
 @patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google.
-At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 monthly queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer.
-And a large number of users do not pay: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023].
-The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing  to the Global South the most disturbing, degrading, and damaging aspects of training AIs [@nix_analysis_2024].
+At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 monthly queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) before OpenAI stareted losing money on that one customer.
+The profitability of AI would look even worse if not for the fact that the cost of labor is contained by outsourcing  to the Global South the most disturbing, degrading, and damaging aspects of training AIs [@nix_analysis_2024].
 
@@
 
-## What do GenAIs do to everyone and everything else? {#what-do-ais-do-to-everyone-else}
-
-### AI is an eminently extractive and exploitative industry
-
-ChatGPT 4.o was announced on May 13, 2024, with great fanfare and a flashy display of the model's new capabilities, including functions to respond vocally to prompts, and to identify images via a mobile device's camera.
-4.o's voice, in particular, caused considerable titillation [@roose_is_2024], modeled as it was after the digital personal assistant of the movie _Her_ (Sam Altman's favorite movie by his own account).
-The background story of 4.o's voice is more sordid: Altman reportedly tried to hire Scarlett Johannson, who played the AI assistant in the movie, to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when she declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways.
-This voice was retired after Ms.
-Johansson threatened legal action [@allyn_scarlett_2024].
-
-Alas, very few have the wealth and social standing of Scarlett Johansson to challenge OpenAI.
-Most people in the world lack that power and the AI industry has not been shy to exploit the imbalance [@roberts_commercial_2016; @germain_google_2023; @weatherbed_adobes_2024].
-
-Besides being extractive, the AI industry is exploitative in darker ways.
-ChatGPT was trained on vast amounts of data scooped from the internet, where many societal problems are amplified.
-Trained on horrible content, it was hard to sell a ChatGPT prone to racist, sexist, obscene outbursts.
-Fortunately for OpenAI, companies like facebook had already hit on a solution of sorts to reduce the cost of moderation.
-The idea was basically to train AIs to detect content that was unpalatable.
-Indeed, they had found that feeding an AI with labeled examples of all the worst that exists on the internet (e.g., violence, hate speech, child sexual abuse) could produce tools that learned to detect various manifestations of such toxicity, to filter it out before it reached users.
-To obtain the essential labeled inputs, OpenAI sent tens of thousands of samples to Kenya, with text that described murder, suicide, bestiality, self-harm, incest, torture, and child sexual abuse situations in graphic detail.
-A sanitized ChatGPT is only possible through the labor (with wages around $1.32 and $2 per hour) of workers who were exposed to the worst of the internet.
-As one worker tasked with reading and labeling text for OpenAI told TIME, "he suffered from recurring visions after reading a graphic description of a man having sex with a dog in the presence of a young child.
-'That was torture,' he said.
-'You will read a number of statements like that all through the week.
-By the time it gets to Friday, you are disturbed from thinking through that picture'" [@perrigo_2_2023].
-
-### Privacy issues
-
-LLM are known to regurgitate verbatim chunks of the training datasets.
-This is called _memorization_, or in more conventional terms, model overfitting--an extreme regression to the mean that excels at predicting data in the training set.
-Overfitting happens when a model has too many parameters or too few data inputs.
-Data "memorization"--the verbatim or quasi-verbatim output of training data--can be as high as 7% in the case of some LLMs [@peng_near_duplicate_2023, p.
-8].
-This means that data cannot be presumed to be private once that they have been used to train an AI.
-And there is besides the issue of data leaks.
-So far, organizations such as the National Archives and Records Administration in the US, the US Congress, and companies like Samsung and Apple have restricted or banned the use of ChatGPT and Copilot for data security reasons [@koebler_national_2024; @singh_after_2024]: no data transferred in the form of a prompt can be presumed to be safe.
-
-### Environmental impacts
-
-Every human activity has an environmental impact, and at a time when we face unprecedented challenges with climate change, we must be discerning in how we consume resources.
-The choice is clear and Microsoft has told us: we need to burn the planet [@smith_our_2024] to create the AIs that will help us save the planet [@nakagawa_accelerating_2023].
-
-Researchers like @crawford_generative_2024 have already sounded the alarm: the environmental costs of AI are soaring, and doing so mostly in secret [@luccioni_power_2024].
-It has been estimated that ChatGPT consumes up to 25 times more energy than a regular web search,, and according to some estimates, the whole of AI may consume twice as much energy as the whole of France by 2030 [@brussels_chatgpt_2024].
-As the examples in the preceding sections illustrate, this is in exchange for results that are not better than a simple search, and often are considerably worse.
-
-### Reproduction of unjust structures of power
-
-When using a Generative AI we must ask whose perspectives the tool will tend to reproduce and amplify.
-In this sense, numerous observers [e.g., @ahuja_we_2023; @stross_tech_2023; @gebru_tescreal_2024] have noted the ideological leanings of various actors in the AI scene, an elite world of tech investors with a political agenda hiding in plain sight^[@thiel_education_2009 has written about how he "no longer believe[s] that freedom and democracy are compatible".
-Sam Bankman-Fried's FTX donated millions to a group with racist ties [@wilson_sam_2024].
-Elon Musk once shared a meme comparing Canada PM Justin Trudeau to Hitler while defending the vandals who terrorized Ottawa in 2022 [@henderson_and_elon_2022].].
-
-Besides concern about the influence of a small group of uber-wealthy people who collectively can muster more than _half a trillion dollars_ to turn their beliefs into reality [@stross_tech_2023], there is still leaves immediate problem that the datasets used to train AIs are, like the technology itself, not politically neutral.
-_Platformed racism_ [@matamoros-fernandez_platformed_2017], for instance, is known to reproduce malignant biases, that can emerge in a grass-roots fashion, but are also assisted by bad-faith actors^[Russian disinformation is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020].
-A U.S.
-Department of State report notes that Russian intelligence exploits "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S.
-politics [@us_department_of_state_gec_2024, p.
-51].].
-Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has already been passed to their AIs^[This is part of the Unfathomable Data problem discussed by @bender_dangers_2021.].
-
-Search engines already tended to reinforce racism [@noble_algorithms_2018]; given the above, no one should expect AI-augmented search engines to do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023].
-In addition, AI is also a tool ripe for abuse.
-Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].].
-In another wretched incident, Google's AI Overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that follows on a years-long Russian-assisted right-wing propaganda campaign to otherize Mr.
-Obama [@tenbarge_glue_2024].
-
-Mishaps like this, and others, are dismissed by Google as attempts "to trip up the technology with uncommon questions" [@tenbarge_glue_2024].
-This dismissal only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do).
-Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring.
-Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes?
-
 ## Concluding remarks
@@
 But those uses have been suffocated, and to some extent discredited, by the current, carefully manufactured hype around Generative AI [@morrone_signals_2024].
-There are now calls to press pause on the hype [e.g., @angwin_press_2024] and universities and colleges would be wise to heed this advice, lest they fall prey to a confidence trick.
-Higher education is often driven by a deeply ingrained belief in innovation (even, or perhaps particularly, for its own sake), and this impulse is compounded by how AI adoption has been drawn as a false dilemma,  triggering a fear of missing out.
+There are now calls to press pause on the hype [e.g., @angwin_press_2024] and universities and colleges would be wise to heed this advice.
+Higher education is often driven by a deeply ingrained belief in innovation (even, or perhaps particularly, for its own sake), and this impulse is compounded by how AIs have been presented as _everything tools_ and their adoption drawn as a false dilemma.
 Other pressures at work include a decades-long history of manufactured crises in education [@cizek_give_1999; @usher_state_2023].
 
-This essay described the numerous issues with AI as an _everything tool_.
+This essay argues that current AI technologies, far from being _everything tools_, are simply exceptionally complex regressions to the mean.
+They cannot be used reliably to retrieve or generate information [@hanlon_llm_2024], and their adoption is inconsistent with values like integrity [@vee_moral_2024], respect for human rights [@nix_analysis_2024], and environmental sustainability [@hogan_fumes_2024].
+We would do our students an unforgivable disservice by teaching them to use a crutch while leading them to believe that they are driving an all-terrain.
+In truth, a university is too expensive to be attractive for learning to depend on Skill-as-a-Service, when there are far cheaper alternatives (@fig-trail-blazing-bundle).
+As well, following the hype would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors.
+A university’s main source of authority is its reputation as a place that preserves and expands knowledge in a principled way: AIs do not and cannot do this.
+If we trade the reputation of universities for expediency, what would be left to sustain them as institutions valuable to societies?
 
-Current AI technologies cannot be used in a way that is consistent with values like integrity, respect for human rights, and environmental sustainability.
-This is well-known: most discussions of AI in education acknowledge that their use is rife with problems.
-For instance, @mollick_instructors_2024 note issues with copyright, exploitation, reproduction of biases, automation bias, and privacy loss.
-These are issues that "educators may want to consider before deciding whether to use these systems".
-This disclaimer highlights the important role of institutions to provide the moral and ethical clarity that may elude individual educators working under desperately under-resourced conditions, or simply willing to win a race to the bottom.
-And yet, I have little illusion that these arguments will prove persuasive.
+```{r ai-bundle, out.width="150%", fig.cap="A trail-blazing bundle: ChatGPT for data science and data analysis in Python"}
+#| label: fig-trail-blazing-bundle
 
-I would still appeal, though, to the university's sense of self-regard and self-preservation. 
-We would do our students an unforgivable disservice by offering them a skill that regressed them to mediocrity and left them at the mercy of predatory corporations.
-"This university uses ChatGPT to teach" is simply a way of saying that we are too cheap to offer a human-centered education.
-But in truth, a university is too expensive an institution to be attractive under such conditions, when there are far less expensive alternatives for an AI-based education ("for only $9.99 proof yourself against layoffs").
-Following the hype would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors.
-A university’s main source of authority is its reputation as a place that preserves and expands knowledge in a principled way.
-If we give this principle away, what would be left to sustain universities as institutions that add value to society?
+ggdraw() + 
+  draw_image(image = glue::glue(here::here(), "/images/trail-blazing-ai-bundle.png", width = 0.8))
+```
 

### 2025-01-26 paper/paper.qmd
Converted the paper into an essay for submission to Critical AI

@@
 
-The use of Generative Artificial Intelligence (AI) in higher education has been a sensational topic since the launch of ChatGPT and similar tools.
+The use of Generative Artificial Intelligence (AI) in higher education has become a sensational topic.
 AI has been presented as an everything tool that can do a fabulous array of tasks, from creating content to grading to providing feedback to students.
@@
 
-Unlike a linear regression model, the predictions of the neural network can be non-linear (@fig-data-vs-preds-02 ).
+Unlike a linear regression model, the predictions of the neural network can be non-linear (@fig-data-vs-preds-02).
 This flexibility is achieved by being a less parsimonious model (it uses more parameters).
